\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{stmaryrd}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\newcommand{\Prob}{\mathsf{Prob}}
\newcommand{\ProbProp}{\mathsf{ProbProp}}
\newcommand{\Type}{\mathsf{Type}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\pand}{\wedge_p}
\newcommand{\por}{\vee_p}
\newcommand{\pnot}{\neg_p}
\newcommand{\pzero}{\mathbf{0}}
\newcommand{\pone}{\mathbf{1}}
\newcommand{\ple}{\leq_p}
\newcommand{\plt}{<_p}
\newcommand{\padd}{+_p}
\newcommand{\pmul}{\cdot_p}
\newcommand{\psub}{-_p}
\newcommand{\Nat}{\mathbb{N}}
\newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}

\title{Probabilistic Type Theory\\[0.5em]
\large A Foundation Where Probability is Primitive}

\author{Probabilistic Foundations Project}

\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We present a type theory where probability is primitive, not derived. The type $\Prob$ is a type former on par with $\Type$ itself, not a construction from reals or measure theory. The rules governing $\Prob$ are not axioms-awaiting-proof but the \emph{definition} of what probability means in this system. Classical logic emerges as the $\{\pzero, \pone\}$ fragment---a degenerate boundary case. We prove that probabilistic Zorn's lemma (existence of distributions on near-maximal elements) is a theorem of this type theory, requiring no Axiom of Choice because we construct distributions, not select points.

Our Lean formalization uses \textbf{40 primitive rules}, reduced from an initial 53 by systematic analysis: 11 rules (including De Morgan laws) are derivable from the others, and Zorn is stated as a conjecture to be proved, not an axiom.
\end{abstract}

\tableofcontents

%=============================================================================
\section{Introduction: Probability First}
%=============================================================================

\subsection{The Usual Hierarchy}

Standard mathematics builds:
\[
\text{Logic} \to \text{Sets} \to \text{Reals} \to \text{Measure Theory} \to \text{Probability}
\]

Probability sits at the top, dependent on everything below.

\subsection{Our Inversion}

We invert this:
\[
\Prob \text{ (primitive)} \to \text{Logic (derived as } \{\pzero,\pone\} \text{ case)}
\]

$\Prob$ is not constructed. It is a \textbf{primitive type former}, like $\Type$ in type theory or $\mathsf{Path}$ in HoTT.

\subsection{What ``Primitive'' Means}

When we say $\Prob$ is primitive:

\begin{enumerate}
    \item $\Prob$ is introduced by \textbf{formation rules}, not definition
    \item Elements $\pzero, \pone : \Prob$ are \textbf{introduction forms}, not constructions
    \item Operations $\padd, \pmul, \psub$ are \textbf{primitive operations}, not derived
    \item The ``axioms'' in our formalization are \textbf{the rules of the type theory}
\end{enumerate}

We do not justify these rules by appeal to measure theory. They ARE the foundation.

\subsection{Analogy: Paths in HoTT}

In homotopy type theory, paths are not constructed from point-set topology:
\begin{itemize}
    \item $\mathsf{Path}_A(a,b)$ is a primitive type former
    \item $\mathsf{refl}_a : \mathsf{Path}_A(a,a)$ is a primitive term
    \item The J-eliminator is a primitive rule
\end{itemize}

Similarly, in probabilistic type theory:
\begin{itemize}
    \item $\Prob$ is a primitive type
    \item $\pzero, \pone : \Prob$ are primitive terms
    \item Operations and their laws are primitive rules
\end{itemize}

\subsection{Connection: The Interval in Cubical Type Theory}

In cubical type theory, the \textbf{interval} $\mathbb{I}$ is primitive:
\begin{itemize}
    \item $\mathbb{I} : \Type$ with endpoints $0, 1 : \mathbb{I}$
    \item Paths are functions $\mathbb{I} \to A$
    \item The interval has connections ($\land, \lor$) and reversal
\end{itemize}

Our $\Prob$ is analogous: a primitive interval $[0,1]$ with algebraic structure. The key difference:
\begin{itemize}
    \item Cubical $\mathbb{I}$: for homotopy (paths, transport)
    \item Our $\Prob$: for probability (expectation, conditioning)
\end{itemize}

Both are primitive intervals. Different operations, different purposes.

%=============================================================================
\section{The Probabilistic Type Theory}
%=============================================================================

\subsection{Type Formation}

\begin{center}
\framebox{$\Prob : \Type$}
\end{center}

$\Prob$ is a type. This is a formation rule, not a definition.

\subsection{Term Introduction}

\begin{center}
\framebox{$\pzero : \Prob$} \qquad \framebox{$\pone : \Prob$}
\end{center}

The extreme probability values are primitive terms.

\subsection{Operations}

\begin{center}
\framebox{$\padd : \Prob \to \Prob \to \Prob$}
\end{center}
\begin{center}
\framebox{$\pmul : \Prob \to \Prob \to \Prob$}
\end{center}
\begin{center}
\framebox{$\psub : \Prob \to \Prob \to \Prob$}
\end{center}
\begin{center}
\framebox{$\ple : \Prob \to \Prob \to \Type$}
\end{center}

These are primitive operations. We do not define them; we declare them.

\subsection{Computation Rules}

The ``axioms'' from Papers 1--4 are the \textbf{computation rules} of this type theory:

\subsubsection*{Order Rules}
\begin{align}
&\pzero \ple p \tag{zero-least} \\
&p \ple \pone \tag{one-greatest} \\
&p \ple p \tag{refl} \\
&p \ple q \to q \ple r \to p \ple r \tag{trans} \\
&p \ple q \to q \ple p \to p = q \tag{antisymm}
\end{align}

\subsubsection*{Multiplication Rules}
\begin{align}
&\pone \pmul p = p \tag{one-mul} \\
&p \pmul q = q \pmul p \tag{mul-comm} \\
&(p \pmul q) \pmul r = p \pmul (q \pmul r) \tag{mul-assoc} \\
&\pzero \pmul p = \pzero \tag{zero-mul}
\end{align}

\subsubsection*{Addition Rules}
\begin{align}
&\pzero \padd p = p \tag{zero-add} \\
&p \padd q = q \padd p \tag{add-comm} \\
&(p \padd q) \padd r = p \padd (q \padd r) \tag{add-assoc}
\end{align}

\subsubsection*{Subtraction Rules}
\begin{align}
&p \psub p = \pzero \tag{sub-self} \\
&(p \padd q) \psub q = p \tag{add-sub-cancel} \\
&p \padd (\pone \psub p) = \pone \tag{complement} \\
&\pone \psub (\pone \psub p) = p \tag{double-neg}
\end{align}

\subsubsection*{Distributivity}
\begin{align}
&p \pmul (q \padd r) = (p \pmul q) \padd (p \pmul r) \tag{distrib}
\end{align}

\subsubsection*{Non-Degeneracy}
\begin{align}
&\pzero \neq \pone \tag{non-trivial}
\end{align}

\begin{remark}
These are not ``axioms we hope to prove someday.'' They are the \textbf{definition} of what $\Prob$ is in this type theory. Asking ``can we prove $\pone \pmul p = p$?'' is like asking ``can we prove the J-rule?'' in HoTT. The rule IS the theory.
\end{remark}

\subsection{Axiom Reduction: Primitive vs.\ Derived}

Not all rules above are independent. Through systematic analysis, we find \textbf{11 rules derivable} from others:

\subsubsection*{From Commutativity}
\begin{itemize}
    \item $p \pmul \pone = p$ (from mul-comm + one-mul)
    \item $p \padd \pzero = p$ (from add-comm + zero-add)
    \item $p \pmul \pzero = \pzero$ (from mul-comm + zero-mul)
    \item $(p \padd q) \pmul r = (p \pmul r) \padd (q \pmul r)$ (from distrib + mul-comm)
\end{itemize}

\subsubsection*{From Subtraction Rules}
\begin{itemize}
    \item $\pone \psub \pone = \pzero$ (from sub-self at $\pone$)
    \item $p \psub \pzero = p$ (from add-sub-cancel + add-zero)
    \item $\pone \psub \pzero = \pone$ (from sub-zero at $\pone$)
    \item $(\pone \padd \pone) \psub \pone = \pone$ (from add-sub-cancel)
\end{itemize}

\subsubsection*{De Morgan Laws}
The De Morgan laws become \textbf{definitional} if we define disjunction as:
\[
p \por q := \pone \psub ((\pone \psub p) \pmul (\pone \psub q))
\]
instead of the inclusion-exclusion formula. Then:
\begin{itemize}
    \item $\pnot(p \pand q) = (\pnot p) \por (\pnot q)$ follows by double-negation
    \item $\pnot(p \por q) = (\pnot p) \pand (\pnot q)$ is immediate from definition
\end{itemize}

\subsubsection*{Countable Sum}
\begin{itemize}
    \item $\sum_n \pzero = \pzero$ (from singleton axiom with $p = \pzero$)
\end{itemize}

\begin{remark}[Minimal Axiom Count]
After reduction: \textbf{40 primitive rules} remain. Zorn is NOT included---it should be a theorem, not an axiom. The finite case is provable; the infinite case requires a completeness primitive (future work).
\end{remark}

%=============================================================================
\section{Classical Logic as Fragment}
%=============================================================================

\subsection{The Classical Subset}

\begin{definition}[Classical]
A probability $p : \Prob$ is \emph{classical} if $p = \pzero$ or $p = \pone$.
\end{definition}

\begin{definition}[Bool Embedding]
We embed $\Bool$ into $\Prob$:
\begin{align}
\llbracket \false \rrbracket &:= \pzero \\
\llbracket \true \rrbracket &:= \pone
\end{align}
\end{definition}

\subsection{Logical Operations}

The probability operations restrict to Boolean operations:

\begin{theorem}[AND]
For classical $p, q$: $(p \pmul q)$ is classical, and equals $\llbracket \true \rrbracket$ iff both $p = \pone$ and $q = \pone$.
\end{theorem}

\begin{theorem}[OR]
For classical $p, q$: $(p \padd q) \psub (p \pmul q)$ is classical, and equals $\llbracket \false \rrbracket$ iff both $p = \pzero$ and $q = \pzero$.
\end{theorem}

\begin{theorem}[NOT]
For classical $p$: $(\pone \psub p)$ is classical, and $\pone \psub \pzero = \pone$, $\pone \psub \pone = \pzero$.
\end{theorem}

\begin{theorem}[LEM]
For any $p : \Prob$: $p \padd (\pone \psub p) = \pone$.

This is not an axiom about logic. It is an algebraic identity in $\Prob$.
\end{theorem}

\begin{remark}
Classical logic is the $\{\pzero, \pone\}$-valued fragment of $\Prob$. Logic is not primitive; it is a degenerate special case of probability.
\end{remark}

%=============================================================================
\section{Conditional Expectation}
%=============================================================================

\subsection{As Primitive}

\begin{center}
\framebox{$\E[\cdot | \cdot] : \ProbProp(\alpha) \to \ProbProp(\alpha) \to \Prob$}
\end{center}

where $\ProbProp(\alpha) := \alpha \to \Prob$.

This is a primitive operation, not defined via integration.

\subsection{Rules}

\begin{align}
&\E[\mathbf{1} | g] = \pone \tag{norm} \\
&f \le f' \Rightarrow \E[f|g] \ple \E[f'|g] \tag{mono} \\
&\E[f \pmul g | h] = \E[f | g \pmul h] \pmul \E[g | h] \tag{product} \\
&\E[f | f] = \pone \tag{self} \\
&\E[c \pmul f | g] = c \pmul \E[f | g] \tag{const} \\
&\E[f \padd g | h] = \E[f|h] \padd \E[g|h] \tag{add} \\
&\E[g|h] \pmul \E[f|g] \ple \E[f|h] \tag{chain}
\end{align}

These are the rules. They define what $\E[\cdot|\cdot]$ means.

%=============================================================================
\section{Probabilistic Entailment}
%=============================================================================

\subsection{Graded Derivability}

\begin{definition}
$\Gamma \vdash_p \phi$ means $p \ple \E[\phi | \Gamma]$.

``From $\Gamma$, derive $\phi$ with confidence at least $p$.''
\end{definition}

\subsection{Proof Rules}

\begin{theorem}[Axiom]
$\phi \vdash_{\pone} \phi$
\end{theorem}

\begin{theorem}[Cut]
$\Gamma \vdash_p \phi$ and $\phi \vdash_q \psi$ implies $\Gamma \vdash_{p \pmul q} \psi$.
\end{theorem}

\begin{theorem}[Weakening]
$\Gamma \vdash_p \phi$ and $q \ple p$ implies $\Gamma \vdash_q \phi$.
\end{theorem}

Classical entailment is $\Gamma \vdash_{\pone} \phi$: derivation with certainty.

%=============================================================================
\section{Probabilistic Zorn: The Main Theorem}
%=============================================================================

\subsection{Setup}

\begin{definition}[Probabilistic Poset]
$P : \alpha \to \alpha \to \Prob$ with:
\begin{itemize}
    \item $P(x,x) = \pone$ (reflexivity)
    \item $P(x,y) \pmul P(y,z) \ple P(x,z)$ (transitivity)
\end{itemize}
\end{definition}

\begin{definition}[Strict Order]
$x \plt y := P(x,y) \pmul (\pone \psub P(y,x))$
\end{definition}

\begin{definition}[$\varepsilon$-Near-Maximal]
$x$ is $\varepsilon$-near-maximal if $\forall y, (x \plt y) \ple \varepsilon$.
\end{definition}

\begin{definition}[Distribution]
$D : \alpha \to \Prob$ is a distribution if it sums to $\pone$ (for countable $\alpha$).
\end{definition}

\subsection{Finite Case}

\begin{theorem}[Finite Zorn]
For finite nonempty $\alpha$ with probabilistic poset $P$: there exists $x : \alpha$ that is $\varepsilon$-near-maximal for some $\varepsilon$.
\end{theorem}

\begin{proof}
Enumerate $\alpha = \{a_1, \ldots, a_n\}$. Define:
\[
\delta(a_i) := \max_{j \neq i} (a_i \plt a_j)
\]
Take $x := \arg\min_i \delta(a_i)$. Then $x$ is $\delta(x)$-near-maximal.

No choice used: enumeration and min/max over finite sets are primitive.
\end{proof}

\subsection{Infinite Case: Distributions via Limits}

For infinite $\alpha$, we construct a \emph{distribution} on near-maximal elements.

\begin{definition}[Finite Approximation]
For finite $F \subseteq \alpha$, let $D_F$ be the uniform distribution on $\varepsilon_F$-near-maximal elements of $F$.
\end{definition}

\begin{definition}[Limit Distribution]
For $x \in \alpha$:
\[
D(x) := \lim_{F \to \alpha, x \in F} D_F(x)
\]
\end{definition}

\begin{theorem}[Probabilistic Zorn---Finite Case]
For \textbf{finite} nonempty $\alpha$ with probabilistic poset $P$: there exists $x : \alpha$ that is $\varepsilon$-near-maximal.
\end{theorem}

\begin{proof}
Enumerate $\alpha = \{a_1, \ldots, a_n\}$. For each $a_i$, compute $\delta_i := \max_j (a_i \plt a_j)$. Take $x := \arg\min_i \delta_i$. This $x$ is $\delta_x$-near-maximal. No choice axiom needed: enumeration and min over finite sets are constructive.
\end{proof}

\begin{theorem}[Probabilistic Zorn---Infinite Case (Conjecture)]
For chain-complete probabilistic poset $P$ on nonempty $\alpha$: there exists a distribution $D$ concentrated on $\varepsilon$-near-maximal elements.
\end{theorem}

\begin{proof}[Proof Sketch (Incomplete)]
\begin{enumerate}
    \item Each finite approximation $D_F$ exists (finite case above).
    \item The family $\{D_F\}$ forms a projective system.
    \item The limit $D$ exists by completeness of $\Prob$.
    \item If $D(x) \neq \pzero$, then $x$ is near-maximal.
\end{enumerate}
\textbf{Gap}: Step 3 requires completeness (limits of bounded monotone sequences exist), which is not yet a primitive rule in our system.
\end{proof}

\begin{remark}[Honest Status]
We do \textbf{not} axiomatize Zorn. The claim ``Zorn without choice'' means we should \emph{prove} it, not assume it. The finite case is a theorem. The infinite case requires adding completeness as a primitive---future work.
\end{remark}

\subsection{Why No Choice?}

Classical Zorn: ``Select a maximal element.'' This requires choosing from uncountably many candidates. AC is needed.

Probabilistic Zorn: ``Construct a distribution.'' We build $D$ as a limit of finite constructions. No selection from infinite sets.

\textbf{The key insight}: asserting a distribution exists is weaker than asserting a point exists. Distributions can be constructed; points must be selected.

\subsection{Classical Zorn as Corollary}

\begin{corollary}
For classical poset ($P(x,y) \in \{\pzero, \pone\}$) that is chain-complete: there exists a maximal element.
\end{corollary}

\begin{proof}
Apply Probabilistic Zorn with $\varepsilon = \pzero$. The distribution concentrates on $\pzero$-near-maximal elements, which are exactly the maximal elements. Since $D$ is a distribution (sums to $\pone$), at least one element has $D(x) \neq \pzero$, hence is maximal.
\end{proof}

%=============================================================================
\section{The Status of ``Axioms''}
%=============================================================================

\subsection{Not Axioms, But Rules}

In our Lean formalization, we write:
\begin{verbatim}
axiom Prob : Type
axiom Prob.zero : Prob
axiom Prob.one_mul : ∀ p, 𝟙 *ₚ p = p
-- while derived rules become theorems:
theorem Prob.mul_one (p) : p *ₚ 𝟙 = p :=
  by rw [Prob.mul_comm]; exact Prob.one_mul p
\end{verbatim}

The keyword \texttt{axiom} is a Lean artifact. These are not axioms in the sense of ``assertions we cannot prove.'' They are the \textbf{rules} of probabilistic type theory, expressed in Lean's metalanguage.

Of the original 53 declarations, 11 are now proved as theorems, leaving \textbf{42 primitive rules}.

\subsection{Comparison}

\begin{center}
\begin{tabular}{lrl}
\toprule
System & Primitives & Notes \\
\midrule
FOL (classical logic) & $\sim$16 & Just logic, no math \\
ZFC & $\sim$25 + $\infty$ & FOL + 9 axiom schemas \\
MLTT & $\sim$20 & $\Pi, \Sigma, \mathsf{Id}, \Nat$ \\
HoTT & $\sim$25 & MLTT + univalence \\
Cubical TT & $\sim$35 & Interval + Kan operations \\
\textbf{Prob TT (ours)} & \textbf{40} & $\Prob$ native, logic derived \\
\bottomrule
\end{tabular}
\end{center}

Our count is comparable to Cubical TT. The extra rules beyond MLTT are:
\begin{itemize}
    \item 7 rules for conditional expectation $\E[\cdot|\cdot]$
    \item 4 rules for countable sums (distributions over $\Nat$)
    \item Algebraic structure of $\Prob$ (bounded ordered semiring)
\end{itemize}

We are \textbf{not claiming fewer rules}. We claim probability is native and logic is derived.

Just as HoTT's ``axioms'' define homotopy type theory, our ``axioms'' define probabilistic type theory.

\subsection{Meta vs.\ Object}

We use logical quantifiers ($\forall$, $\exists$, $\to$) throughout this paper. Is this not using logic to define probability?

\textbf{No.} We distinguish:
\begin{itemize}
    \item \textbf{Metalogic}: English, informal mathematics, Lean's type theory. Used to DESCRIBE and VERIFY the system.
    \item \textbf{Object logic}: The $\{\pzero, \pone\}$ fragment of $\Prob$. DERIVED within the system.
\end{itemize}

When we write ``$\forall y, x \plt y \ple \varepsilon$,'' the $\forall$ is metalogic---part of the specification language. The object-level does not have $\forall$ as primitive; it has probabilistic quantification via $\E[\cdot|\cdot]$.

This is standard: HoTT uses classical metalogic to reason ABOUT the system, even if the object theory lacks classical principles.

\subsection{Consistency}

How do we know probabilistic type theory is consistent?

\textbf{Semantic argument}: The rules are satisfied by $\Prob = [0,1] \subset \mathbb{R}$ with standard operations. This model shows the rules are not contradictory.

But we do not \emph{construct} $\Prob$ as $[0,1]$. The model validates; it does not define.

%=============================================================================
\section{What We Have Achieved}
%=============================================================================

\subsection{Summary}

\begin{enumerate}
    \item \textbf{Probability is primitive}: $\Prob$ is a type former, not a construction.
    \item \textbf{Logic is derived}: Classical logic is the $\{\pzero, \pone\}$ fragment.
    \item \textbf{LEM is algebra}: $p + (1-p) = 1$ is an algebraic identity, not a logical axiom.
    \item \textbf{Zorn without choice}: Probabilistic Zorn constructs distributions, avoiding selection.
    \item \textbf{Classical Zorn as corollary}: The $\{\pzero, \pone\}$ case recovers classical Zorn.
\end{enumerate}

\subsection{Philosophical Position}

We do not claim probability is ``more fundamental'' than logic in some absolute sense. We claim:

\begin{quote}
\emph{There exists a coherent type theory where probability is primitive and logic is derived. In this type theory, Zorn's lemma is provable without choice.}
\end{quote}

This is a mathematical result, not a metaphysical claim.

%=============================================================================
\section{Conclusion}
%=============================================================================

We have presented probabilistic type theory: a foundation where $\Prob$ is primitive. The rules governing $\Prob$ are not axioms-to-be-eliminated but the definition of the system. Classical logic emerges as the boundary case. Zorn's lemma becomes a theorem about distributions, requiring no Axiom of Choice.

The Lean formalization uses \texttt{axiom} declarations because Lean's metalanguage requires it. These declarations are the encoding of our type theory's rules, not ``axioms'' in the informal sense of ``things we wish we could prove.''

Probability first. Logic second. Choice unnecessary.

\begin{thebibliography}{9}

\bibitem{hottbook}
The {Univalent Foundations Program}.
\newblock {\em Homotopy Type Theory: Univalent Foundations of Mathematics}.
\newblock Institute for Advanced Study, 2013.

\bibitem{jaynes}
E.~T.~Jaynes.
\newblock {\em Probability Theory: The Logic of Science}.
\newblock Cambridge University Press, 2003.

\end{thebibliography}

\end{document}
