\documentclass[11pt,a4paper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{stmaryrd}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage[capitalize]{cleveref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,calc,positioning,shapes.geometric}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue,
  pdfauthor={Christopher Albert},
  pdftitle={Conditioning by Fibers: A Chain-Rule Semantics for Probability, Many-Valued Logic, and Impossible Evidence}
}

\emergencystretch=3em

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\cred}{\mathsf{cred}}
\newcommand{\Cond}{\mathsf{Cond}}
\newcommand{\Adm}{\mathsf{Adm}}
\newcommand{\Sol}{\mathsf{Sol}}
\newcommand{\collapse}{\kappa}
\newcommand{\negc}{\mathord{\sim}}
\newcommand{\conjc}{\otimes}
\newcommand{\disjc}{\sqcup}
\newcommand{\half}{\tfrac12}
\newcommand{\quarter}{\tfrac14}
\newcommand{\three}{\mathbf{3}}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\val}[1]{\llbracket #1 \rrbracket}
\newcommand{\Lean}[1]{\texttt{#1}}
\newcommand{\seq}{\vdash}

\title{Conditioning by Fibers:\\
\large A Chain-Rule Semantics for Probability, Many-Valued Logic, and
Impossible Evidence}
\author{Christopher Albert\\
\small Institute for Theoretical and Computational Physics\\
\small Graz University of Technology}
\date{June 2026}

\begin{document}

\maketitle

\begin{center}
\small
Transparenzhinweis: Mit Unterstützung von Claude- und GPT-Modellen erstellt; Inhalt, Auswahl und fachliche Prüfung verantwortet der Autor.
\end{center}

\begin{abstract}
This paper isolates a small formal core for reasoning with impossible
evidence.  Values usually live in $[0,1]$ with complement
$\negc c=1-c$, product conjunction $a\conjc b=ab$, and De Morgan
disjunction $a\disjc b=a+b-ab$.  The primitive conditional object is
the admissibility judgment
\[
  \Adm(c;j,e)\quad\Longleftrightarrow\quad c e=j.
\]
The displayed set
\[
  \Cond(j,e)=\{c\in[0,1]:ce=j\}
\]
is the external fiber of that relation.  It is a singleton $\{j/e\}$
when $e>0$ and $j\le e$; it is all of $[0,1]$ when $j=e=0$; it is
empty when $j>e$.  Impossible evidence therefore does not imply
everything.  It imposes no equation.

The paper proves three consequences of that small observation.  First,
the product De Morgan triplet collapses to the Kleene lattice
$\{0,\half,1\}$, and the collapse matches LP consequence with
positivity on $[0,1]$, and K$_3$ consequence with certainty.  Second,
that bridge cannot be extended to conditioning: no truth function on
$\{0,\half,1\}$ reproduces min-copula conditioning after collapse.
Third, adding a total internal conditional with modus ponens,
conditional proof, and contraction is impossible; the product residuum
is forced at positive antecedents and violates contraction.  Liar and
Russell equations become solution sets rather than detonators:
$c=1-c$ has the unique solution $\half$, while the truth-teller has
the same shape as zero-evidence conditioning, the whole interval.

The useful contribution is the separation principle: truth values,
dependence data, and conditional inference are distinct inputs.  The
same fiber construction works on the continuum, on the Boolean carrier,
and on three-valued variants once the carrier and conjunction are
fixed.  A non-singleton fiber is propagated: subsequent rules carry the
admissible values instead of selecting a quotient at zero evidence.
Unknown dependence is handled the same way: with the joint confined to
its Fr\'echet interval, the conditional propagates as an interval that
is either robust or dependence-sensitive against a threshold.
Consequence and conditioning are then one inverse-image-inclusion
operator; a finite measure on valuations grades it, with ordinary
probability as the single-measure case and logical entailment as the
all-measures corner, so the universal quantifier is a special case of a
measure.  Finite
toy models give a first formalized cut at a foundation: dependence
contexts, proof provenance, hypothetical branches without explosion, a
generative connective calculus, and the $\sqrt2$ number-theoretic seed.
The formal results are machine-checked in Lean 4 with Mathlib; theorem
names are cited in the text.
\end{abstract}

\tableofcontents

\section{Contribution and Claims}

The paper keeps one mechanism and the obstructions around it.  The
novelty is not the product t-norm, the Kleene lattice, primitive
conditional probability, conditioning along fibers, set-valued
probability, fuzzy probability logic, conditional-event algebra, or
paraconsistency; those are established.  The claim is the verified
conjunction of nine statements:
\[
\begin{array}{ll}
\text{(N1)} &
\Adm_V(c;j,e)\Longleftrightarrow c\conjc e=j
\text{ as a carrier-uniform fiber semantics};\\[1mm]
\text{(N2)} &
\text{propagation of non-singleton fibers, with }
\Cond(0,0)=[0,1]\text{ as no ex falso};\\[1mm]
\text{(N3)} & \text{LP/K$_3$ collapse bridges for full formulas};\\[1mm]
\text{(N4)} & \text{no truth-functional bridge for conditioning};\\[1mm]
\text{(N5)} & \text{scalar admissibility proof theory (sound, closed-complete},\\
  & \text{plus a generative connective and first-order fragment)};\\[1mm]
\text{(N6)} & \text{Curry block explaining why no internal arrow is added};\\[1mm]
\text{(N7)} & \text{value, dependence, and conditioning as separate inputs:}\\
  & \text{the joint is supplied, and unknown dependence propagates an interval};\\[1mm]
\text{(N8)} & \text{a first formalized cut of the foundation layers}\\
  & \text{(dependence contexts, robust collapse, provenance, branches)};\\[1mm]
\text{(N9)} & \text{logic, probability, and imprecise probability as one}\\
  & \text{measured aggregation; the universal quantifier is its}\\
  & \text{all-measures corner (\Lean{entails\_iff\_cond\_one}).}
\end{array}
\]
Consequence (N3) and conditioning (N4) are the two instances of one
inverse-image-inclusion operator; a measure on valuations grades it,
with entailment as its all-measures corner (N9).  The separation theorem
is that only the value-level instance resists collapse:
\[
\boxed{\text{truth-functional consequence can collapse;
conditional inference cannot.}}
\]
The underlying thesis is elementary:
\[
\boxed{\text{inference constrains admissible values;
impossible evidence constrains none.}}
\]
Its force comes from the no-go results, N4 and N6.  Adjacent material,
arithmetic coding, bootstrapping analogies, and graded numerical
analysis, does not support this package and belongs in separate venues.

The title phrase \emph{conditioning by fibers} is literal.  Fix the
evidence value $e$.  Multiplication by $e$, or the chosen conjunction
with $e$, is a map from candidate conditionals to joint values.  A
conditional is admissible exactly when it lies in the preimage of the
given joint value.  At positive coherent evidence, probability theory
has a singleton fiber and writes a quotient.  Boolean and many-valued
variants restrict the carrier; the same preimage construction still
applies.  A truth table would have to choose one value.  The singular
fiber at zero evidence is a set of admissible values.  Logical
consequence is the same inclusion read over valuations rather than
values; one operator covers both, and a later section makes it explicit.

\paragraph{What the system is for.}
Cred is an audit calculus for uncertain inference.  Its practical use is
to expose hidden semantic commitments and the false precision they
produce.  A step from marginals to a conditional silently fixes a joint
policy, a way of reading the conditional off the joint, and a semantic
frame for the verdict.  Cred names those commitments and computes the
admissible set they leave open (\Lean{Audit.InferenceCommitments}).  The
practical target is false precision: a single confident number that
depends on an unstated joint, when the honest answer is an interval.
Cred audits such a step; it does not replace probabilistic, fuzzy, or
soft-logic systems and does not subsume them.  Those systems carry
inference and learning machinery Cred has no analogue for.  Cred adds
the Lean-checked factorization of the layers and the no-go and
uniqueness results that pin each layer down.

\begin{figure}[h]
\centering
\begin{tikzpicture}[>=Latex,node distance=4.2mm,
    lyr/.style={draw,rounded corners,fill=blue!6,align=center,
                inner sep=3pt,text width=0.62\linewidth}]
  \node[lyr] (v) {value algebra: $[0,1]$ with $\negc,\conjc,\disjc$};
  \node[lyr,below=of v] (w) {valuation / world space: $V=\mathrm{Atom}\to[0,1]$};
  \node[lyr,below=of w] (f) {$\forall$ consequence over valuations + designation};
  \node[lyr,below=of f] (d) {supplied dependence: the joint $j=\cred(A\wedge B)$};
  \node[lyr,below=of d] (c) {conditioning fiber: $\Cond(j,e)=\{c:c\conjc e=j\}$};
  \node[lyr,below=of c] (m) {optional measure on valuations: probability / imprecise};
  \draw[->] (v)--(w); \draw[->] (w)--(f); \draw[->] (f)--(d);
  \draw[->] (d)--(c); \draw[->] (c)--(m);
\end{tikzpicture}
\caption{The layers Cred factors.  Each is a separate input, with its own Lean
anchors (generic \texttt{Consequence}; \texttt{ExactJointContext}; \texttt{Cond};
\texttt{FinMeasure}).  Classical, Kleene, and fuzzy logics live at the consequence
layer; probability and imprecise probability enter only at the measure layer.}
\label{fig:layer-stack}
\end{figure}

\section{The Setting}

\subsection{Values and Operations}

\begin{definition}[Credence value]
A credence is a real number $c\in[0,1]$.
\end{definition}

The basic operations are
\[
  \negc c=1-c,\qquad
  a\conjc b=ab,\qquad
  a\disjc b=\negc(\negc a\conjc\negc b)=a+b-ab.
\]
The product is the product t-norm; the disjunction is its De Morgan
dual.  These operations are standard in many-valued and fuzzy logic
\cite{zadeh1965,goguen1969,hajek1998,klement2000,cintula2011}.
They are also familiar probabilistic expressions: if $A$ and $B$ are
independent, then
\[
  P(A\cap B)=P(A)P(B),\qquad
  P(A\cup B)=P(A)+P(B)-P(A)P(B).
\]
The paper never assumes independence by default.  The product
$a\conjc b$ is an algebraic operation on values.  A joint credence
$j=\cred(A\land B)$ is separate data.

\begin{figure}[h]
\centering
\begin{tikzpicture}[x=11cm,y=1.2cm]
  \draw[->] (0,0) -- (1.03,0) node[right] {$c$};
  \draw (0,0.08) -- (0,-0.08) node[below] {$0$};
  \draw (0.5,0.08) -- (0.5,-0.08) node[below] {$\half$};
  \draw (1,0.08) -- (1,-0.08) node[below] {$1$};
  \draw[very thick,blue] (0,0) -- (1,0);
  \draw[very thick,red,-{Latex[length=3mm]}] (0.13,0.42) -- (0.87,0.42);
  \draw[very thick,red,-{Latex[length=3mm]}] (0.87,0.42) -- (0.13,0.42);
  \node[above] at (0.5,0.48) {$\negc c=1-c$};
  \draw[fill=black] (0.5,0) circle (0.012);
  \node[above] at (0.5,0.02) {unique fixed point};
\end{tikzpicture}
\caption{The complement reverses the unit interval and fixes only
$\half$.  Lean: \Lean{Cred.Credence.neg\_fixed\_point\_unique}.}
\label{fig:unit-interval}
\end{figure}

The verified algebraic facts are routine but important:
\[
\begin{gathered}
\negc\negc c=c,\quad
a\conjc b=b\conjc a,\quad
(a\conjc b)\conjc c=a\conjc(b\conjc c),\\
a\conjc 1=a,\quad a\conjc 0=0,\quad
\negc(a\conjc b)=\negc a\disjc\negc b.
\end{gathered}
\]
Conjunction is idempotent only at the Boolean boundary:
\[
  c\conjc c=c \quad\Longleftrightarrow\quad c\in\{0,1\}.
\]
Lean: \Lean{Cred.Credence.conj\_idempotent\_iff}.  Thus intermediate
credences are not hidden truth values.

\subsection{Algebraic Form}

The construction is carrier-independent.  Let $V$ be a value carrier
with a conjunction-like operation $\conjc$, bottom $0$, and top $1$.
Given evidence $e\in V$ and joint value $j\in V$, define the primitive
admissibility judgment
\[
  \Adm_V(c;j,e)\quad:\Longleftrightarrow\quad c\conjc e=j.
\]
This judgment needs equality and conjunction.  It does not need sets.
After the surrounding mathematics has predicates or sets, one may
write the fiber.  For fixed evidence $e$, put
\[
  m_e(c)=c\conjc e.
\]
Then
\[
  \Cond_V(j,e)=\{c\in V:\Adm_V(c;j,e)\}=m_e^{-1}(j).
\]
The word fiber is used only in this ordinary mathematical sense:
preimage under the multiplication-by-$e$ map.  The paper uses this set
notation as metalanguage.

For the Boolean carrier $\mathbf 2=\{0,1\}$ with $\conjc=\wedge$:
\[
  \Cond_{\mathbf 2}(j,1)=\{j\},\qquad
  \Cond_{\mathbf 2}(0,0)=\{0,1\},\qquad
  \Cond_{\mathbf 2}(1,0)=\varnothing.
\]
Classical conditioning on true fixes the value.  Conditioning on
false leaves the conditional unconstrained unless the joint value is
impossible.

For the Kleene carrier $\three=\{0,\half,1\}$, the same schema applies
with the chosen three-valued conjunction:
\[
  \Cond_{\three}(j,e)=\{c\in\three:c\wedge e=j\}
\]
for the usual strong-Kleene/Godel conjunction $\wedge=\min$.
There is a closure caveat.  The literal set
$\{0,\half,1\}\subset[0,1]$ is not closed under product, since
$\half\cdot\half=\quarter$.  A three-valued product reading is
therefore a quotient of the continuum, while an internal three-valued
logic uses its own closed conjunction, usually $\min$.

\subsection{Conditioning as a Relation}

Conditional probability is usually introduced by division:
\[
  P(A\mid B)=\frac{P(A\land B)}{P(B)},\qquad P(B)>0.
\]
H\'ajek argues that this ratio is not a general analysis of
conditional probability \cite{hajek2003conditional}.  R\'enyi and
Popper instead treat conditional probability as primitive
\cite{renyi1955,popper1959,krauss1968,dubins1975}.  De Finetti's
conditional-event tradition makes the same point in a betting idiom:
if the antecedent fails, the bet is void rather than true
\cite{definetti1936,egre2020part1,egre2021}.  Current work on
indicative conditionals still turns on this boundary between truth,
probability, and undefinedness \cite{egre2025}.

Here the primitive is the chain rule alone:
\[
  \cred(A\mid B)\,\cred(B)=\cred(A\land B).
\]
Given joint value $j$ and evidence value $e$, define
\[
  \Adm(c;j,e)\quad:\Longleftrightarrow\quad c e=j.
\]
The associated fiber is
\[
  \Cond(j,e)=\{c\in[0,1]:c e=j\}.
\]
This is the weak-definition move: keep the identity that survives the
singular case, rather than the quotient that requires $e>0$.  The
analogy stops at uniqueness.  At $e=j=0$, the fiber is genuinely
underdetermined.

\begin{theorem}[Admissible-set trichotomy]
\label{thm:trichotomy}
For $j,e\in[0,1]$:
\[
\Cond(j,e)=
\begin{cases}
\{j/e\}, & e>0\ \text{and}\ j\le e,\\[2mm]
[0,1], & e=0,\ j=0,\\[1mm]
\varnothing, & j>e.
\end{cases}
\]
\end{theorem}

\noindent
Lean:
\Lean{Cred.Credence.cond\_singleton\_of\_pos},
\Lean{Cred.Credence.cond\_zero\_zero\_univ},
\Lean{Cred.Credence.cond\_nonempty\_iff}.

\begin{proof}
If $e>0$, the equation $ce=j$ has the unique solution $c=j/e$.
Since $c\le1$, existence requires $j\le e$.
If $e=0$, the equation is $0=j$.  For $j=0$ every $c$ solves it.
For $j>0$ none does.  The last case is included in $j>e$.
\end{proof}

When the fiber is not a singleton, later reasoning need not stop.
The next step carries the set of admissible values.  A positive
evidence step carries a singleton and therefore behaves like ordinary
Bayes division.  A zero-evidence step carries the whole interval.  A
downstream rule then maps that interval, intersects it with new
constraints, or leaves it unchanged.  This is the same discipline used
by interval and credal-set methods: compute with the set until extra
structure selects a value.  Here the set is the fiber of the
chain-rule equation itself, not a prior ignorance model.
The price is explicit: conditioning is no longer a total function at
singular evidence.  Later rules must carry sets of possible values.
This bookkeeping is the mechanism.  It replaces division by zero and
ex falso with propagation of the remaining admissible values.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=5]
  \fill[blue!7] (0,0) -- (1,0) -- (1,1) -- cycle;
  \draw[->] (0,0) -- (1.08,0) node[right] {$e$};
  \draw[->] (0,0) -- (0,1.08) node[above] {$j$};
  \draw[thick] (0,0) -- (1,1) node[above right] {$j=e$};
  \draw[fill=black] (0,0) circle (0.008);
  \node[below left] at (0,0) {$(0,0)$};
  \node[align=center] at (0.66,0.30) {singleton\\$\{j/e\}$};
  \node[align=center] at (0.28,0.76) {empty\\$j>e$};
  \node[align=left] at (0.08,0.11) {$[0,1]$};
  \draw[thick,red] (0,0) circle (0.03);
\end{tikzpicture}
\caption{The chain rule as geometry.  The coherent region is
$j\le e$.  Positive evidence gives a point.  The origin gives the
whole vertical fiber of possible conditionals.}
\label{fig:cond-geometry}
\end{figure}

This is the whole non-explosion mechanism.  From impossible evidence
one obtains no arbitrary conclusion because no conclusion is selected.
The equation has lost rank.

\subsection{Why no scalar joint is general}
One might hope to compute the joint truth-functionally, $j=F(a,b)$.
Two choices are natural: the product $F=ab$ and the minimum
$F=\min(a,b)$.  Read as joints, the product is independence and the
minimum is maximum positive dependence.  The product trivializes
conditioning, since $ab/b=a$ leaves the evidence irrelevant; and among
symmetric idempotent scalar joints satisfying the copula axioms the
minimum is forced (\Lean{Cred.min\_copula\_unique},
\Lean{Cred.truth\_functional\_forces\_min}).  So a single
truth-functional $F(a,b)$ is never the general dependence: it is either
independence or the comonotone extreme.  The joint
$j=\cred(A\wedge B)$ must therefore be supplied as data, free in the
Fr\'echet interval $[\max(a+b-1,0),\min(a,b)]$, and the conditional is
the fiber $\Cond(j,e)$.  Worlds and measures, introduced later, sit at
the level that fixes $j$; the value layer leaves it open.

\section{Binary and Continuous Cred}
\label{sec:binary-continuous}

Write $c$ for the value of ``$A$ given $B$'', $e$ for the value of $B$,
and $j$ for the value of ``$A$ and $B$''.  The basic relation is
$ce=j$.  Reading it on the Boolean carrier and on $[0,1]$ separates two
ideas that ordinary conditional probability folds together: loss of
constraint at impossible evidence, and freedom in the joint.

\subsection{Binary}
On $\mathbf 2=\{0,1\}$ the joint is Boolean conjunction $j=A\land B$,
fixed by the two truth values.  With $e=B$ the equation $ce=j$ gives:
\begin{itemize}[nosep]
\item $B=1$: then $c\cdot1=j=A$, so $c=A$.
\item $B=0$: then $c\cdot0=0$ holds for $c=0$ and $c=1$, so $c$ is
unconstrained.
\end{itemize}
\begin{center}
\begin{tabular}{@{}cccc@{}}
\toprule
$A$ & $B=e$ & $j=A\land B$ & admissible $c=(A\mid B)$\\
\midrule
$0$ & $0$ & $0$ & $0$ or $1$\\
$1$ & $0$ & $0$ & $0$ or $1$\\
$0$ & $1$ & $0$ & $0$\\
$1$ & $1$ & $1$ & $1$\\
\bottomrule
\end{tabular}
\end{center}
True evidence fixes the conditional to the value of $A$; false evidence
leaves it free.  This is the binary form of no ex falso.  The Boolean
carrier carries no dependence: $A$ and $B$ already determine $j$, so two
true statements cannot be marked independent, correlated, or irrelevant.

\subsection{Continuous}
On $[0,1]$ the marginals no longer fix the joint.  For values $a,b$ the
admissible joint ranges over the Fr\'echet interval
\[
  \max(a+b-1,0)\le j\le\min(a,b),
\]
which is the dependence freedom.  At $a=0.6$, $b=0.7$ it is
$0.3\le j\le0.6$.  Independence is $j=ab=0.42$, giving the conditional
$j/b=0.6$: learning $B$ does not move $A$.  Maximum positive dependence
is $j=\min(a,b)=0.6$, giving $0.6/0.7\approx0.857$: $B$ supports $A$.
Minimal overlap is $j=0.3$.  The supplied joint selects the dependence;
the marginal values alone do not.

\subsection{The chain rule reads both ways}
Standard conditional probability computes the conditional from the
joint: $j,e\Rightarrow c=j/e$ for $e>0$.  The same equation also
computes the joint from a conditional taken as primitive:
$c,e\Rightarrow j=ce$.  With $c=0.8$, $e=0.5$ the joint is $0.4$.  So
the conditional is first-class: it need not be defined as a quotient.
It is not self-sufficient, since $j=ce$ still needs the evidence.  The
same $c=0.8$ with $e=0.1$ gives $j=0.08$.

The direction that division loses is the singular one:
$e=0$, $j=0\Rightarrow c\in[0,1]$.  Division writes $0/0$; a residuated
implication defaults to $1$; the chain rule records the rank loss as the
whole fiber, by the admissible-set trichotomy.  The constraint
$ce=j$ is therefore read in both useful directions,
$j,e\Rightarrow c$ and $c,e\Rightarrow j$, which makes the relation more
symmetric than the ratio definition while keeping the joint as supplied
dependence data.

\section{A Running Computation}

Let $a=\cred(A)$ and $b=\cred(B)$.  Suppose the joint is the
Fr\'echet upper bound
\[
  j=\min(a,b).
\]
For $b>0$ the conditional is
\[
  U(a,b)=\frac{\min(a,b)}{b}
  =
  \begin{cases}
  a/b, & a<b,\\
  1, & b\le a.
  \end{cases}
\]
This is the min-copula conditional.  It is not the only possible
conditional.  It is a clean test case because it turns dependence into
a visible formula.

Two interior pairs show the problem:
\[
  U(0.3,0.7)=\frac37,\qquad
  U(0.7,0.3)=1.
\]
Both inputs collapse to the same qualitative pair
$(\half,\half)$, but their conditionals collapse to different values.

\section{The Kleene Collapse}

Define
\[
\collapse(c)=
\begin{cases}
0,&c=0,\\
1,&c=1,\\
\half,&0<c<1.
\end{cases}
\]
The image is the Kleene lattice $\three=\{0,\half,1\}$.
On $\three$, negation fixes $\half$, and conjunction/disjunction
become min/max.

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=1.4cm]
  \node[circle,draw,minimum size=8mm] (z) {$0$};
  \node[circle,draw,minimum size=8mm,right=of z] (h) {$\half$};
  \node[circle,draw,minimum size=8mm,right=of h] (o) {$1$};
  \draw[thick] (z) -- node[below] {interior class} (h) -- (o);
  \draw[->,bend left=40,red] (z) to node[above] {$\negc$} (o);
  \draw[->,bend left=40,red] (o) to node[below] {$\negc$} (z);
  \draw[->,loop above,red] (h) to (h);
  \node[below=1.0cm of h,align=center] {$\collapse^{-1}(0)=\{0\}$,\quad
  $\collapse^{-1}(\half)=(0,1)$,\quad
  $\collapse^{-1}(1)=\{1\}$};
\end{tikzpicture}
\caption{The Kleene collapse keeps the two boundaries and identifies
all interior values.}
\label{fig:collapse}
\end{figure}

\begin{theorem}[Collapse homomorphism]
\label{thm:collapse}
For all $a,b\in[0,1]$,
\[
  \collapse(\negc a)=\negc\collapse(a),\qquad
  \collapse(a\conjc b)=\collapse(a)\wedge\collapse(b),\qquad
  \collapse(a\disjc b)=\collapse(a)\vee\collapse(b).
\]
\end{theorem}

\noindent
Lean: \Lean{Cred.collapse\_neg},
\Lean{Cred.collapse\_conj},
\Lean{Cred.collapse\_disj}.

This collapse is rigid.

\begin{theorem}[Three-element quotient]
\label{thm:three-quotient}
Any surjective map $\phi:[0,1]\to\three$ that preserves complement
and product sends $\half$ to $\half$ and, up to swapping $0$ and $1$,
sends the boundary values to the boundary values.  The Kleene quotient
is therefore the unique non-trivial three-element quotient of this
signature.
\end{theorem}

\noindent
Lean: \Lean{Cred.three\_element\_quotient\_unique}.

A sharper rigidity result holds if multiplication is allowed by
arbitrary real scalars rather than only by unit-interval values.

\begin{theorem}[No finite real-multiplicative quotient]
\label{thm:real-quotient}
No non-trivial finite quotient compatible with complement and
unrestricted real multiplication exists.
\end{theorem}

\noindent
Lean: \Lean{Cred.RealCongruence.no\_nontrivial\_finite\_quotient}.

The proof is a scaling argument.  A finite quotient makes powers of
$1/2$ cycle.  Unrestricted multiplication cancels the earlier power,
forcing $1\sim(1/2)^p$.  Complement then gives
$0\sim 1-(1/2)^p$, a positive value.  That collapses every value to
$0$.

\section{Consequence: Positivity and Certainty}

Let formulas be built from atoms using $\negc,\conjc,\disjc$.
For a valuation $v:\alpha\to[0,1]$, write $\val{\varphi}_v$ for
evaluation.

There are two natural designation thresholds:
\[
  \text{positive: }\val{\varphi}_v>0,\qquad
  \text{certain: }\val{\varphi}_v=1.
\]
LP designates $\half$ and $1$; K$_3$ designates only $1$.

\begin{theorem}[Formula bridge]
\label{thm:formula-bridge}
For every list of premises $\Gamma$ and conclusion $\varphi$:
\[
\Gamma\models_{\mathrm{LP}}\varphi
\quad\Longleftrightarrow\quad
\forall v\,
\bigl((\forall\gamma\in\Gamma,\ \val{\gamma}_v>0)
\Rightarrow \val{\varphi}_v>0\bigr),
\]
and
\[
\Gamma\models_{\mathrm{K3}}\varphi
\quad\Longleftrightarrow\quad
\forall v\,
\bigl((\forall\gamma\in\Gamma,\ \val{\gamma}_v=1)
\Rightarrow \val{\varphi}_v=1\bigr).
\]
\end{theorem}

\noindent
Lean: \Lean{Cred.lp\_formula\_bridge},
\Lean{Cred.k3\_formula\_bridge}.

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=2.0cm,>=Latex]
  \node[draw,rounded corners,align=center] (r) {$[0,1]$ formulas\\
  $\negc,\conjc,\disjc$};
  \node[draw,rounded corners,align=center,right=of r] (k) {Kleene formulas\\
  min/max};
  \node[draw,rounded corners,align=center,below=of r] (p) {positivity\\
  $>0$};
  \node[draw,rounded corners,align=center,below=of k] (lp) {LP designation\\
  $\{\half,1\}$};
  \node[draw,rounded corners,align=center,below=1.2cm of p] (c) {certainty\\
  $=1$};
  \node[draw,rounded corners,align=center,below=1.2cm of lp] (k3) {K$_3$ designation\\
  $\{1\}$};
  \draw[->,thick] (r) -- node[above] {$\collapse$} (k);
  \draw[<->,thick] (p) -- node[above] {LP bridge} (lp);
  \draw[<->,thick] (c) -- node[above] {K$_3$ bridge} (k3);
  \draw[->] (r) -- (p);
  \draw[->] (k) -- (lp);
  \draw[->] (p) -- (c);
  \draw[->] (lp) -- (k3);
\end{tikzpicture}
\caption{The propositional bridge.  Collapse commutes with formula
evaluation, and the two designation policies become positivity and
certainty on the real interval.}
\label{fig:bridge}
\end{figure}

Classical logic is still present.  Restrict valuations to $\{0,1\}$.
Then positivity and certainty coincide with ordinary Boolean truth.
\begin{quote}\scriptsize
\begin{tabular}{@{}l@{}}
Lean anchors: \\
\Lean{Cred.crisp\_embedding}\\
\Lean{Cred.crisp\_certainty\_iff\_classical}\\
\Lean{Cred.crisp\_positivity\_iff\_classical}
\end{tabular}
\end{quote}

\section{Inference as Inverse-Image Inclusion}
\label{sec:inference-principle}

Conditioning and consequence are one construction.  Each fixes
constraints, forms the set of points those constraints admit, and
accepts a conclusion exactly when it holds across the whole set.

Fix the value object $([0,1],\conjc,\le)$ and read a threshold $t$ as
the up-set ${\uparrow}t=[t,1]$.  A \emph{constraint} on a point set $X$
is a pair $(f,S)$ with $f:X\to[0,1]$ and $S\subseteq[0,1]$: it holds at a
point $x$ when the score $f(x)$ lands in the allowed set $S$, so its
admissible set is the preimage $f^{-1}(S)$.  For a family $\Gamma$ of
constraints, the admissible set is the intersection
$\mathrm{Mod}(\Gamma)=\bigcap_{(f,S)\in\Gamma}f^{-1}(S)$, the points that
meet every constraint.  Inference is inclusion of admissible sets:
\[
\boxed{\;\Gamma\vdash(g,T)\quad:\Longleftrightarrow\quad
  \mathrm{Mod}(\Gamma)\subseteq g^{-1}(T).\;}
\]
Adding premises shrinks the admissible set, and the conclusions that
hold across it grow in step; this antitone pairing between constraints
and points is a Galois connection, and its closure is the ordinary
shape of semantic consequence.  Cred's two
inferences are its valuation-level and value-level instances, and the
two words are not synonyms.  A \emph{valuation} is a whole assignment of
values to the atoms, one possible world; a \emph{value} is a single
number.  Entailment ranges over worlds: it asks what holds in every
valuation that designates the premises.  Conditioning fixes the two
numbers $j$ and $e$ and ranges over the candidate values $c$ of one
conditional.  Same operator, different point space.
\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
 & points $X$ & evaluation $f$ & target $S$ \\
\midrule
consequence $\Gamma\models_t\varphi$ & valuations $v$ &
  $v\mapsto\val{\varphi}_v$ & ${\uparrow}t$ \\
conditioning $\Cond(j,e)$ & values $c\in[0,1]$ &
  $c\mapsto c\conjc e$ & $\{j\}$ \\
\bottomrule
\end{tabular}
\end{center}
For consequence the admissible set is
$\bigcap_{\gamma\in\Gamma}\val{\gamma}^{-1}({\uparrow}t)$ and the query
is $\val{\varphi}^{-1}({\uparrow}t)$; this is the formula bridge and
threshold consequence.  For conditioning the admissible set is the
fiber $m_e^{-1}(j)=\Cond(j,e)$ of the chain rule, and a threshold query
$c\ge t$ is warranted exactly when $\Cond(j,e)\subseteq{\uparrow}t$,
which is the robust and dependence-sensitive split below.  The value
carrier plays both roles: it is the codomain of every evaluation and,
for conditioning, the point set itself.

\begin{figure}[h]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[>=Latex]
  \node[draw,rounded corners,fill=gray!10,align=center] (op)
    {one operator\\$\mathrm{Mod}(\Gamma)\subseteq g^{-1}(T)$};
  \node[draw,rounded corners,fill=blue!8,align=center,
        below left=12mm and 2mm of op] (ent)
    {consequence\\points: valuations $v$ (whole worlds)\\$f=\val{\cdot}_v$,\quad $S={\uparrow}t$};
  \node[draw,rounded corners,fill=red!8,align=center,
        below right=12mm and 2mm of op] (con)
    {conditioning\\points: values $c$ (numbers; $j,e$ fixed)\\$f=(\,\cdot\,)\conjc e$,\quad $S=\{j\}$};
  \draw[->] (op) -- (ent);
  \draw[->] (op) -- (con);
  \draw[<->,thick,gray] (con) -- node[below,align=center,font=\footnotesize]
    {chain-rule cut} (ent);
\end{tikzpicture}%
}
\caption{Two fibrations of one consequence operator.  Entailment runs
over valuations (whole worlds), conditioning over values (single
numbers); the chain-rule cut pushes a
value fiber forward into the valuation consequence, turning evidence
above $t$ and a conditional at least $\ell$ into a joint above
$\ell\conjc t$.}
\label{fig:two-fibrations}
\end{figure}

The three shapes of an admissible set are the same on both sides.  A
singleton forces one answer: positive evidence $c=j/e$, or premises
that pin the conclusion.  The full set forces nothing: zero evidence
gives $\Cond(0,0)=[0,1]$, and a glut leaves a surviving counter-valuation.
The empty set is incoherence: $j>e$ on the value side, no countermodel on
the valuation side, which is reductio.  No explosion is then not an extra
axiom.  Classical ex falso is the step ``the admissible set is empty,
therefore everything''; the principle declines it by reporting the empty
and full sets faithfully instead of collapsing either to anything.

One instance does not reduce to the other.  The conditioning evaluation
$m_e=(\,\cdot\,)\conjc e$ is a genuine map on values, but the next
section shows it does not descend to a connective on the collapsed
values: no truth function on $\{0,\half,1\}$ reproduces it.  The two
fibrations share a closure operator while the value fibration resists
internalization as an object-language conditional.  Unification at the
level of consequence; separation at the level of connectives.

\subsection{Probability over valuations: $\forall$ as a measure}
\label{sec:measure-layer}

Inclusion is itself a special case.  Reading
$\mathrm{Mod}(\Gamma)\subseteq\mathrm{Mod}(\varphi)$ as ``the conclusion
holds at every premise point'' is a yes/no verdict.  Put a finite
measure $\mu$ on the point space and grade it:
\[
  P_\mu(\varphi\mid\Gamma)
  =\frac{\mu\bigl(\mathrm{Mod}(\Gamma)\cap\mathrm{Mod}(\varphi)\bigr)}
        {\mu\bigl(\mathrm{Mod}(\Gamma)\bigr)}\in[0,1].
\]
The aggregation operator chooses the layer.  One measure $\mu$ makes
$P_\mu(\varphi\mid\Gamma)$ ordinary conditional probability over the
valuation space.  A set $\mathbb{M}$ of measures makes
$\inf_{\mu\in\mathbb{M}}P_\mu(\varphi\mid\Gamma)$ a lower probability
(imprecise probability).  Quantifying over \emph{all} measures recovers
logic: the universal quantifier is the all-measures corner, and
entailment is conditional probability one under every measure,
\[
  \mathrm{Mod}(\Gamma)\subseteq\mathrm{Mod}(\varphi)
  \iff
  \forall\mu,\ \mu(\mathrm{Mod}(\Gamma))>0\Rightarrow
  P_\mu(\varphi\mid\Gamma)=1,
\]
equivalently $\mathrm{Mod}(\Gamma)\subseteq\mathrm{Mod}(\varphi)\iff
\forall\mu,\ P_\mu(\mathrm{Mod}(\Gamma))\le P_\mu(\mathrm{Mod}(\varphi))$
(\Lean{entails\_iff\_cond\_one}, \Lean{entails\_iff\_dominated} in
\texttt{Cred.Probability}, over a finite valuation space).

The chain rule reappears as the aggregate identity
\[
  P_\mu(\varphi\mid\Gamma)\cdot\mu(\mathrm{Mod}(\Gamma))
  =\mu\bigl(\mathrm{Mod}(\Gamma)\cap\mathrm{Mod}(\varphi)\bigr),
\]
which is the value equation $c\conjc e=j$ with $c=P_\mu(\varphi\mid\Gamma)$,
$e=\mu(\mathrm{Mod}(\Gamma))$, and joint
$j=\mu(\mathrm{Mod}(\Gamma)\cap\mathrm{Mod}(\varphi))$
(\Lean{cond\_chain\_rule}); zero-measure evidence leaves it unconstrained
(\Lean{zero\_evidence\_unconstrained}), the measure-layer image of
$\Cond(0,0)=[0,1]$.

\begin{figure}[h]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[>=Latex,node distance=6mm]
  \node[draw,rounded corners,fill=gray!10,align=center] (agg)
    {conditional aggregate\\$P_\mu(\varphi\mid\Gamma)
      =\mu(\mathrm{Mod}\Gamma\cap\mathrm{Mod}\varphi)/\mu(\mathrm{Mod}\Gamma)$};
  \node[draw,rounded corners,fill=red!8,align=center,below=14mm of agg] (prob)
    {one measure $\mu$\\probability $P_\mu(\varphi\mid\Gamma)\in[0,1]$};
  \node[draw,rounded corners,fill=orange!12,align=center,left=12mm of prob] (imp)
    {set $\mathbb{M}$ of measures\\$\inf_{\mathbb{M}}$: lower probability};
  \node[draw,rounded corners,fill=blue!8,align=center,right=12mm of prob] (log)
    {all measures, $=1$\\$\mathrm{Mod}\Gamma\subseteq\mathrm{Mod}\varphi$: entailment};
  \draw[->] (agg) -- (prob);
  \draw[->] (agg) -| (imp);
  \draw[->] (agg) -| (log);
\end{tikzpicture}%
}
\caption{One conditional aggregate, three layers.  A single measure
gives probability; a credal set gives lower (imprecise) probability;
quantifying over all measures with verdict $1$ gives logical entailment.
The universal quantifier is the all-measures corner of the measure.}
\label{fig:three-layers}
\end{figure}

Probability proper is therefore values plus a measure on valuations, and
that measure does two jobs at once: it grades consequence, and it
\emph{determines} the joint $\mu(A\cap B)$ that the scalar layer leaves
supplied.  Three layers, one operator: aggregate the conclusion over the
premises' admissible set; $\forall$ is logic, $\int\mu$ is probability,
$\inf_{\mathbb{M}}$ is imprecise probability, and the chain rule
$c\conjc e=j$ relates the aggregates throughout.

\section{Why the Conditional Does Not Collapse}
\label{sec:no-cond-bridge}

Consequence is the inclusion $\mathrm{Mod}(\Gamma)\subseteq
\val{\varphi}^{-1}({\uparrow}t)$ of the previous section, and it
descends to the Kleene quotient.  Conditioning is the same inclusion on
the value fibration, and the next theorem shows it does \emph{not}
descend: the evaluation $m_e$ is no truth function on the collapsed
values.

\begin{theorem}[No truth-functional conditional bridge]
\label{thm:no-cond-bridge}
There is no function
$f:\three\times\three\to\three$ such that for every $a,b\in[0,1]$
with $b>0$,
\[
  \collapse\!\left(\frac{\min(a,b)}{b}\right)
  =f(\collapse(b),\collapse(a)).
\]
\end{theorem}

\noindent
Lean: \Lean{Cred.no\_truthfunctional\_cond\_bridge}.

\begin{proof}
Take $(a_1,b_1)=(0.3,0.7)$ and $(a_2,b_2)=(0.7,0.3)$.
All four inputs are interior, so
\[
  (\collapse(b_1),\collapse(a_1))
  =(\half,\half)
  =(\collapse(b_2),\collapse(a_2)).
\]
But
\[
  \collapse\!\left(\frac{\min(0.3,0.7)}{0.7}\right)
  =\collapse(3/7)=\half,
\]
while
\[
  \collapse\!\left(\frac{\min(0.7,0.3)}{0.3}\right)
  =\collapse(1)=1.
\]
Thus $f(\half,\half)$ would have to be both $\half$ and $1$.
\end{proof}

\begin{figure}[h]
\centering
\begin{tikzpicture}[>=Latex,node distance=2.2cm]
  \node[draw,rounded corners,align=center] (a) {$(0.3,0.7)$\\
  $U=3/7$};
  \node[draw,rounded corners,align=center,right=of a] (b) {$(0.7,0.3)$\\
  $U=1$};
  \node[draw,rounded corners,align=center,below=1.2cm of a] (ca) {collapsed input\\
  $(\half,\half)$};
  \node[draw,rounded corners,align=center,below=1.2cm of b] (cb) {collapsed input\\
  $(\half,\half)$};
  \node[draw,rounded corners,align=center,below=1.2cm of ca] (oa) {output\\
  $\half$};
  \node[draw,rounded corners,align=center,below=1.2cm of cb] (ob) {output\\
  $1$};
  \draw[->,thick] (a) -- node[left] {$\collapse$} (ca);
  \draw[->,thick] (b) -- node[right] {$\collapse$} (cb);
  \draw[->,thick] (ca) -- (oa);
  \draw[->,thick] (cb) -- (ob);
  \draw[red,thick] (oa) -- node[below] {same input, different output} (ob);
\end{tikzpicture}
\caption{The obstruction to a truth-functional conditional on the
collapsed values.}
\label{fig:no-conditional}
\end{figure}

The obstruction is a small Lewis-style triviality result
\cite{lewis1976,vanfraassen1976,adams1975,mcgee1985}.  It differs
from the classical probability-of-conditionals problem in one respect:
the failure is visible in a finite quotient.  The quotient has thrown
away the ratio $a/b$, and no truth table can recover it.

The bridge does hold at boundaries:
\[
  a=0,\quad a=1,\quad b=1,\quad b\le a.
\]
Lean: \Lean{Cred.cond\_bridge\_boundary}.  These are exactly the
cases where the min-copula conditional lands on a boundary value or
where evidence is certain.

\section{Zero Evidence}

At zero evidence the chain rule is
\[
  c\cdot0=0.
\]
The equation holds for every $c$.  Three consequences share this one
root.

\begin{theorem}[Zero-evidence triple]
\label{thm:zero-triple}
The identity $c\cdot0=0$ yields:
\begin{enumerate}[label=(\roman*),nosep]
\item $\Cond(0,0)=[0,1]$;
\item LP explosion has a countermodel, witnessed by $\half$;
\item no collapsed truth function can recover the conditional value at
zero evidence.
\end{enumerate}
\end{theorem}

\noindent
Lean: \Lean{Cred.zero\_evidence\_duality},
\Lean{Cred.zero\_evidence\_duality\_cond}.

At zero evidence the constraint loses rank, and no value is selected.
Impossible evidence is not false evidence plus an implication rule.  One can add a convention after the fact, for example
``choose $1$'' or ``choose $\half$'', but the chain rule itself has no
reason to choose.

This locates the framework relative to imprecise probability
\cite{walley1991,troffaes2014,debock2015,wheeler2021}.  A set of
possible posteriors appears when the supplied constraint fails to
determine a point.  The set is not psychological indecision.  It is the
fiber of an equation.

Unknown dependence is a second source of a set, and it does not need
zero evidence.  Suppose the marginals $a,b$ are given but the joint is
not.  Coherence confines the joint to the Fr\'echet interval
$[\max(a+b-1,0),\,\min(a,b)]$, so for $b>0$ the admissible conditionals
are the image
\[
  \left\{\frac{j}{b}:\max(a+b-1,0)\le j\le\min(a,b)\right\}
  =\left[\frac{\max(a+b-1,0)}{b},\,\frac{\min(a,b)}{b}\right].
\]
A conclusion at a demanded threshold holds for every admissible joint
when this whole interval lies above the threshold, and is
dependence-sensitive when the interval straddles it.  The no-go theorem
of the previous section shows the collapse to $\three$ discards the
ratio $a/b$, so the order of work is fixed: carry the joint, or its
interval, check it against the threshold, and collapse only the
conclusions the quotient cannot distort.

\section{Self-Reference as Equations}

Let $\Phi:[0,1]\to[0,1]$.  A self-referential sentence governed by
$c=\Phi(c)$ has meaning
\[
  \Sol(\Phi)=\{c\in[0,1]:\Phi(c)=c\}.
\]

\begin{proposition}[Basic solution sets]
\label{prop:solution-sets}
\[
  \Sol(\negc)=\{\half\},\qquad
  \Sol(\mathrm{id})=[0,1],\qquad
  \Sol(\mathrm{id})=\Cond(0,0).
\]
\end{proposition}

\noindent
\begin{quote}\scriptsize
\begin{tabular}{@{}l@{}}
Lean anchors: \\
\Lean{Cred.Credence.solutions\_neg\_eq\_singleton}\\
\Lean{Cred.Credence.solutions\_id\_eq\_univ}\\
\Lean{Cred.Credence.truth\_teller\_same\_shape\_as\_zero\_evidence}
\end{tabular}
\end{quote}

The liar and the scalar Russell equation have the same form:
\[
  c=1-c.
\]
The unique solution is $c=\half$.  Classical logic sees no Boolean
solution.  The credence algebra sees one fixed point.

The truth-teller is
\[
  c=c.
\]
Every value solves it.  Its solution set is the same as zero-evidence
conditioning.  That is not a coincidence in the formal system: both
are unconstrained equations.

\section{Why the Conditional Must Stay External}

One might try to internalize conditioning as a binary connective
$a\Rightarrow b$.  Product logic does this by residuation
\cite{hajek1998,esteva2001,galatos2007}:
\[
  a\Rightarrow_{\Pi} b=
  \begin{cases}
    1, & a=0,\\
    \min(1,b/a), & a>0.
  \end{cases}
\]
This operation is forced by modus ponens plus conditional proof.
Let
\[
\begin{aligned}
\mathrm{MP}(f)&:\quad f(a,b)\conjc a\le b,\\
\mathrm{CP}(f)&:\quad c\conjc a\le b\Rightarrow c\le f(a,b),\\
\mathrm{Contraction}(f)&:\quad f(a,f(a,b))\le f(a,b).
\end{aligned}
\]

\begin{theorem}[Curry block]
\label{thm:curry-block}
No total function $f:[0,1]^2\to[0,1]$ satisfies
\[
  \mathrm{MP}(f)\ \wedge\ \mathrm{CP}(f)\ \wedge\
  \mathrm{Contraction}(f).
\]
\end{theorem}

\noindent
Lean: \Lean{Cred.Credence.curry\_block}.

\begin{proof}
MP and CP give full residuation.  Hence for $a>0$,
\[
  f(a,b)=\min(1,b/a).
\]
Choose $a=\half$ and $b=\quarter$.  Then
\[
  f(\half,\quarter)=\half,\qquad
  f(\half,f(\half,\quarter))=f(\half,\half)=1.
\]
Contraction would require $1\le\half$, contradiction.
\end{proof}

This explains the design decision.  The object language has negation,
conjunction, and disjunction.  Conditioning is a side judgment, a
constraint on supplied joint and evidence values.  Adding a total
internal conditional restores the very structural package that creates
Curry trouble.

\section{Threshold Consequence}

For $t\in[0,1]$, define
\[
  \Gamma\models_t\varphi
  \quad\Longleftrightarrow\quad
  \forall v\,
  \left(
  \forall\gamma\in\Gamma,\ t\le\val{\gamma}_v
  \right)
  \Rightarrow
  t\le\val{\varphi}_v.
\]
This consequence relation has reflexivity, monotonicity, and cut.
\begin{quote}\scriptsize
\begin{tabular}{@{}l@{}}
Lean anchors: \\
\Lean{Cred.threshold\_reflexivity}\\
\Lean{Cred.threshold\_monotonicity}\\
\Lean{Cred.threshold\_cut}
\end{tabular}
\end{quote}

Two thresholds are sharp:
\[
\text{explosion countermodel exists}
\quad\Longleftrightarrow\quad
0<t\le\half,
\]
and
\[
\text{excluded middle holds at threshold }t
\quad\Longleftrightarrow\quad
t\le\frac34.
\]
\begin{quote}\scriptsize
\begin{tabular}{@{}l@{}}
Lean anchors: \\
\Lean{Cred.threshold\_explosion\_countermodel\_iff}\\
\Lean{Cred.threshold\_excluded\_middle\_iff}
\end{tabular}
\end{quote}

\begin{figure}[h]
\centering
\begin{tikzpicture}[x=11cm,y=1cm]
  \draw[->] (0,0) -- (1.05,0) node[right] {$t$};
  \draw[very thick,blue] (0,0) -- (0.5,0);
  \draw[very thick,green!60!black] (0,0.32) -- (0.75,0.32);
  \draw (0.5,0.08) -- (0.5,-0.08) node[below] {$\half$};
  \draw (0.75,0.40) -- (0.75,0.24) node[below] {$\frac34$};
  \draw (1,0.08) -- (1,-0.08) node[below] {$1$};
  \node[above] at (0.25,0.05) {explosion countermodels};
  \node[above] at (0.38,0.37) {excluded middle};
  \node[align=center] at (0.625,-0.45) {non-explosive\\and excluded-middle};
\end{tikzpicture}
\caption{Sharp threshold facts.  For $\half<t\le 3/4$, explosion has
no threshold countermodel while excluded middle still holds.}
\label{fig:thresholds}
\end{figure}

The interval $\half<t\le3/4$ is the non-explosive band that still
validates excluded middle: contradiction licenses nothing, yet excluded
middle stays designated.  It is a status calculus for graded
consequence, not a foundation for mathematics.

\section{Toy Models and Worked Examples}
\label{sec:toy-models}

The architecture is finite and checkable.  Each subsection below is a
Lean example module with rational or Boolean data and concrete theorems.

\subsection{Pointwise truth is not connection}
Worlds $W=\{w_0,w_1,w_2,w_3\}$ carry Boolean propositions
$A,B,C:W\to\{0,1\}$.
\begin{center}
\begin{tabular}{@{}ccccc@{}}
\toprule
world & $A$ & $B$ & $C$ & $A\wedge B$\\
\midrule
$w_0$ & $1$ & $1$ & $1$ & $1$\\
$w_1$ & $1$ & $0$ & $1$ & $0$\\
$w_2$ & $0$ & $1$ & $1$ & $0$\\
$w_3$ & $0$ & $0$ & $1$ & $0$\\
\bottomrule
\end{tabular}
\end{center}
Entailment is truth-set inclusion,
$A\models B\iff\{w:A_w=1\}\subseteq\{w:B_w=1\}$
(\Lean{Cred.Examples.FiniteWorlds.entails\_iff\_subset}).
At $w_0$ all three are true, yet $A\not\models B$ because $w_1$ is a
countermodel (\Lean{Cred.Examples.FiniteWorlds.A\_not\_entails\_B}),
while $A\models C$
(\Lean{Cred.Examples.FiniteWorlds.A\_entails\_C}); pointwise agreement
does not fix the global relation
(\Lean{Cred.Examples.FiniteWorlds.pointwise\_does\_not\_determine}).
Dependence is invisible pointwise too.  The weightings
$\mu_1=(\tfrac14,\tfrac14,\tfrac14,\tfrac14)$ and
$\mu_2=(\tfrac12,0,0,\tfrac12)$ give equal marginals
$P(A)=P(B)=\tfrac12$ but different joints
$P(A\wedge B)=\tfrac14$ and $\tfrac12$
(\Lean{Cred.Examples.FiniteWorlds.joint\_not\_determined\_by\_marginals}).

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1]
  \foreach \x/\name in {0/w_0, 2/w_1, 4/w_2, 6/w_3} {
    \node at (\x,2.3) {$\name$};
  }
  % C: all worlds
  \draw[rounded corners,draw=green!55!black,thick,fill=green!8]
    (-0.7,-0.5) rectangle (6.7,0.55);
  \node[green!45!black,anchor=west] at (6.8,0) {$C=\{w_0,w_1,w_2,w_3\}$};
  % A: w0,w1
  \draw[rounded corners,draw=blue!70,thick,fill=blue!12]
    (-0.55,0.7) rectangle (2.55,1.7);
  \node[blue!70,anchor=west] at (6.8,1.2) {$A=\{w_0,w_1\}$};
  % B: w0,w2
  \draw[rounded corners,draw=red!70,thick,fill=red!10]
    (-0.45,1.85) rectangle (4.45,2.85);
  \node[red!70,anchor=west] at (6.8,2.35) {$B=\{w_0,w_2\}$};
  \foreach \x/\name in {0/w_0, 2/w_1, 4/w_2, 6/w_3} {
    \fill (\x,0) circle (0.06);
    \node[below] at (\x,-0.1) {$\name$};
  }
\end{tikzpicture}
\caption{Truth sets in $W=\{w_0,\dots,w_3\}$ for $A=[1,1,0,0]$,
$B=[1,0,1,0]$, $C=[1,1,1,1]$.  $A\subseteq C$, so $A\models C$, but
$A\not\subseteq B$: world $w_1\in A\setminus B$ is the countermodel that
breaks $A\models B$ even though both hold at $w_0$.}
\label{fig:finite-worlds}
\end{figure}

\subsection{Robust conditioning from a Fr\'echet interval}
Take $a=\tfrac35$, $b=\tfrac{7}{10}$; the example anchors are in
\texttt{Cred.Examples.RobustConditioning}.  The joint is confined to
$[\tfrac{3}{10},\tfrac35]$ (\Lean{frechet\_lower},
\Lean{frechet\_upper}); the independence joint is $\tfrac{21}{50}$
(\Lean{product\_joint}).  For $b>0$ the admissible conditional ranges
over the image interval
\[
  \left[\frac{3/10}{7/10},\,\frac{3/5}{7/10}\right]
  =\left[\tfrac37,\tfrac67\right]
\]
(\Lean{cond\_interval}; the general image map is
\Lean{cond\_interval\_of\_pos} in \texttt{Cred.Dependence}).  The whole
interval clears the threshold $\tfrac13$, so that conclusion holds for
every admissible joint (\Lean{robust\_at\_low\_threshold}); the threshold
$\tfrac45$ lands inside it, so that conclusion is dependence-sensitive
(\Lean{sensitive\_at\_high\_threshold}).  The three-valued summary is
\Lean{collapseIntervalToThree} into \Lean{RobustStatus} (both in
\texttt{Cred.Dependence}), recording robustness or sensitivity.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1,x=11cm]
  % joint axis: from 0 to 1, but show the relevant window 0.25..0.65
  % map joint value j to x via x=j
  \def\jlo{0.3}   % 3/10
  \def\jprod{0.42} % 21/50
  \def\jhi{0.6}   % 3/5
  % joint number line
  \draw[->] (0.22,0) -- (0.7,0) node[right] {joint $j=\cred(A\wedge B)$};
  \draw[very thick,blue!70] (\jlo,0) -- (\jhi,0);
  \foreach \x in {\jlo,\jprod,\jhi} { \fill (\x,0) circle (0.6pt); }
  \node[below=4pt] at (\jlo,0) {$\tfrac{3}{10}$};
  \node[below=4pt] at (\jprod,0) {$\tfrac{21}{50}$};
  \node[below=4pt] at (\jhi,0) {$\tfrac{3}{5}$};
  \node[above=2pt,align=center] at (\jlo,0.05) {\footnotesize\L uk.};
  \node[above=2pt,align=center] at (\jprod,0.05) {\footnotesize product};
  \node[above=2pt,align=center] at (\jhi,0.05) {\footnotesize min};
  \node[blue!70,above=16pt] at (\jprod,0) {admissible joint $[\tfrac{3}{10},\tfrac35]$};
  % conditional axis below: cond = j/(7/10)
  \draw[->] (0.22,-1.3) -- (0.7,-1.3) node[right] {conditional $j/b$};
  \draw[very thick,red!70] ({3/7*0.7},-1.3) -- ({6/7*0.7},-1.3);
  \foreach \v in {3,6} { \fill ({\v/7*0.7},-1.3) circle (0.6pt); }
  \node[below=4pt] at ({3/7*0.7},-1.3) {$\tfrac{3}{7}$};
  \node[below=4pt] at ({6/7*0.7},-1.3) {$\tfrac{6}{7}$};
  \node[red!70,above=14pt] at ({4.5/7*0.7},-1.3) {induced $[\tfrac37,\tfrac67]$};
  % divide-by-b arrows
  \draw[->,gray] (\jlo,-0.12) -- ({3/7*0.7},-1.18);
  \draw[->,gray] (\jhi,-0.12) -- ({6/7*0.7},-1.18);
  \node[gray,right] at (0.5,-0.72) {\footnotesize $\div\,b=\tfrac{7}{10}$};
\end{tikzpicture}
\caption{Fr\'echet band and its induced conditional interval for
$a=\tfrac35$, $b=\tfrac{7}{10}$.  The joint is confined to
$[\tfrac{3}{10},\tfrac35]$ with the three named couplings as marked
points.  Dividing by $b$ maps the band onto the conditional interval
$[\tfrac37,\tfrac67]$.}
\label{fig:frechet-interval}
\end{figure}

\subsection{Fuzzy t-norms as fixed dependence policies}
A t-norm used as the joint is one coupling policy, not the general
joint.  With $a=\tfrac35$, $b=\tfrac{7}{10}$ the admissible joint ranges
over the Fr\'echet interval $[\tfrac{3}{10},\tfrac35]$; three named
t-norms pick three points of it.
\begin{center}
\begin{tabular}{@{}lccl@{}}
\toprule
policy (t-norm) & joint $j$ & conditional $j/b$ & dependence \\
\midrule
\L ukasiewicz $\max(a{+}b{-}1,0)$ & $\tfrac{3}{10}$ & $\tfrac37$ & minimal overlap \\
product $a\cdot b$ & $\tfrac{21}{50}$ & $\tfrac35=a$ & independence \\
minimum $\min(a,b)$ & $\tfrac35$ & $\tfrac67$ & maximum positive \\
\bottomrule
\end{tabular}
\end{center}
Lean fixes the three joint values \Lean{frechet\_lower} ($\tfrac{3}{10}$),
\Lean{product\_joint} ($\tfrac{21}{50}$), and \Lean{frechet\_upper}
($\tfrac35$), again in \texttt{Cred.Examples.RobustConditioning}.
Each is a valid coupling, and the product policy makes the evidence
irrelevant ($j/b=a$).  None is the general joint: the supplied
$j=\cred(A\wedge B)$ may be any point of $[\tfrac{3}{10},\tfrac35]$, and
only that supplied value fixes the conditional.

\subsection{Robust three-valued collapse}
An interval verdict collapses to a \Lean{RobustStatus} through
\Lean{collapseIntervalToThree} (\texttt{Cred.Dependence}), which records
whether a conclusion survives every admissible joint.  Four worked
cases, with anchors in \texttt{Cred.Examples.RobustCollapse}:
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
interval & verdict & anchor \\
\midrule
$[\tfrac12,\tfrac12]$ & robust half &
  \Lean{collapse\_interval\_interior\_robust\_half}\\
$[\tfrac37,\tfrac67]$ & dependence-sensitive &
  \Lean{collapse\_interval\_boundary\_sensitive}\\
$[0,1]$ & underdetermined &
  \Lean{zero\_evidence\_collapse\_underdetermined}\\
$[1,0]$ & incoherent &
  \Lean{empty\_fiber\_collapse\_incoherent}\\
\bottomrule
\end{tabular}
\end{center}
The collapse happens after the dependence reasoning, not before: a
ternary status is a summary of an interval of joints, never a place
where dependence is invented.

\subsection{Semantic entailment is not proof use}
A toy proof object carries named assumptions, and \Lean{usedAssumptions}
extracts the set actually used; soundness is \Lean{provenance\_sound}
(\texttt{Cred.ProofTheory}).  A proof of the valid formula
$B\disjc\negc B$ uses no assumption (\Lean{entails\_but\_unused} in
\texttt{Cred.Examples.ProofProvenance}), even though it is valid, so any
premise entails it.  Which theorem depends on which is therefore a
proof-provenance question, not an entailment one.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    >=Stealth,node distance=10mm,
    pnasm/.style={draw,rounded corners,fill=gray!12,inner sep=3pt},
    pnstep/.style={draw,fill=blue!8,inner sep=3pt},
    pnconcl/.style={draw,thick,fill=green!12,inner sep=3pt}]
  \node[pnasm] (A) {assume $A$};
  \node[pnasm,right=16mm of A] (D) {assume $D$};
  \node[pnstep,below=of A] (s1) {$A\disjc B$};
  \node[pnstep,below=of D] (s2) {$D\conjc E$};
  \node[pnconcl,below=14mm of $(s1)!0.5!(s2)$] (g) {$B\disjc\negc B$ (valid)};
  \draw[->] (A) -- (s1);
  \draw[->] (D) -- (s2);
  % conclusion is valid: no edge into it from assumptions
  \node[align=center,right=18mm of g] {valid formula:\\ \emph{no} assumption feeds it};
\end{tikzpicture}
\caption{Proof-provenance DAG.  Edges record which nodes a derivation
actually consumes.  The assumptions $A,D$ feed their own consequences,
but the valid conclusion $B\disjc\negc B$ has no incoming edge: its
\Lean{usedAssumptions} set is empty, so any premise trivially entails it
without being used.}
\label{fig:provenance-dag}
\end{figure}

\subsection{Local contradiction without explosion}
A branch whose assumptions contain both $A$ and $\negc A$ is locally
contradictory, yet the generative calculus, which has no explosion rule,
does not derive an unrelated atom $U$.  The worked witness is
\Lean{local\_contradiction\_no\_explosion\_example}
(\texttt{Cred.Examples.Branches}); the general statement is
\Lean{local\_contradiction\_no\_explosion} (\texttt{Cred.ProofTheory}).
This is the proof-theoretic image of $\Cond(0,0)=[0,1]$.

\subsection{The $\sqrt2$ dependency chain}
The number-theoretic core of the irrationality proof is a benchmark for
hypothetical branching; the anchors are in \texttt{Cred.Math} and
\texttt{Cred.Examples.Sqrt2Branch}.  From coprime $p,q$ with $p^2=2q^2$
and $q\neq0$: $p$ is even (\Lean{sqrt2\_core\_even\_p}), hence $q$ is
even (\Lean{sqrt2\_core\_even\_q}), contradicting coprimality
(\Lean{not\_coprime\_if\_both\_even}); the chain closes at
\Lean{sqrt2\_core\_contradiction}.  The recorded step list
\Lean{sqrt2\_dependency\_chain} localizes the contradiction to the
coprimality step: the false assumption is the rational representation,
and inconsistency enters at one named step rather than spreading by
explosion.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    >=Stealth,
    nd/.style={draw,rounded corners,fill=blue!8,inner sep=3.5pt,
               minimum height=8mm,align=center},
    bad/.style={draw,thick,fill=red!12,inner sep=3.5pt,
                minimum height=8mm,align=center}]
  \node[nd] (r) {$\sqrt2\in\mathbb{Q}$};
  \node[nd,right=8mm of r] (c) {coprime\\$p,q$};
  \node[nd,right=8mm of c] (sq) {$p^2=2q^2$};
  \node[nd,below=8mm of sq] (pe) {$p$ even};
  \node[nd,left=8mm of pe] (qe) {$q$ even};
  \node[bad,left=8mm of qe] (bot) {$\bot$};
  \draw[->] (r) -- (c);
  \draw[->] (c) -- (sq);
  \draw[->] (sq) -- (pe);
  \draw[->] (pe) -- (qe);
  \draw[->,red!70,thick] (qe) -- node[above,font=\footnotesize,red!70]{vs.\ coprime} (bot);
\end{tikzpicture}
\caption{The $\sqrt2$ dependency chain.  Each node depends only on its
predecessor.  The contradiction enters at the single red step where
``$p,q$ both even'' meets the coprimality assumption; the false
assumption is the rational representation $\sqrt2\in\mathbb{Q}$ at the
head of the chain.}
\label{fig:sqrt2-chain}
\end{figure}

\section{Foundational Status}

The real interval is a semantic carrier, not an object-language
ontology.  Using $[0,1]\subset\Real$ supplies order, topology, and
multiplication for the model; the Boolean and three-valued fragments use
$\mathbf 2$ and the Kleene carrier.  The conditioning rule is the same
equation $c\conjc e=j$ on whichever carrier is chosen, and
$\Cond_V(j,e)$ is its fiber in the metatheory, not an object-language
set former.

\section{Proof Theory}

The proof layer has two judgment forms.  The scalar form proves
admissibility:
\[
  \Sigma\seq\Adm(c;j,e).
\]
The formula form proves labelled sequents:
\[
  \Gamma;\Sigma\seq_k\varphi,
\]
where $k$ is positive, certain, or a threshold $t$.

The scalar rules are deliberately small:
\[
\frac{\Adm(c;j,e)\in\Sigma}{\Sigma\seq\Adm(c;j,e)}
\quad
\frac{c\conjc e=j}{\Sigma\seq\Adm(c;j,e)}
\quad
\frac{}{\Sigma\seq\Adm(c;0,0)}
\]
and, for positive evidence,
\[
\frac{0<e\quad j\le e}{\Sigma\seq
  \Adm\!\left((j/e);j,e\right)}.
\]
Weakening and cut are structural:
\[
\frac{\Sigma\seq A\qquad A,\Sigma\seq B}{\Sigma\seq B}.
\]

Lean verifies scalar soundness and closed completeness:
\[
  \seq\Adm(c;j,e)
  \quad\Longleftrightarrow\quad
  c\in\Cond(j,e).
\]
It also verifies the boundary facts:
\[
  \seq\Adm(c;0,0)\quad\text{for every }c,
\]
and
\[
  0<e,\quad
  \seq\Adm(c_1;j,e),\quad
  \seq\Adm(c_2;j,e)
  \quad\Longrightarrow\quad
  c_1=c_2.
\]
\begin{quote}\scriptsize
\begin{tabular}{@{}l@{}}
Lean anchors: \\
\Lean{Cred.Credence.AdmDerivation.closed\_iff\_mem\_cond}\\
\Lean{Cred.Credence.zeroEvidence\_derivable}\\
\Lean{Cred.Credence.closed\_unique\_of\_pos}
\end{tabular}
\end{quote}

The formula calculus keeps conditioning external.  Its chain rule is a
cut between an admissibility side judgment and a formula bound:
\[
\frac{
  \Gamma;\Sigma\seq_t E
  \qquad
  \Sigma\seq\Adm(c;J,E)
  \qquad
  \ell\le c
}{
  \Gamma;\Sigma\seq_{\ell\conjc t} J
}.
\]
There is no rule introducing an implication formula $E\to J$.
Lean proves soundness for the labelled calculus and for kernel
certificates:
\[
  \Gamma;\Sigma\seq_k\varphi
  \quad\Longrightarrow\quad
  \text{every model of }\Gamma,\Sigma\text{ satisfies }k(\varphi).
\]
The cut-free fragment embeds into the full calculus, and cut is
semantically admissible.  Completeness holds for the closed threshold,
certainty, and positivity fragments because the corresponding semantic
consequence relation is available as a proof rule.  This is the exact
proof-theoretic status: sound external conditioning, no internal arrow,
and no ex-falso derivation.

\section{Relation to Existing Work}

Most ingredients here pre-exist.  The product t-norm, the Kleene
lattice, primitive conditional probability, conditioning over fibers,
set-valued probability, and the copula bounds are all established.
Cred's addition is the Lean-checked factorization of value, dependence,
and conditioning into separate layers, with the no-go and uniqueness
results that fix each one.  The section reads as an audit of where the
existing commitments sit, governed by one rule: Cred audits an inference
step, it does not replace the systems below.

The table places Cred against neighbouring traditions along the layers
it factors.  The full positioning, with Lean anchors and overclaim
risks, is in the repository's literature positioning matrix.

\begin{table}[h]
\centering
\resizebox{\textwidth}{!}{%
\footnotesize
\begin{tabular}{@{}lllll@{}}
\toprule
tradition & conjunction / joint & conditional & zero evidence & aggregation \\
\midrule
classical logic & $\wedge$ & material $\supset$ / entailment & vacuous & $\forall$ worlds \\
Kleene / LP / K$_3$ & $\min$ & designation, no internal $\to$ & n/a & $\forall$ + designation \\
product fuzzy logic & product (as joint) & residuum $j/e$ & residuum $=1$ & truth-functional \\
G\"odel / \L ukasiewicz & $\min$ / $\max(a{+}b{-}1,0)$ & residuum & residuum $=1$ & truth-functional \\
probabilistic worlds & event $\cap$ & ratio $j/e$ & undefined $0/0$ & one measure \\
imprecise / credal & joint interval & conditional interval & full interval & set of measures \\
de Finetti conditionals & event $\cap$ & conditional event & void & coherence \\
\textbf{native Cred} & \textbf{supplied $j$} & \textbf{fiber $\Cond(j,e)$} & \textbf{$[0,1]$} & \textbf{$\forall$ / measure} \\
\bottomrule
\end{tabular}%
}
\caption{Cred against neighbouring traditions.  The product is one coupling policy,
not the joint; the conditional is the external fiber, not a residuum; zero evidence
gives the whole interval, not a default value or an undefined ratio; and
$\forall$ consequence is the all-measures corner of the measure aggregation.}
\label{tab:comparison}
\end{table}

\paragraph{Fuzzy and many-valued logic.}
Product logic uses the product t-norm and its residuum
\cite{hajek1998,esteva2001,cintula2011}.  Probabilistic fuzzy logics
and coherent conditional-probability logics also use many-valued
languages to reason about probability operators
\cite{hajek1995,flaminio2009,godo2006}.  More generally, fuzzy logics
choose a carrier and a conjunction-like operation, then study the
induced connectives.  The present system uses that algebraic data for
a different purpose.  It forms the external fiber
$\{c:c\conjc e=j\}$ and propagates that set when it is not a singleton.
It therefore does not identify conditioning with the product residuum.
Explicitly, for evidence $e$ and joint $j$ the residuum returns one
value while the fiber is a set,
\[
\begin{aligned}
  e\to_\Pi j&=
  \begin{cases}1,&e\le j,\\ j/e,&e>j,\end{cases}\\
  \Cond(j,e)&=\{c:c\conjc e=j\}.
\end{aligned}
\]
At positive coherent evidence both give the value $j/e$
(\Lean{residuum\_vs\_fiber\_positive}); at $e=j=0$ the residuum returns
$1$ while $\Cond(0,0)=[0,1]$ (\Lean{residuum\_vs\_fiber\_zero}, in
\texttt{Cred.Native}).  The Curry block shows
why the residuated package cannot be internalized: it is incompatible
with contraction on this carrier.

\paragraph{Paraconsistent and three-valued logic.}
LP and K$_3$ arise through the collapse to
$\{0,\half,1\}$ \cite{kleene1952,asenjo1966,priest1979,belnap1977}.
The new claim is the exact formula-level bridge:
LP is positivity, K$_3$ is certainty.  The bridge is then shown to fail
for conditioning.

\paragraph{Connectives and consequence.}
Non-classical logic already separates an object-language conditional
from a metalanguage consequence relation \cite{priest2008,anderson1975}.
Classical logic can hide the distinction because $A\models B$ coincides
with validity of $A\supset B$.  That coincidence is not structural.  It
fails or changes in many non-classical settings.  Cred uses the
separation as a design rule: conditioning is the external judgment
$\Adm(c;j,e)$, not a new truth-functional connective.  The additional
claim is algebraic.  The external relation is fixed by $c\conjc e=j$,
and its zero-evidence fiber cannot be represented by any collapsed
truth function.

\paragraph{Conditionals.}
The de Finettian line studies trivalent truth conditions and their
proof theory \cite{egre2020part1,egre2021,egre2025}.  This paper takes
a different route.  It does not add a conditional truth table.  It
treats conditional credence as the admissibility relation determined
by the chain rule.

\paragraph{Cox and Jaynes.}
Cox derives the sum and product rules from desiderata on a real
plausibility, and Jaynes builds probability as extended logic on that
base \cite{cox1946,cox1961,jaynes2003}.  Cox's product rule
$p(AB\mid C)=p(A\mid BC)\,p(B\mid C)$ is the chain rule, Cred's closest
classical relative.  Halpern's counterexample shows the derivation needs
a dense domain and smoothness; a finite plausibility can satisfy the
functional equations without being probability \cite{halpern1999}.  Cred
takes that conservative reading as a license.  It keeps the chain rule
$c\conjc e=j$ as primitive, drops the universality and smoothness axioms,
and refuses the division step, so $e=0$ leaves the conditional
unconstrained rather than ill-posed.

\paragraph{Copulas and the dependence bounds.}
Sklar's theorem factors every joint as a copula applied to the margins,
so the copula carries all dependence \cite{sklar1959}.  The
Fr\'echet--Hoeffding bounds $W\le C\le M$ frame the joint between the
countermonotone and comonotone corners, with the product copula $\Pi$ at
independence \cite{nelsen2006,dhaene2002}.  Cred fixes $\conjc$ to $\Pi$
and uses $W$ and $M$ as the admissible range for the supplied joint; the
min copula $M$ is the unique idempotent one, which is why $\min$ is the
boundary case and product the interior case.  When dependence matters,
Cred supplies the joint rather than reading it off one truth function.

\paragraph{Probability and imprecision.}
Standard conditional probability uses
division at positive evidence, and H\'ajek gives reasons not to treat
that ratio as the whole concept \cite{hajek2003conditional}.  R\'enyi
and Popper instead take
conditional probability as primitive
\cite{renyi1955,popper1959,krauss1968,dubins1975}.  Disintegration and
Markov-kernel accounts condition by families of measures over fibers or
level sets \cite{chang1997,fritz2020}.  De Finetti-style conditional
events make the false-antecedent case void, and coherence theory studies
compounds of such conditionals
\cite{definetti1936,gilio2013,gilio2014,flaminio2020conditionals}.
Imprecise probability studies sets of probabilities as first-class
objects \cite{walley1991,troffaes2014}.  Natural and regular extension
handle lower-probability-zero updating, and full conditional
probabilities can force imprecision
\cite{debock2015,wheeler2021}.  These schools already supply most
ingredients: positive-evidence division, primitive conditionals,
conditioning over fibers, void antecedents, and set-valued updating.
The word ``fiber'' is therefore not new here.  Cred puts a smaller
invariant under these ingredients.  The set $\Cond(j,e)$ is the
preimage of one scalar equation.  It becomes wide exactly when evidence
is zero.  Its width records loss of constraint, not subjective
indecision.

\paragraph{Conditional-event algebras.}
Conditional-event algebras are richer object theories: they build
compound conditionals and probabilities over them.  Cred is narrower.
It keeps the scalar value, the supplied joint, and the admissible
fiber.  The same chain-rule fiber works on $[0,1]$, on the Boolean
carrier, and on three-valued variants.

\paragraph{Practical probabilistic and soft-logic systems.}
PSL, MLN, ProbLog, Bayesian networks, and credal networks are deployed
inference and learning engines
\cite{bach2017,richardson2006,deraedt2007,pearl1988,cozman2000}.  Cred
provides none of their machinery: no hinge-loss density, no weight
learning, no weighted model counting, no bound optimization.  It does
not subsume them.  The next subsections audit their commitments in Cred
vocabulary, which locates Cred's design choices against theirs without
any claim of dominance.

\subsection{Old ingredients and the one new layer}
\label{sec:old-vs-new}

Cred reuses standing constructions and adds one thing: the Lean-checked
factorization with its no-go and uniqueness theorems.
\Cref{tab:old-vs-new} keeps the reuse and the addition in separate
columns.  Namespaces
are stated once: anchors live in \texttt{Cred.Credence},
\texttt{Cred.Collapse}, \texttt{Cred.Native}, and \texttt{Cred.Audit}.

\begin{table}[h]
\centering
\resizebox{\textwidth}{!}{%
\footnotesize
\begin{tabular}{@{}p{2.2cm}p{2.9cm}p{2.4cm}p{4.2cm}p{2.9cm}@{}}
\toprule
component & existing literature & Cred reuse & Cred addition & anchor \\
\midrule
continuous values $[0,1]$ & Zadeh, Goguen \cite{zadeh1965,goguen1969} & yes, as credence & reading as credence, not membership & \Lean{Credence} \\
t-norm logics & H\'ajek, Esteva--Godo \cite{hajek1998,esteva2001} & product $\conjc$, De Morgan dual & $\conjc$ barred from fixing $\cred(A\wedge B)$ & \Lean{conj} \\
LP / K$_3$ & Kleene, Priest \cite{kleene1952,priest1979} & three-value tables & exact bridge: LP positivity, K$_3$ certainty & \Lean{kleeneCongruence} \\
possible-world probability & Pearl \cite{pearl1988} & chain-rule shape & conditioning primitive, not a ratio & \Lean{Adm} \\
Cox / Jaynes & \cite{cox1946,jaynes2003,halpern1999} & chain rule as primitive & no division, no universality axiom & \Lean{conditioning\_mk} \\
de Finetti & \cite{definetti1936,renyi1955,popper1959} & conditional-first, void at zero & free at $e=0$ as a theorem & \Lean{conditioning\_}\allowbreak\Lean{zero\_any} \\
imprecise / credal & Walley \cite{walley1991,augustin2014} & interval-valued conditional & one scale, imprecision only at the conditional & \Lean{cond\_zero\_}\allowbreak\Lean{zero\_univ} \\
PSL & Bach et al. \cite{bach2017} & none (contrast) & product not \L ukasiewicz conjunction & n/a \\
the $\Cond$ fiber & Frechet bounds \cite{gilio2021,nelsen2006} & Frechet trichotomy & preimage of one scalar equation & \Lean{cond\_}\allowbreak\Lean{singleton\_}\allowbreak\Lean{of\_pos} \\
supplied joint & copulas \cite{sklar1959} & $W,M$ as bounds & joint is an input, not a rule & \Lean{ExactJoint}\allowbreak\Lean{Context} \\
all-measures entailment & probabilistic worlds \cite{pearl1988} & $\forall$ over valuations & entailment as the all-measures corner & \Lean{entails\_iff\_}\allowbreak\Lean{cond\_one} \\
Lean-checked factorization & n/a & n/a & the layered package itself, machine-verified & \Lean{Inference}\allowbreak\Lean{Commitments} \\
\bottomrule
\end{tabular}%
}
\caption{Old ingredients versus the one new layer.  The fourth column is
deliberately thin: most rows reuse a standing construction, and Cred's
addition is the verified factorization and the obstruction results, not
a new operation.}
\label{tab:old-vs-new}
\end{table}

\subsection{Mapping practical systems to Cred commitments}
\label{sec:system-mapping}

Each deployed system fixes a joint policy and a conditioning policy that
Cred can name.  \Cref{tab:system-mapping} records the mapping.  It is not
a subsumption claim: the systems carry inference and learning Cred does
not.  The mapping only states which commitment each system makes in
Cred's vocabulary.  ProbLog assumes independence of base facts, so its
fact joint is the product, with dependence only through rules
\cite{sato1995,deraedt2007,fierens2015}.  PSL relaxes conjunction with
the \L ukasiewicz t-norm rather than the product \cite{bach2017,kimmig2012}.
MLN normalizes a log-linear measure over crisp worlds, with correlation
from learned weights \cite{richardson2006}.  Imprecise and credal
methods carry sets of measures \cite{walley1991,cozman2000,maua2020}.

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{@{}p{2.3cm}p{3.7cm}p{2.6cm}p{3.6cm}@{}}
\toprule
system & joint commitment & conditioning & Cred reading \\
\midrule
ProbLog & independence of base facts & Bayesian, $0/0$ at zero & \Lean{JointPolicy.product} \\
PSL & \L ukasiewicz conjunction & relaxed rule penalty & non-product t-norm joint \\
MLN & normalized measure over worlds & Bayesian, $0/0$ on hard evidence & measure frame, \Lean{usesMeasure} \\
imprecise / credal & set of factorizing joints & lower/upper extension & credal frame, \Lean{usesCredal} \\
\bottomrule
\end{tabular}
\caption{Practical systems and their commitment in Cred vocabulary.  The
right column names the Cred ledger field, not a reduction of the system
to Cred.}
\label{tab:system-mapping}
\end{table}

\subsection{A false-precision audit}
\label{sec:false-precision}

Take marginals $\cred(A)=\tfrac35$ and $\cred(B)=\tfrac{7}{10}$.  Two
analysts each report a single confident $\cred(A\mid B)$, differing only
in an unstated joint.  Hidden independence sets the joint to
$\tfrac35\conjc\tfrac{7}{10}=\tfrac{21}{50}$, so the conditional is
$\tfrac{21}{50}\big/\tfrac{7}{10}=\tfrac35$: the posterior equals the
prior and the evidence did nothing
(\Lean{hidden\_independence\_value}).  Hidden maximal dependence sets the
joint to $\min(\tfrac35,\tfrac{7}{10})=\tfrac35$, so the conditional is
$\tfrac35\big/\tfrac{7}{10}=\tfrac67$ (\Lean{hidden\_min\_conditional}).

The audit refuses both single numbers.  The joint is known only to its
Fr\'echet band $[\tfrac{3}{10},\tfrac35]$, so the conditional is the band
$[\tfrac37,\tfrac67]$ (\Lean{cred\_audit\_interval}).  Both point
estimates sit inside it (\Lean{hidden\_points\_in\_band}).  A threshold
test at $\tfrac45$ is dependence-sensitive: $\tfrac35<\tfrac45<\tfrac67$,
so the verdict flips with the unstated joint
(\Lean{threshold\_four\_fifths\_sensitive}).  A test at $\tfrac13$ is
robust: $\tfrac13$ sits at or below the whole band
(\Lean{threshold\_one\_third\_robust}).  Anchors are in
\texttt{Cred.Examples.FalsePrecision}.

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{@{}lll@{}}
\toprule
reading & hidden joint & reported $\cred(A\mid B)$ \\
\midrule
hidden independence & $\tfrac{21}{50}$ & $\tfrac35$ \\
hidden maximal dependence & $\tfrac35$ & $\tfrac67$ \\
Cred audit & Fr\'echet band $[\tfrac{3}{10},\tfrac35]$ & band $[\tfrac37,\tfrac67]$ \\
\midrule
threshold $\tfrac45$ & either & dependence-sensitive (flips) \\
threshold $\tfrac13$ & either & robust \\
\bottomrule
\end{tabular}
\caption{The false-precision audit.  Two confident point estimates
straddle the threshold $\tfrac45$; the audit reports the band and flags
the threshold as dependence-sensitive.}
\label{tab:false-precision}
\end{table}

\subsection{How to run a Cred audit}
\label{sec:cookbook}

A step from marginals to a conditional is audited in four moves.

\begin{enumerate}[leftmargin=*,nosep]
\item Identify the value layer.  Decide whether the numbers are
  credences in $[0,1]$, crisp truth values, or a measure over worlds, and
  whether a downstream collapse to three values is intended.
\item Name the joint policy.  State whether $\cred(A\wedge B)$ is the
  product, the min, a Fr\'echet interval, or an explicitly supplied
  value (\Lean{JointPolicy}).  An unstated policy is the usual source of
  false precision.
\item Check robust versus dependence-sensitive.  Compute the conditional
  band the joint policy allows and test the claim against it.  A verdict
  that survives the whole band is robust; one that flips is
  dependence-sensitive (\Lean{Audit.audit\_robust\_or\_sensitive}).
\item Check zero-evidence handling.  When $\cred(B)=0$, the chain rule
  constrains nothing, so the conditional is the full interval; report
  that, not a default value or an undefined ratio
  (\Lean{cond\_zero\_zero\_univ}).
\end{enumerate}

\section{Toward a Foundation for Mathematics}
\label{sec:foundations}

The scalar kernel is not yet a foundation.  A foundation must keep four
structures separate: the value of a formula, semantic consequence over
all valuations, the supplied dependence (a joint or family of joints),
and proof provenance (which assumptions a derivation uses).  Ordinary
prose collapses these; the toy modules above formalize a first cut of
each missing layer.  The table records status against the Lean
development; anchors are bare names in the \texttt{Cred.Dependence},
\texttt{Cred.ProofTheory}, \texttt{Cred.Examples}, and \texttt{Cred.Math}
namespaces.

\begin{center}
\begin{tabular}{@{}p{0.27\linewidth}p{0.40\linewidth}p{0.27\linewidth}@{}}
\toprule
Layer & Formalized now (anchor) & Remaining \\
\midrule
Dependence contexts &
exact/coherent/interval contexts and joint families, with the
interval-to-fiber image (\Lean{Cred.Dependence}) &
credal networks over many propositions \\[2pt]
Robust collapse &
\Lean{collapseIntervalToThree} into \Lean{RobustStatus} &
status-preserving rule propagation \\[2pt]
Proof provenance &
\Lean{usedAssumptions}, \Lean{provenance\_sound} &
full DAG over the foundation language \\[2pt]
Hypothetical branches &
\Lean{local\_contradiction\_no\_explosion} &
$T+R$ over genuine theories \\[2pt]
Generative calculus &
local connective rules (conjunction introduction and elimination,
disjunction introduction, double negation) with \Lean{generative\_sound},
no semantic-oracle rule &
quantifiers, equality, induction; syntactic completeness \\[2pt]
Mathematics seed &
parity, divisibility, the $\sqrt2$ core contradiction
(\Lean{Cred.Math}, \Lean{Cred.Examples.Sqrt2Branch}) &
full real-number development \\
\bottomrule
\end{tabular}
\end{center}

Two honesty markers.  The labelled calculus of the previous section is
sound and closed-complete because the semantic consequence relation is
admitted as a proof rule; that is an interface result, not a deep
syntactic completeness theorem.  The new generative fragment
(\Lean{Cred.ProofTheory.Generative}) removes the oracle rule for the
connective fragment and proves soundness directly, but it does not yet
cover quantifiers, equality, or induction.  Classical mathematics is
recovered at certainty on crisp structures; probability enters when the
joint context is supplied or learned; ternary and binary views are
robust summary layers, not places where dependence is invented.

\section{Lean Artifact}

All formal claims cited above are present in the current Lean
development under `lean/Cred`.  The core files are:
\[
\begin{array}{ll}
\text{values} & \Lean{Cred/Core/Value.lean},\\
\text{conditioning} & \Lean{Cred/Cond/Admissible.lean},\\
\text{admissibility proof theory} & \Lean{Cred/Cond/Proof.lean},\\
\text{collapse} & \Lean{Cred/Collapse/Hom.lean},\\
\text{LP/K$_3$ bridge} & \Lean{Cred/Bridge/LPK3.lean},\\
\text{conditional no-go} & \Lean{Cred/Bridge/CondBridge.lean},\\
\text{Curry block} & \Lean{Cred/Bridge/Curry.lean},\\
\text{labelled sequents} & \Lean{Cred/Sequent.lean},\\
\text{kernel certificates} & \Lean{Cred/Kernel.lean},\\
\text{cut and completeness} & \Lean{Cred/CutElim.lean}, \Lean{Cred/Completeness.lean},\\
\text{solution sets} & \Lean{Cred/Fixpoint.lean},\\
\text{thresholds} & \Lean{Cred/Threshold.lean},\\
\text{dependence contexts} & \Lean{Cred/Dependence/*.lean},\\
\text{generative calculus} & \Lean{Cred/ProofTheory/*.lean},\\
\text{toy models} & \Lean{Cred/Examples/*.lean}.
\end{array}
\]

The intended verification command is:
\[
  \texttt{cd lean \&\& lake build}.
\]
The manuscript itself is built by:
\begin{quote}\small\ttfamily
cd singlepaper\\
latexmk -pdf -interaction=nonstopmode -halt-on-error paper.tex
\end{quote}

\section{Conclusion}

The algebra is small:
\[
  \negc c=1-c,\qquad a\conjc b=ab,\qquad
  a\disjc b=a+b-ab,\qquad \Adm(c;j,e)\Longleftrightarrow ce=j.
\]
The lesson is also small.  A conditional is not a truth value waiting
to be tabulated.  It is a constraint that may determine one value,
many values, or none.  The set $\Cond(j,e)$ is the fiber of that
constraint, once the surrounding mathematics permits set notation.

At positive evidence, the chain rule recovers ordinary division.  At
zero evidence, it refuses to manufacture information.  The Kleene
collapse explains why positivity gives LP and certainty gives K$_3$.
The no-go theorem explains why conditioning does not extend to
that collapse.  The Curry block explains why the system should not add
an internal arrow to repair this.

The formal core is therefore not a new truth table.  Consequence and
conditioning are one inference, the inclusion of the admissible set the
constraints cut out; the separation is that only the value fibration
resists collapse to a truth-functional connective.

The use is an audit.  Cred reads an uncertain inference step, names the
joint policy and the conditioning policy it commits to, and reports the
admissible set instead of a single number.  The practical target is
false precision: a confident point estimate whose value rests on an
unstated joint.  When the joint is unknown, the verdict is an interval,
and the audit flags any threshold the interval straddles
(\Lean{Audit.audit\_robust\_or\_sensitive}).  Cred audits a step; it
does not replace the probabilistic, fuzzy, and soft-logic systems that
do inference and learning.  Its contribution over the prior art is the
Lean-checked factorization of value, dependence, and conditioning, with
the no-go and uniqueness results that fix each layer.

\bibliographystyle{alpha}
\bibliography{../cred,refs}

\end{document}
