\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{proof}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{stmaryrd}

\newtheorem{definition}{Definition}[section]
\newtheorem{remark}{Remark}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}

% Basic notation
\newcommand{\ctx}{\;\mathsf{ctx}}
\newcommand{\type}{\;\mathsf{type}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\Type}{\mathsf{Type}}
\newcommand{\Nat}{\mathbb{N}}
\newcommand{\at}[1]{\,@\,#1}

\title{ProbTT: Type Theory with Primitive Probability}

\author{Probabilistic Foundations Project}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We define a type theory where probability is primitive. Judgments have the form $\Gamma \vdash a : A \at{p}$, meaning ``$a$ inhabits $A$ with weight $p$.'' The weight $p$ is itself a term of type $\Prob$. When weights are restricted to $\{0,1\}$, we recover Martin-L\"of Type Theory. We give precise formation, introduction, elimination, and computation rules, and sketch the categorical semantics in Markov categories.
\end{abstract}

\tableofcontents

%=============================================================================
\section{Syntax}
%=============================================================================

\subsection{Pre-terms and Pre-types}

We define raw syntax first, then typing rules.

\begin{definition}[Raw Terms]
\begin{align*}
p, q, r &::= 0 \mid 1 \mid p + q \mid p \cdot q \mid p - q \mid x && \text{(probability terms)} \\
a, b, c &::= x \mid \lambda x.a \mid a\,b \mid (a,b) \mid \pi_1 a \mid \pi_2 a \mid \mathsf{refl} \mid J(\ldots) && \text{(terms)} \\
A, B, C &::= \Prob \mid A \to_p B \mid \Sigma_p(x:A).B \mid \mathsf{Id}_A(a,b) \mid \Type_i && \text{(types)}
\end{align*}
\end{definition}

\begin{remark}
The subscript $p$ on $\to_p$ and $\Sigma_p$ is a \emph{weight annotation}---a term of type $\Prob$. This is the key novelty: type formers carry probability weights.
\end{remark}

\subsection{Judgments}

\begin{definition}[Judgment Forms]
ProbTT has five judgment forms:
\begin{enumerate}
    \item $\Gamma \ctx$ \hfill (context validity)
    \item $\Gamma \vdash A \type$ \hfill (type validity)
    \item $\Gamma \vdash a : A$ \hfill (typing)
    \item $\Gamma \vdash a \equiv b : A$ \hfill (definitional equality)
    \item $\Gamma \vdash p : \Prob$ \hfill (probability term)
\end{enumerate}
\end{definition}

\begin{remark}[No @ Annotation on Judgments]
Unlike the sketch, we do NOT write $\Gamma \vdash a : A \at{p}$. Instead, the weight $p$ appears \emph{inside} the type: $A \to_p B$ means ``function that produces output with weight $p$.'' This avoids the question ``what does the annotation mean?''---the weight is part of the type structure.
\end{remark}

%=============================================================================
\section{The Probability Type}
%=============================================================================

\subsection{Formation}

\begin{gather}
\infer[\textsc{Prob-Form}]
  {\Gamma \vdash \Prob \type}
  {\Gamma \ctx}
\end{gather}

$\Prob$ is a type in any valid context.

\subsection{Introduction}

\begin{gather}
\infer[\textsc{Prob-Zero}]
  {\Gamma \vdash 0 : \Prob}
  {\Gamma \ctx}
\qquad
\infer[\textsc{Prob-One}]
  {\Gamma \vdash 1 : \Prob}
  {\Gamma \ctx}
\\[1em]
\infer[\textsc{Prob-Add}]
  {\Gamma \vdash p + q : \Prob}
  {\Gamma \vdash p : \Prob & \Gamma \vdash q : \Prob}
\\[1em]
\infer[\textsc{Prob-Mul}]
  {\Gamma \vdash p \cdot q : \Prob}
  {\Gamma \vdash p : \Prob & \Gamma \vdash q : \Prob}
\\[1em]
\infer[\textsc{Prob-Sub}]
  {\Gamma \vdash p - q : \Prob}
  {\Gamma \vdash p : \Prob & \Gamma \vdash q : \Prob}
\end{gather}

\subsection{Equations}

$\Prob$ satisfies the equations of a bounded semiring with subtraction:
\begin{align}
p + q &\equiv q + p & p \cdot q &\equiv q \cdot p \\
(p + q) + r &\equiv p + (q + r) & (p \cdot q) \cdot r &\equiv p \cdot (q \cdot r) \\
p + 0 &\equiv p & p \cdot 1 &\equiv p \\
p \cdot 0 &\equiv 0 & p \cdot (q + r) &\equiv p \cdot q + p \cdot r \\
p - p &\equiv 0 & p + (1 - p) &\equiv 1 \\
1 - (1 - p) &\equiv p & (p + q) - q &\equiv p
\end{align}

These are definitional equalities (computation rules).

\subsection{Elimination}

There is no general eliminator for $\Prob$. We cannot ``case split'' on whether $p = 0$ or $p = 1$ in general. This is crucial: probability is not Boolean.

However, in the \emph{semantics}, $\Prob$ is interpreted as $[0,1]$, and type-checking may use real-number reasoning.

%=============================================================================
\section{Weighted Function Types}
%=============================================================================

The key innovation: functions carry a \emph{weight} indicating the probability of producing a valid output.

\subsection{Formation}

\begin{gather}
\infer[\textsc{Fun-Form}]
  {\Gamma \vdash (x : A) \to_p B \type}
  {\Gamma \vdash A \type & \Gamma, x : A \vdash B \type & \Gamma, x : A \vdash p : \Prob}
\end{gather}

Read $(x : A) \to_p B$ as: ``given $x : A$, produce $b : B$ with weight $p(x)$.''

When $B$ does not depend on $x$, write $A \to_p B$.

\subsection{Introduction}

\begin{gather}
\infer[\textsc{Fun-Intro}]
  {\Gamma \vdash \lambda x.b : (x : A) \to_p B}
  {\Gamma, x : A \vdash b : B & \Gamma, x : A \vdash p : \Prob}
\end{gather}

The weight $p$ is determined by the function's specification, not inferred from the body.

\subsection{Elimination}

\begin{gather}
\infer[\textsc{Fun-Elim}]
  {\Gamma \vdash f\,a : B[a/x] \at{p[a/x]}}
  {\Gamma \vdash f : (x : A) \to_p B & \Gamma \vdash a : A}
\end{gather}

\begin{remark}
Here we use $\at{p[a/x]}$ as an annotation on the \emph{result}, not on the judgment. This means: the term $f\,a$ has type $B[a/x]$, but carries weight $p[a/x]$.

We need to formalize what ``carries weight'' means. Two options:
\begin{enumerate}
    \item \textbf{Refinement types}: $f\,a : \{y : B[a/x] \mid \mathsf{weight}(y) = p[a/x]\}$
    \item \textbf{Weighted types}: A separate judgment $\Gamma \vdash a : A \at{p}$ where $\at{p}$ is sugar for a weighted type.
\end{enumerate}

We take option 2 for clarity.
\end{remark}

\subsection{Weighted Judgment (Derived)}

\begin{definition}[Weighted Typing]
Define $\Gamma \vdash a : A \at{p}$ as sugar for:
\[
\Gamma \vdash a : A \times_{\mathsf{wt}} p
\]
where $A \times_{\mathsf{wt}} p$ is a type of pairs $(a, p)$ with $a : A$ and $p : \Prob$, subject to coherence.
\end{definition}

Actually, this gets complicated. Let's take a different approach.

%=============================================================================
\section{Alternative: Judgments with Weights}
%=============================================================================

We now take the weights seriously as part of the judgment.

\subsection{Judgment Forms (Revised)}

\begin{definition}[Weighted Judgments]
\begin{enumerate}
    \item $\Gamma \ctx$ \hfill (context validity)
    \item $\Gamma \vdash A \type$ \hfill ($A$ is a type)
    \item $\Gamma \vdash a : A \at{p}$ \hfill ($a$ has type $A$ with weight $p$)
    \item $\Gamma \vdash a \equiv b : A \at{p}$ \hfill ($a$ equals $b$ at $A$ with weight $p$)
\end{enumerate}
where $p$ is a \textbf{closed term} of type $\Prob$.
\end{definition}

\begin{remark}[Weight Semantics]
$\Gamma \vdash a : A \at{p}$ means: in context $\Gamma$, the term $a$ is a valid element of $A$ with ``probability mass'' $p$. Think of this as:
\begin{itemize}
    \item $p = 1$: $a$ definitely inhabits $A$
    \item $p = 0$: $a$ does not inhabit $A$ (vacuously true)
    \item $0 < p < 1$: $a$ partially inhabits $A$
\end{itemize}
The categorical semantics makes this precise: weighted typing corresponds to morphisms in a Markov category.
\end{remark}

\subsection{Structural Rules}

\begin{gather}
\infer[\textsc{Weaken}]
  {\Gamma \vdash a : A \at{q}}
  {\Gamma \vdash a : A \at{p} & q \leq p}
\\[1em]
\infer[\textsc{Var}]
  {\Gamma, x : A, \Delta \vdash x : A \at{1}}
  {\Gamma \vdash A \type}
\end{gather}

Variables have weight 1: if $x : A$ is in context, then $x$ definitely has type $A$.

\subsection{Prob Rules (Weighted)}

\begin{gather}
\infer[\textsc{Prob-Intro}]
  {\Gamma \vdash p : \Prob \at{1}}
  {\Gamma \vdash p : \Prob \text{ (by syntax)}}
\end{gather}

Every well-formed probability term has weight 1 (probability values are certain).

\subsection{Function Rules (Weighted)}

\begin{gather}
\infer[\textsc{Fun-Form}]
  {\Gamma \vdash (x : A) \to B \type}
  {\Gamma \vdash A \type & \Gamma, x : A \vdash B \type}
\\[1em]
\infer[\textsc{Fun-Intro}]
  {\Gamma \vdash \lambda x.b : (x : A) \to B \at{p}}
  {\Gamma, x : A \vdash b : B \at{p}}
\\[1em]
\infer[\textsc{Fun-Elim}]
  {\Gamma \vdash f\,a : B[a/x] \at{p \cdot q}}
  {\Gamma \vdash f : (x : A) \to B \at{p} & \Gamma \vdash a : A \at{q}}
\\[1em]
\infer[\textsc{Fun-Beta}]
  {\Gamma \vdash (\lambda x.b)\,a \equiv b[a/x] : B[a/x] \at{p}}
  {\Gamma, x : A \vdash b : B \at{p} & \Gamma \vdash a : A \at{1}}
\end{gather}

\begin{remark}[Weight Multiplication]
In \textsc{Fun-Elim}, the output weight is $p \cdot q$: the product of the function's weight and the argument's weight. This models:
\begin{itemize}
    \item Uncertain function applied to certain argument: weight $p$
    \item Certain function applied to uncertain argument: weight $q$
    \item Both uncertain: weight $p \cdot q$
\end{itemize}
\end{remark}

\subsection{Pair Rules (Weighted)}

\begin{gather}
\infer[\textsc{Pair-Form}]
  {\Gamma \vdash \Sigma(x : A).B \type}
  {\Gamma \vdash A \type & \Gamma, x : A \vdash B \type}
\\[1em]
\infer[\textsc{Pair-Intro}]
  {\Gamma \vdash (a, b) : \Sigma(x : A).B \at{p \cdot q}}
  {\Gamma \vdash a : A \at{p} & \Gamma \vdash b : B[a/x] \at{q}}
\\[1em]
\infer[\textsc{Pair-Fst}]
  {\Gamma \vdash \pi_1\,t : A \at{p}}
  {\Gamma \vdash t : \Sigma(x : A).B \at{p}}
\\[1em]
\infer[\textsc{Pair-Snd}]
  {\Gamma \vdash \pi_2\,t : B[\pi_1 t/x] \at{p}}
  {\Gamma \vdash t : \Sigma(x : A).B \at{p}}
\end{gather}

Pair introduction multiplies weights (joint probability). Projection preserves weight.

\subsection{Identity Rules (Weighted)}

\begin{gather}
\infer[\textsc{Id-Form}]
  {\Gamma \vdash \mathsf{Id}_A(a, b) \type}
  {\Gamma \vdash a : A \at{1} & \Gamma \vdash b : A \at{1}}
\\[1em]
\infer[\textsc{Id-Intro}]
  {\Gamma \vdash \mathsf{refl}_a : \mathsf{Id}_A(a, a) \at{p}}
  {\Gamma \vdash a : A \at{p}}
\\[1em]
\infer[\textsc{Id-Elim}]
  {\Gamma \vdash J(P, d, e) : P(b, e) \at{p \cdot q}}
  {\begin{array}{c}
    \Gamma, x : A, y : \mathsf{Id}_A(a, x) \vdash P(x, y) \type \\
    \Gamma \vdash d : P(a, \mathsf{refl}_a) \at{p} \\
    \Gamma \vdash e : \mathsf{Id}_A(a, b) \at{q}
  \end{array}}
\end{gather}

\begin{remark}
Identity formation requires weight 1 for both endpoints---we only form $\mathsf{Id}_A(a,b)$ when $a$ and $b$ definitely exist. But a proof of identity can have lower weight.
\end{remark}

%=============================================================================
\section{Universes}
%=============================================================================

\begin{gather}
\infer[\textsc{Univ-Form}]
  {\Gamma \vdash \Type_i \type}
  {\Gamma \ctx}
\\[1em]
\infer[\textsc{Univ-Cumul}]
  {\Gamma \vdash A : \Type_{i+1} \at{1}}
  {\Gamma \vdash A : \Type_i \at{1}}
\\[1em]
\infer[\textsc{Prob-Univ}]
  {\Gamma \vdash \Prob : \Type_0 \at{1}}
  {\Gamma \ctx}
\end{gather}

$\Prob$ lives in the lowest universe. Types-as-terms have weight 1 (types are certain).

%=============================================================================
\section{MLTT as Fragment}
%=============================================================================

\begin{theorem}[MLTT Embedding]
Restrict to judgments where all weights are $1$. Then:
\begin{enumerate}
    \item $\Gamma \vdash a : A \at{1}$ becomes $\Gamma \vdash a : A$
    \item Weight multiplication $1 \cdot 1 = 1$ is trivial
    \item All ProbTT rules reduce to MLTT rules
    \item Weakening with $q \leq 1$ where $q = 1$ is trivial
\end{enumerate}
\end{theorem}

\begin{proof}
By inspection of each rule. When $p = q = 1$:
\begin{itemize}
    \item \textsc{Fun-Elim}: weight $1 \cdot 1 = 1$
    \item \textsc{Pair-Intro}: weight $1 \cdot 1 = 1$
    \item \textsc{Id-Elim}: weight $1 \cdot 1 = 1$
\end{itemize}
The rules become exactly MLTT rules with weight annotations erased.
\end{proof}

%=============================================================================
\section{Categorical Semantics}
%=============================================================================

\subsection{Markov Categories}

\begin{definition}[Markov Category]
A \emph{Markov category} is a symmetric monoidal category $(\mathcal{C}, \otimes, I)$ with:
\begin{enumerate}
    \item A commutative comonoid structure on each object: $\mathsf{copy}_A : A \to A \otimes A$, $\mathsf{del}_A : A \to I$
    \item Compatibility with the monoidal structure
\end{enumerate}
\end{definition}

Examples: $\mathsf{Stoch}$ (measurable spaces with Markov kernels), $\mathsf{FinStoch}$ (finite sets with stochastic matrices).

\subsection{Interpretation}

\begin{itemize}
    \item $\llbracket \Prob \rrbracket = [0,1]$
    \item $\llbracket A \to B \at{p} \rrbracket$ = morphism $A \to B$ in the Markov category, scaled by $p$
    \item $\llbracket \Gamma \vdash a : A \at{p} \rrbracket$ = morphism $\llbracket \Gamma \rrbracket \to \llbracket A \rrbracket$ with ``mass'' $p$
    \item Contexts $\Gamma$ interpret as products
    \item Weight multiplication = composition of scaled morphisms
\end{itemize}

\begin{theorem}[Soundness]
If $\Gamma \vdash a : A \at{p}$ is derivable, then $\llbracket a \rrbracket : \llbracket \Gamma \rrbracket \to \llbracket A \rrbracket$ is a valid morphism in the Markov category with appropriate scaling.
\end{theorem}

\begin{proof}
By induction on derivations. The key cases:
\begin{itemize}
    \item \textsc{Fun-Elim}: Composition of morphisms, scaling factors multiply
    \item \textsc{Pair-Intro}: Product of morphisms, independent probabilities multiply
    \item \textsc{Weaken}: Scaling by smaller factor is valid (sub-probability)
\end{itemize}
\end{proof}

%=============================================================================
\section{Logic from Probability}
%=============================================================================

\subsection{Propositions as Types with Weight}

In MLTT, propositions are types (Curry-Howard). In ProbTT:
\begin{center}
\emph{Propositions are types, proofs have weights.}
\end{center}

\begin{definition}[Probabilistic Proposition]
A proposition $P$ is a type. A proof of $P$ with confidence $p$ is a term $a : P \at{p}$.
\end{definition}

\subsection{Logical Connectives}

\begin{align}
P \wedge Q &:= \Sigma(\_:P).Q && \text{(conjunction = pair)} \\
P \vee Q &:= P + Q && \text{(disjunction = sum, needs sum types)} \\
P \Rightarrow Q &:= P \to Q && \text{(implication = function)} \\
\neg P &:= P \to \mathbf{0} && \text{(negation = function to empty)} \\
\forall(x:A).P(x) &:= (x:A) \to P(x) && \text{(universal = dependent function)} \\
\exists(x:A).P(x) &:= \Sigma(x:A).P(x) && \text{(existential = dependent pair)}
\end{align}

\subsection{Weight Interpretation}

\begin{itemize}
    \item $a : P \wedge Q \at{p \cdot q}$ means: proof of $P$ with weight $p$, proof of $Q$ with weight $q$, joint weight $p \cdot q$
    \item $f : P \Rightarrow Q \at{p}$ means: given proof of $P$, produce proof of $Q$ with weight $p$
    \item $a : \forall(x:A).P(x) \at{p}$ means: for all $x$, prove $P(x)$ with weight $p$ (uniform bound)
\end{itemize}

\subsection{Classical Logic}

When weights are $\{0, 1\}$:
\begin{itemize}
    \item $a : P \at{1}$ = definite proof of $P$
    \item $a : P \at{0}$ = no proof (vacuous)
    \item Weight multiplication = conjunction
    \item This gives classical propositional reasoning
\end{itemize}

%=============================================================================
\section{Open Problems}
%=============================================================================

\begin{enumerate}
    \item \textbf{Sum Types}: How do weighted sum types work? $A + B \at{p}$ where the weight might differ for left vs.\ right injection.

    \item \textbf{Normalization}: Does ProbTT normalize? The weight annotations don't affect computation, so this should follow from MLTT normalization.

    \item \textbf{Decidability}: Type checking requires comparing elements of $\Prob$. This is decidable if $\Prob$ is interpreted as rationals, undecidable for reals.

    \item \textbf{Weight Inference}: Can we infer weights from term structure, or must they be explicit?

    \item \textbf{Continuous Distributions}: For measure-theoretic probability, we need integration. Can we add $\int : (A \to \Prob) \to \Prob$ as a primitive?

    \item \textbf{Conditioning}: How do we express $\Gamma \vdash a : A \at{p \mid q}$ (conditional probability)?
\end{enumerate}

%=============================================================================
\section{Summary}
%=============================================================================

ProbTT extends type theory with:
\begin{enumerate}
    \item Primitive type $\Prob$ with bounded semiring structure
    \item Weighted judgments $\Gamma \vdash a : A \at{p}$
    \item Weight multiplication in elimination rules
    \item MLTT as the $\{0,1\}$-fragment
    \item Semantics in Markov categories
\end{enumerate}

The key insight: \textbf{logic is certain reasoning} (weight 1), \textbf{probability is uncertain reasoning} (weight in $(0,1)$). ProbTT unifies both.

\end{document}
