% Cred Part 2: From Algebra to Reasoning
% Publication-quality paper
%
% References to Lean formalization:
%   lean/Cred/Valuation.lean
%   lean/Cred/Consequence.lean
%   lean/Cred/Update.lean
%   lean/Cred/Predicate.lean

\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
    pdfauthor={Anonymous},
    pdftitle={Cred Part 2: From Algebra to Reasoning}
}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

% Custom commands (same as Part 1)
\newcommand{\cred}{\mathsf{cred}}
\newcommand{\Cred}{\mathsf{Cred}}
\newcommand{\negc}{\mathord{\sim}}
\newcommand{\conjc}{\otimes}
\newcommand{\disjc}{\sqcup}
\newcommand{\condc}[2]{({#1} \mid {#2})}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Nat}{\mathbb{N}}

\title{\textbf{Cred Part~2: Two Views of Credence Reasoning}\\[0.5em]
\large Probability-Style and Logic-Style Interpretations}

\author{Anonymous for blind review}
\date{February 2026}

\begin{document}

\maketitle

\begin{abstract}
Part~1 established \emph{Cred} as a constraint algebra on $[0,1]$: values,
complement, the product operation, and the chain rule for conditioning.
The key design choice: the joint credence $\cred(A \land B)$ is an
\emph{external parameter}, not fixed by the algebra.
This makes Cred compatible with both probability and many-valued logic.
%
Part~1's central finding is that maximum positive dependence---the joint
$\min(\cred(A), \cred(B))$---is the \emph{unique} fixed formula satisfying
all requirements for proper reasoning \emph{under the Part~1 hypotheses}:
truth-functionality (a fixed joint $j(a,b)$), Bayes consistency (symmetry of
the induced joint), copula-style regularity (2-increasing / Fr\'echet upper
bound), non-trivial conditioning (evidence matters), and idempotence.
Independence ($\cred(A) \cdot \cred(B)$) gives \emph{trivial} conditioning
where evidence has no effect; other choices fail Bayes consistency.
%
This paper develops both interpretations of Cred.
The \emph{probability-style view} supplies joints from empirical data:
credences propagate via Bayesian and Jeffrey conditionalization,
uncertainty remains graded, and zero-evidence underdetermination connects
to imprecise probability.
The \emph{logic-style view} uses the minimum joint (maximum dependence) and
collapses to three values; designated-value consequence gives K$_3$
(designating $\{1\}$), LP (designating $\{1, \tfrac{1}{2}\}$), or graded
thresholds.
The minimum is not arbitrary---it is \emph{forced} by requiring truth-functional
composition compatible with proper Bayesian inference.
%
Explosion fails non-vacuously for LP and for graded consequence at
thresholds $0 < t \leq \tfrac{1}{2}$; K$_3$ has explosion only vacuously.
The two views are complementary: probability-style answers ``how confident
should I be?''; logic-style answers ``is this valid to conclude?''
We extend to first order via graded predicates with inf/sup quantifiers,
showing that the Russell predicate yields credence~$\tfrac{1}{2}$ rather
than a paradox.
The main definitions and theorems are machine-checked in Lean~4.
\end{abstract}

%=============================================================================
\section{Introduction}
\label{sec:intro}
%=============================================================================

Part~1~\cite{part1} established the distinction between two traditions:
\emph{degrees of truth} (compositional, logic tradition) and
\emph{degrees of belief} (non-compositional, probability tradition).
The former asks ``Is this valid to conclude?''; the latter asks
``How confident should I be?''

\subsection*{Key results from Part~1}

Part~1~\cite{part1} introduced \emph{Cred}, a constraint algebra on $[0,1]$
with the chain rule $\condc{A}{B} \conjc \cred(B) = \cred(A \land B)$ as its
sole structural axiom for conditioning.
The key design choice: the joint credence $\cred(A \land B)$ is an
\emph{external parameter}, not fixed by the algebra.
When $\cred(B) = 0$, the chain rule is satisfied by any value---conditioning
is unconstrained, avoiding explosion.

Part~1's central finding is that maximum positive dependence---the joint
$\min(\cred(A), \cred(B))$---is the \emph{unique} fixed formula satisfying
truth-functionality, Bayes consistency, non-trivial conditioning, and
idempotence (Part~1, \texttt{min\_copula\_unique}).
Independence gives trivial conditioning; RM3 and G\"odel fail Bayes consistency.
Part~1 also established: a surjective homomorphism from Cred to the Kleene
lattice $\{0, \tfrac{1}{2}, 1\}$; and the unique negation fixed point at
$\tfrac{1}{2}$.

\subsection*{This paper: two views of Cred}

This paper asks: how do we \emph{use} Cred's algebra for reasoning?
The answer is that Cred admits two complementary interpretations---one
probability-style, one logic-style---and the choice depends on what question
you are asking:

\begin{description}
\item[Probability-style view.]
The chain rule \emph{is} inference.
Credences propagate through reasoning: given any two of (joint, evidence,
conditional), the chain rule determines the third.
Uncertainty remains graded---there is no final yes/no verdict.
Update rules (Bayesian and Jeffrey conditionalization) revise entire
credence assignments when evidence arrives.
When evidence has credence zero, the conditional is unconstrained,
yielding interval-valued posteriors that connect to imprecise
probability~\cite{walley1991}.

\item[Logic-style view.]
Fix the joint to the minimum (maximum positive dependence)---the unique
truth-functional choice with non-trivial Bayes-consistent conditioning.
Collapse credences to discrete values via Part~1's homomorphism to the
Kleene lattice $\{0, \tfrac{1}{2}, 1\}$.
Designating $\{1\}$ gives K$_3$ (paracomplete); designating
$\{1, \tfrac{1}{2}\}$ gives LP (paraconsistent).
Graded consequence on $[0,1]$ parametrizes validity by a threshold~$t$:
an inference is valid when every valuation making premises $\geq t$
also makes the conclusion $\geq t$.
\end{description}

\noindent
The two views answer different questions.
Probability-style asks ``how confident should I be?''---the output is a
credence in $[0,1]$.
Logic-style asks ``is this valid to conclude?''---the output is yes or no,
determined by whether the credence meets the designation criterion.
Neither view is ``correct''; they are complementary tools for different
purposes.
The collapse homomorphism is the bridge: it converts probability-style
reasoning to logic-style judgments.

\begin{center}
\small
\begin{tabular}{lll}
\toprule
\textbf{View} & \textbf{Core mechanism} & \textbf{Section} \\
\midrule
Probability-style & Chain rule / update rules & Sections~\ref{sec:valuations}, \ref{sec:update} \\
Logic-style & Collapse / designated values & Section~\ref{sec:consequence} \\
Comparison & Two views contrasted & Section~\ref{sec:comparison} \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\noindent
The paper also extends to first order via graded predicates
(Section~\ref{sec:predicates})---functions from a domain to $[0,1]$,
generalizing fuzzy set membership~\cite{zadeh1965}---with inf/sup
quantifiers satisfying duality.
The self-referential equation $c = \negc c$ (as would arise from
unrestricted comprehension) yields a fixed point ($\tfrac{1}{2}$) rather
than a paradox.

\textbf{Contributions.}
\begin{enumerate}[label=(\arabic*)]
\item The two-views framing: Cred can be interpreted probability-style
(credences propagate, uncertainty remains graded) or logic-style
(collapse to discrete values, validity via designation).
The views are complementary, with the collapse as the bridge
(Section~\ref{sec:comparison}).
\item Probability-style reasoning: joint-parametrized valuations,
Bayesian/Jeffrey update, Fr\'echet bounds constraining but not determining
the joint, and zero-evidence underdetermination connecting to imprecise
probability (Sections~\ref{sec:valuations}, \ref{sec:update}).
\item Logic-style reasoning: designated-value consequence (K$_3$, LP) via
the three-valued collapse, and graded consequence on $[0,1]$.
Explosion fails non-vacuously for LP and graded consequence at
$0 < t \leq \tfrac{1}{2}$; K$_3$ has explosion only vacuously
(Section~\ref{sec:consequence}).
\item Graded predicates with inf/sup quantifiers, quantifier duality,
classical agreement for crisp predicates, and unrestricted comprehension
yielding fixed points rather than paradoxes
(Section~\ref{sec:predicates}).
\item Lean~4 formalization of the above (Section~\ref{sec:lean}).
\end{enumerate}

\textbf{Notation.}
We use the notation of Part~1: $\negc$ for complement ($1-c$),
$\conjc$ for product conjunction ($c_1 \cdot c_2$),
$\disjc$ for De~Morgan disjunction, and
$\condc{A}{B}$ for the conditional credence of $A$ given~$B$.
New notation: $\models_D$ and $\models_t$ for semantic consequence under
designation~$D$ or threshold~$t$; specific instances include
$\models_{\mathrm{K}_3}$ (K$_3$ consequence, designating~$\{1\}$) and
$\models_{\mathrm{LP}}$ (LP consequence, designating~$\{1, \tfrac{1}{2}\}$).
For graded predicates: $\cred(\forall x.\; P(x)) = \inf_x P(x)$ and
$\cred(\exists x.\; P(x)) = \sup_x P(x)$.
As in Part~1, roman-font names ($\mathrm{neg}$, $\mathrm{conj}$,
$\mathrm{disj}$) denote the three-valued (Kleene) operations;
$\negc$, $\conjc$, $\disjc$ denote Cred's $[0,1]$ operations.

%=============================================================================
\section{Valuations}
\label{sec:valuations}
%=============================================================================

A \emph{valuation} assigns credences to propositions---it is the bridge
between the propositional world and the algebraic world.
In many-valued logic~\cite{priest2008}, a valuation maps propositions
to a set of truth values; here the truth values are the full
interval $[0,1]$.

\begin{definition}[Complement-Preserving Valuation]
\label{def:cp-valuation}
Given a type $\alpha$ of propositions with a propositional negation
$\sigma : \alpha \to \alpha$, a \emph{complement-preserving valuation}
is a function $v : \alpha \to [0,1]$ satisfying
\[
v(\sigma(A)) = \negc v(A) = 1 - v(A)
\]
for all $A \in \alpha$.
Lean: \texttt{CpValuation}.
Note that $\sigma$ is not required to be involutive
($\sigma(\sigma(A)) = A$); involutivity is an additional hypothesis
where needed.
\end{definition}

Complement preservation is the minimal constraint connecting the
propositional negation to the algebraic complement.

\subsection{Independent Valuations (Trivial Conditioning)}
\label{subsec:independent}

The simplest valuation computes the joint as the product of marginals---this
corresponds to assuming all propositions are probabilistically independent.
\textbf{Warning}: this makes conditioning trivial, so evidence has no effect
on conclusions.
Independent valuations are a degenerate case, not suitable for interesting
reasoning.

\begin{definition}[Independent Valuation]
\label{def:indep-valuation}
An \emph{independent valuation} is a
complement-preserving valuation equipped with propositional conjunction
$\wedge$ and disjunction $\vee$ such that
\[
v(A \wedge B) = v(A) \conjc v(B), \qquad
v(A \vee B) = v(A) \disjc v(B).
\]
Lean: \texttt{IndepValuation}.
\end{definition}

An independent valuation is a homomorphism from the propositional algebra
to the Cred algebra for all three connectives
(Lean: \texttt{indep\_is\_conj\_homomorphism}).
The product $v(A) \cdot v(B)$ treats propositions as probabilistically
independent---the joint equals the product of marginals.

\textbf{Independence makes conditioning trivial.}
Under independence, $v(A \mid B) = v(A)$---evidence has no effect.
This is because the joint $v(A \land B) = v(A) \cdot v(B)$ factors:
$v(A \mid B) = v(A) \cdot v(B) / v(B) = v(A)$.
This is a degenerate case: no information flows from evidence to conclusion.
For interesting reasoning, we need dependence between propositions.

\begin{example}[Rain and Wet Street]
\label{ex:rain}
Let $R$ = ``it rained'' and $W$ = ``the street is wet,'' with
$v(R) = 0.7$ and $v(W) = 0.9$.
A product-copula valuation gives
$v(R \wedge W) = 0.7 \times 0.9 = 0.63$---the product treats $R$ and $W$
as independent evidence.
But rain causes wet streets, so in practice one expects
$v(R \wedge W) > 0.63$.
\end{example}

\begin{proposition}[Non-Idempotence]
\label{prop:idempotence}
Under an independent valuation, $v(A \wedge A) = v(A)^2 \neq v(A)$
whenever $v(A) \notin \{0, 1\}$.
In particular, if $v(A) = \tfrac{1}{2}$, then
$v(A \wedge A) = \tfrac{1}{4} \neq \tfrac{1}{2}$.
\end{proposition}
\begin{proof}
$v(A \wedge A) = v(A) \conjc v(A) = v(A)^2$.
Part~1's idempotence characterization (Lean: \texttt{conj\_idempotent\_iff})
shows $c \conjc c = c$ iff $c \in \{0, 1\}$.
Lean: \texttt{IndepValuation.conj\_not\_idempotent}.
\end{proof}

This is not a defect but a feature: the product treats two occurrences
of $A$ as independent evidence, so $A \wedge A$ is ``weaker'' than~$A$.
Real propositions are not independent of themselves---this motivates
the alternatives below.

\subsection{Kleene Valuations (Logic-Style Canonical Choice)}
\label{subsec:kleene}

Part~1's uniqueness theorem identifies the \emph{only} truth-functional
joint that yields Bayes-consistent, non-trivial conditioning:
the minimum $j(A,B) = \min(v(A), v(B))$ (maximum positive dependence).

\begin{definition}[Kleene Valuation]
\label{def:kleene-valuation}
A \emph{Kleene valuation} is a complement-preserving valuation with
$v(A \wedge B) = \min(v(A), v(B))$ and
$v(A \vee B) = \max(v(A), v(B))$.
\end{definition}

\begin{proposition}[Kleene Valuations are the Logic-Style Canonical Choice]
\label{prop:kleene-canonical}
Among all truth-functional joint formulas:
\begin{enumerate}[label=(\roman*),nosep]
\item The minimum is \emph{idempotent}: $v(A \wedge A) = \min(v(A), v(A)) = v(A)$.
\item Conditioning is \emph{non-trivial}: $v(A \mid B) = \min(v(A), v(B))/v(B)
  = \min(v(A)/v(B), 1) \neq v(A)$ in general.
\item The system is \emph{Bayes-consistent}: the induced joint is symmetric,
  so the chain rule holds in both directions.
\end{enumerate}
No other fixed formula achieves all three.
\end{proposition}
\begin{proof}
(i) and (ii) are immediate from the definitions.
(iii) follows from Part~1's \texttt{prod\_resid\_bayes\_consistent\_real}.
Part~1's \texttt{min\_copula\_unique} (uniqueness) shows no other fixed formula satisfies
all four requirements (truth-functionality, Bayes consistency, non-triviality,
idempotence).
\end{proof}

This is why K$_3$, LP, and RM3 all use the Kleene lattice operations:
the minimum conjunction is not an arbitrary choice---it is \emph{forced}
by the requirement that truth-functional composition be compatible with
proper Bayesian inference.

Part~1's \texttt{min\_copula\_unique} proves that idempotence plus the
2-increasing property mathematically forces $j = \min$---see Part~1 for details.

\subsection{Joint-Parametrized Valuations (General Case)}
\label{subsec:joint-param}

Independent valuations assume a fixed dependence structure (none).
The general case supplies the joint as external data---this is how
probability works, and it is also how Cred can instantiate different
logical systems.

\textbf{Key insight.}
By parameterizing the joint, Cred becomes a \emph{superset} of both traditions:
\begin{itemize}[nosep]
\item Setting $j(A,B) = v(A) \cdot v(B)$ gives independence (probability-like,
  but trivial conditioning).
\item Setting $j(A,B) = \min(v(A), v(B))$ gives maximum positive dependence
  (this is the Kleene conjunction used by K$_3$, LP, and RM3).
\item Setting $j(A,B) = \max(v(A) + v(B) - 1, 0)$ gives maximum negative
  dependence (the \L ukasiewicz t-norm).
\item Supplying an arbitrary joint from empirical data or domain knowledge
  gives full probability-style reasoning.
\end{itemize}
The joint-parametrized framework unifies these: the dependence structure
is a \emph{choice}, not a commitment.

\begin{definition}[Joint-Parametrized Valuation]
\label{def:joint-valuation}
A \emph{joint-parametrized valuation} is a complement-preserving
valuation equipped with a joint function
$j : \alpha \times \alpha \to [0,1]$ satisfying
$j(A, B) \leq v(A)$ and $j(A, B) \leq v(B)$ (marginal bounds).
Lean: \texttt{JointValuation}.
Symmetry ($j(A,B) = j(B,A)$) is not required; when modelling a
commutative conjunction, one may additionally impose symmetry.
\end{definition}

Unlike copula-based valuations, the joint $j(A,B)$ is not determined by
the marginals $v(A)$ and $v(B)$---it can be any value in the
Fr\'echet-Hoeffding interval (Part~1, \texttt{frechet\_upper}, \texttt{frechet\_lower}):
\[
\max(v(A) + v(B) - 1, 0) \;\leq\; j(A, B) \;\leq\; \min(v(A), v(B)).
\]
The upper bound follows from the marginal-bound axioms; the lower bound
requires complement non-negativity as an additional hypothesis
(Lean: \texttt{joint\_frechet\_upper}, \texttt{joint\_frechet\_lower}).

This is how probability works: the joint $P(A \land B)$ is determined
by the measure, not by the marginals alone.
Cred allows the same approach without requiring a measure.

\begin{example}[Rain and Wet Street, Continued]
\label{ex:rain-joint}
Continuing Example~\ref{ex:rain}: with $v(R) = 0.7$, $v(W) = 0.9$,
one might supply $j(W, R) = 0.65$ from data or domain knowledge.
This gives $v(W \mid R) = j(W,R) / v(R) = 0.65 / 0.7 \approx 0.93$:
knowing it rained raises the credence of a wet street from $0.9$
to $0.93$.
The Fr\'echet bounds require
$\max(0.7 + 0.9 - 1, 0) = 0.6 \leq j(W,R) \leq \min(0.7, 0.9) = 0.7$,
and $0.65$ is within this interval.
\end{example}

When $v(B) > 0$, the chain rule has a unique solution; we write
$v(A \mid B) = j(A, B) / v(B)$ for the conditional credence
(Lean: \texttt{JointValuation.conditioning}, which constructs the
value via Part~1's \texttt{conditioning\_mk}; uniqueness follows
separately from \texttt{conditioning\_unique}).

\subsection{Stochastic Independence Between Domains}
\label{subsec:independence}

Can we partition propositions into domains with different dependence
structures?
Part~1 (\S6.3) shows that independence within a domain forces classical
$\{0,1\}$ credences (unless idempotence is abandoned), and independence
between domains trivializes cross-domain inference.
The unique meeting point extends: any domain with meaningful intermediate
credences and idempotence must use maximum dependence internally
(Part~1: \texttt{independence\_within\_forces\_classical},
\texttt{cross\_world\_independence\_trivial}).

\subsection{Collapse Composition}

Composing a Cred valuation with Part~1's collapse homomorphism yields a
three-valued (Kleene) valuation.

\begin{proposition}[Collapse Preserves Connectives]
\label{prop:collapse-val}
Let $v$ be a product-copula valuation and $\kappa = \mathrm{collapse} \circ v$.
Then $\kappa$ preserves negation, conjunction, and disjunction:
\begin{align*}
\kappa(\sigma(A)) &= \mathrm{neg}(\kappa(A)), \\
\kappa(A \wedge B) &= \mathrm{conj}(\kappa(A), \kappa(B)), \\
\kappa(A \vee B) &= \mathrm{disj}(\kappa(A), \kappa(B)).
\end{align*}
Lean: \texttt{collapse\_val\_neg}, \texttt{collapse\_val\_conj},
\texttt{collapse\_val\_disj}.
\end{proposition}
\begin{proof}
Each equation follows by rewriting with the valuation's compatibility
axiom and then applying Part~1's collapse homomorphism theorem
(\texttt{collapse\_neg}, \texttt{collapse\_conj}, \texttt{collapse\_disj}).
\end{proof}

This is the mechanism by which Cred connects to K$_3$, LP, and RM3:
the collapse maps any Cred valuation to a Kleene valuation, and
the choice of designated values then determines the consequence relation.

\subsection{Comparison}

\begin{center}
\small
\begin{tabular}{llllll}
\toprule
\textbf{System} & \textbf{Target} & \textbf{Joint} & \textbf{Idemp.?} & \textbf{Bayes?} & \textbf{Conditioning} \\
\midrule
Classical logic & $\{0, 1\}$ & $\min$ & Yes & Yes & Material \\
Probability & $[0,1]$ & Measure & Yes & Yes & Bayes (non-trivial) \\
Product fuzzy & $[0,1]$ & Product & No & Yes & Trivial ($a$) \\
Cred (indep.) & $[0,1]$ & Product & No & Yes & Trivial ($a$) \\
\textbf{Cred (Kleene)} & $[0,1]$ & $\min$ & Yes & Yes & \textbf{Non-trivial} \\
Cred (joint) & $[0,1]$ & External & By choice & Yes & Chain rule \\
G\"odel fuzzy & $[0,1]$ & $\min$ & Yes & No & Not Bayes-cons.\ \\
\L ukasiewicz & $[0,1]$ & $\max(a{+}b{-}1,0)$ & No & No & Not Bayes-cons.\ \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\noindent
The table highlights the \emph{unique meeting point}: only maximum positive
dependence ($\min$) yields a truth-functional joint that is simultaneously
idempotent, Bayes-consistent, and non-trivial.
Independence (product) is Bayes-consistent but trivial; G\"odel and
\L ukasiewicz are truth-functional but not Bayes-consistent.

The constant valuation $v(A) = \tfrac{1}{2}$ for all $A$, with
the identity function as propositional negation ($\sigma(A) = A$ for
all $A$), is complement-preserving because
$\tfrac{1}{2} = 1 - \tfrac{1}{2}$ (the negation fixed point from Part~1).
This is a degenerate example (every proposition equals its own negation),
used as a technical witness.
Its collapse to three values gives a constant-$\tfrac{1}{2}$
three-valued valuation.
(The K$_3$ no-tautology proof constructs the constant-$\tfrac{1}{2}$
function directly at the three-valued level, without using the
collapse.)
Lean: \texttt{CpValuation.constant\_half}.

%=============================================================================
\section{Consequence Relations (Logic-Style View)}
\label{sec:consequence}
%=============================================================================

This section develops the logic-style view: collapse credences to discrete
values and determine validity via designated values.
A \emph{consequence relation}~\cite{priest2008} specifies which inferences
are valid: premises entail conclusion $B$ when every valuation
accepting the premises also accepts $B$.
The key result is that explosion (from a contradiction, anything follows)
fails non-vacuously for LP and graded consequence at
$t \leq \tfrac{1}{2}$, while K$_3$ has explosion only vacuously.

\subsection{Designated Values}

The Kleene lattice $\{0, \tfrac{1}{2}, 1\}$ admits three standard
consequence relations, differing only in which values are
\emph{designated} (accepted as ``holding'').

\begin{definition}[K$_3$, LP Designation]
\label{def:designated}
K$_3$ designates $\{1\}$ (only certainty counts).
LP designates $\{1, \tfrac{1}{2}\}$ (certainty and the
intermediate value both count).
RM3 uses the same designation as LP ($\{1, \tfrac{1}{2}\}$) with
a distinct implication truth table; RM3-valid entailments satisfy the
variable-sharing property (premises and conclusion share a
propositional variable)~\cite{priest2008}.
Lean: \texttt{isDesignatedK3}, \texttt{isDesignatedLP}.
(RM3 designation is not formalized; the relevance condition is
syntactic.)
\end{definition}

\begin{definition}[Three-Valued Consequence]
\label{def:three-val-consequence}
Premises $A_1, \ldots, A_n$ entail conclusion $B$ under designation
$D$ when every three-valued valuation $v$ satisfying
$v(A_i) \in D$ for all $i$ also satisfies $v(B) \in D$:
\[
A_1, \ldots, A_n \models_D B \iff
\forall v.\;(\forall i.\; D(v(A_i))) \Rightarrow D(v(B)).
\]
Lean: \texttt{threeValConsequence}.
As in the Lean formalization, $v$ ranges over all functions
$\alpha \to \{0, \tfrac{1}{2}, 1\}$, not just homomorphic
(truth-table-respecting) valuations.
\end{definition}

\subsection{K$_3$: Paracomplete}

A logic is \emph{paracomplete} if it lacks some classically valid
tautologies~\cite{priest2008}.
K$_3$ (Kleene's strong three-valued logic~\cite{kleene1952}) designates
only~$1$ and is maximally paracomplete: it has no tautologies at all.

\begin{proposition}[K$_3$ Has No Tautologies]
\label{prop:k3-no-taut}
Under K$_3$ consequence (for nonempty proposition types),
no formula is a tautology:
$\not\models_{\mathrm{K}_3} A$ for any $A$.
\end{proposition}
\begin{proof}
The constant-$\tfrac{1}{2}$ valuation assigns every formula the value
$\tfrac{1}{2}$, which is not K$_3$-designated.
Lean: \texttt{k3\_no\_tautology}.
\end{proof}

\begin{proposition}[K$_3$: Excluded Middle Fails]
\label{prop:k3-em}
There exists a value $c$ such that $\mathrm{disj}(c, \mathrm{neg}(c))$ is not
K$_3$-designated.
It follows from Proposition~\ref{prop:k3-no-taut} that excluded middle
is not a K$_3$ tautology.
\end{proposition}
\begin{proof}
At $c = \tfrac{1}{2}$: $\mathrm{disj}(\tfrac{1}{2}, \tfrac{1}{2}) = \max(\tfrac{1}{2}, \tfrac{1}{2}) = \tfrac{1}{2}$,
which is not in $\{1\}$.
Lean: \texttt{k3\_excluded\_middle\_fails}.
\end{proof}

\subsection{LP: Paraconsistent}

A logic is \emph{paraconsistent} if a contradiction does not entail
every proposition---i.e., explosion fails~\cite{priest2008}.
LP (the Logic of Paradox) designates $\{1, \tfrac{1}{2}\}$: both
certainty and the intermediate value count as ``holding.''

\begin{proposition}[LP: Excluded Middle Holds]
\label{prop:lp-em}
For every three-valued value $c$, $\mathrm{disj}(c, \mathrm{neg}(c))$ is
LP-designated.
\end{proposition}
\begin{proof}
Case analysis: $\mathrm{disj}(0, 1) = 1$; $\mathrm{disj}(\tfrac{1}{2}, \tfrac{1}{2}) =
\tfrac{1}{2}$; $\mathrm{disj}(1, 0) = 1$.
All are in $\{1, \tfrac{1}{2}\}$.
Lean: \texttt{lp\_excluded\_middle}.
\end{proof}

\begin{proposition}[LP: Explosion Fails]
\label{prop:lp-no-explosion}
There exists a three-valued function $v$ and propositions $A$, $B$
such that both $v(A)$ and $\mathrm{neg}(v(A))$ are LP-designated but
$v(B)$ is not.
\end{proposition}
\begin{proof}
Set $v(A) = \tfrac{1}{2}$ and $v(B) = 0$.
Then $\mathrm{neg}(\tfrac{1}{2}) = \tfrac{1}{2}$, so both $v(A)$ and
$\mathrm{neg}(v(A))$ are LP-designated, but $v(B) = 0$ is not.
Lean: \texttt{lp\_no\_explosion}.
\end{proof}

\subsection{Graded Consequence}

On the full $[0,1]$, we can define consequence without collapsing to
three values.

\begin{definition}[Graded Consequence]
\label{def:graded-consequence}
For a threshold $t \in [0,1]$, \emph{graded consequence} holds when
every valuation making all premises $\geq t$ also makes the conclusion
$\geq t$:
\[
A_1, \ldots, A_n \models_t B \iff
\forall v \colon \alpha \to [0,1].\; (\forall i.\; v(A_i) \geq t) \Rightarrow v(B) \geq t.
\]
Lean: \texttt{gradedConsequence}.
Here $v$ ranges over all functions $\alpha \to [0,1]$, not just
complement-preserving valuations.
This makes the consequence relation weaker (more functions must be
checked) but the no-explosion existential easier (the witness need
not be complement-preserving); see Theorem~\ref{thm:no-explosion}.
\end{definition}

\begin{proposition}[Graded at $t=1$ Is Strict]
\label{prop:graded-strict}
Graded consequence at $t = 1$ holds iff every valuation making all
premises equal to~$1$ makes the conclusion equal to~$1$.
Lean: \texttt{graded\_at\_one\_is\_strict}.
\end{proposition}

\subsection{No-Explosion Theorem}

The key structural result: Cred's unconstrained conditioning at zero
forces explosion to fail.

\begin{example}[Explosion Fails]
\label{ex:no-explosion}
In classical logic, if both ``it rained'' and ``it did not rain'' hold,
then every proposition follows---including ``the moon is made of cheese.''
In Cred at threshold $t = 0.3$: set $v(R) = 0.5$, so
$\negc v(R) = 0.5$; both exceed $0.3$.
But $v(\text{moon-cheese}) = 0$, which is below $0.3$.
The contradiction in $R$ does not force acceptance of an unrelated
proposition.
\end{example}

\begin{theorem}[No Explosion]
\label{thm:no-explosion}
For any threshold $0 < t \leq \tfrac{1}{2}$, graded consequence at~$t$
does not have explosion: there exist propositions $A$, $B$ and a
function $v$ with $v(A) \geq t$, $1 - v(A) \geq t$, and $v(B) < t$.
The condition is on the algebraic complement $1 - v(A)$, which is
strictly weaker than requiring $v(\sigma(A)) \geq t$ for a
complement-preserving valuation; since any complement-preserving
witness is also a valid unrestricted witness, the result implies
a~fortiori that explosion fails for complement-preserving valuations
as well.
\end{theorem}
\begin{proof}
Set $v(A) = \tfrac{1}{2}$, $v(B) = 0$.
Then $1 - v(A) = \tfrac{1}{2}$,
so both $v(A)$ and $1 - v(A)$ equal~$\tfrac{1}{2} \geq t$.
But $v(B) = 0 < t$.
Lean: \texttt{no\_explosion\_at\_half}, \texttt{graded\_no\_explosion}.
\end{proof}

\begin{remark}[Algebraic Basis]
The algebraic ingredient is Part~1's
\texttt{conditioning\_zero\_any}: when evidence has credence~$0$, the
chain rule $c \cdot 0 = 0$ is satisfied by any $c \in [0,1]$
(re-exported in Consequence.lean as
\texttt{unconstrained\_blocks\_explosion}).
The consequence-level Lean proofs (\texttt{lp\_no\_explosion},
\texttt{graded\_no\_explosion}) construct direct witness valuations
without invoking the conditioning machinery; the connection to
\texttt{conditioning\_zero\_any} is conceptual, not proof-level.
Informally, classical consequence (two-valued designation on
$\{0,1\}$) has explosion; it arises from a setting where conditioning
at zero is resolved by the measure, not left unconstrained.
\end{remark}

\subsection{Comparison}

\begin{center}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{System} & \textbf{Designated} & \textbf{Tautologies?}
  & \textbf{Excl.\ middle?} & \textbf{Explosion?} \\
\midrule
Classical & $\{1\}$ (two-valued) & Yes$^\|$ & Yes$^\|$ & Yes \\
K$_3$ & $\{1\}$ & No & No & Holds vacuously$^\S$ \\
LP & $\{1, \tfrac{1}{2}\}$ & Yes$^\|$ & Yes$^\|$ & No \\
RM3$^\dagger$ & $\{1, \tfrac{1}{2}\}$+relev. & Yes$^\|$ & Yes$^\|$ & No \\
Graded ($t \leq \tfrac{1}{2}$) & $\{c \geq t\}$ & $^\ddagger$ & $^\ddagger$ & No \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\smallskip
\noindent
$^\|$\,Tautology and excluded-middle entries for Classical, LP, and
RM3 hold under the standard restriction to homomorphic
(truth-table-respecting) valuations.
Definition~\ref{def:three-val-consequence} quantifies over all
functions $\alpha \to \{0, \tfrac{1}{2}, 1\}$; under that definition,
no formula is a tautology in any of these systems (the constant-$0$
function defeats every formula).
The K$_3$ no-tautology result and the LP/graded no-explosion
results are proved under the all-functions definition.\\
$^\S$\,K$_3$ has explosion vacuously: no valuation makes both $v(A)$
and $\mathrm{neg}(v(A))$ K$_3$-designated, so the premise of explosion
is never fulfilled.\\
$^\dagger$\,RM3 properties are well-known~\cite{priest2008}; they are not
formalized in Lean.\\
$^\ddagger$\,Tautologies and excluded middle for graded consequence depend
on the threshold and are not formalized.
At $t = 0$, trivially every formula is a tautology (all credences
are $\geq 0$ by definition); at $t > 0$, the constant-$0$ valuation
assigns every formula value~$0 < t$, so no formula is a tautology.
At the algebraic level, excluded middle holds whenever
$t \leq \tfrac{3}{4}$ for complement-preserving valuations
(Part~1, \texttt{certainty\_ge\_three\_quarters}; not connected to
graded consequence at the Lean level): expanding the De~Morgan
definition, $v(A) \disjc \negc v(A) = 1 - v(A)(1 - v(A)) \geq
\tfrac{3}{4}$ for all $v(A) \in [0,1]$, since
$v(A)(1 - v(A)) \leq \tfrac{1}{4}$.
This does not extend to unrestricted graded consequence, where $v$
need not be complement-preserving.

%=============================================================================
\section{Comparing the Two Views}
\label{sec:comparison}
%=============================================================================

\begin{center}
\small
\begin{tabular}{lll}
\toprule
\textbf{Aspect} & \textbf{Probability-style} & \textbf{Logic-style} \\
\midrule
Output of inference & Credence in $[0,1]$ & Yes/no (designated or not) \\
Uncertainty & Propagates through reasoning & Collapses at final judgment \\
What it answers & ``How confident should I be?'' & ``Is this valid to conclude?'' \\
Core mechanism & Chain rule / update & Designated-value checking \\
At zero evidence & Underdetermined (interval) & N/A (collapse applies before) \\
Explosion & N/A (no binary validity) & Fails for LP, graded $t \leq \tfrac{1}{2}$ \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\noindent
The views are complementary: probability-style outputs graded confidence,
logic-style outputs binary validity.
The collapse homomorphism bridges them.

%=============================================================================
\section{Update Rules}
\label{sec:update}
%=============================================================================

This section develops the probability-style view in detail.
The chain rule is the inference engine: it constrains how credences
relate, and update rules apply this constraint dynamically.

\subsection{The Chain Rule as Inference}

The chain rule $\condc{A}{B} \cdot \cred(B) = \cred(A \land B)$ relates
three quantities: the conditional credence, the evidence credence, and
the joint credence.
Given any two, the third is determined (when evidence is positive) or
constrained (when evidence is zero).
This is Bayesian-style inference without requiring additivity or a
measure space.

\textbf{From evidence and joint to conditional.}
Given $v(B) > 0$ and $j(A,B) = v(A \land B)$, the chain rule gives
$v(A \mid B) = j(A,B) / v(B)$.
This is how conditioning works: evidence and joint determine the
conditional.

\textbf{From conditional and evidence to joint.}
If we know $v(A \mid B)$ and $v(B)$, then $j(A,B) = v(A \mid B) \cdot v(B)$.
This is how the chain rule propagates credences forward through a
reasoning chain.

\textbf{Fr\'echet bounds constrain but do not determine.}
The Fr\'echet-Hoeffding bounds (Part~1) constrain which joints are coherent
with the marginals, but do not select a unique joint.

\subsection{Bayesian Conditionalization}

\begin{definition}[Bayesian Update]
\label{def:bayesian-update}
Given a joint-parametrized valuation $v$ and evidence $B$ with
$v(B) > 0$, the \emph{Bayesian posterior} is
\[
v'(A) = \frac{j(A, B)}{v(B)}
\]
where $j(A, B)$ is the joint credence.
Lean: \texttt{bayesianUpdate} (single proposition),
\texttt{bayesianUpdateValuation} (returns a function
$\alpha \to \mathrm{Credence}$, not necessarily complement-preserving).
\end{definition}

The Bayesian update is the conditioning structure from Part~1 applied
propositionwise: $v'(A) \cdot v(B) = j(A, B)$ at every~$A$.
The posterior is well-defined ($j(A,B)/v(B) \in [0,1]$) because
$j(A,B) \leq v(B)$ by the marginal bound on the second argument.
Note that $v'$ need not be a complement-preserving valuation.

\begin{example}[Rain and Wet Street: Update]
\label{ex:rain-update}
From Example~\ref{ex:rain-joint}: $v(R) = 0.7$, $v(W) = 0.9$,
$j(W, R) = 0.65$.
You observe that it rained ($B = R$).
The Bayesian posterior for $W$ is
$v'(W) = j(W,R) / v(R) = 0.65 / 0.7 \approx 0.93$:
your credence in a wet street rises from $0.9$ to $0.93$.
\end{example}

\begin{proposition}[Chain-Rule Preservation]
\label{prop:chain-rule-preservation}
The Bayesian posterior satisfies the chain rule by construction:
$v'(A) \cdot v(B) = j(A, B)$.
Lean: \texttt{bayesianUpdate\_chain\_rule},
\texttt{bayesianUpdateValuation\_chain\_rule}.
\end{proposition}
\begin{proof}
Immediate from the definition via \texttt{conditioning\_mk}.
\end{proof}

\begin{proposition}[Uniqueness]
\label{prop:bayesian-unique}
The Bayesian update is the unique credence satisfying the chain rule
with the given evidence and joint.
Lean: \texttt{bayesianUpdate\_unique} (via Part~1's
\texttt{conditioning\_unique}).
\end{proposition}

\subsection{Jeffrey Conditionalization}

Jeffrey~\cite{jeffrey1965} generalized Bayesian conditionalization to
uncertain evidence.
Instead of learning $B$ with certainty, one learns that $B$ has a new
credence $q$:
\[
v'(A) = v(A \mid B) \cdot q + v(A \mid \negc B) \cdot (1 - q).
\]
This is a weighted mixture of the conditional-on-$B$ and
conditional-on-$\negc B$ valuations.
Determining $v(A \mid B)$ and $v(A \mid \sigma(B))$ from a joint-parametrized
valuation requires $v(B) > 0$ and $v(\sigma(B)) > 0$ (i.e.,
$v(B) < 1$, by complement preservation), respectively.
Here $v(A \mid B) = j(A, B) / v(B)$ as defined in
Section~\ref{sec:valuations}.
(We write $\negc B$ for $\sigma(B)$ when the propositional reading is
clear from context.)

\begin{definition}[Jeffrey Update]
\label{def:jeffrey-update}
Given conditional credences $c_B = v(A \mid B)$ and
$c_{\negc B} = v(A \mid \negc B)$ in $[0,1]$, and a new evidence
credence $q \in [0,1]$, the Jeffrey update is
\[
v'(A) = c_B \cdot q + c_{\negc B} \cdot (1 - q).
\]
Lean: \texttt{jeffreyUpdate}.
\end{definition}

\begin{proposition}[Jeffrey Bounds]
\label{prop:jeffrey-bounds}
The Jeffrey update satisfies $0 \leq v'(A) \leq 1$.
\end{proposition}
\begin{proof}
Since $c_B, c_{\negc B} \in [0,1]$ and $q, (1-q) \geq 0$ with
$q + (1-q) = 1$, the expression $c_B \cdot q + c_{\negc B} \cdot (1-q)$
is a convex combination and thus in $[0,1]$.
Lean: verified by the type constraints of \texttt{jeffreyUpdate}.
\end{proof}

\begin{proposition}[Jeffrey at Extreme Evidence]
\label{prop:jeffrey-bayesian}
At $q = 1$, the Jeffrey update returns~$c_B$.
When $c_B = v(A \mid B)$, this is the Bayesian posterior.
At $q = 0$, it returns~$c_{\negc B}$.
Lean: \texttt{jeffreyUpdate\_certain}, \texttt{jeffreyUpdate\_zero}.
\end{proposition}

\subsection{Zero-Evidence Underdetermination}

When $v(B) = 0$, the chain rule $\condc{A}{B} \cdot 0 = 0$ is satisfied
by any conditional (Part~1, \texttt{conditioning\_zero\_any}).
The conditional $v(A \mid B)$ is therefore underdetermined, and since both
Bayesian and Jeffrey update take this conditional as input, neither
produces a unique posterior at zero evidence.

\begin{proposition}[Underdetermination]
\label{prop:underdetermination}
For any credence $c$, there exists a conditioning structure with
evidence~$0$ and conditional~$c$.
That is, the conditional input to Bayesian and Jeffrey update is
unconstrained when evidence is zero.
Lean: \texttt{bayesian\_underdetermined\_at\_zero},
\texttt{jeffrey\_underdetermined\_at\_zero} (both re-export
\texttt{conditioning\_zero\_any}).
\end{proposition}

This connects Cred to \emph{imprecise probability}~\cite{walley1991}:
when conditioning at zero is underdetermined, the honest response
is to report the full interval $[0,1]$ as the set of compatible
posteriors, rather than selecting a single value.
Levi~\cite{levi1980} argued that such indeterminacy is a genuine
epistemic state, not a failure of the formalism.

\subsection{Comparison}

\begin{center}
\small
\begin{tabular}{llll}
\toprule
\textbf{Update rule} & \textbf{Evidence type} &
  \textbf{Requirements beyond Cred} & \textbf{Zero evidence?} \\
\midrule
Bayesian & $B$ certain & Full joints, $v(B) > 0$ & Underdetermined \\
Jeffrey & $B$ has credence $q$ & Full joints, $v(B) > 0$, $v(\negc B) > 0^\star$ & Underdetermined \\
Disintegration & $B$ certain & Measure space & Resolved \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\smallskip
\noindent
$^\star$\,$v(\negc B) > 0$ is needed to determine the conditional input
$v(A \mid \negc B)$, not for the Jeffrey formula itself.

\smallskip
\noindent
The chain rule constrains what updates are coherent; it does not select one.
Each update rule requires additional structure beyond the algebra.

%=============================================================================
\section{Graded Predicates and Quantifiers}
\label{sec:predicates}
%=============================================================================

The extension from propositional to first-order requires predicates
varying over a domain and quantifiers aggregating over that domain.
A graded predicate---a function $P : X \to [0,1]$---generalizes
Zadeh's fuzzy set membership~\cite{zadeh1965}: $P(x)$ is the degree
to which $x$ has property~$P$.

\subsection{Graded Predicates}

\begin{definition}[Graded Predicate]
\label{def:graded-predicate}
A \emph{graded predicate} over a domain $X$ is a function
$P : X \to [0,1]$.
The value $P(x)$ is the degree to which $x$ satisfies~$P$.
Lean: \texttt{GradedPredicate}.
\end{definition}

Crisp predicates (range $\subseteq \{0, 1\}$) are the special case
corresponding to classical sets.

The Cred operations extend pointwise (expanding the De~Morgan
definition for disjunction):
\begin{align*}
(\negc P)(x) &= 1 - P(x) &\text{(complement)} \\
(P \conjc Q)(x) &= P(x) \cdot Q(x) &\text{(conjunction)} \\
(P \disjc Q)(x) &= P(x) + Q(x) - P(x) \cdot Q(x) &\text{(disjunction)}
\end{align*}
Lean: \texttt{GradedPredicate.neg}, \texttt{GradedPredicate.conj},
\texttt{GradedPredicate.disj}.

All Part~1 algebraic laws hold pointwise:

\begin{proposition}[Pointwise Laws]
\label{prop:pointwise}
For graded predicates $P$, $Q$, $R$:
\begin{enumerate}[label=(\roman*)]
\item $\negc\negc P = P$ (double negation),
\item $P \conjc Q = Q \conjc P$ and $P \disjc Q = Q \disjc P$ (commutativity),
\item $(P \conjc Q) \conjc R = P \conjc (Q \conjc R)$ and
      $(P \disjc Q) \disjc R = P \disjc (Q \disjc R)$ (associativity),
\item $\negc(P \conjc Q) = \negc P \disjc \negc Q$ and
      $\negc(P \disjc Q) = \negc P \conjc \negc Q$ (De~Morgan).
\end{enumerate}
Lean: \texttt{GradedPredicate.neg\_neg}, \texttt{GradedPredicate.conj\_comm},
\texttt{GradedPredicate.conj\_assoc}, \texttt{GradedPredicate.disj\_comm},
\texttt{GradedPredicate.disj\_assoc},
\texttt{GradedPredicate.de\_morgan\_conj},
\texttt{GradedPredicate.de\_morgan\_disj}.
\end{proposition}
\begin{proof}
Each law follows from the corresponding Part~1 law applied at every point.
\end{proof}

\subsection{Quantifiers}

\begin{definition}[Quantifiers]
\label{def:quantifiers}
For a graded predicate $P : X \to [0,1]$ over a nonempty domain~$X$:
\begin{align*}
\cred(\forall x.\; P(x)) &= \inf_{x \in X} P(x), \\
\cred(\exists x.\; P(x)) &= \sup_{x \in X} P(x).
\end{align*}
Lean: \texttt{credInf} / \texttt{forall'}, \texttt{credSup} / \texttt{exists'}.
\end{definition}

The infimum and supremum are well-defined because credence values are
bounded in $[0,1]$.
All results in this subsection require $X$ to be nonempty
(Lean: \texttt{[Nonempty} $\alpha$\texttt{]}).
The universal quantifier is bounded above by any instance
(Lean: \texttt{forall\_le}); the existential quantifier is bounded below
by any instance (Lean: \texttt{le\_exists}).

\begin{theorem}[Quantifier Duality]
\label{thm:quantifier-duality}
For $P$ over a nonempty domain:
\[
1 - \inf_x P(x) = \sup_x (1 - P(x)).
\]
Equivalently, in credence notation:
$\negc\cred(\forall x.\; P(x)) = \cred(\exists x.\; \negc P(x))$,
where $\negc$ on the left is complement of a value and $\negc P$ on the
right is pointwise predicate negation (Definition~\ref{def:graded-predicate}).
Lean: \texttt{quantifier\_duality\_val} (formalizes the real-valued
identity; the credence-level reading follows by the definitions of
\texttt{credInf} and \texttt{credSup}).
\end{theorem}
\begin{proof}
For the $\geq$ direction: for each $x$, $\inf P \leq P(x)$, so
$1 - P(x) \leq 1 - \inf P$.
Taking the supremum gives $\sup(1 - P) \leq 1 - \inf P$.
For the $\leq$ direction: for each $x$, $1 - P(x) \leq \sup(1 - P)$,
so $P(x) \geq 1 - \sup(1 - P)$ for all $x$,
hence $\inf P \geq 1 - \sup(1 - P)$, i.e., $1 - \inf P \leq \sup(1 - P)$.
\end{proof}

\subsection{Classical Agreement}

\begin{proposition}[Classical Agreement]
\label{prop:classical-agreement}
For crisp predicates (range $\subseteq \{0, 1\}$):
\begin{enumerate}[label=(\roman*)]
\item $\inf_x P(x) = 0$ iff some $x$ has $P(x) = 0$
  (Lean: \texttt{crisp\_inf\_zero\_iff}),
\item $\sup_x P(x) = 1$ iff some $x$ has $P(x) = 1$
  (Lean: \texttt{crisp\_sup\_one\_iff}).
\end{enumerate}
\end{proposition}
\begin{proof}
For (i): if $P(x) = 0$ for some $x$, then $\inf P \leq P(x) = 0$,
and $\inf P \geq 0$ since all credences are non-negative.
Conversely, if $\inf P = 0$ but $P(x) \neq 0$ for all $x$, then
$P(x) = 1$ for all $x$ (by crispness); since the domain is nonempty,
$\inf P = 1$, contradiction.
Symmetric for (ii).
\end{proof}

These are the ``boundary witness'' cases of classical agreement.
The complementary cases ($\inf P = 1$ iff $P(x) = 1$ for all $x$;
$\sup P = 0$ iff $P(x) = 0$ for all $x$) hold by similar arguments
but are not formalized.
Together, they show that for crisp predicates, $\inf = \forall$ and
$\sup = \exists$ in the classical sense.

\subsection{Unrestricted Comprehension}

\begin{theorem}[Negation Fixed-Point Resolution]
\label{thm:russell}
If a credence $c$ satisfies $c = \negc c$, then $c = \tfrac{1}{2}$.
\end{theorem}
\begin{proof}
The equation $c = 1 - c$ has a unique solution at $c = \tfrac{1}{2}$
(Part~1, \texttt{neg\_fixed\_point\_unique}).
Lean: \texttt{russell\_fixed\_point}.
\end{proof}

In particular, any self-referential predicate $R$ satisfying
$R(R) = \negc R(R)$ would have $R(R) = \tfrac{1}{2}$.
Russell's predicate $R = \{x : x \notin x\}$ applied to itself gives
$R(R) = \negc R(R) = \tfrac{1}{2}$: the set half-contains itself.
The algebraic core---$c = \negc c$ implies $c = \tfrac{1}{2}$---is
formalized in Lean; the self-application mechanism (unrestricted
comprehension producing the equation $R(R) = \negc R(R)$) requires
set-theoretic infrastructure not formalized here.
This connects Cred to paraconsistent set theory~\cite{brady2006,weber2010},
where comprehension is retained and contradictions are tolerated.
In Cred, the ``contradiction'' has a precise numerical value:
$\tfrac{1}{2}$, the unique negation fixed point.

The relationship between Cred and t-norm fuzzy logics (G\"odel, product,
\L ukasiewicz) is developed in Part~1, \S6.
The key result: the product t-norm is the unique continuous t-norm compatible
with chain-rule conditioning (unique solution when evidence is positive), and
when the joint is set to the minimum (maximum positive dependence), Cred
coincides with Kleene valuations---the foundation of K$_3$, LP, and RM3.

%=============================================================================
\section{Related Work}
\label{sec:related}
%=============================================================================

For related work on the core algebra, primitive conditioning, t-norm fuzzy
logics, and copulas, see Part~1, \S7.

\textbf{Adams' Probability Logic.}
Adams~\cite{adams1965,adams1998} showed that conditional probability
propagates uncertainty through valid classical inferences: if all premises
have probability $\geq 1 - \varepsilon$, then the conclusion has
probability $\geq 1 - n\varepsilon$ (for $n$ premises).
Whether analogous bounds hold for Cred valuations without additivity
is an open problem.

\textbf{Jeffrey Conditionalization.}
Jeffrey~\cite{jeffrey1965,jeffrey1983} generalized Bayesian updating to
uncertain evidence via a mixture formula.
Section~\ref{sec:update} implements Jeffrey update on Cred valuations.

\textbf{Imprecise Probability.}
Walley~\cite{walley1991} developed statistical reasoning with sets of
probability measures rather than a single one.
Levi~\cite{levi1980} argued that indeterminate credences are genuine
epistemic states.
Zero-evidence underdetermination connects directly to imprecise probability.

\textbf{Paraconsistent Set Theory.}
Brady~\cite{brady2006} and Weber~\cite{weber2010} developed set theories
with unrestricted comprehension.
Graded predicates (Section~\ref{sec:predicates}) offer a complementary
approach: the Russell predicate yields $\tfrac{1}{2}$.

%=============================================================================
\section{Lean Formalization}
\label{sec:lean}
%=============================================================================

In addition to Part~1's \texttt{Cred.Basic} module, the formalization
includes:
\texttt{Cred/Valuation.lean} (valuations, collapse composition,
Fr\'echet bounds);
\texttt{Cred/Consequence.lean} (K$_3$/LP designation, no-explosion
theorems);
\texttt{Cred/Update.lean} (Bayesian/Jeffrey update);
\texttt{Cred/Predicate.lean} (graded predicates, quantifier duality,
Russell fixed point).

Key theorems: \texttt{k3\_no\_tautology}, \texttt{lp\_no\_explosion},
\texttt{graded\_no\_explosion}, \texttt{bayesianUpdate\_unique},
\texttt{quantifier\_duality\_val}, \texttt{russell\_fixed\_point}.
All are machine-checked in Lean~4 with Mathlib 4.16.0.
Verification: \texttt{cd lean \&\& lake build}.

%=============================================================================
\section{Conclusion}
\label{sec:conclusion}
%=============================================================================

Part~2 developed two complementary views of Cred:
probability-style (chain rule as inference, Bayesian/Jeffrey update,
zero-evidence underdetermination) and logic-style (collapse to Kleene lattice,
designated-value consequence).
Explosion fails non-vacuously for LP and graded consequence at
$t \leq \tfrac{1}{2}$.
Graded predicates extend to first order with quantifier duality;
unrestricted comprehension yields $\tfrac{1}{2}$ rather than paradoxes.

\textbf{What Part~2 does not provide.}
Proof theory and self-hosting remain for Part~3.

\textbf{Future work.}
Proof theory, Adams-type bounds without additivity, and connections to
probabilistic programming.

%=============================================================================
% Bibliography
%=============================================================================

\begin{thebibliography}{99}

\bibitem{adams1965}
Adams, E.W.
\newblock The logic of conditionals.
\newblock \emph{Inquiry}, 8:166--197, 1965.

\bibitem{adams1998}
Adams, E.W.
\newblock \emph{A Primer of Probability Logic}.
\newblock CSLI Publications, Stanford, 1998.

\bibitem{anderson1975}
Anderson, A.R. and Belnap, N.D.
\newblock \emph{Entailment: The Logic of Relevance and Necessity}, Vol.~I.
\newblock Princeton University Press, 1975.

\bibitem{brady2006}
Brady, R.
\newblock \emph{Universal Logic}.
\newblock CSLI Lecture Notes, vol.~109. CSLI Publications, 2006.

\bibitem{definetti1936}
De Finetti, B.
\newblock La logique de la probabilit\'e.
\newblock In \emph{Actes du Congr\`es International de Philosophie Scientifique},
IV:1--9. Hermann, Paris, 1936.

\bibitem{definetti1937}
De Finetti, B.
\newblock La pr\'evision: ses lois logiques, ses sources subjectives.
\newblock \emph{Annales de l'Institut Henri Poincar\'e}, 7(1):1--68, 1937.

\bibitem{dubois2001}
Dubois, D. and Prade, H.
\newblock Possibility theory, probability theory and multiple-valued logics:
A clarification.
\newblock \emph{Annals of Mathematics and Artificial Intelligence}, 32:35--66,
2001.

\bibitem{egre2021}
\'Egr\'e, P., Rossi, L., and Sprenger, J.
\newblock De Finettian logics of indicative conditionals, Part~I:
Trivalent semantics and validity.
\newblock \emph{Journal of Philosophical Logic}, 50:187--213, 2021.

\bibitem{esteva2000}
Esteva, F., Godo, L., H\'ajek, P., and Navara, M.
\newblock Residuated fuzzy logics with an involutive negation.
\newblock \emph{Archive for Mathematical Logic}, 39(2):103--124, 2000.

\bibitem{goguen1969}
Goguen, J.A.
\newblock The logic of inexact concepts.
\newblock \emph{Synthese}, 19(3--4):325--373, 1969.

\bibitem{godel1932}
G\"odel, K.
\newblock Zum intuitionistischen Aussagenkalk\"ul.
\newblock \emph{Anzeiger der Akademie der Wissenschaften in Wien},
69:65--66, 1932.

\bibitem{gottwald2001}
Gottwald, S.
\newblock \emph{A Treatise on Many-Valued Logics}.
\newblock Studies in Logic and Computation, vol.~9. Research Studies Press, 2001.

\bibitem{hajek1998}
H\'ajek, P.
\newblock \emph{Metamathematics of Fuzzy Logic}.
\newblock Trends in Logic, vol.~4. Kluwer, 1998.

\bibitem{jeffrey1965}
Jeffrey, R.C.
\newblock \emph{The Logic of Decision}.
\newblock McGraw-Hill, 1965.

\bibitem{jeffrey1983}
Jeffrey, R.C.
\newblock Bayesianism with a human face.
\newblock In Earman, J., editor, \emph{Testing Scientific Theories},
pages 133--156. University of Minnesota Press, 1983.

\bibitem{kleene1952}
Kleene, S.C.
\newblock \emph{Introduction to Metamathematics}.
\newblock North-Holland, Amsterdam, 1952.

\bibitem{klement2000}
Klement, E.P., Mesiar, R., and Pap, E.
\newblock \emph{Triangular Norms}.
\newblock Trends in Logic, vol.~8. Kluwer, 2000.

\bibitem{levi1980}
Levi, I.
\newblock \emph{The Enterprise of Knowledge}.
\newblock MIT Press, 1980.

\bibitem{lukasiewicz1920}
\L ukasiewicz, J.
\newblock O logice tr\'ojwarto\'sciowej.
\newblock \emph{Ruch Filozoficzny}, 5:170--171, 1920.

\bibitem{nelsen2006}
Nelsen, R.B.
\newblock \emph{An Introduction to Copulas}.
\newblock Springer Series in Statistics, 2nd edition. Springer, 2006.

\bibitem{part1}
Anonymous.
\newblock Cred: A credence algebra with primitive conditioning.
\newblock Working paper, 2026.

\bibitem{popper1959}
Popper, K.R.
\newblock \emph{The Logic of Scientific Discovery}.
\newblock Hutchinson, London, 1959.

\bibitem{priest2008}
Priest, G.
\newblock \emph{An Introduction to Non-Classical Logic: From If to Is}.
\newblock Cambridge University Press, 2nd edition, 2008.

\bibitem{renyi1955}
R\'enyi, A.
\newblock On a new axiomatic theory of probability.
\newblock \emph{Acta Mathematica Academiae Scientiarum Hungaricae},
6:285--335, 1955.

\bibitem{sklar1959}
Sklar, A.
\newblock Fonctions de r\'epartition \`a $n$ dimensions et leurs marges.
\newblock \emph{Publications de l'Institut de Statistique de l'Universit\'e de Paris}, 8:229--231, 1959.

\bibitem{ramsey1931}
Ramsey, F.P.
\newblock Truth and probability.
\newblock In Braithwaite, R.B., editor, \emph{The Foundations of Mathematics
and Other Logical Essays}, pages 156--198. Routledge and Kegan Paul, 1931.

\bibitem{walley1991}
Walley, P.
\newblock \emph{Statistical Reasoning with Imprecise Probabilities}.
\newblock Chapman and Hall, 1991.

\bibitem{weber2010}
Weber, Z.
\newblock Transfinite numbers in paraconsistent set theory.
\newblock \emph{Review of Symbolic Logic}, 3(1):71--92, 2010.

\bibitem{zadeh1965}
Zadeh, L.A.
\newblock Fuzzy sets.
\newblock \emph{Information and Control}, 8:338--353, 1965.

\end{thebibliography}

\end{document}
