\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
    pdfauthor={Anonymous},
    pdftitle={Universal Bootstrapping}
}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}
\usepackage{tikz}
\usetikzlibrary{positioning}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\cred}{\mathsf{cred}}
\newcommand{\Cond}{\mathsf{Cond}}
\newcommand{\Sol}{\mathsf{Sol}}
\newcommand{\negc}{\mathord{\sim}}
\newcommand{\conjc}{\otimes}
\newcommand{\disjc}{\sqcup}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Bool}{\mathbb{B}}

\title{\textbf{Universal Bootstrapping}\\[0.5em]
\large Seeds, Substrates, and Self-Hosting Fixed Points}

\author{Anonymous for blind review}
\date{June 2026}

\begin{document}

\maketitle

\begin{abstract}
A compiler can compile its own source. The binary that results was built by an earlier binary, which was built by one before it, back to a first compiler written and translated by hand. The chain never bottoms out in nothing. We read this pattern as a schema: a substrate, a seed, a transition rule, and a validator, with a self-hosting fixed point where the artifact reproduces itself. The same schema recurs in proof kernels checking their own logic, in AI systems proposing their own training signal, and in accounts of life's origin. We formalize the schema in Lean as \texttt{SeededSystem} and prove that no run starts without a seed, that fixed points self-host, and that the irreducible-commitment system embeds. The formal core is one theorem family. The cross-domain readings are structured analogies, not consequences of it. The thesis: seed minimization is not origin minimization, and nothing bootstraps from nothing.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Take a C compiler written in C. To run it you need a C compiler, which is the
thing you are trying to build. The escape is historical. Some earlier compiler,
written in assembly or fed through another language, produced the first working
binary; that binary compiled the next version of the source; and the source has
been improved ever since while every new binary is built by the previous one.
Thompson's \emph{Reflections on Trusting Trust} \cite{thompson1984} turns this
chain into a warning: a compiler can carry behavior present in no version of its
source, passed binary to binary, because the source is only ever compiled by an
existing binary. Wheeler's diverse double-compiling \cite{wheeler2005} answers
the warning, but it does not remove the seed. It checks the seed against a second
independent one. The first compiler still came from outside.

Strip the example to its moving parts. There is a space of artifacts: source
files, binaries, states. There is a \emph{seed}: the first artifact, supplied
from outside the system. There is a \emph{transition}: the operation that folds
the current artifact and its substrate into the next one, here ``compile this
source with this binary.'' And there is a \emph{validator}: the check that a step
is admissible, that the new binary is a correct compiler. Run the transition from
the seed and you get an orbit. The interesting states are the ones where the
transition reproduces the input: a compiler whose source, compiled by itself,
yields a binary equivalent to itself. Such a state \emph{self-hosts}.

\begin{definition}[Seeded system, informal]
\label{def:seeded-informal}
A seeded system is a substrate of states $S$, a predicate $\mathit{Seed}$ marking
the artifacts you may start from, a transition $T : S \to S$ with the substrate
folded in, a validator $V$ on steps, and a correctness relation $E$. A state $s$
self-hosts when $E\,s\,(T\,s)$: the transition reproduces it up to $E$.
\end{definition}

This schema is the formal core of the paper. As a theorem, the Lean structure
\texttt{Cred.SeededSystem} carries exactly the five components of
\Cref{def:seeded-informal}, and \texttt{run} iterates $T$ from a seed. Two facts
about it set the tone. First, \texttt{no\_empty\_bootstrap}: if no state
satisfies $\mathit{Seed}$, no state is a seed, so there is nothing to start a run
from. The missing information has not vanished; it has moved into the substrate.
Second, \texttt{selfHosts\_of\_fixed}: an $E$-fixed point of $T$ self-hosts. The
compiler that recompiles to itself is the fixed point of the build operator.

The schema also has content beyond the compiler. As a theorem, the
irreducible-commitment structure of Part~4 embeds into it: \texttt{toSeeded}
turns any \texttt{Credence.CommitmentSystem} into a \texttt{SeededSystem} whose
transition is the refinement operator and whose correctness relation is equality,
and \texttt{fp\_selfHosts} and \texttt{deg\_selfHosts} show that both the
convergence target and the degenerate absorbing element self-host. The bootstrap
schema and the prior-and-seed schema are one structure seen twice.

\subsection*{The bottleneck}

Bootstrapping has a characteristic shape, and it is not minimization of the seed
alone. A self-hosting compiler is small as source, but it expresses a rich
language; the seed binary that first produced it was a separate, simpler tool;
and what comes out is again a full compiler. The pattern is
\emph{complex-prior $\to$ compressed-seed $\to$ complex-self-host}. A large prior
capability gets squeezed through a narrow opening, the seed, and re-expands into
a system that maintains itself. The narrow opening is real and worth finding. It
is what \texttt{run} starts from, and what \texttt{no\_empty\_bootstrap} says
cannot be empty. But the prior on either side of it is not removed by making the
seed small.

\begin{example}[Reading the bottleneck across domains]
\label{ex:bottleneck}
As structured analogies, not as instances of the theorem: a proof kernel checks
proofs, including, after Lean4Lean \cite{carneiro2024}, a proof of its own
checker's soundness, but the trusted kernel it runs on is the seed. An
AI-assisted search like FunSearch \cite{romera2024} proposes programs that
improve a problem it was set, yet the model, the evaluator, and the task are
supplied. Accounts of life's origin, hypercycles \cite{eigen1971},
autocatalytic sets \cite{kauffman1993}, protocell synthesis \cite{szostak2001},
describe chemistry that maintains and copies itself, but always from a stocked
environment. In each, a complex prior passes through a compressed seed into a
self-maintaining system. These illustrate the schema. They are not proved by it.
\end{example}

\subsection*{Thesis and guardrails}

The organizing claim, as philosophical synthesis: nothing bootstraps from
nothing, and seed minimization is not origin minimization. You can make the seed
explicit, hold it as small as the task allows, and check it for self-consistency.
You cannot make it zero. Shrinking the seed redistributes the assumed complexity
into the substrate and the transition; it does not abolish it. This continues the
irreducible-commitment thesis of Part~4 and the regress horn of the
M\"unchhausen trilemma \cite{albert1968}, now read through self-hosting fixed
points.

The guardrails matter, so we state them at the start. This paper has one formal
core and a set of structured analogies built around it. The formal core is the
Lean development: \texttt{SeededSystem} with \texttt{no\_empty\_bootstrap},
\texttt{SelfHosts} and \texttt{selfHosts\_of\_fixed}, the
\texttt{CommitmentSystem} embedding, and the earlier Cred results it rests on,
no-ex-falso and the Curry block among them. Everything proved is proved about
that schema and its instances. The compiler, proof-kernel, AI-design, and
biological readings are analogies that share the schema's shape. They are not
theorems.

So we do not claim, and the schema does not show, that all complexity shares a
single origin. A common seed-transition-validator structure recurring across
domains is a claim about structure, not about causal history. We do not claim
that biology is designed; the origin-of-life work we cite is about self-organizing
chemistry, and the schema is neutral on how a seed arises. We do not claim that
emergence is solved; that more is different \cite{anderson1972} is the open
question the schema organizes, not one it answers. And we do not claim that an AI
can bootstrap from nothing; \texttt{no\_empty\_bootstrap} says the opposite, and
the FunSearch reading respects it. What the paper does claim is narrow and, we
think, useful: one schema, substrate plus seed plus transition plus validator,
with a self-hosting fixed point, recurs across these domains and is worth writing
down once, formally, with its seed made explicit.

The rest of the paper develops the schema (\Cref{sec:intro} fixes the
vocabulary), states the Lean core as theorems, and then walks the four
analogies in turn, marking each claim as theorem, structured analogy, or
philosophical synthesis as it appears.

\section{The formal core: seeded systems}

Take a compiler that compiles its own source. Feed it the binary $b$; it
runs the source through that binary and emits a new binary; when the new
binary equals $b$, the compiler hosts itself. Strip away the C and the
machine code and a skeleton remains: a space of artifacts, a starting
artifact you are handed rather than derive, a step that folds the substrate
into a transition, a check on each step, and a notion of when two artifacts
count as the same. That skeleton is machine-checked. Everything in this
section is a formal theorem about it, verified in Lean; the compiler is the
motivating case, not a claim the schema proves. We tag each statement by
kind. The definitions and theorems below are the \textbf{formal theorem}
layer. The compiler reading is a \textbf{structured analogy}.

\subsection{The schema}

\begin{definition}[Seeded system]
A \emph{seeded system} bundles five components: a state space $S$, a
predicate $\mathrm{Seed} : S \to \mathrm{Prop}$ marking allowed seeds, a
transition $T : S \to S$ with the substrate folded in, a step validator
$V : S \to S \to \mathrm{Prop}$, and a correctness relation
$E : S \to S \to \mathrm{Prop}$. In Lean this is the structure
\texttt{SeededSystem} with fields \texttt{S}, \texttt{Seed}, \texttt{T},
\texttt{V}, \texttt{E}.
\end{definition}

The five fields name the object ($S$), the operation ($T$), the constraint
($V$ and the allowed seeds), and the consequence we track ($E$). As a
structured analogy: $S$ is the artifact format, $\mathrm{Seed}$ marks a
trusted starting binary, $T$ is one compile pass, $V$ checks a pass, and $E$
is observational equivalence. Nothing in the structure forces that reading.
The same skeleton accepts a proof kernel (states are derivations, $T$ a
reduction, $E$ definitional equality), or an autocatalytic set (states are
molecular populations, $T$ one reaction round). These are analogies; the
schema does not assert them.

\begin{definition}[Bootstrap run]
The run of length $n$ from a seed $s_0$ iterates the transition:
$\mathrm{run}\, s_0\, n = T^{[n]}\, s_0$. In Lean, \texttt{SeededSystem.run}
with \texttt{run\_zero} and \texttt{run\_succ} unfolding the iteration.
\end{definition}

A run needs something to iterate from. There is no run without a seed, and
the schema makes that literal.

\begin{theorem}[No empty bootstrap]
If the system has no seed, then nothing is a seed. Formally
\texttt{SeededSystem.no\_empty\_bootstrap}: from $\neg\, \exists s,\,
\mathrm{Seed}\, s$ it follows that $\forall s,\, \neg\, \mathrm{Seed}\, s$.
\end{theorem}

The statement is near-trivial as logic, and that is the point. A bootstrap
never starts from nothing. The information you do not supply as a seed has
not vanished; it has moved into the substrate folded into $T$. As
\textbf{philosophical synthesis}: this is the irreducible-commitment thesis
in its thinnest form. The bottleneck is unavoidable, not because the theorem
is deep, but because the schema refuses to hide where the starting datum
goes. The theorem does not show that all complexity shares one origin, that
biology was designed, or that a system can lift itself from nothing. It
shows the opposite of the last: a seed is required.

\begin{figure}[t]
\centering
\begin{tikzpicture}[>=stealth,
  box/.style={draw, rounded corners, font=\small, minimum height=8mm, inner sep=4pt}]
  \node[box] (seed) {seed $s_0$};
  \node[box, right=12mm of seed] (T) {transition $T$};
  \node[box, right=12mm of T] (V) {validator $V$};
  \node[box, right=12mm of V] (sh) {self-host $E(s,\,Ts)$};
  \draw[->] (seed) -- (T);
  \draw[->] (T) -- (V);
  \draw[->] (V) -- (sh);
\end{tikzpicture}
\caption{The seed--transition--validator pipeline of a \texttt{SeededSystem}. A
run iterates $T$ from a seed; the validator $V$ gates each step; a self-hosting
fixed point satisfies $E(s, Ts)$. There is no empty seed
(\texttt{no\_empty\_bootstrap}).}
\label{fig:schema}
\end{figure}

\subsection{Self-hosting}

\begin{definition}[Self-hosting]
A state $s$ \emph{self-hosts} when the transition reproduces it up to
correctness: $\mathrm{SelfHosts}\, s := E\, s\, (T\, s)$. In Lean,
\texttt{SeededSystem.SelfHosts}.
\end{definition}

Read $T\, s$ as $\mathrm{compile}(s)$. Then $\mathrm{SelfHosts}\, s$ says
$\mathrm{compile}(s)$ reproduces $s$ up to $E$: the compiler recompiles to
itself. That is the structured analogy. The theorem is about $E$ and $T$,
not about compilers.

The simplest route to self-hosting is a fixed point.

\begin{theorem}[Fixed points self-host]
If $E\, s\, s$ holds and $T\, s = s$, then $\mathrm{SelfHosts}\, s$.
Formally \texttt{SeededSystem.selfHosts\_of\_fixed}.
\end{theorem}

When $T$ already returns $s$, rewriting $T\, s$ to $s$ turns the target
$E\, s\, (T\, s)$ into $E\, s\, s$, which the reflexivity hypothesis
supplies. Fixed points are also stationary.

\begin{theorem}[Fixed points are stationary]
If $T\, s = s$, then $\mathrm{run}\, s\, n = s$ for every $n$. Formally
\texttt{SeededSystem.run\_const\_of\_fixed}, by induction on $n$.
\end{theorem}

A run from a fixed point never moves. Once the artifact reproduces itself,
further passes change nothing observable. The schema is inhabited: the
one-state system \texttt{trivialSelfHost} (state \texttt{Unit}, identity
transition, equality as $E$) self-hosts by \texttt{rfl}, so self-hosting is
not a vacuous predicate.

Self-hosting is not self-justifying, and the gap between them is the hinge of the
whole problem. Operationally, a fixed point self-hosts and the schema is
inhabited; \texttt{selfHosting\_available} exhibits it, and it needs no axioms.
At the meta level the matching move fails. A credence-valued certification that,
at a self-application point, reports its own non-certification cannot take a crisp
value: it is pinned to one half (\texttt{self\_validation\_not\_crisp}), neither
$0$ nor $1$ (\texttt{no\_crisp\_self\_validation}). As a theorem, a system can host
itself but cannot crisply certify itself. This is the Tarski boundary in graded
form, and it is why the reflection of Section~\ref{sec:kernels} climbs a tower
rather than resting at one fixed point: the operational self-host is available,
the meta self-justification is not.

The half is not only a ceiling, it is a consistent value. Non-crisp
self-certification is available: a faithful self-certification, one that reports
its own value, exists and equals one half
(\texttt{faithful\_self\_validation\_exists}), where a classical system has only
the liar contradiction. The half is also the cap, by an exact trade: any
self-certification above one half is unfaithful, it no longer reports its own
value (\texttt{self\_validation\_unfaithful\_of\_ne\_half}). So a system can
certify itself, but never to certainty. Certainty is still available for
\emph{parts}: a strictly lower level is certified with full confidence by the
tower, the crisp fragment by \texttt{crisp\_certainty\_iff\_classical}, and any
single accepted proof by \texttt{checkBool\_true\_sound}. The cap at one half
falls only on the total self-applied statement; every part that does not loop
back through the self-application point is certified crisply.

\subsection{The commitment system embeds}

Part~4 studied an irreducible-commitment structure on credences: a
refinement operator with a fixed point reached from every seed and a
degenerate absorbing element. That structure is the metric instance of the
schema, and the embedding is a theorem.

\begin{theorem}[Commitment systems are seeded systems]
Every \texttt{CommitmentSystem} $C$ yields a seeded system
$C.\mathtt{toSeeded}$ with $S = \mathrm{Credence}$, transition $T = C.op$,
correctness $E = {=}$, and every credence an allowed seed. Formally
\texttt{Credence.CommitmentSystem.toSeeded}.
\end{theorem}

Under this embedding the refinement operator becomes the transition,
equality becomes correctness, and the two distinguished credences self-host.

\begin{theorem}[Fixed point and degenerate element self-host]
For every commitment system $C$, both $C.\mathit{fp}$ and $C.\mathit{deg}$
self-host in $C.\mathtt{toSeeded}$. Formally
\texttt{Credence.CommitmentSystem.fp\_selfHosts} and
\texttt{deg\_selfHosts}, each via \texttt{selfHosts\_of\_fixed}: the fixed
point from the self-consistency equation $\mathit{op}\, \mathit{fp} =
\mathit{fp}$, the degenerate element from its absorption $\mathit{op}\,
\mathit{deg} = \mathit{deg}$.
\end{theorem}

Two states self-host for opposite reasons. The fixed point is the value the
orbit certifies from any seed: the converged commitment that reproduces
itself. The degenerate element is the absorbing zero of Cromwell's rule,
unrecoverable once entered, which reproduces itself by going nowhere. The
schema records both as $E\, s\, (T\, s)$ without collapsing the difference;
$V$ and $\mathrm{Seed}$, left trivial in this embedding, are where a richer
instance would separate a legitimate fixed point from a stuck one.

As \textbf{philosophical synthesis}: the same seed-transition-validator
skeleton recurs across the compiler, the proof kernel, the converging prior,
and the autocatalytic origin story. The recurrence is a schema match, not a
shared cause. The formal layer establishes only what the Lean names above
assert. The analogies illustrate the schema; they are not proved by it.

\section{Compiler bootstrapping}
\label{sec:bootstrapping}

A C compiler is itself a C program. To build it you need a C compiler. That circle is the cleanest worked example of the seeded-system schema, so we enter through it before generalizing. The schema columns are fixed by \texttt{SeededSystem} in the Lean development: a substrate $S$, a seed predicate $\mathrm{Seed}$, a transition $T$, a validator $V$, and a correctness relation $E$. Compiler bootstrapping instantiates all five, and it does so twice, under two opposed seed strategies. We present both as a \emph{structured analogy}: the schema illustrates the practice, and the practice illustrates the schema. Neither proves the other.

\subsection{Two seed strategies}

\paragraph{The large inherited seed.}
The fast and practical route reuses what already exists. An installed GNU toolchain, GCC and its runtime, compiles the next compiler, which compiles the one after. The substrate is the source tree plus the host operating system and libraries. The seed is the whole prior toolchain, a large binary nobody reads byte by byte. The transition is one compile-and-link pass. The validator is the build's own machinery: the test suite, and the standard three-stage check in which the compiler rebuilds itself and the stage-two and stage-three binaries are compared for equality. Correctness is observational equivalence of the produced compilers.

This route is cheap and it works. Its weakness is the size of the seed. A binary you cannot audit can carry a payload that the source does not mention. Thompson made the point exact: a compiler can be taught to recognize when it is compiling itself, reinsert a backdoor into its own next generation, and recognize when it is compiling the login program, inserting a second backdoor there~\cite{thompson1984}. Remove both patterns from the source and recompile with the infected binary: the source is clean, the binary is not, and the compromise reproduces through every later generation. As a \emph{structured analogy}, this is the self-hosting fixed point of the schema turned hostile. \texttt{SeededSystem.SelfHosts} says $E\,s\,(T\,s)$: the state reproduces itself up to correctness. Trusting trust is a self-hosting state whose reproduction the source-level validator cannot see.

\paragraph{The minimal audited seed.}
The opposed route shrinks the seed until a human can read it. A Stage0-style path starts from a few hundred bytes of hand-written hex, small enough to check by eye. That seed builds a minimal assembler. The assembler builds a slightly richer one. A chain of small, individually auditable stages climbs from there to a C compiler capable of building a full toolchain. The substrate is the bare machine and its instruction set. The seed is the hex monitor. The transition is each stage building the next. The validator is human inspection at the bottom, where the artifact is small, then conventional testing once the stages grow.

This route is slow and laborious. Its strength is that the seed is auditable in full. Where the inherited route trusts a large opaque binary, the Stage0 route trusts only what a person has read. The two strategies trade the same currency in opposite directions: build speed against seed size and auditability. That trade is the whole content of the choice; the schema does not pick a winner.

\subsection{Diverse double-compiling}

Even a clean source does not certify a binary, because the binary built it. Wheeler's diverse double-compiling breaks the circle without demanding that anyone read the seed~\cite{wheeler2005}. Compile the compiler's source twice: once with the suspect compiler, once with a second, independently produced compiler. The two stage-one outputs differ, since they came from different builders. Use each to recompile the source a second time. If the source is the only common cause, the two stage-two binaries are bit-for-bit identical. A divergence exposes a builder that injected something the source did not contain.

As a \emph{structured analogy} this is the validator $V$ doing work the correctness relation $E$ alone cannot. Self-hosting checks that a compiler reproduces \emph{itself}; double-compiling checks reproduction against a \emph{diverse} witness, so a backdoor that survives self-comparison still has to survive comparison with a compiler that never carried it. The disclaimer matters here. Double-compiling does not certify from nothing. It relocates trust: from one binary onto the independence of two builders and the readability of the shared source. If both builders share the infected ancestor, the diversity is fake and the check passes a compromise. The seed never reaches zero. It moves.

\subsection{Mapping onto the schema}

Table~\ref{tab:bootstrap} reads both strategies through the five columns of \texttt{SeededSystem}.

\begin{table}[t]
\centering
\small
\begin{tabular}{p{0.15\linewidth}p{0.36\linewidth}p{0.36\linewidth}}
\toprule
Column & Inherited seed (GCC) & Audited seed (Stage0) \\
\midrule
substrate $S$ &
source tree, host OS, libraries &
bare machine, instruction set \\[4pt]
seed $\mathrm{Seed}$ &
full prior toolchain (large, opaque) &
hand-written hex monitor (small, read) \\[4pt]
transition $T$ &
one compile-and-link pass &
each stage builds the next \\[4pt]
validator $V$ &
test suite, three-stage rebuild, double-compiling &
human inspection low, testing high \\[4pt]
correctness $E$ &
observational equivalence of compilers &
observational equivalence of compilers \\
\bottomrule
\end{tabular}
\caption{The two bootstrap strategies as instances of \texttt{SeededSystem}.
The columns are the schema fields $S$, $\mathrm{Seed}$, $T$, $V$, $E$. The
mapping is a structured analogy: it illustrates the schema, it is not proved
by it.}
\label{tab:bootstrap}
\end{table}

The one Lean fact behind the analogy is small and exact. \texttt{SeededSystem.no\_empty\_bootstrap} states that if no seed exists, no state satisfies the seed predicate: with nothing allowed to start from, no run begins. Read against the table, it formalizes a single observation. As a \emph{theorem} it says only this; the gloss that follows is \emph{philosophical synthesis}. The seed that looks empty is never empty. When a build appears to start from nothing, the missing information has moved into the substrate: the loader that places the first bytes, the firmware that runs the loader, the instruction set the hex assumes, the host that hosts the cross-compiler. \texttt{no\_empty\_bootstrap} marks the place where that information has to live. It does not say what fills it, and it does not claim any particular toolchain is trustworthy.

The minimal positive case is equally plain. \texttt{trivialSelfHost} is a one-state system with identity transition and equality as correctness, and \texttt{trivialSelfHost\_selfHosts} shows its state recompiles to itself: $\mathrm{compile}(b)\simeq b$. As a \emph{theorem} it certifies only that the schema is inhabited by a genuine self-hosting fixed point. It is a witness, not a model of GCC. The compiler readings stay in the prose, by the same guardrail that keeps the analogies out of the Lean file.

The honest conclusion is the one the irreducible-commitment thesis draws elsewhere, now in a setting where it can be inspected to the byte. A toolchain's worth is not that it has no opening commitment. It is that the commitment is made explicit, kept minimal, and validated. The inherited seed hides a large commitment behind a fast build. The audited seed exposes a small one and pays in labor. Double-compiling does not abolish the commitment; it diversifies the witness that checks it. Across all three, the seed is irreducible. The only question is how much of it you can read.

\section{Proof kernels and the trusted seed}
\label{sec:kernels}

Take a Lean development with ten thousand lines of proof. What do you trust when \texttt{lake build} reports success? Not the ten thousand lines. You trust the kernel: a small typechecker that re-derives every proof term against the inference rules of the underlying type theory. The tactics, the elaborator, the library, all of it can be wrong and the worst outcome is a failed check. A proof the kernel accepts is sound to the degree the kernel is. This is the de Bruijn criterion, named for the Automath system that first separated proof construction from proof checking \cite{debruijn1970}: let an arbitrarily clever, arbitrarily untrusted front end produce candidate proofs, and certify each one with a checker small enough to audit by hand.

The criterion is the proof-theoretic face of the seeded-system schema. As a structured analogy, the kernel is the seed: a small trusted base from which everything else is licensed. The transition $T$ is the elaboration of a high-level script into a low-level proof term. The validator $V$ is the kernel's typecheck of that term. Correctness $E$ is ``the term inhabits the stated type.'' The schema does not prove that any particular kernel is correct; it names the slots that a trusting discipline must fill. The Cred skeleton fixes those slots abstractly. \texttt{SeededSystem} carries the state space \texttt{S}, the seed predicate \texttt{Seed}, the transition \texttt{T}, the validator \texttt{V}, and the correctness relation \texttt{E}; \texttt{run} iterates \texttt{T} from a seed. As a theorem, \texttt{no\_empty\_bootstrap} states the obvious-but-load-bearing fact that with no seed there is nothing to start a run from. The kernel cannot be elided. The missing trust does not vanish when you delete it; it moves into whatever you check the kernel against.

\subsection{Self-hosting checkers}

A compiler written in its own language can compile itself. A proof checker can do the analogous thing: check a proof, encoded in the system's own logic, that the checker is correct. Lean4Lean reimplements Lean's kernel inside Lean and proves the reimplementation sound against a formal specification of the type theory \cite{carneiro2024}. As a structured analogy, this is the proof-kernel reading of self-hosting, and it maps onto one definition in the Cred core. A state \emph{self-hosts} when the transition reproduces it up to correctness: \texttt{SelfHosts s := E s (T s)}, the checker that recompiles to itself. As a theorem, \texttt{selfHosts\_of\_fixed} shows that any \texttt{E}-fixed point of the transition self-hosts, and \texttt{run\_const\_of\_fixed} shows such a point is stationary: every run from it stays put. The minimal witness \texttt{trivialSelfHost}, one state with the identity transition, inhabits the schema so the definition is not vacuous.

Self-hosting buys a sharper audit, not a smaller seed. When a checker certifies a proof of its own correctness, the certificate is only as good as the kernel that checked it: the very kernel whose correctness is at issue. The argument is not circular in a vicious way, because an independent checker, or a second compiler built from different sources, can re-run it; this is the diverse double-compiling answer to the trusting-trust attack, where a malicious compiler reproduces its own backdoor through self-compilation \cite{thompson1984,wheeler2005}. As a structured analogy, double-compiling is a second seed used to cross-check the first. Two seeds, audited against each other, beat one seed trusted blind. Neither construction removes the seed. It distributes trust across more than one and makes disagreement detectable.

\subsection{Stratified reflection}

The other route to enlarging a trusted base is to climb. Level $n+1$ certifies level $n$. A theory $T$ strong enough for arithmetic cannot prove its own consistency \cite{godel1931}. It can be extended: adjoin the soundness of $T$, the reflection principle ``if $T$ proves $\varphi$ then $\varphi$,'' and call the result $T_1$, which proves the consistency $T$ could not. Iterate, past the finite stages into the transfinite, indexed by ordinals. Turing studied the first such progressions \cite{turing1939} and Feferman developed the transfinite recursive progression in full \cite{feferman1962}. This is the reflection analogue of self-hosting: not a single fixed point that recompiles to itself, but a tower in which each floor certifies the one below.

The tie to Cred is direct. As a theorem, the commitment system embeds into the schema via \texttt{Credence.CommitmentSystem.toSeeded}, taking the refinement operator as the transition and equality as correctness; \texttt{fp\_selfHosts} and \texttt{deg\_selfHosts} exhibit the fixed point and the degenerate element as self-hosting states. The reflection tower and the commitment orbit are two readings of one structure: an operator iterated from a seed toward a value that certifies itself. As philosophical synthesis, the irreducible commitment recurs here in its stratified form. Each level trusts the level below, and the bottom level trusts nothing but itself.

That bottom level is where reflection meets its limit. Each adjoined soundness statement is an arithmetic sentence, so the new theory is again incomplete by the same theorem that forced the first extension \cite{godel1931}. No level certifies itself at its own level: Goedel and Tarski bound same-level self-certification, and the tower only ever pushes the boundary up, never closes it. As philosophical synthesis, this is the bottleneck the whole series turns on. The trusted seed shrinks as you stratify, double-compile, and minimize the kernel, but it never reaches zero. A system that claims no commitment at the base has only hidden the base it is standing on.

\paragraph{The second-incompleteness boundary is the Curry boundary.} Goedel's
second theorem is reached through Loeb's: a theory proves its own consistency only
by proving the derivability conditions, and that derivation runs through an
internal conditional closed under modus ponens, conditional proof, and
contraction. As a theorem, Cred has no such conditional. \texttt{curry\_block}
shows no operation on credences satisfies all three, and
\texttt{loeb\_internal\_arrow\_blocked} records that the internal-arrow form of
the Loeb conditions is therefore impossible. So the place where a naive system
would derive its own consistency, and on the verge of inconsistency explode, is
exactly the place Curry is blocked. The graded reading agrees: a self-certifying
provability value sits at the interior (\texttt{self\_validation\_not\_crisp}),
not at certainty. The boundary of the second incompleteness theorem and the
boundary of Curry are one boundary.

\subsection{The link to Cred's aim}

Cred's own no-explosion discipline is a seed of this kind, and the schema says where it sits. As a theorem, the chain-rule conditional refuses ex falso: conditioning on credence $0$ leaves the conditional unconstrained (\texttt{conditioning\_zero\_any}), so impossible evidence licenses nothing rather than everything. As a theorem, the product residuum blocks Curry: contraction is incompatible with the pinned-down implication (\texttt{prod\_resid\_no\_contraction}, \texttt{curry\_block}), so the self-reference that would explode a naive system instead has a fixed-point solution set. These results are the validator $V$ of Cred's own bootstrap. They say which steps the system will accept, and the schema makes explicit that the seed behind them, the choice to take conditioning as primitive rather than as division, is itself a commitment the formalism does not derive. The aim is to keep that seed small and stated: a kernel of accepted rules, checked, with the one unavoidable commitment in plain view rather than smuggled into the substrate.

As philosophical synthesis, a common seed-transition-validator schema recurs across compiler bootstrapping, proof kernels, stratified reflection, and Cred's own no-explosion core. The recurrence is a pattern worth naming, not a proof that these domains share an origin. The schema does not show that all complexity comes from one source, that emergence is solved, or that any system can bootstrap from nothing. It shows the opposite: every bootstrap names a seed, and the work is to make that seed minimal, explicit, and open to audit.

\section{The bottleneck: small seeds, large priors}
\label{sec:bottleneck}

\begin{figure}[t]
\centering
\begin{tikzpicture}[>=stealth,
  big/.style={draw, rounded corners, font=\small, minimum width=30mm, minimum height=14mm},
  smallb/.style={draw, rounded corners, font=\small, minimum width=14mm, minimum height=8mm}]
  \node[big] (prior) {complex prior system};
  \node[smallb, right=16mm of prior] (seed) {seed};
  \node[big, right=16mm of seed] (host) {complex self-host};
  \draw[->] (prior) -- node[above, font=\footnotesize] {compress} (seed);
  \draw[->] (seed) -- node[above, font=\footnotesize] {bootstrap} (host);
\end{tikzpicture}
\caption{The bottleneck. A large prior system produces a small seed, which
bootstraps a large self-hosting system. Seed minimization is not origin
minimization: the small artifact still came from somewhere large.}
\label{fig:bottleneck}
\end{figure}

A self-hosting compiler ships as a few hundred kilobytes of source. You read it
in an afternoon. The validator is small, the seed is small, the transition rule
is small. Then ask where the seed came from. It came from a host compiler,
written by people, trained in a discipline, running on a fabrication stack that
took decades and continents to build. The artifact you hold is tiny. The system
that produced it is not.

This section states the organizing claim of the part, as philosophical
synthesis: \emph{seed minimization is not origin minimization}. The seed of a
bootstrap can be made arbitrarily small as an object while the prior system that
emitted it stays large. Shrinking the artifact does not shrink the history. The
two quantities are independent, and conflating them is the bottleneck illusion.

\subsection{What the formal core actually fixes}

The formal core supplies the bottleneck, not the largeness behind it. As a
theorem, \texttt{SeededSystem} bundles five things and nothing else: a state
space \texttt{S}, an allowed-seed predicate \texttt{Seed}, a transition
\texttt{T}, a validator \texttt{V}, and a correctness relation \texttt{E}. The
run is iteration of \texttt{T} from a seed. The narrowest fact in the file is
the one that names the bottleneck directly.

\begin{theorem}[No empty bootstrap]
\label{thm:noempty}
In \texttt{SeededSystem}, \texttt{no\_empty\_bootstrap} states: if no state
satisfies \texttt{Seed}, then no run can begin. There is nothing to iterate
\texttt{T} on.
\end{theorem}

Read this carefully, because it says less than it seems to and that is the
point. It does not say the seed is uncaused. It says the run needs the seed as
an input the run cannot manufacture. The transition \texttt{T} folds the
substrate in; whatever \texttt{T} draws on at each step is already present
before the first step. As a theorem, \texttt{selfHosts\_of\_fixed} adds that an
\texttt{E}-fixed point of \texttt{T} reproduces itself: \texttt{SelfHosts s :=
E s (T s)}, the compiler that recompiles to itself. The commitment system
embeds, and both its convergence fixed point and its degenerate element
self-host, by \texttt{fp\_selfHosts} and \texttt{deg\_selfHosts}. These are
statements about a state once it exists. None of them is a statement about where
the state came from. The formalization is deliberately silent on origin, and
that silence is load-bearing: it is the gap the synthesis names rather than
fills.

\paragraph{Minimality is relative to the substrate.} Because \texttt{T} folds the
substrate in, shrinking the explicit seed need not shrink the commitment: it can
move into the substrate, the decoder, the transition, the validator, or the
selector. As a theorem, \texttt{commitment\_conservation} makes the relocation
exact. Carry the substrate $b$ in the state instead of in the transition, and the
run on the original state is unchanged: $(\mathrm{withSubstrate}\,T)^{n}(b,s) =
(b, (T\,b)^{n}\,s)$. The explicit seed grew to include $b$, the substrate shrank to
nothing, and nothing observable moved; what is conserved is the pair (substrate,
seed), not the size of the seed alone. So every claim of a minimal seed is
relative to a fixed substrate. As structured analogy the operational cases read
the same way: a Stage0 hex seed is tiny only against a fixed instruction set,
loader, and stage manuals; a minimal proof kernel is tiny only against a fixed
syntax, semantics, value algebra, and metatheory; an AI checkpoint is a seed only
against fixed data, architecture, and objectives. The Bayesian cases too: a flat
prior is minimal only against a base measure, a Jeffreys prior canonical only
against a Fisher geometry, a Solomonoff prior universal only against a chosen
machine, and a credal prior does not remove the seed but exposes a set of seeds.
This is the companion to fixed-substrate coherence: there the substrate is held
fixed and the data grouping does not matter; here the question is what the fixed
substrate already committed.

\subsection{Two sizes, one artifact}

Separate two measurements of a bootstrap. Call the first the \emph{seed size}:
the length of the artifact you hand to \texttt{T}, the kilobytes of source, the
bits of the prior, the base sequence of the first replicator. Call the second
the \emph{prior size}: the system that had to be in place for that seed to be a
usable seed at all. Theorem~\ref{thm:noempty} constrains the first to be
nonempty. It says nothing about the second.

The gap between them is large in every concrete case, and the cases are
structured analogies, not consequences of the schema.

\begin{example}[Trusting trust]
\label{ex:trust}
Thompson's compiler attack \cite{thompson1984} hides a few lines in a compiler
binary; the binary then reinserts them into every recompilation, including
recompilations of itself. As a structured analogy, the seed (the planted source
delta) is small and the validator (read the source) passes, yet the prior system
(the already-compromised host binary) carries the payload the source never
shows. Wheeler's diverse double-compiling \cite{wheeler2005} defeats the attack
precisely by enlarging the prior on purpose: introduce a second, independent
compiler so the host is no longer a single trusted point. The defense does not
shrink the seed. It admits that trust lives in the prior and audits the prior.
\end{example}

\begin{example}[Proof kernels]
\label{ex:kernel}
A proof checker earns trust by having a small kernel: de Bruijn's criterion
\cite{debruijn1970}, a core you can read in full. Lean's kernel re-checked by an
independent implementation, Lean4Lean \cite{carneiro2024}, sharpens this: the
checked artifact is minimal, but the confidence rests on the metatheory, the
type theory, the second implementation, the mathematical culture that vetted
them. As a structured analogy, the kernel is the small seed; the trusted
foundation is the large prior. Feferman's transfinite progressions of theories
\cite{feferman1962} and Turing's ordinal logics \cite{turing1939} make the same
shape exact: each level certifies the one below, and the tower never bottoms out
in nothing. G\"odel's second theorem \cite{godel1931} is the reason the prior
cannot be folded into the seed: the system cannot prove its own consistency from
inside.
\end{example}

\begin{example}[Origins of order]
\label{ex:origins}
The first replicator is short. The chemistry that made a short replicator
selectable is not. As a structured analogy, Eigen's hypercycles
\cite{eigen1971}, Kauffman's autocatalytic sets \cite{kauffman1993}, and
Szostak's protocells \cite{szostak2001} each exhibit a small self-reproducing
seed riding on a large prior: an energy gradient, a monomer supply, a
compartment, a prebiotic environment that took a planet to assemble. The seed
is the sequence. The prior is the world that made the sequence count as a seed.
\end{example}

We state the disclaimer the guardrails require, in the prose where it bites.
None of Examples~\ref{ex:trust}--\ref{ex:origins} is proved by
\texttt{SeededSystem}. The schema illustrates them; they do not follow from it.
The recurrence of a seed-transition-validator shape across these domains is a
descriptive pattern, not a derivation. In particular: the schema does not show
that biology is designed, it does not solve emergence, and it does not license
the claim that any system bootstraps from nothing. The biological cases are
analogies to a fixed-point structure, not assertions that life was engineered.

\subsection{Solomonoff: minimal description, chosen machine}

The cleanest formal witness that small artifact and large prior are independent
is the universal prior. Solomonoff's induction \cite{solomonoff1964} weights a
hypothesis by the length of the shortest program that outputs it, which looks
like a recipe for origin minimization: prefer the smallest description. As
philosophical synthesis, read where the largeness hides. The description length
is measured relative to a fixed universal machine. The minimal program is small;
the machine that interprets it is a complete choice of computational basis, and
different machines shift every length by a constant the data cannot remove. The
prior is small in description and total in commitment.

\begin{remark}[Where the prior sits]
\label{rem:machine}
This is the bottleneck in its purest form. The seed (the shortest program) is
minimized by construction. The prior (the reference machine) is presupposed by
construction. Jaynes's maximum-entropy priors \cite{jaynes2003} repeat the
shape: entropy is maximized subject to constraints, and the constraints, the
choice of what to hold fixed, are the irreducible commitment the optimization
does not supply. Minimizing the seed sharpens the artifact and relocates the
commitment into the machine. It does not abolish it.
\end{remark}

This mirrors the formal silence exactly. \texttt{no\_empty\_bootstrap} fixes
that a seed is needed; it leaves the substrate folded into \texttt{T}
unexamined. Solomonoff fixes the shortest program; it leaves the machine
unexamined. Same gap, two vocabularies.

\subsection{Emergence: more is different, not more from nothing}

One reading of a small seed yielding a large system is emergence, and the
synthesis must mark its limit. Anderson's \emph{more is different}
\cite{anderson1972} argues that each level of organization has laws not reducible
to the level below: a few rules of local interaction give rise to phases,
symmetries, and structure the rules do not name. As philosophical synthesis,
this is the right slogan for the bottleneck and the wrong slogan for origin.
Emergence is large behavior from compact rules \emph{within} a substrate already
rich enough to support it. The compact rules are the seed. The substrate, the
many degrees of freedom, the prior conditions for a phase to form, is the large
prior. Emergence does not produce the substrate; it organizes one.

So the disclaimer stands. A small seed generating rich behavior is a real and
common pattern, witnessed by FunSearch and LLM-driven mathematical discovery
\cite{romera2024}, where a compact search loop over a large pretrained prior
finds genuinely new constructions. As a structured analogy, the loop is the
seed and the model plus its training corpus is the prior; the discovery is
emergent over that prior, not over nothing. The schema does not claim that
emergence is solved, that all complexity shares a single origin, or that an
intelligence can bootstrap itself from an empty start. It claims, descriptively,
that the seed-transition-validator shape recurs, and that in every recurrence
the prior outweighs the seed.

\subsection{The thesis, stated as synthesis}

Gather the threads. As philosophical synthesis, the bottleneck thesis is this:
every bootstrap admits a small seed and presupposes a large prior, and the two
sizes are not the same measurement. You can compress the artifact toward its
minimal description, and the compiler, the kernel, the prior, the replicator all
reward the effort with auditability. What you cannot do is compress the prior to
zero by compressing the seed. The missing information moves into the substrate;
it does not vanish. That movement is exactly what \texttt{no\_empty\_bootstrap}
records inside the formal core and what Solomonoff's chosen machine records
inside the universal prior.

The honest engineering response follows from the descriptive pattern, not from
any theorem. Make the seed small, because a small seed is readable and a
readable seed can be checked. Then make the prior explicit, because an unstated
prior is where trusting-trust hides, where the reference machine smuggles in a
basis, where a foundation borrows consistency it cannot prove. Seed minimization
buys auditability. It does not buy a beginning from nothing, and the part claims
no more than that.

\section{AI-assisted design as recursive bootstrapping}
\label{sec:ai-bootstrap}

A program-search system mutates candidate programs, runs them against a scoring
function, and keeps what scores higher. FunSearch is one such loop: a language
model proposes code, an evaluator checks each proposal, and the surviving
programs feed the next round of proposals \cite{romera2024}. The system improved
on known constructions in extremal combinatorics. Nothing in the loop creates a
proposal out of nothing. The model was trained, the evaluator was written, and
the first prompt was supplied. The loop refines; it does not originate.

That structure is the seeded-system schema of Part~4, read in a new domain. We
make the mapping explicit, then say plainly what it does and does not license.

\subsection{The schema, and the AI-design reading of it}

Recall the formal skeleton. \emph{As a theorem}, a \texttt{SeededSystem} fixes a
state space $S$, a seed predicate $\mathrm{Seed}$, a transition $T$, a step
validator $V$, and a correctness relation $E$. A run iterates the transition,
$\mathrm{run}(s_0, n) = T^{[n]}(s_0)$. A state self-hosts when the transition
reproduces it up to correctness: $\mathrm{SelfHosts}(s) := E\,s\,(T\,s)$. Two
facts pin the schema down. First, \texttt{no\_empty\_bootstrap}: if no seed
exists, no state is a seed, so there is nothing to start a run from. The missing
information moved into the substrate; it did not vanish. Second,
\texttt{selfHosts\_of\_fixed}: an $E$-fixed point of the transition self-hosts,
and by \texttt{run\_const\_of\_fixed} every run from it stays put.

\begin{example}[The AI-design reading]
\emph{As a structured analogy}, take the program-search loop. The state $s$ is a
design: an algorithm, a hardware layout, a successor model's weights and code.
The transition $T$ is one assisted-design step, model proposal plus selection,
with the substrate folded in. The validator $V$ is the evaluator: tests,
benchmarks, held-out scores, and in the strongest case a machine-checked proof.
The correctness relation $E$ asks whether the new design meets the standard the
old one set. A loop that produces designs which pass the same validator that
seeded it is a self-hosting state: the artifact reproduces its own
acceptability. This reading \emph{illustrates} the schema. It is not proved by
it. The Lean facts hold over the abstract $(S,\mathrm{Seed},T,V,E)$ and carry no
claim about any particular AI system.
\end{example}

The validator gates each step, and that is the load-bearing part. Refinement
without a validator is drift. The schema separates the two cleanly: $T$ proposes,
$V$ accepts or rejects, $E$ records what acceptance was supposed to mean. A
recursive technological loop, AI helping design the next AI, is many turns of
this single step, each turn gated by whatever $V$ the designers chose.

\subsection{The seed is the substrate, and it is not small}

\emph{As philosophical synthesis}, the irreducible-commitment thesis says a
bootstrap never starts from nothing. \texttt{no\_empty\_bootstrap} is the formal
shadow of that thesis: remove the seed and the run has no first state. For
AI-assisted design the seed is the inherited substrate, and here the synthesis
must be stated against a tempting misreading.

The loop does not create itself from nothing. Compute, training data, the
optimizer, the objective, and the evaluator are all supplied before the first
proposal. They are the seed in the sense of Part~4: an input the system cannot
derive inside itself, only inherit and make explicit. Unlike the prior of
Part~4, which one can write on a line, this seed is large and often opaque. A
trillion-token corpus and a learned reward model are still a commitment; nobody
can fully read them. Holding the seed small, the management discipline the
schema recommends, is hard precisely when the seed is a learned artifact.

So the claim and the disclaimers, kept apart on purpose:
\begin{itemize}
  \item The paper claims a common seed--transition--validator schema recurs
        across these domains, and that AI-assisted design instantiates it.
  \item The paper does not claim the schema lets an AI bootstrap from nothing.
        \texttt{no\_empty\_bootstrap} says the opposite: no seed, no run.
  \item The paper does not claim the schema proves AI design must converge,
        improve, or stay safe. \texttt{selfHosts\_of\_fixed} describes fixed
        points if one is reached; it does not promise one is.
\end{itemize}

\subsection{Trusting trust, for learned components}

One concern transfers directly. Thompson's compiler attack hides a backdoor in
the binary that compiles the compiler, so the malicious behavior survives even
though no source shows it \cite{thompson1984}. The fix is to break the
self-reference from outside: compile the compiler with a second, independent
compiler and check the results agree \cite{wheeler2005}. In the schema, the
attack is a self-hosting state whose validator $V$ was the very component the
attacker controlled. Self-hosting certifies reproduction, not honesty.

The learned case is worse in one respect. A backdoored compiler is a fixed
program; a diverse second compiler can in principle reproduce it. A learned
component, model weights and a learned evaluator, has no readable source to
double-compile. If the evaluator $V$ is itself a trained artifact inherited from
the seed, a design can pass it while doing something its authors never intended,
and there is no clean second $V$ to disagree. The proof-kernel discipline points
the other way: keep $V$ small, fixed, and independently checkable, so that
correctness rests on something a person can read rather than on a learned oracle.
The schema does not resolve this. It locates the trust precisely, in $V$ and in
the seed, which is where any answer will have to act.

\section{The origins boundary}
\label{sec:origins}

The schema recurs in places where no one designed anything. A bacterium copies its genome; the copy must pass enough fidelity checks to function, or the lineage ends. There is a state (the cell), a transition (replication), and a validator (selection on viability). Nothing in that loop requires an author. This section walks the schema to its hardest edge, the origin of life, and states plainly what the recurrence does and does not license. Two warnings frame everything below. First, the seed-transition-validator pattern here is a \emph{structured analogy}, not a theorem; the Lean core proves facts about commitment systems, not about chemistry. Second, and more important, the recurrence of the pattern is \emph{descriptive}. It carries no inference to a designer. Section~\ref{sec:origins} rejects that inference explicitly.

\begin{example}[Replication as a seeded transition]
\label{ex:replication}
Read $\mathsf{S}$ as the space of cell states, $\mathsf{T}$ as one round of replication-with-metabolism, and $\mathsf{V}$ as the predicate ``the daughter is viable.'' The orbit $\mathsf{run}(s_0, n)$ is $n$ generations from an ancestor $s_0$. This is the metric commitment system one abstraction up: \texttt{Credence.CommitmentSystem.toSeeded} sends the refinement operator to $\mathsf{T}$ and equality to the correctness relation $\mathsf{E}$. The biological reading replaces equality by ``equally viable,'' which is weaker, so the embedding is an illustration of the skeleton, not an instance of the theorem.
\end{example}

As a theorem, the skeleton says something narrow and exact. \texttt{SeededSystem.no\_empty\_bootstrap} states that if no seed exists, no run starts: $\neg(\exists s,\ \mathsf{Seed}\,s)$ forces $\forall s,\ \neg\,\mathsf{Seed}\,s$, and \texttt{run} has nothing to fold. \texttt{SeededSystem.selfHosts\_of\_fixed} states that a transition fixed point reproduces itself under $\mathsf{E}$, so $\mathsf{SelfHosts}\,s := \mathsf{E}\,s\,(\mathsf{T}\,s)$ holds. These are claims about an abstract state space. They do not mention molecules, fitness, or time. When we read $\mathsf{T}$ as replication we leave the proved territory and enter analogy.

\subsection*{Where does the first seed come from?}

The schema needs a seed. Push the replication loop backward and the question sharpens: what supplied the first replicator? Three research programs answer without invoking an agent.

As a structured analogy, natural selection is the validator with no validator above it. Variation perturbs the transition; differential survival is $\mathsf{V}$; the population is the state. The loop closes on itself the way a self-hosting compiler does, and \texttt{trivialSelfHost} shows the schema admits a genuine self-reproducing fixed point with the identity transition. The point of the analogy is structural: selection certifies its own output the way a kernel certifies a proof, not by appeal to a higher court.

\begin{remark}[Candidate near-seedless transitions]
\label{rem:candidates}
Three accounts narrow the seed toward chemistry. Eigen's hypercycle \cite{eigen1971} couples replicating molecules into a cycle that sustains itself above an error threshold; the cycle is the transition, the threshold is the validator, and the seed is a population of short templates, not an organism. Kauffman's autocatalytic sets \cite{kauffman1993} make collective self-production a generic property of a sufficiently rich chemical network: once the catalytic closure forms, the set produces itself, a chemical $\mathsf{SelfHosts}$. Szostak's protocells \cite{szostak2001} supply the missing container, a vesicle whose growth and division give selection something to act on. None of the three claims a seedless start. Each shrinks the seed to a chemical configuration that physics and chemistry can plausibly reach.
\end{remark}

So the seed does not vanish; it relocates into the substrate. This is the same move the schema makes everywhere. A self-hosting compiler hides the bootstrap binary in the substrate it was built with; the missing information moved, it did not disappear. The \texttt{no\_empty\_bootstrap} fact is the formal shadow of that relocation: you cannot run from nothing, but you can run from a seed thin enough that chemistry, not an author, supplies it. Whether the prebiotic seed is in fact reachable is an empirical question for chemistry. The schema does not settle it and does not pretend to.

\subsection*{The non-sequitur, stated and rejected}

A reader might run the recurrence the wrong way: the pattern looks engineered, therefore something engineered it. That inference does not follow, and we reject it.

As philosophical synthesis, the recurrence of seed-transition-validator across replication, abiogenesis, and autocatalysis is a fact about \emph{description}, about the vocabulary that fits self-producing systems. A vocabulary fitting a system says nothing about the system's cause. The validator role is filled by natural selection and by error thresholds, both of which are blind. The transition role is filled by chemistry and by copying, neither of which deliberates. The seed role is filled by a molecular configuration, not an intention. Selection and chemistry supply transition and validator without an agent; Remark~\ref{rem:candidates} names concrete mechanisms for each role. The schema therefore gives intelligent-design inference no foothold. Reading design into the pattern mistakes a description of structure for evidence of a designer, which is a category error, not a discovery.

The paper makes a bounded claim and refuses three larger ones. It claims: a common seed-transition-validator schema recurs across these domains, including the biological ones. It does \emph{not} claim that the schema proves all complexity shares a single origin; the domains share a description, not a derivation. It does \emph{not} claim that biology is designed; the candidate transitions in Remark~\ref{rem:candidates} are explicitly seedless of any agent. It does \emph{not} claim that emergence is solved; whether a chemical autocatalytic set is reachable under early-Earth conditions is open, and the schema is silent on it. Anderson's slogan that more is different \cite{anderson1972} marks the gap we leave open: new organization at one level is not deducible from the level below, and our skeleton describes such organization without explaining its arrival.

\subsection*{What survives at the boundary}

Strip the analogies and one formal residue remains. \texttt{no\_empty\_bootstrap} and \texttt{selfHosts\_of\_fixed} are about an abstract state space; they hold whether $\mathsf{T}$ is a compiler, a proof kernel, or a replicator. The biological reading borrows their shape and nothing more. The honest statement at the origins boundary is therefore modest. The seed is irreducible: every running system we examined had one, and the formal core says a run needs one. Where the seed comes from is, in the biological case, a question for chemistry, and the leading answers (\cite{eigen1971}, \cite{kauffman1993}, \cite{szostak2001}) place it in the substrate rather than in an agent. The schema points to the boundary and stops there. It does not cross into a designer, and pretending it could would betray both the mathematics and the biology.

\section{Case studies and positioning}
\label{sec:cases}

Start with a compiler that has no compiler. To build the first C compiler you
need a working C compiler, which does not yet exist. The way out is a seed: a
smaller compiler, written in assembly or in an earlier language, good enough to
compile the real one. Thompson's reflections on trusting trust \cite{thompson1984}
turn this on its head. A compiler can learn to miscompile, hide the trick in its
own binary, and pass it forward through every later self-build, so the malicious
seed survives even after its source is removed. Wheeler's diverse
double-compiling \cite{wheeler2005} answers with a second, independent compiler
that validates the first by reproducing its output. The shape is fixed:
substrate (source plus build environment), seed (a prior compiler), transition
(compilation), validator (the diverse rebuild), and a self-host condition (the
compiler recompiles to itself).

That same shape is what the Lean core formalizes. As a theorem, the structure
\texttt{SeededSystem} bundles a state space \texttt{S}, a seed predicate
\texttt{Seed}, a transition \texttt{T}, a step validator \texttt{V}, and a
correctness relation \texttt{E}. A run is iterated transition,
\texttt{run s0 n}${} = \texttt{T}^{[n]}\, \texttt{s0}$, and a state self-hosts
when the transition reproduces it up to correctness, $\texttt{SelfHosts}\, s :=
\texttt{E}\, s\, (\texttt{T}\, s)$. Two facts pin the schema down. As a theorem,
\texttt{no\_empty\_bootstrap} states that with no seed there is nothing to start
a run from: the missing information moves into the substrate, it does not vanish.
As a theorem, \texttt{selfHosts\_of\_fixed} states that an \texttt{E}-fixed point
of the transition self-hosts, the compiler that recompiles to itself.

The irreducible-commitment structure of Part~4 embeds as the metric instance.
As a theorem, \texttt{Credence.CommitmentSystem.toSeeded} sends the refinement
operator to the transition and equality to correctness; \texttt{fp\_selfHosts}
and \texttt{deg\_selfHosts} then certify that both the fixed point and the
degenerate absorbing element self-host. The schema is inhabited by a genuine
self-hosting state, \texttt{trivialSelfHost}: one state, identity transition,
equality as correctness, so \texttt{trivialSelfHost\_selfHosts} holds by
reflexivity. Earlier parts supply the rest of the formal core: the no-ex-falso
witness \texttt{labelled\_no\_ex\_falso}, the Curry block
\texttt{curry\_block}, and reflection through the kernel certificates. These are
the only rows the schema proves.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}\begin{tabular}{@{}p{0.13\linewidth}p{0.13\linewidth}p{0.13\linewidth}p{0.15\linewidth}p{0.18\linewidth}p{0.09\linewidth}@{}}
\toprule
Substrate & Seed & Transition & Validator & Self-host condition & Status \\
\midrule
Stage-0 compiler
  & hand-written core
  & compile
  & test suite
  & recompiles to itself
  & analogy \\
GNU/GCC
  & prior GCC binary
  & bootstrap build
  & 3-stage comparison
  & stage 2 = stage 3
  & analogy \\
Proof kernel
  & trusted checker
  & elaborate, check
  & re-check in Lean
  & checks its own export
  & analogy \\
Cred toy model
  & \texttt{Seed} state
  & \texttt{T} (refine)
  & \texttt{V} step
  & \texttt{E}\,$s$\,(\texttt{T}\,$s$)
  & \textbf{formal} \\
AI design loop
  & base model + spec
  & propose, evaluate
  & scored evaluator
  & passes own metric
  & analogy \\
Biological origin
  & autocatalytic set
  & replication
  & selection
  & set reproduces itself
  & analogy only \\
\bottomrule
\end{tabular}
\caption{The seed--transition--validator schema across six substrates. Only the
Cred toy model is a formal theorem; the other rows are structured analogies that
illustrate the schema, not consequences of it. The proof-kernel row is the meta
case: its self-host is bounded by Goedel and Tarski, the checker cannot certify
its own soundness at its own level, unlike the compiler's operational self-host.}
\label{tab:cases}
\end{table}

\Cref{tab:cases} reads down the same five columns. As a structured analogy, the
GNU/GCC bootstrap fills them with a prior binary, a three-stage build, and the
stage-2-equals-stage-3 fixed-point check that GCC actually runs. As a structured
analogy, a Lean-style proof kernel reads as substrate (the elaborated terms),
seed (a trusted checker), transition (elaborate and check), and validator (an
independent re-check, as in Lean4Lean \cite{carneiro2024}); the self-host
condition is that the kernel certifies its own exported proof. As a structured
analogy, an AI design loop fits too: a base model and a specification seed the
search, generation is the transition, and a scored evaluator validates, as in
FunSearch \cite{romera2024}, where an LLM proposes programs that a fixed scorer
ranks. These analogies illustrate the schema. They are not proved by the Lean
core, and we mark each such row accordingly.

The biological row is the one to read with most care. As a structured analogy,
the origin of self-reproduction has the same five slots: an autocatalytic set as
seed \cite{kauffman1993}, replication as transition, selection as validator, and
a set that reproduces itself as the self-host condition, with the hypercycle
\cite{eigen1971} and protocell synthesis \cite{szostak2001} as the concrete
biochemistry. The schema describes the bookkeeping. It does not explain how the
first such set arose, and it does not make the system designed. We mark this row
analogy only.

\begin{remark}[What the schema does and does not buy]
The recurrence of one bookkeeping pattern across six substrates is a description,
not a unification theorem. Anderson's point that more is different
\cite{anderson1972} cuts the other way from any claim that a single schema
explains complexity: new levels bring laws the lower level does not contain. The
schema tracks where the opening commitment sits; it says nothing about what fills
the slots in any particular domain.
\end{remark}

\paragraph{What the paper claims.} As philosophical synthesis: a common
seed--transition--validator schema recurs across compiler bootstrapping, proof
kernels, AI-assisted design, biological origins, and the Cred credence model;
every instance carries an irreducible opening commitment that the downstream
machinery cannot generate, because the machinery runs only after the seed is in
place. As a theorem: the schema \texttt{SeededSystem} is inhabited and
non-degenerate, \texttt{no\_empty\_bootstrap} and \texttt{selfHosts\_of\_fixed}
hold, the commitment system of Part~4 embeds with both distinguished elements
self-hosting, and the no-ex-falso witness, the Curry block, and reflection
through the kernel certificates hold in Cred.

\paragraph{What the paper does not claim.} The schema does not prove that all
complexity shares a single origin: a shared bookkeeping pattern is not a shared
cause, and Anderson's caution \cite{anderson1972} stands. It does not assert that
biology is designed: the biological row is a structured analogy for where the
seed sits, not a design inference. It does not solve emergence: the slots are
labels, not a mechanism for how any level produces the next. It does not claim
that an AI, a compiler, or any system can bootstrap from nothing:
\texttt{no\_empty\_bootstrap} states the opposite, and Thompson's attack
\cite{thompson1984} shows the seed is exactly what must be trusted or
independently checked. Only the Cred row is a formal theorem; every other row is
an analogy and is marked as such in \cref{tab:cases}.

\paragraph{The schema is bookkeeping, not the deep result.} The formal bootstrap
theorems are deliberately elementary. No seed gives no run
(\texttt{no\_empty\_bootstrap}); a fixed point gives self-hosting
(\texttt{selfHosts\_of\_fixed}); for a fixed Bayesian model, sequential update
equals batch update. Their role is not a deep mathematical result but an
accounting discipline, like double-entry bookkeeping: elementary, yet useful
because it catches a hidden host tool in a trusted compiler path, a hidden
model-volume commitment in a Bayesian comparison, a hidden internal conditional
in a proof system, a hidden metatheory in a foundation, or a hidden validator
assumption in a self-hosting claim. The nontrivial mathematical content of Cred
lies elsewhere: admissible conditioning and no-ex-falso through zero-evidence
underdetermination, the LP/K3 collapse bridges, the no-go against any
truth-functional conditional bridge, the Curry block on an internal conditional,
the provability and reflection boundaries, and the real-free checker soundness
bridge. \Cref{tab:claim-hierarchy} separates the layers so the schema is not
read as carrying weight it does not.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
Claim & Status & Depth \\
\midrule
No seed, no run & elementary theorem & bookkeeping \\
Fixed point gives self-hosting & elementary theorem & bookkeeping \\
Sequential = batch, fixed Bayes model & coherence theorem & bookkeeping \\
No truth-functional conditional bridge & substantive theorem & Cred content \\
No-ex-falso via zero-evidence & substantive theorem & Cred content \\
Curry internal-conditional block & substantive theorem & Cred content \\
Real-free checker soundness & formal artifact & Cred content \\
Compiler / AI / biology readings & structured analogy & expository \\
\bottomrule
\end{tabular}
\caption{The claim hierarchy. The bootstrap schema rows are elementary
bookkeeping; the substantive mathematical content is the Cred-specific theorems;
the cross-substrate readings are analogies. The paper claims each row only at its
stated depth.}
\label{tab:claim-hierarchy}
\end{table}

\section{Conclusion}
\label{sec:conclusion}

One schema, six substrates, a single formal core. Across probability, logic,
foundations, compilers, proof kernels, AI design loops, and the origin of
self-reproduction, the same five slots recur: a substrate, a seed, a transition,
a validator, and a self-host condition under which the transition reproduces the
state up to correctness. As a theorem, that schema is the Lean structure
\texttt{SeededSystem}, inhabited by \texttt{trivialSelfHost}, constrained by
\texttt{no\_empty\_bootstrap} and \texttt{selfHosts\_of\_fixed}, and large enough
to embed the irreducible-commitment system of Part~4 with both its fixed point
and its degenerate element self-hosting. As philosophical synthesis, the seed is
irreducible: it cannot be derived inside the system it grounds, only made
explicit, held small, and tested for self-consistency. The other rows are honest
analogies, marked as analogies, illustrating the schema without being proved by
it. The contribution is the schema, the one formal row that anchors it, and the
discipline of not claiming more.

\bibliographystyle{plain}
\bibliography{../cred}

\end{document}
