% Cred Part 3: Positioning Relative to Many-Valued Traditions
% Publication-quality paper
%
% References to Lean formalization: lean/Cred/Basic.lean, lean/Cred/Update.lean,
%                                   lean/Cred/Predicate.lean

\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
    pdfauthor={Anonymous},
    pdftitle={Cred Part 3: Positioning Relative to Many-Valued Traditions}
}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\cred}{\mathsf{cred}}
\newcommand{\Cred}{\mathsf{Cred}}
\newcommand{\negc}{\mathord{\sim}}
\newcommand{\conjc}{\otimes}
\newcommand{\disjc}{\sqcup}
\newcommand{\condc}[2]{({#1} \mid {#2})}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Nat}{\mathbb{N}}

\title{\textbf{Cred Part~3: Positioning Relative to Many-Valued Traditions}}

\author{Anonymous for blind review}
\date{February 2026}

\begin{document}

\maketitle

\begin{abstract}
Parts~1--2 established \emph{Cred} as a credence algebra on $[0,1]$ with
primitive conditioning governed by the chain rule and showed how to build
reasoning layers (valuations, consequence relations, updates) on this
foundation.  Part~3 positions Cred relative to the existing literature on
many-valued logic, primitive conditional probability, and fixed-point semantics.

We identify three distinguishing features of Cred: (1)~the chain rule as the
sole constraint on conditioning, with evidence-zero conditioning left
underdetermined rather than forced to~$1$; (2)~the product De~Morgan triplet
for algebraic operations, yielding a collapse homomorphism to the Kleene
algebra but not to G\"odel or \L ukasiewicz; and (3)~negation fixed points at
$\tfrac12$ that dissolve self-referential paradox without explosion.

We show that Cred's conditioning diverges from both RM3 and product-residuated
implication precisely at the evidence-zero boundary: residuation forces
$0 \to b = 1$ (ex falso), while Cred's chain rule imposes no constraint.
This positioning is consistent with de~Finetti's void conditional,
Popper--R\'enyi primitive conditional probability, and the imprecise probability
tradition's treatment of null events.

All technical claims are machine-checked in Lean~4; we provide a table mapping
theorems to their formal proofs.
\end{abstract}

%=============================================================================
\section{Introduction}
\label{sec:intro}
%=============================================================================

Part~1 presented the core algebra of Cred: credence values in $[0,1]$, the
complement $\negc c = 1 - c$, the product conjunction $c_1 \conjc c_2 = c_1
\cdot c_2$, the De~Morgan disjunction $c_1 \disjc c_2 = 1 - (1-c_1)(1-c_2)$,
and primitive conditioning via the chain rule
$\cred(A \mid B) \cdot \cred(B) = \cred(A \land B)$.
The key design choice was treating the evidence-zero case as underdetermined:
when $\cred(B) = 0$, the chain rule becomes $c \cdot 0 = 0$, satisfied by any
$c$, so conditioning imposes no constraint.

Part~2 added the reasoning layers: valuations that assign credences to
propositions, consequence relations with various threshold semantics, Bayesian
and Jeffrey update rules, and graded predicates with quantifiers as inf/sup.
Together, Parts~1--2 provide a complete framework for graded reasoning.

Part~3 addresses the question: \emph{how does Cred relate to existing
traditions?}  The goal is twofold: to prevent reinventing known results under
new names, and to clarify which aspects of Cred are inherited from prior work
and which represent genuine design choices.

We identify three main traditions that intersect with Cred:
\begin{enumerate}[label=(\roman*),nosep]
\item \textbf{Mathematical fuzzy logic}: the t-norm logics (product, G\"odel,
  \L ukasiewicz) and their algebraic semantics (BL-algebras, residuated
  lattices)~\cite{hajek1998,klement2000,galatos2007}.
\item \textbf{Primitive conditional probability}: R\'enyi, Popper, and
  van~Fraassen's approach where conditioning is primitive rather than defined
  by division~\cite{renyi1955,popper1959,vanfraassen1976}.
\item \textbf{Fixed-point semantics for self-reference}: Kripke's theory of
  truth, Gupta--Belnap revision theory, and de~Finetti's three-valued
  conditional events~\cite{kripke1975,gupta1993,definetti1936}.
\end{enumerate}

The main results of Part~3 are:
\begin{itemize}[nosep]
\item Cred's conditioning with the min-joint agrees with product-residuated
  implication for positive evidence and diverges precisely at zero
  (Sections~\ref{sec:product}--\ref{sec:zero}).
\item The negation fixed point at $\tfrac12$ is algebraic (forced by
  $c = \negc c$) and should be distinguished from proof-theoretic fixed points
  (Section~\ref{sec:fixed}).
\item No collapse homomorphism exists for G\"odel or \L ukasiewicz algebras,
  making the Kleene target unique (Section~\ref{sec:product}).
\item Cred's evidence-zero underdetermination aligns with de~Finetti's void,
  measure-theoretic disintegration, and imprecise probability
  (Section~\ref{sec:zero}).
\end{itemize}

\paragraph{Notation.}
We write $\negc c$ for negation, $c_1 \conjc c_2$ for conjunction,
$c_1 \disjc c_2$ for disjunction, and $\condc{A}{B}$ for the conditional
credence of $A$ given $B$.  The half value is $\tfrac12$.

%=============================================================================
\section{Background: The Many-Valued Traditions}
\label{sec:background}
%=============================================================================

This section reviews the relevant prior work.  The goal is to establish a
common vocabulary and to identify the specific technical results that Cred
builds on or departs from.

%-----------------------------------------------------------------------------
\subsection{Mathematical Fuzzy Logic}
\label{sec:fuzzy}
%-----------------------------------------------------------------------------

H\'ajek's \emph{Metamathematics of Fuzzy Logic}~\cite{hajek1998} organizes the
``big three'' continuous t-norm logics: \L ukasiewicz, G\"odel, and product.
Each is determined by a \emph{triangular norm} (t-norm): a commutative,
associative, monotone binary operation on $[0,1]$ with identity~$1$.

\begin{definition}[T-norms]
The three basic continuous t-norms are:
\begin{align*}
  \text{G\"odel (min):} \quad & a \land_G b = \min(a, b) \\
  \text{Product:} \quad & a \cdot b \\
  \text{\L ukasiewicz:} \quad & a \land_L b = \max(a + b - 1, 0)
\end{align*}
\end{definition}

Each t-norm determines a unique \emph{residuated implication} via the adjunction:
\[
a \to b = \sup\{c \in [0,1] : a \otimes c \leq b\}.
\]
The product-residuated (Goguen) implication is:
\[
a \to_P b = \begin{cases} 1 & \text{if } a \leq b \\ b/a & \text{otherwise} \end{cases}
= \min(b/a, 1) \text{ for } a > 0.
\]
Crucially, every residuated implication satisfies $0 \to b = 1$ for all $b$:
since $0 \otimes c = 0 \leq b$ for all $c$, the supremum is~$1$.
This is \emph{ex falso quodlibet} built into the residuation
framework~\cite{hajek1998}.

The algebraic semantics of t-norm logics uses \emph{BL-algebras} and their
special cases (MV-algebras for \L ukasiewicz, product algebras,
G\"odel algebras)~\cite{galatos2007,cintula2011}.

\emph{Pavelka-style} fuzzy logic~\cite{pavelka1979i,pavelka1979ii,pavelka1979iii}
introduces graded provability: theorems come with provability degrees, and
completeness is formulated in terms of these degrees.
This provides a baseline for any graded proof theory; whenever we discuss
graded consequence, we should be explicit about whether we mean a semantic
threshold relation (as in Part~2) or a graded proof system.

%-----------------------------------------------------------------------------
\subsection{Primitive Conditional Probability}
\label{sec:primitive}
%-----------------------------------------------------------------------------

R\'enyi~\cite{renyi1955,renyi1970} proposed an alternative axiomatization of
probability where the conditional probability $P(A \mid B)$ is primitive,
governed by axioms including the chain rule
$P(A \mid B) \cdot P(B \mid C) = P(A \land B \mid C)$.
Popper~\cite{popper1959} developed a similar approach.
In both systems, conditioning is not defined as a ratio $P(A \land B)/P(B)$
but taken as primitive, with the ratio emerging as a theorem when $P(B) > 0$.

Van~Fraassen~\cite{vanfraassen1976,vanfraassen1983} showed that the primitive
approach connects naturally to relevant logic: both avoid the inference
``from impossibility, anything follows.''  The key observation is that the
chain rule $P(A \mid B) \cdot P(B) = P(A \land B)$ at $P(B) = 0$ becomes
$c \cdot 0 = 0$, which is satisfied by any $c$---the conditional is undefined
rather than forced to~$1$.

%-----------------------------------------------------------------------------
\subsection{Conditional Events and Coherence}
\label{sec:coherence}
%-----------------------------------------------------------------------------

De~Finetti~\cite{definetti1936,definetti1937} proposed that a conditional event
$A \mid B$ is three-valued:
\begin{itemize}[nosep]
\item \emph{true} when $A \land B$,
\item \emph{false} when $\lnot A \land B$,
\item \emph{void} when $\lnot B$.
\end{itemize}
The void case is not assigned a truth value; conditioning on a false antecedent
yields no determinate value rather than vacuous truth.

Calabrese~\cite{calabrese1987} formalized conditional events algebraically as
``Boolean fractions.''
\'Egr\'e, Rossi, and Sprenger~\cite{egre2021} develop the modern logic of
indicative conditionals using de~Finetti tables, providing proof theory and
algebraic semantics for systems (DF/TT) that avoid vacuous truth.

The coherence tradition~\cite{gilio2013,gilio2014,coletti2002} extends this to
conditional random quantities, developing conjunction and iterated conditioning
operations while maintaining explicit treatment of probability-zero cases.

%-----------------------------------------------------------------------------
\subsection{Fixed-Point Semantics}
\label{sec:kripke}
%-----------------------------------------------------------------------------

Kripke's theory of truth~\cite{kripke1975} uses monotone operators and
fixed points to give consistent semantics for self-referential truth predicates.
The construction yields a partial valuation where the liar sentence is
undefined (has no truth value), rather than leading to contradiction.

Gupta and Belnap's revision theory~\cite{gupta1993} models truth as evolving
under revisions: starting from an arbitrary hypothesis about truth, one
applies a revision operator repeatedly, and stable patterns emerge.
This gives a dynamical semantics for self-reference.

Many-valued approaches to paradox~\cite{bergmann2008} assign graded truth
values to paradoxical sentences.  In a three-valued setting (e.g., K$_3$,
LP, RM3), the liar sentence is assigned the middle value $\tfrac12$.
This is related to but distinct from Cred's algebraic fixed point.

%=============================================================================
\section{Cred and Product Logic}
\label{sec:product}
%=============================================================================

Cred's algebraic operations are the product De~Morgan triplet: product
conjunction $c_1 \conjc c_2 = c_1 \cdot c_2$, probabilistic disjunction
$c_1 \disjc c_2 = c_1 + c_2 - c_1 c_2$, and standard complement
$\negc c = 1 - c$.  These are exactly the operations of product fuzzy
logic~\cite{klement2000}.

When one fixes the joint credence to be $\min(a, b)$ (the Fr\'echet upper
bound, maximal positive dependence), the induced conditional for positive
evidence is:
\[
\condc{b}{a} = \frac{\min(a, b)}{a} = \min(b/a, 1) \quad (a > 0).
\]
This is exactly the product-residuated implication.

%-----------------------------------------------------------------------------
\subsection{Agreement for Positive Evidence}
\label{sec:agreement}
%-----------------------------------------------------------------------------

For positive evidence, Cred conditioning (with the min-joint) and
product-residuated implication agree and are both Bayes-consistent.

\begin{theorem}[Bayes Consistency]
\label{thm:bayes}
The product-residuated implication is Bayes-consistent:
\[
(a \to_P b) \cdot a = (b \to_P a) \cdot b = \min(a, b).
\]
\textup{[Lean: \texttt{prod\_resid\_bayes\_consistent\_real}]}
\end{theorem}

\begin{proof}
By the formula for product residuation,
$(a \to_P b) \cdot a = \min(b/a, 1) \cdot a = \min(b, a)$.
By symmetry of min, $(b \to_P a) \cdot b = \min(a, b)$.
\end{proof}

\begin{remark}
Not all implications are Bayes-consistent.  The RM3 material conditional
$a \to_{RM3} b = \max(1-a, b)$ fails:
\[
(a \to_{RM3} b) \cdot a \neq (b \to_{RM3} a) \cdot b
\]
in general (counterexample: $a = \tfrac12$, $b = 0$).
\textup{[Lean: \texttt{rm3\_not\_bayes\_consistent}]}
\end{remark}

%-----------------------------------------------------------------------------
\subsection{Divergence at Zero}
\label{sec:diverge}
%-----------------------------------------------------------------------------

The divergence between Cred and product logic occurs precisely at the
evidence-zero boundary.

\begin{theorem}[Cred Blocks Ex Falso]
\label{thm:no-ex-falso}
In Cred, conditioning on evidence with credence~$0$ is unconstrained: for any
credence $c$, there exists a valid conditioning structure with
$\condc{b}{0} = c$.
\textup{[Lean: \texttt{cred\_no\_ex\_falso}, \texttt{conditioning\_zero\_any}]}
\end{theorem}

\begin{proof}
The chain rule at zero evidence is $c \cdot 0 = 0$, which holds for any $c$.
Thus any credence value satisfies the constraint.
\end{proof}

\begin{remark}
Product-residuated implication gives $0 \to_P b = 1$ for all $b$; this is
\emph{ex falso}.  Cred replaces this with ``unconstrained''---any value is
consistent with the chain rule.  This is Cred's primary divergence from
standard residuated logics.
\end{remark}

%-----------------------------------------------------------------------------
\subsection{Why Kleene Collapse Works}
\label{sec:collapse}
%-----------------------------------------------------------------------------

Part~1 proved that Cred admits a surjective homomorphism to the Kleene
three-valued algebra (RM3 for $\{\negc, \conjc, \disjc\}$), but no analogous
homomorphism exists for G\"odel or \L ukasiewicz algebras.

\begin{theorem}[No G\"odel Collapse]
\label{thm:no-godel}
There is no function $\phi: [0,1] \to \{0, \tfrac12, 1\}$ that preserves
G\"odel negation $\negc_G c = \begin{cases} 1 & c = 0 \\ 0 & c > 0 \end{cases}$.
\textup{[Lean: \texttt{no\_godel\_collapse}]}
\end{theorem}

\begin{proof}
G\"odel negation has no fixed point: $\negc_G 0 = 1 \neq 0$,
$\negc_G \tfrac12 = 0 \neq \tfrac12$, $\negc_G 1 = 0 \neq 1$.
But $\phi(\tfrac12)$ must satisfy $\phi(\negc \tfrac12) = \negc_G \phi(\tfrac12)$;
since $\negc \tfrac12 = \tfrac12$ in Cred, this requires
$\phi(\tfrac12) = \negc_G \phi(\tfrac12)$, a fixed point---contradiction.
\end{proof}

\begin{theorem}[No \L ukasiewicz Collapse]
\label{thm:no-luk}
There is no function $\phi: [0,1] \to \{0, \tfrac12, 1\}$ that preserves both
standard negation and \L ukasiewicz conjunction
$a \land_L b = \max(a + b - 1, 0)$.
\textup{[Lean: \texttt{no\_luk\_collapse}]}
\end{theorem}

\begin{proof}
Negation preservation forces $\phi(\tfrac12) = \tfrac12$ (since
$\negc \tfrac12 = \tfrac12$).
For \L ukasiewicz conjunction, $\tfrac12 \land_L \tfrac12 = 0$, so we need
$\phi(\tfrac12 \land_L \tfrac12) = \phi(0) = 0$ and
$\phi(\tfrac12) \land_L \phi(\tfrac12) = \tfrac12 \land_L \tfrac12 = 0$.
But pick $c = \sqrt{\tfrac12}$; then $c \conjc c = \tfrac12$ (Cred product),
so $\phi(c \conjc c) = \phi(\tfrac12) = \tfrac12$.
Meanwhile $\phi(c) \land_L \phi(c) \in \{0, 1\}$ (since $v \land_L v$ for
$v \in \{0, \tfrac12, 1\}$ is in $\{0, 1\}$, not $\tfrac12$)---contradiction.
\end{proof}

%=============================================================================
\section{Self-Reference: Fixed Points Without Triviality}
\label{sec:fixed}
%=============================================================================

Cred's most visible self-reference result is the negation fixed point.

%-----------------------------------------------------------------------------
\subsection{Negation Fixed Point}
\label{sec:liar}
%-----------------------------------------------------------------------------

\begin{theorem}[Liar Fixed Point]
\label{thm:liar}
The equation $c = \negc c$ has exactly one solution: $c = \tfrac12$.
\textup{[Lean: \texttt{liar\_fixed\_point}, \texttt{neg\_fixed\_point\_unique}]}
\end{theorem}

\begin{proof}
If $c = \negc c = 1 - c$, then $2c = 1$, so $c = \tfrac12$.
Conversely, $\negc \tfrac12 = 1 - \tfrac12 = \tfrac12$.
\end{proof}

This result is \emph{algebraic}: it follows directly from the negation
operation on $[0,1]$ and does not require any construction involving truth
predicates or diagonalization.

%-----------------------------------------------------------------------------
\subsection{Product Fixed Points}
\label{sec:conj-fixed}
%-----------------------------------------------------------------------------

\begin{theorem}[Conjunction Idempotence]
\label{thm:idemp}
The equation $c \conjc c = c$ holds if and only if $c \in \{0, 1\}$.
\textup{[Lean: \texttt{conj\_idempotent\_iff}]}
\end{theorem}

\begin{proof}
$c^2 = c$ iff $c(c-1) = 0$ iff $c \in \{0, 1\}$.
\end{proof}

This shows that intermediate credences do \emph{not} satisfy Boolean
idempotence.  The product conjunction distinguishes between ``partial
confidence in $A$'' and ``partial confidence in $A \land A$'': these differ
when $0 < \cred(A) < 1$.

%-----------------------------------------------------------------------------
\subsection{Russell Fixed Point}
\label{sec:russell}
%-----------------------------------------------------------------------------

The Russell predicate $R(x) = \negc(x(x))$ applied to itself yields
$R(R) = \negc(R(R))$, which is exactly the liar equation.

\begin{theorem}[Russell Fixed Point]
\label{thm:russell}
If $c = \negc c$ (the Russell equation), then $c = \tfrac12$.
\textup{[Lean: \texttt{russell\_fixed\_point}]}
\end{theorem}

This is a direct corollary of \cref{thm:liar}.

%-----------------------------------------------------------------------------
\subsection{Comparison to Kripke and Revision Theory}
\label{sec:compare-fixed}
%-----------------------------------------------------------------------------

Cred's fixed-point result should be distinguished from the proof-theoretic
fixed points in Kripke~\cite{kripke1975} and Gupta--Belnap~\cite{gupta1993}:

\begin{itemize}[nosep]
\item \textbf{Kripke}: constructs a partial valuation where the liar is
  \emph{undefined} (has no truth value in the minimal fixed point).
  The construction uses monotone operators on partial valuations.
\item \textbf{Gupta--Belnap}: models truth as evolving under revision; the liar
  cycles rather than settling to a single value.
\item \textbf{Cred}: assigns the liar the definite value $\tfrac12$.
  This is purely algebraic---no construction of truth predicates or
  revision sequences is needed.
\end{itemize}

The three approaches are complementary.  Cred's algebraic result shows that
\emph{if} one commits to $[0,1]$-valued credences with standard negation,
\emph{then} self-referential negation is forced to~$\tfrac12$.
Kripke and revision theory analyze what happens in proof-theoretic settings
where truth predicates are constructed internally.

%=============================================================================
\section{The Evidence-Zero Boundary}
\label{sec:zero}
%=============================================================================

The evidence-zero boundary is where Cred diverges most clearly from both
classical probability (where $P(A \mid B)$ is undefined when $P(B) = 0$) and
residuated logic (where $0 \to b = 1$).

%-----------------------------------------------------------------------------
\subsection{Probabilistic: Disintegration}
\label{sec:disintegration}
%-----------------------------------------------------------------------------

Standard measure theory handles conditioning on null events via
\emph{regular conditional probabilities} and \emph{disintegration}
\cite{chang1997}.  Given a probability space and a sub-sigma-algebra,
disintegration provides conditional probabilities that are unique only
\emph{almost surely}---different choices agree on sets of measure~$1$ but
may differ on null sets.

The category-theoretic perspective (Markov categories~\cite{fritz2020})
treats conditioning structurally: the conditional is characterized by a
universal property rather than division, making explicit which uniqueness
guarantees follow from the structure.

%-----------------------------------------------------------------------------
\subsection{Imprecise: Set-Valued Conditionals}
\label{sec:imprecise}
%-----------------------------------------------------------------------------

The imprecise probability tradition~\cite{walley1991,troffaes2014} takes a
different approach: when conditioning would be undefined or non-unique, return
a \emph{set} or \emph{interval} of admissible values rather than a single
number.  This makes explicit that the evidence does not fully determine the
posterior.

Cred's underdetermination at zero aligns naturally with this perspective:
the chain rule constrains the conditional to the \emph{entire} interval
$[0,1]$ when evidence has credence~$0$.

%-----------------------------------------------------------------------------
\subsection{Cred's Stance: Underdetermination as Design}
\label{sec:underdetermination}
%-----------------------------------------------------------------------------

Cred makes evidence-zero underdetermination a \emph{first-class design
feature} rather than an exception to be patched.

\begin{theorem}[Bayesian Underdetermined at Zero]
\label{thm:bayes-zero}
When evidence has credence~$0$, Bayesian conditionalization is underdetermined:
any credence value is consistent with the chain rule.
\textup{[Lean: \texttt{bayesian\_underdetermined\_at\_zero}]}
\end{theorem}

\begin{theorem}[Jeffrey Underdetermined at Zero]
\label{thm:jeffrey-zero}
When evidence has credence~$0$, Jeffrey conditionalization depends on an
underdetermined conditional and is therefore itself underdetermined.
\textup{[Lean: \texttt{jeffrey\_underdetermined\_at\_zero}]}
\end{theorem}

This is intentional: impossible evidence (credence~$0$) provides no constraint
on posterior belief.  The philosophy is ``inference constrains'': evidence
narrows possibilities; impossible evidence cannot narrow anything.

%-----------------------------------------------------------------------------
\subsection{Path Dependence and the Borel--Kolmogorov Connection}
\label{sec:path}
%-----------------------------------------------------------------------------

Part~1 establishes algebraically that the conditional's limit as evidence
$b \to 0$ is \emph{path-dependent}: it varies with the ratio $a/b$
(equivalently, the approach angle in polar coordinates), not with $b$ alone.
This is not merely underdetermination---it is a structured family of limits
depending on how the credences approach zero together.

This path dependence has the same structure as the \emph{Borel--Kolmogorov
paradox}~\cite{borel1909,kolmogorov1933} in measure theory: conditioning on
a null event is not uniquely determined without specifying the limiting
procedure.  Regular conditional probabilities are defined only almost
surely---they may differ on null sets.  The algebraic path dependence in
Cred is the same phenomenon: the chain rule does not select a unique limit
at zero; it selects an angle-dependent family of limits.

The totalization convention $\cred(A \mid 0) = 1$ (ex falso) is known in
probability logics~\cite{adams1975} and has been recognized as problematic
in product-fuzzy encodings of conditional probability~\cite{hajek1998}.
Cred's path-dependence result shows this convention is not forced by the
algebra---it must be added as an extra axiom.

%-----------------------------------------------------------------------------
\subsection{De Finetti's Void as Discrete Case}
\label{sec:void}
%-----------------------------------------------------------------------------

De~Finetti's three-valued conditional events~\cite{definetti1936} assign
\emph{void} to $A \mid B$ when $B$ is false.  In the graded setting, this
corresponds to evidence credence~$0$: when $\cred(B) = 0$, the conditional
$\condc{A}{B}$ is not forced to any particular value.

The \'Egr\'e--Rossi--Sprenger systems (DF/TT)~\cite{egre2021} formalize this
in a proof-theoretic setting, providing rules that avoid vacuous truth.
Cred's chain-rule approach achieves the same effect algebraically: there is
no rule that forces the conditional to~$1$ when evidence is impossible.

%=============================================================================
\section{Synthesis and Non-Claims}
\label{sec:synthesis}
%=============================================================================

%-----------------------------------------------------------------------------
\subsection{What Cred Adds: The Synthesis Statement}
\label{sec:what-adds}
%-----------------------------------------------------------------------------

The literature review suggests a clean synthesis:

\begin{quote}
\textbf{Synthesis.} Cred isolates the chain rule as the sole algebraic
constraint on conditioning and combines it with the product De~Morgan triplet
on $[0,1]$.  It is compatible with both (i)~truth-functional many-valued logics
(when a fixed joint such as min is chosen) and (ii)~probability-style semantics
(when joints are supplied from empirical data).  Its distinguishing choice is
to leave the evidence-zero boundary underdetermined by default, in line with
de~Finetti conditional events and imprecise probability, rather than forcing
ex falso via residuation.
\end{quote}

The synthesis makes clear which parts of Cred are inherited:
\begin{itemize}[nosep]
\item The product De~Morgan triplet is standard in fuzzy logic~\cite{klement2000}.
\item The chain rule as primitive comes from R\'enyi--Popper~\cite{renyi1955,popper1959}.
\item The three-valued collapse is related to K$_3$/LP/RM3~\cite{priest2008}.
\item The fixed-point treatment of self-reference is related to
  Kripke~\cite{kripke1975}.
\end{itemize}

What Cred adds is the deliberate \emph{combination} of these elements with a
specific stance on evidence-zero: underdetermination as a first-class feature.

%-----------------------------------------------------------------------------
\subsection{What Cred Does NOT Claim}
\label{sec:non-claims}
%-----------------------------------------------------------------------------

To prevent overclaiming, we explicitly note what Cred does \emph{not} provide:

\begin{enumerate}[label=(\arabic*),nosep]
\item \textbf{A graded proof theory.}  Part~2 provides consequence relations
  (threshold semantics), but no proof calculus with graded derivability in the
  Pavelka sense.  This is future work.
\item \textbf{A resolution of G\"odel incompleteness.}  The $\tfrac12$ fixed
  point for the liar is \emph{algebraic}.  G\"odel incompleteness is
  \emph{proof-theoretic} and applies to any sufficiently strong formal system,
  including any formalization of Cred.
\item \textbf{A complete foundation for mathematics.}  Cred is an algebra, not
  a set theory.  Building graded set theory (comprehension, extensionality)
  requires additional choices, as explored in~\cite{takeuti1984}.
\item \textbf{A claim that residuated logics are wrong.}  Residuation is a
  mathematically coherent and well-developed framework.  Cred simply makes a
  different design choice at zero, suited to epistemic/probabilistic contexts
  where ``from nothing, nothing follows'' is preferred to ``from nothing,
  anything follows.''
\end{enumerate}

%-----------------------------------------------------------------------------
\subsection{Open Directions}
\label{sec:open}
%-----------------------------------------------------------------------------

Part~3 opens several directions for future work:

\begin{itemize}[nosep]
\item \textbf{Graded proof theory}: develop a proof calculus for Cred with
  graded provability degrees, following the Pavelka
  approach~\cite{pavelka1979i}.
\item \textbf{Set-valued updates}: formalize update mechanisms that return
  intervals when evidence has low credence, connecting to imprecise
  probability~\cite{walley1991}.
\item \textbf{Graded set theory}: work through comprehension and extensionality
  in a Cred-valued setting, using the many-valued set theory literature
  \cite{takeuti1984} as a guide.
\item \textbf{Asymptotic proofs}: investigate what ``proof'' means when
  credence approaches but never reaches~$1$.
\end{itemize}

%=============================================================================
\section{Lean Formalization}
\label{sec:lean}
%=============================================================================

The main results of Parts~1--3 are machine-checked in Lean~4 using mathlib.
\Cref{tab:lean} maps the key theorems to their Lean identifiers.

\begin{table}[h]
\centering
\caption{Lean formalization of key results}
\label{tab:lean}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Result} & \textbf{Lean identifier} & \textbf{File} \\
\midrule
\multicolumn{3}{l}{\emph{Core algebra (Part 1)}} \\
Negation fixed point & \texttt{liar\_fixed\_point} & Basic.lean \\
Fixed point uniqueness & \texttt{neg\_fixed\_point\_unique} & Basic.lean \\
Conjunction idempotence & \texttt{conj\_idempotent\_iff} & Basic.lean \\
Non-distributivity & \texttt{conj\_disj\_not\_distrib} & Basic.lean \\
\midrule
\multicolumn{3}{l}{\emph{Conditioning (Part 1)}} \\
Conditioning at zero & \texttt{conditioning\_zero\_any} & Basic.lean \\
Zero forces joint zero & \texttt{conditioning\_zero\_forces\_joint\_zero} & Basic.lean \\
Conditioning uniqueness & \texttt{conditioning\_unique} & Basic.lean \\
\midrule
\multicolumn{3}{l}{\emph{Comparison (Part 1)}} \\
Cred blocks ex falso & \texttt{cred\_no\_ex\_falso} & Basic.lean \\
RM3 has ex falso & \texttt{rm3\_ex\_falso} & Basic.lean \\
Bayes consistency & \texttt{prod\_resid\_bayes\_consistent\_real} & Basic.lean \\
RM3 not Bayes consistent & \texttt{rm3\_not\_bayes\_consistent} & Basic.lean \\
No G\"odel collapse & \texttt{no\_godel\_collapse} & Basic.lean \\
No \L ukasiewicz collapse & \texttt{no\_luk\_collapse} & Basic.lean \\
\midrule
\multicolumn{3}{l}{\emph{Updates (Part 2)}} \\
Bayesian underdetermined & \texttt{bayesian\_underdetermined\_at\_zero} & Update.lean \\
Jeffrey underdetermined & \texttt{jeffrey\_underdetermined\_at\_zero} & Update.lean \\
\midrule
\multicolumn{3}{l}{\emph{Predicates (Part 2)}} \\
Russell fixed point & \texttt{russell\_fixed\_point} & Predicate.lean \\
\bottomrule
\end{tabular}
\end{table}

%=============================================================================
% Bibliography
%=============================================================================

\begin{thebibliography}{99}

\bibitem{adams1975}
Adams, E.W.
\newblock \emph{The Logic of Conditionals}.
\newblock Reidel, 1975.

\bibitem{bergmann2008}
Bergmann, M.
\newblock \emph{An Introduction to Many-Valued and Fuzzy Logic}.
\newblock Cambridge University Press, 2008.

\bibitem{borel1909}
Borel, \'E.
\newblock \emph{\'El\'ements de la th\'eorie des probabilit\'es}.
\newblock Hermann, Paris, 1909.

\bibitem{calabrese1987}
Calabrese, P.
\newblock An algebraic synthesis of the foundations of logic and probability.
\newblock \emph{Information Sciences}, 42(2):187--237, 1987.

\bibitem{chang1997}
Chang, J.T. and Pollard, D.
\newblock Conditioning as disintegration.
\newblock \emph{Statistica Neerlandica}, 51(3):287--317, 1997.

\bibitem{cintula2011}
Cintula, P., H\'ajek, P., and Noguera, C., editors.
\newblock \emph{Handbook of Mathematical Fuzzy Logic}, Vols.~1--2.
\newblock College Publications, 2011.

\bibitem{coletti2002}
Coletti, G. and Scozzafava, R.
\newblock \emph{Probabilistic Logic in a Coherent Setting}.
\newblock Trends in Logic 15. Kluwer/Springer, 2002.

\bibitem{definetti1936}
de Finetti, B.
\newblock La logique de la probabilit\'e.
\newblock \emph{Actes du Congr\`es International de Philosophie Scientifique},
4:31--39, 1936.

\bibitem{definetti1937}
de Finetti, B.
\newblock La pr\'evision: ses lois logiques, ses sources subjectives.
\newblock \emph{Annales de l'Institut Henri Poincar\'e}, 7(1):1--68, 1937.

\bibitem{egre2021}
\'Egr\'e, P., Rossi, L., and Sprenger, J.
\newblock De Finettian logics of indicative conditionals {Part II}: Proof theory and algebraic semantics.
\newblock \emph{Journal of Philosophical Logic}, 50:215--247, 2021.

\bibitem{fritz2020}
Fritz, T.
\newblock A synthetic approach to {Markov} kernels, conditional independence and theorems on sufficient statistics.
\newblock \emph{Advances in Mathematics}, 370:107239, 2020.

\bibitem{galatos2007}
Galatos, N., Jipsen, P., Kowalski, T., and Ono, H.
\newblock \emph{Residuated Lattices: An Algebraic Glimpse at Substructural Logics}.
\newblock Elsevier, 2007.

\bibitem{gilio2013}
Gilio, A. and Sanfilippo, G.
\newblock Conjunction, disjunction and iterated conditioning of conditional events.
\newblock In Kruse, R. et al., editors, \emph{Synergies of Soft Computing and Statistics},
pages 399--407. Springer, 2013.

\bibitem{gilio2014}
Gilio, A. and Sanfilippo, G.
\newblock Conditional random quantities and compounds of conditionals.
\newblock \emph{Studia Logica}, 102(4):709--729, 2014.

\bibitem{gupta1993}
Gupta, A. and Belnap, N.
\newblock \emph{The Revision Theory of Truth}.
\newblock MIT Press, 1993.

\bibitem{hajek1998}
H\'ajek, P.
\newblock \emph{Metamathematics of Fuzzy Logic}.
\newblock Kluwer, 1998.

\bibitem{klement2000}
Klement, E.P., Mesiar, R., and Pap, E.
\newblock \emph{Triangular Norms}.
\newblock Kluwer, 2000.

\bibitem{kolmogorov1933}
Kolmogorov, A.N.
\newblock \emph{Grundbegriffe der Wahrscheinlichkeitsrechnung}.
\newblock Springer, Berlin, 1933.
\newblock English translation: \emph{Foundations of the Theory of Probability},
Chelsea, 1950.

\bibitem{kripke1975}
Kripke, S.
\newblock Outline of a theory of truth.
\newblock \emph{Journal of Philosophy}, 72(19):690--716, 1975.

\bibitem{pavelka1979i}
Pavelka, J.
\newblock On fuzzy logic {I}: Many-valued rules of inference.
\newblock \emph{Zeitschrift f\"ur mathematische Logik und Grundlagen der Mathematik}, 25:45--52, 1979.

\bibitem{pavelka1979ii}
Pavelka, J.
\newblock On fuzzy logic {II}: Enriched residuated lattices and semantics of propositional calculi.
\newblock \emph{Zeitschrift f\"ur mathematische Logik und Grundlagen der Mathematik}, 25:119--134, 1979.

\bibitem{pavelka1979iii}
Pavelka, J.
\newblock On fuzzy logic {III}: Semantical completeness of some many-valued propositional calculi.
\newblock \emph{Zeitschrift f\"ur mathematische Logik und Grundlagen der Mathematik}, 25:447--464, 1979.

\bibitem{popper1959}
Popper, K.R.
\newblock \emph{The Logic of Scientific Discovery}.
\newblock Hutchinson, London, 1959.

\bibitem{priest2008}
Priest, G.
\newblock \emph{An Introduction to Non-Classical Logic}.
\newblock Cambridge University Press, 2nd edition, 2008.

\bibitem{renyi1955}
R\'enyi, A.
\newblock On a new axiomatic theory of probability.
\newblock \emph{Acta Mathematica Academiae Scientiarum Hungaricae}, 6:285--335, 1955.

\bibitem{renyi1970}
R\'enyi, A.
\newblock \emph{Foundations of Probability}.
\newblock Holden-Day, 1970.

\bibitem{takeuti1984}
Takeuti, G. and Titani, S.
\newblock Fuzzy logic and fuzzy set theory.
\newblock \emph{Journal of Symbolic Logic}, 49(3):851--866, 1984.

\bibitem{troffaes2014}
Troffaes, M.C.M. and de Cooman, G.
\newblock \emph{Lower Previsions}.
\newblock Wiley, 2014.

\bibitem{vanfraassen1976}
van Fraassen, B.C.
\newblock Probabilities of conditionals.
\newblock In Harper, W.L. and Hooker, C.A., editors, \emph{Foundations of Probability Theory, Statistical Inference, and Statistical Theories of Science},
Vol.~1, pages 261--308. Reidel, 1976.

\bibitem{vanfraassen1983}
van Fraassen, B.C.
\newblock Gentlemen's wagers: Relevant logic and probability.
\newblock \emph{Philosophical Studies}, 43:47--61, 1983.

\bibitem{walley1991}
Walley, P.
\newblock \emph{Statistical Reasoning with Imprecise Probabilities}.
\newblock Chapman and Hall, 1991.

\end{thebibliography}

\end{document}
