% Cred: A Credence Algebra with Primitive Conditioning
% Publication-quality paper
%
% References to Lean formalization: lean/Cred/Basic.lean

\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
    pdfauthor={Anonymous},
    pdftitle={Cred: A Credence Algebra with Primitive Conditioning}
}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

% Custom commands
\newcommand{\cred}{\mathsf{cred}}
\newcommand{\Cred}{\mathsf{Cred}}
\newcommand{\negc}{\mathord{\sim}}
\newcommand{\conjc}{\otimes}
\newcommand{\disjc}{\sqcup}
\newcommand{\condc}[2]{({#1} \mid {#2})}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Nat}{\mathbb{N}}

\title{\textbf{Cred: A Credence Algebra with Primitive Conditioning}\\[0.5em]
\large Formalization and Connections to Relevant Logic}

\author{Anonymous for blind review}
\date{February 2026}

\begin{document}

\maketitle

\begin{abstract}
We present \emph{Cred}, a weakening of probability theory that retains values in
$[0,1]$ and the complement but drops the measure space and sigma-additivity.
The sole structural axiom is the chain rule: $\cred(A \mid B) \cdot \cred(B) =
\cred(A \land B)$.
When evidence is positive, division emerges as the unique solution; when evidence
has credence zero, the constraint is satisfied by any value---the conditional is
\emph{unconstrained}, and no division is needed.
Every probability space satisfies Cred's axioms, but Cred requires neither a
sigma-algebra nor additivity.
We establish a surjective homomorphism from Cred to the three-valued Kleene
algebra---the shared foundation of K$_3$, LP, and RM3---and prove this collapse
target is unique: no analogous homomorphism exists for \L ukasiewicz, G\"odel,
or classical algebras.
Applying the chain rule from both directions (Bayes' theorem) selects the
product residuated implication as the unique standard Bayes-consistent binary arrow.
Negation has a unique fixed point at $\tfrac{1}{2}$.
All core results are machine-checked in Lean~4.
\end{abstract}

%=============================================================================
\section{Introduction}
\label{sec:intro}
%=============================================================================

A \emph{credence} is a real number in $[0,1]$ representing degree of belief---the
standard representation in formal epistemology~\cite{ramsey1931,definetti1937,jaynes2003}.
The value $0$ represents impossibility, $1$ represents certainty, and intermediate
values represent partial confidence.

Probability theory builds on credences by introducing a sigma-algebra of events,
a measure, and sigma-additivity.
Conditional probability is then derived via division:
$P(A \mid B) = P(A \land B)/P(B)$ when $P(B) > 0$; when $P(B) = 0$, the ratio
is undefined.
R\'enyi~\cite{renyi1955,renyi1970} and Popper~\cite{popper1959} proposed an
alternative: treat conditioning as \emph{primitive}, governed by the chain rule
$P(A \mid B) \cdot P(B) = P(A \land B)$ without invoking division in the
definition.
Van Fraassen~\cite{vanfraassen1983} showed that this primitive approach
connects naturally to relevant logic: both avoid forcing ``from impossibility,
anything follows.''

Cred takes this further: drop the measure entirely.
Keep only values in $[0,1]$, the complement $\negc c = 1 - c$, and the chain
rule $\cred(A \mid B) \cdot \cred(B) = \cred(A \land B)$ as the sole structural
axiom for conditioning.
This is strictly weaker than probability---every probability space satisfies
Cred's axioms, but Cred requires neither a sigma-algebra nor additivity.
When $\cred(B) = 0$, the chain rule becomes $\cred(A \mid B) \cdot 0 = 0$,
satisfied by \emph{any} value of $\cred(A \mid B)$: the conditional is
unconstrained.
Division emerges as the unique solution when evidence is positive; at zero,
no division is needed.

The product $c_1 \cdot c_2$ and its De Morgan dual $c_1 + c_2 - c_1 c_2$ are
available as algebraic operations on $[0,1]$---the same product that appears in
the chain rule.
These form the \emph{product De Morgan triplet}, well-known in fuzzy
logic~\cite{klement2000,hajek1998}.
But where product fuzzy logic uses residuated implication (forcing $0 \to y = 1$,
i.e.\ ex falso), Cred uses chain-rule conditioning (leaving the conditional
unconstrained at zero).

\textbf{Contributions.}
\begin{enumerate}[label=(\arabic*)]
\item A weakening of probability that retains the chain rule as primitive
conditioning, where division is derived (unique when evidence $> 0$) and
conditioning is unconstrained at evidence zero (Section~\ref{sec:conditioning}).
\item A surjective homomorphism from Cred to the Kleene algebra (underlying
K$_3$, LP, and RM3) for $\{\negc, \conjc, \disjc\}$, with impossibility results
for alternative targets and machine-checked proof (Section~\ref{sec:connections}).
\item Bayes consistency---the chain rule applied from both directions---selects
the product residuated implication as the unique standard Bayes-consistent
binary arrow; Bayes-consistent arrows are exactly the copula-derived arrows
(Section~\ref{sec:connections}).
\item Fixed-point analysis: $\tfrac{1}{2}$ is the unique fixed point of negation
(Section~\ref{sec:fixed-points}).
\item Lean~4 formalization of the core algebra (Section~\ref{sec:lean}).
\end{enumerate}

\textbf{Outline.}
Section~\ref{sec:algebra} defines the credence algebra.
Section~\ref{sec:conditioning} develops primitive conditioning.
Section~\ref{sec:fixed-points} treats fixed points.
Section~\ref{sec:connections} relates Cred to other systems.
Section~\ref{sec:lean} describes the Lean formalization.

%=============================================================================
\section{The Credence Algebra}
\label{sec:algebra}
%=============================================================================

\subsection{Primitive Operations}

\begin{definition}[Credence]
\label{def:credence}
A \emph{credence} is a real number $c \in [0, 1]$.
The set of credences is $\Cred = [0, 1] \subset \Real$.
\end{definition}

\begin{definition}[Negation]
\label{def:neg}
The \emph{negation} of $c$ is $\negc c = 1 - c$.
\end{definition}

Negation is the standard complement: if we assign credence $0.7$ to proposition $A$,
we assign credence $0.3$ to its negation.

\begin{definition}[Product]
\label{def:conj}
The \emph{product} of $c_1$ and $c_2$ is $c_1 \conjc c_2 = c_1 \cdot c_2$.
\end{definition}

The product is multiplication on $[0,1]$---the same operation that appears in the
chain rule (Section~\ref{sec:conditioning}).
It equals the joint probability $P(A \land B)$ when $A$ and $B$ are independent,
hence the alternative name \emph{independence product}.
For dependent propositions, the joint credence must be supplied as an external
parameter; $\conjc$ does not claim to compute the joint of arbitrary propositions.

\begin{definition}[Disjunction]
\label{def:disj}
The \emph{disjunction} of $c_1$ and $c_2$ is defined via De Morgan duality:
\[
c_1 \disjc c_2 = \negc(\negc c_1 \conjc \negc c_2) = c_1 + c_2 - c_1 \cdot c_2
\]
\end{definition}

The triple $(\conjc, \disjc, \negc)$ is the \emph{product De Morgan triplet}:
conjunction is the product t-norm, disjunction its dual t-conorm (the
``probabilistic sum''), and negation the standard
complement~\cite{klement2000,hajek1998}.
These operations are truth-functional---determined entirely by the operand
values---which is what makes the collapse homomorphism to RM3 possible
(Section~\ref{sec:connections}).

\begin{remark}[Credence vs.\ Probability]
\label{rem:not-probability}
Cred's core consists of three ingredients: values in $[0,1]$, the complement
$\negc c = 1 - c$, and the chain rule for conditioning.
Probability adds a sigma-algebra, a measure, and sigma-additivity; every
probability space therefore satisfies Cred's axioms, making probability a
special case.
In probability, the joint $P(A \land B)$ is determined by the measure---not
by the product of marginals (unless events are independent).
Cred makes no such determination: $c_1 \conjc c_2$ is multiplication on
$[0,1]$, not a claim about the actual joint of arbitrary propositions.
The actual joint is an external input to conditioning.
Cred's disjunction $c_1 \disjc c_2 = c_1 + c_2 - c_1 c_2$ is the De Morgan
dual of the product; probability's inclusion-exclusion
$P(A \lor B) = P(A) + P(B) - P(A \land B)$ uses the actual joint instead.
\end{remark}

\subsection{Algebraic Properties}

\begin{theorem}[Algebraic Properties]
\label{thm:algebraic}
The following hold for all $c, c_1, c_2, c_3 \in [0,1]$:
\begin{enumerate}[label=(\alph*)]
\item $\negc\negc c = c$ \hfill (involution)
\item $\negc 0 = 1$, $\negc 1 = 0$ \hfill (boundary)
\item $c_1 \conjc c_2 = c_2 \conjc c_1$, $(c_1 \conjc c_2) \conjc c_3 = c_1 \conjc (c_2 \conjc c_3)$ \hfill (comm., assoc.)
\item $c \conjc 1 = c$, $c \conjc 0 = 0$ \hfill (identity, annihilator)
\item $c_1 \disjc c_2 = c_2 \disjc c_1$, $(c_1 \disjc c_2) \disjc c_3 = c_1 \disjc (c_2 \disjc c_3)$ \hfill (comm., assoc.)
\item $c \disjc 0 = c$, $c \disjc 1 = 1$ \hfill (identity, annihilator)
\item $\negc(c_1 \conjc c_2) = \negc c_1 \disjc \negc c_2$ \hfill (De Morgan)
\end{enumerate}
\end{theorem}
\begin{proof}
Routine verification.
Lean: \texttt{neg\_neg}, \texttt{conj\_comm}, \texttt{conj\_assoc},
\texttt{disj\_comm}, \texttt{disj\_assoc}, \texttt{de\_morgan\_conj}.
\end{proof}

The structure $([0,1], \conjc, \disjc, \negc, 0, 1)$ consists of two dual
commutative monoids ($([0,1], \conjc, 1)$ and $([0,1], \disjc, 0)$) connected
by an order-reversing involution $\negc$ satisfying De Morgan's laws---a
bounded De Morgan algebra in which the lattice absorption and idempotence
laws fail (since $\conjc$ is not idempotent; see below).

\subsection{Non-Idempotence and Non-Distributivity}

Unlike the two-element Boolean algebra $\{0,1\}$, the credence algebra is neither
idempotent nor distributive.

\begin{theorem}[Non-Idempotence]
\label{thm:idempotence}
$c \conjc c = c$ if and only if $c \in \{0, 1\}$.
\end{theorem}
\begin{proof}
$c^2 = c$ iff $c(c-1) = 0$ iff $c \in \{0,1\}$.
In particular, $\tfrac{1}{2} \conjc \tfrac{1}{2} = \tfrac{1}{4} \neq \tfrac{1}{2}$.
Lean: \texttt{conj\_idempotent\_iff}.
\end{proof}

The product $c \conjc c = c^2$ is the credence of two \emph{independent} copies
of a proposition with credence $c$---not the credence of ``$A \land A$'' (which
equals $\cred(A)$).

\begin{theorem}[Non-Distributivity]
\label{thm:nondistrib}
The independence product does not distribute over disjunction.
\end{theorem}
\begin{proof}
Counterexample at $c_1 = c_2 = c_3 = \tfrac{1}{2}$:
\begin{align*}
\text{LHS} &= \tfrac{1}{2} \conjc (\tfrac{1}{2} \disjc \tfrac{1}{2})
= \tfrac{1}{2} \cdot (\tfrac{1}{2} + \tfrac{1}{2} - \tfrac{1}{4})
= \tfrac{1}{2} \cdot \tfrac{3}{4} = \tfrac{3}{8} \\
\text{RHS} &= (\tfrac{1}{2} \conjc \tfrac{1}{2}) \disjc (\tfrac{1}{2} \conjc \tfrac{1}{2})
= \tfrac{1}{4} \disjc \tfrac{1}{4}
= \tfrac{1}{4} + \tfrac{1}{4} - \tfrac{1}{16} = \tfrac{7}{16}
\end{align*}
Lean: \texttt{conj\_disj\_not\_distrib}.
\end{proof}

This failure of distributivity distinguishes Cred from Boolean algebra and is
essential for modeling uncertainty.

%=============================================================================
\section{Primitive Conditioning}
\label{sec:conditioning}
%=============================================================================

\subsection{The Chain Rule Constraint}

\begin{definition}[Conditioning Structure]
\label{def:conditioning}
A \emph{conditioning structure} for parameters $(\text{joint}, \text{evidence}) \in [0,1]^2$
consists of a conditional credence $\text{condCred} \in [0,1]$ satisfying:
\[
\text{condCred} \conjc \text{evidence} = \text{joint}
\]
In propositional notation: $\condc{A}{B} \conjc \cred(B) = \cred(A \land B)$.
\end{definition}

The \texttt{joint} parameter represents $\cred(A \land B)$ and is supplied
as input---it is not computed from marginal credences.
This parametric design parallels how Boolean algebra defines operations on
truth values without specifying which propositions have which values.

\subsection{Boundary Behavior}

\textbf{Evidence = 1.}
The chain rule gives $\cred(A \mid 1) \conjc 1 = \cred(A \land B)$,
so $\cred(A \mid 1) = \cred(A \land B)$.
Conditioning on certainty returns the joint credence.

\textbf{Evidence = 0.}
The chain rule gives $\cred(A \mid 0) \conjc 0 = 0$.
Since $c \conjc 0 = 0$ for all $c \in [0,1]$, this equation is satisfied by any value.
The conditional is \emph{unconstrained}.

\begin{theorem}[Underdetermination at Zero]
\label{thm:noexfalso}
When evidence $= 0$:
\begin{enumerate}[label=(\alph*)]
\item If a conditioning structure exists, then joint $= 0$.
\item For joint $= 0$, any $c \in [0,1]$ satisfies the chain rule.
\end{enumerate}
\end{theorem}
\begin{proof}
(a) If $\text{condCred} \conjc 0 = \text{joint}$, then $\text{joint} = 0$.
(b) For any $c$, we have $c \conjc 0 = 0$, so the equation $c \conjc 0 = 0$ holds.
Lean: \texttt{conditioning\_zero\_forces\_joint\_zero}, \texttt{conditioning\_zero\_any}.
\end{proof}

\begin{corollary}[No Forced Ex Falso]
The chain rule does not force $\cred(A \mid 0) = 1$.
\end{corollary}

This is the central result: conditioning on impossible evidence is unconstrained
by the chain rule.
The constraint-based definition means division is derived (as the unique solution
when evidence $> 0$) rather than primitive; at zero, no division is needed.
To obtain ``from impossibility, anything follows,'' one must add an axiom
$\cred(A \mid 0) = 1$; it does not follow from conditioning alone.

\subsection{Existence and Uniqueness}

\begin{theorem}[Existence]
\label{thm:existence}
If $\cred(B) > 0$ and $\cred(A \land B) \leq \cred(B)$, then there exists a
conditioning structure with $\cred(A \mid B) = \cred(A \land B)/\cred(B)$.
\end{theorem}
\begin{proof}
The value $\cred(A \land B)/\cred(B)$ lies in $[0,1]$ (by the hypotheses) and
satisfies the chain rule.
Lean: \texttt{conditioning\_mk}.
\end{proof}

\begin{theorem}[Uniqueness]
\label{thm:uniqueness}
If $\cred(B) > 0$, the conditional credence is uniquely determined.
\end{theorem}
\begin{proof}
Suppose $c_1 \cdot e = c_2 \cdot e = j$ where $e = \cred(B) > 0$.
Since $e > 0$ we may cancel: $c_1 = j/e = c_2$.
Lean: \texttt{conditioning\_unique}.
\end{proof}

The defining constraint uses only multiplication.
Division appears in the existence proof as a construction technique, not in the
definition itself---it is derived, not primitive.
Probability satisfies the same chain rule: $P(A \mid B) \cdot P(B) = P(A \land B)$.
Cred's conditioning is not new machinery---it is probability's chain rule without
the surrounding measure theory.

\subsection{Worked Example}

\begin{example}
\label{ex:conditioning}
Suppose $\cred(B) = 0.6$ and $\cred(A \land B) = 0.4$.
The joint credence is an input---it might come from empirical data, domain expertise,
or a probability model.
The chain rule gives:
\[
\cred(A \mid B) = \frac{0.4}{0.6} = \frac{2}{3}
\]
Verification: $\frac{2}{3} \cdot 0.6 = 0.4 = \cred(A \land B)$.

If instead $\cred(B) = 0$, then necessarily $\cred(A \land B) = 0$, and any
value $c \in [0,1]$ satisfies $c \cdot 0 = 0$.
The conditional is unconstrained.
\end{example}

%=============================================================================
\section{Fixed Points}
\label{sec:fixed-points}
%=============================================================================

\begin{theorem}[Negation Fixed Point]
\label{thm:liar}
$\negc \tfrac{1}{2} = \tfrac{1}{2}$.
\end{theorem}
\begin{proof}
$1 - \tfrac{1}{2} = \tfrac{1}{2}$.
Lean: \texttt{liar\_fixed\_point}.
\end{proof}

\begin{theorem}[Uniqueness]
\label{thm:fixedunique}
$\negc c = c$ if and only if $c = \tfrac{1}{2}$.
\end{theorem}
\begin{proof}
$1 - c = c$ implies $c = \tfrac{1}{2}$.
Lean: \texttt{neg\_fixed\_point\_unique}.
\end{proof}

If a self-negating sentence $L$ satisfies $L \leftrightarrow \neg L$, requiring
$\cred(L) = \negc \cred(L)$ gives $\cred(L) = \tfrac{1}{2}$.
This provides a natural algebraic value for such sentences.

\begin{remark}[Scope]
This fixed-point result is algebraic, not philosophical.
It does not resolve the liar paradox or address the strengthened liar
(``This sentence is not true'').
The extensive literature on semantic paradoxes offers multiple approaches:
Kripke's groundedness~\cite{kripke1975} assigns no truth value to the liar at
minimal fixed points; Priest's dialetheism~\cite{priest1979} treats it as both
true and false.
Cred's $\tfrac{1}{2}$ is simply the unique fixed point of complement on $[0,1]$.
\end{remark}

%=============================================================================
\section{Connections to Other Systems}
\label{sec:connections}
%=============================================================================

\subsection{Three-Valued Collapse}
\label{sec:collapse}

Define the collapse function:
\[
\text{collapse}(c) = \begin{cases}
0 & \text{if } c = 0 \\
\tfrac{1}{2} & \text{if } 0 < c < 1 \\
1 & \text{if } c = 1
\end{cases}
\]

\begin{theorem}[Collapse Homomorphism]
\label{thm:collapse}
The collapse is a surjective homomorphism from $(\Cred, \negc, \conjc, \disjc)$
to the three-valued algebra $(\{0, \tfrac{1}{2}, 1\}, \mathrm{neg}, \mathrm{conj}, \mathrm{disj})$:
\begin{align*}
\mathrm{collapse}(\negc c) &= \mathrm{neg}(\mathrm{collapse}(c)) \\
\mathrm{collapse}(c_1 \conjc c_2) &= \mathrm{conj}(\mathrm{collapse}(c_1), \mathrm{collapse}(c_2)) \\
\mathrm{collapse}(c_1 \disjc c_2) &= \mathrm{disj}(\mathrm{collapse}(c_1), \mathrm{collapse}(c_2))
\end{align*}
where the three-valued operations are $\mathrm{neg}(c) = 1-c$,
$\mathrm{conj} = \min$, $\mathrm{disj} = \max$.
\end{theorem}
\begin{proof}
We verify each operation by case analysis on whether each $c_i$ is $0$, $1$,
or in the open interval $(0,1)$.

\emph{Negation.}
If $c = 0$ then $\negc c = 1$, and $\mathrm{neg}(0) = 1$.
If $c = 1$ then $\negc c = 0$, and $\mathrm{neg}(1) = 0$.
If $0 < c < 1$ then $0 < 1 - c < 1$, so both $c$ and $\negc c$ collapse to
$\tfrac{1}{2}$, and $\mathrm{neg}(\tfrac{1}{2}) = \tfrac{1}{2}$.

\emph{Conjunction.}
If $c_1 = 0$ or $c_2 = 0$: $c_1 \conjc c_2 = 0$ (absorber), and
$\min(v_1, v_2) = 0$ whenever either $v_i = 0$.
If $c_1 = 1$: $c_1 \conjc c_2 = c_2$ (identity), and $\min(1, v_2) = v_2$.
Symmetrically for $c_2 = 1$.
The remaining case is $0 < c_1, c_2 < 1$.
Then $0 < c_1 c_2$ (product of positives) and $c_1 c_2 < c_1 < 1$
(since $c_2 < 1$), so $c_1 \conjc c_2 \in (0,1)$ collapses to $\tfrac{1}{2}$,
and $\min(\tfrac{1}{2}, \tfrac{1}{2}) = \tfrac{1}{2}$.

\emph{Disjunction.}
If $c_1 = 1$ or $c_2 = 1$: $c_1 \disjc c_2 = 1$, and $\max(v_1, v_2) = 1$.
If $c_1 = 0$: $c_1 \disjc c_2 = c_2$, and $\max(0, v_2) = v_2$.
If $0 < c_1, c_2 < 1$: the disjunction is $c_1 + c_2 - c_1 c_2$.
This is positive (since $c_1 c_2 < c_1 + c_2$) and strictly less than $1$
(since $c_1 + c_2 - c_1 c_2 = 1 - (1-c_1)(1-c_2) < 1$), so it collapses to
$\tfrac{1}{2} = \max(\tfrac{1}{2}, \tfrac{1}{2})$.

Surjectivity: $\mathrm{collapse}(0) = 0$, $\mathrm{collapse}(\tfrac{1}{2}) = \tfrac{1}{2}$,
$\mathrm{collapse}(1) = 1$.

Lean: \texttt{collapse\_neg}, \texttt{collapse\_conj}, \texttt{collapse\_disj}.
\end{proof}

The three-valued algebra $(\{0, \tfrac{1}{2}, 1\}, \min, \max, 1{-}c)$ is the
\emph{Kleene algebra}---the three-element De Morgan chain~\cite{kleene1952}.
This algebra underlies three well-known logics that share the same truth tables for
negation, conjunction, and disjunction:
strong Kleene logic K$_3$~\cite{kleene1952},
Priest's Logic of Paradox LP~\cite{priest1979}, and
RM3~\cite{meyer1972}.
These logics differ only in their consequence relations (which truth values are
designated for valid inference), not in the algebra.
Since Cred is an algebra without a consequence relation, the collapse connects it
to all three simultaneously.

The three-valued operations are:
\begin{center}
\begin{tabular}{c|c}
$c$ & $\mathrm{neg}(c)$ \\
\hline
0 & 1 \\
$\tfrac{1}{2}$ & $\tfrac{1}{2}$ \\
1 & 0
\end{tabular}
\quad
\begin{tabular}{c|ccc}
$\mathrm{conj}$ & 0 & $\tfrac{1}{2}$ & 1 \\
\hline
0 & 0 & 0 & 0 \\
$\tfrac{1}{2}$ & 0 & $\tfrac{1}{2}$ & $\tfrac{1}{2}$ \\
1 & 0 & $\tfrac{1}{2}$ & 1
\end{tabular}
\quad
\begin{tabular}{c|ccc}
$\mathrm{disj}$ & 0 & $\tfrac{1}{2}$ & 1 \\
\hline
0 & 0 & $\tfrac{1}{2}$ & 1 \\
$\tfrac{1}{2}$ & $\tfrac{1}{2}$ & $\tfrac{1}{2}$ & 1 \\
1 & 1 & 1 & 1
\end{tabular}
\end{center}

\subsubsection*{Conditioning and Implication}

\textbf{Scope of the collapse.}
The homomorphism of Theorem~\ref{thm:collapse} covers $\{\negc, \conjc, \disjc\}$.
Implication is a separate question: the Kleene algebra has no canonical arrow,
and the three logics (K$_3$, LP, RM3) that share it each define implication
differently.

\textbf{Conditioning is ternary, not binary.}
Cred conditioning takes three inputs: evidence, joint, and the conclusion.
The chain rule constrains the conclusion given the other two.
For evidence $> 0$, the conditional is $\text{joint}/\text{evidence}$, which
requires joint $\leq$ evidence.

\begin{center}
\begin{tabular}{cc|c}
\textit{evidence} & \textit{joint} & $\cred(A \mid B)$ \\
\hline
0 & 0 & $*$ \\
$\tfrac{1}{2}$ & 0 & 0 \\
$\tfrac{1}{2}$ & $\tfrac{1}{4}$ & $\tfrac{1}{2}$ \\
$\tfrac{1}{2}$ & $\tfrac{1}{2}$ & 1 \\
1 & 0 & 0 \\
1 & $\tfrac{1}{2}$ & $\tfrac{1}{2}$ \\
1 & 1 & 1
\end{tabular}
\end{center}
{\footnotesize $*$ = unconstrained (any value satisfies chain rule).
Only entries with joint $\leq$ evidence are valid.}

\medskip
\textbf{Conditioning as a binary arrow.}
To compare with standard implications, fix the joint to the Kleene conjunction
$\min(a, b)$.
For evidence $a > 0$, the chain rule gives
$\text{cond}(a, b) = \text{joint}/a = \min(a, b)/a$.
If $b \leq a$ this is $b/a$; if $b > a$ this is $a/a = 1$.
In both cases:
\[
\text{cond}(a, b) \;=\; \min(b/a,\; 1)
\]
This is exactly the \emph{product residuated implication} of H\'ajek's
$\Pi$-logic~\cite{hajek1998}.

\textbf{Three-way comparison on $\{0, \tfrac{1}{2}, 1\}$.}

\begin{center}
\begin{tabular}{c|ccc}
\multicolumn{4}{c}{\textbf{RM3}: $\max(\negc a, b)$} \\
& 0 & $\tfrac{1}{2}$ & 1 \\
\hline
0 & 1 & 1 & 1 \\
$\tfrac{1}{2}$ & $\tfrac{1}{2}$ & $\tfrac{1}{2}$ & 1 \\
1 & 0 & $\tfrac{1}{2}$ & 1
\end{tabular}
\quad
\begin{tabular}{c|ccc}
\multicolumn{4}{c}{\textbf{G\"odel/Resid.}: $\min(b/a, 1)$} \\
& 0 & $\tfrac{1}{2}$ & 1 \\
\hline
0 & 1 & 1 & 1 \\
$\tfrac{1}{2}$ & 0 & 1 & 1 \\
1 & 0 & $\tfrac{1}{2}$ & 1
\end{tabular}
\quad
\begin{tabular}{c|ccc}
\multicolumn{4}{c}{\textbf{Cred}: joint $= \min(a,b)$} \\
& 0 & $\tfrac{1}{2}$ & 1 \\
\hline
0 & $*$ & $*$ & $*$ \\
$\tfrac{1}{2}$ & 0 & 1 & 1 \\
1 & 0 & $\tfrac{1}{2}$ & 1
\end{tabular}
\end{center}
\smallskip
{\footnotesize $*$ = unconstrained. Rows with evidence $> 0$ are uniquely
determined.}

\smallskip
RM3 and G\"odel/residuated agree when evidence $\in \{0, 1\}$ but differ at
evidence $= \tfrac{1}{2}$: RM3 gives $\tfrac{1}{2} \to 0 = \tfrac{1}{2}$,
while G\"odel/residuated gives $0$.
Cred conditioning (with Kleene joint) matches G\"odel/residuated for evidence
$> 0$ and replaces the evidence $= 0$ row with~$*$.

Lean: \texttt{godel\_impl}, \texttt{godel\_impl\_ne\_rm3\_impl},
\texttt{cred\_no\_ex\_falso\_godel}.

\textbf{G\"odel coincidence.}
On the finite set $\{0, \tfrac{1}{2}, 1\}$, the product residuated implication
$\min(b/a, 1)$ and the G\"odel implication coincide (both have the same
truth table above).
On $[0,1]$ they diverge: G\"odel implication is $1$ when $a \leq b$ and $b$
otherwise, while the product residuated implication is $\min(b/a, 1)$.

\textbf{Full $[0,1]$ structure.}
On $[0,1]$, conditioning is parametric in the joint---a family of arrows
indexed by the joint credence, not one specific binary arrow.
The product residuated implication is the specific instantiation with
joint $= \min(a, b)$ (the Kleene conjunction).
Product fuzzy logic commits to this choice; Cred leaves the joint free.

\subsubsection*{Bayes Consistency}

Applying the chain rule from both directions gives Bayes' theorem:
\[
\cred(A \mid B) \cdot \cred(B) = \cred(A \land B) = \cred(B \land A) = \cred(B \mid A) \cdot \cred(A).
\]
A binary arrow $f(a, b)$ used as conditioning is \emph{Bayes-consistent} when
the joint it induces is symmetric: $f(a, b) \cdot a = f(b, a) \cdot b$ for all
$a, b$.

\begin{proposition}[Product Residuated is Bayes-Consistent]
\label{prop:resid-bayes}
The product residuated implication $f(a, b) = \min(b/a,\; 1)$ (with
the standard convention $f(0, b) = 1$, i.e.\ ex falso---the convention that
Cred's chain-rule conditioning replaces with ``unconstrained'') is
Bayes-consistent: $f(a, b) \cdot a = \min(a, b) = f(b, a) \cdot b$.
\end{proposition}
\begin{proof}
If $b \leq a$ then $f(a, b) \cdot a = (b/a) \cdot a = b = \min(a, b)$.
If $a < b$ then $f(a, b) \cdot a = 1 \cdot a = a = \min(a, b)$.
By symmetry of $\min$, $f(b, a) \cdot b = \min(a, b)$ as well.
\end{proof}

\begin{proposition}[G\"odel is Not Bayes-Consistent on the Continuum]
\label{prop:godel-not-bayes}
The G\"odel implication ($f(a,b) = 1$ if $a \leq b$, else $b$) fails
Bayes consistency on $[0,1]$.
\end{proposition}
\begin{proof}
Take $a = \tfrac{2}{5}$, $b = \tfrac{3}{5}$.
Then $f(a, b) = 1$ (since $a \leq b$), so $f(a, b) \cdot a = \tfrac{2}{5}$.
But $f(b, a) = a = \tfrac{2}{5}$ (since $b > a$), so
$f(b, a) \cdot b = \tfrac{2}{5} \cdot \tfrac{3}{5} = \tfrac{6}{25} \neq \tfrac{2}{5}$.
\end{proof}

\begin{proposition}[RM3 is Not Bayes-Consistent]
\label{prop:rm3-not-bayes}
The RM3 implication $\max(\negc a, b)$ fails Bayes consistency even on
$\{0, \tfrac{1}{2}, 1\}$.
\end{proposition}
\begin{proof}
Take $a = \tfrac{1}{2}$, $b = 0$.
Then $f(a, b) = \max(\tfrac{1}{2}, 0) = \tfrac{1}{2}$, so
$f(a, b) \cdot a = \tfrac{1}{4}$.
But $f(b, a) = \max(1, \tfrac{1}{2}) = 1$, so $f(b, a) \cdot b = 0$.
Since $\tfrac{1}{4} \neq 0$, RM3 is not Bayes-consistent.
Lean: \texttt{rm3\_not\_bayes\_consistent}.
\end{proof}

\begin{remark}
On $\{0, \tfrac{1}{2}, 1\}$, the G\"odel implication coincides with the product
residuated implication (Section~\ref{sec:collapse}), so it is Bayes-consistent
on this finite set.
Lean: \texttt{godel\_bayes\_consistent}.
The failure occurs only on the full interval $[0,1]$.
\end{remark}

\begin{remark}[Copula Characterization]
\label{rem:copula}
A binary arrow $f$ is Bayes-consistent if and only if the induced joint
$J(a,b) = f(a,b) \cdot a$ is symmetric.
A symmetric function $J : [0,1]^2 \to [0,1]$ satisfying $J(a,0) = 0$ and
$J(a,1) = a$ is a (symmetric) copula~\cite{nelsen2006}.
Grzegorzewski~\cite{grzegorzewski2013} defined \emph{probabilistic
implications} $I_C(a,b) = C(a,b)/a$ from a copula $C$ and computed three
cases: the Fr\'echet upper bound $C(a,b) = \min(a,b)$ gives the product
residuated (Goguen) implication; the independence copula $C(a,b) = ab$
gives the trivial arrow $b$; the Fr\'echet lower bound
$C(a,b) = \max(a+b-1, 0)$ gives a third arrow.
Among the standard logic implications, only the product residuated is
copula-derived (from the Fr\'echet upper bound); RM3's $\max(\negc a, b)$
and G\"odel's implication are not copula-derived, which is why they fail
Bayes consistency.
\end{remark}

\begin{remark}[Dependence Interpretation]
The Fr\'echet upper bound is the unique copula of maximum positive
dependence (comonotonicity).
Setting joint $= \min(a,b)$ in Cred's chain rule therefore corresponds to
conditioning under maximum positive dependence between propositions.
The product $a \cdot b$ (Cred's $\conjc$) is the independence copula;
conditioning under independence gives the trivial arrow $b$.
Thus Cred's two operations play complementary roles: $\conjc$ captures
independence, while chain-rule conditioning with the Kleene joint captures
comonotonicity.
\end{remark}

\begin{proposition}[Fr\'echet-Hoeffding Bounds]
\label{prop:frechet}
If the chain rule holds in both directions---$\condc{A}{B} \cdot \cred(B)
= \cred(A \land B)$ and $\condc{B}{A} \cdot \cred(A) = \cred(A \land B)$---then
the joint is bounded above:
\[
\cred(A \land B) \;\leq\; \min(\cred(A),\; \cred(B)).
\]
If additionally all four cells $\cred(A \land B)$,
$\cred(A) - \cred(A \land B)$,
$\cred(B) - \cred(A \land B)$, and
$1 - \cred(A) - \cred(B) + \cred(A \land B)$ are non-negative, then
\[
\max(\cred(A) + \cred(B) - 1,\; 0) \;\leq\; \cred(A \land B).
\]
Together: the joint lies in the Fr\'echet-Hoeffding
interval~\cite{nelsen2006}.
\end{proposition}
\begin{proof}
Upper bound: each conditioning structure forces
$\text{joint} \leq \text{evidence}$ (since $\text{condCred} \leq 1$).
Applying this from both directions gives
$\cred(A \land B) \leq \cred(A)$ and $\cred(A \land B) \leq \cred(B)$.
Lower bound: the complement cell
$1 - \cred(A) - \cred(B) + \cred(A \land B) \geq 0$ rearranges to
$\cred(A \land B) \geq \cred(A) + \cred(B) - 1$.
Conversely, any joint in the Fr\'echet-Hoeffding interval admits
conditioning structures in both directions (by division when marginals
are positive).
Lean: \texttt{frechet\_upper}, \texttt{frechet\_lower},
\texttt{frechet\_conditioning\_exists}.
\end{proof}

\begin{remark}[Freedom of the Joint]
\label{rem:freedom}
Cred leaves the joint free: any value in the Fr\'echet-Hoeffding interval
$[\max(a + b - 1, 0),\; \min(a, b)]$ is a valid joint credence, and
chain-rule conditioning produces a Bayes-consistent arrow for each choice.
The Fr\'echet upper bound $\min(a, b)$ yields the product residuated
(Goguen) implication $\min(b/a, 1)$---the implication of H\'ajek's
product logic $\Pi$~\cite{hajek1998}.
The independence copula $a \cdot b$ (Cred's $\conjc$) yields the trivial
arrow $b$.
The Fr\'echet lower bound $\max(a + b - 1, 0)$ yields a third arrow.
Fixing a specific copula (dependence structure) selects a specific arrow;
Cred's parametric conditioning encompasses all of them.
\end{remark}

\subsection{Boolean Restriction}

Restricting Cred to $\{0, 1\}$, the operations $\{\negc, \conjc, \disjc\}$ become:
\begin{center}
\begin{tabular}{c|c}
$c$ & $\negc c$ \\
\hline
0 & 1 \\
1 & 0
\end{tabular}
\qquad
\begin{tabular}{c|cc}
$\conjc$ & 0 & 1 \\
\hline
0 & 0 & 0 \\
1 & 0 & 1
\end{tabular}
\qquad
\begin{tabular}{c|cc}
$\disjc$ & 0 & 1 \\
\hline
0 & 0 & 1 \\
1 & 1 & 1
\end{tabular}
\end{center}
These are exactly the two-element Boolean algebra operations $(\neg, \land, \lor)$.

On $\{0, 1\}$, conditioning with Kleene joint matches the classical material
implication for evidence $> 0$: both give $1 \to 0 = 0$, $1 \to 1 = 1$,
$0 \to b = 1$ classically.
Indeed, all three arrows---material, G\"odel, and product residuated---coincide
on Boolean values.
The sole difference is at evidence $= 0$: material implication gives
$0 \to b = 1$; conditioning is unconstrained.
To recover classical logic from Boolean Cred, one must \emph{add} the axiom
$\cred(A \mid 0) = 1$.

\subsection{Subalgebra vs.\ Quotient}
\label{sec:scope}

Cred relates to classical logic and to the Kleene algebra in structurally
different ways: embedding vs.\ quotient.

\textbf{Boolean subalgebra.}
The set $\{0, 1\} \subset [0,1]$ is closed under all Cred operations:
if $c_1, c_2 \in \{0,1\}$, then $\negc c_1$, $c_1 \conjc c_2$, and
$c_1 \disjc c_2$ are all in $\{0,1\}$ (the truth tables are exactly
$\neg$, $\land$, $\lor$; see Section~\ref{sec:connections}).
No quotient map is needed: if you start with only $0$s and $1$s, all
operations keep you there.

\begin{proposition}[No Collapse to Classical Logic]
\label{prop:no-classical}
There is no function $\varphi : [0,1] \to \{0,1\}$ preserving negation:
$\varphi(1-c) = 1 - \varphi(c)$ for all $c$.
\end{proposition}
\begin{proof}
At $c = \tfrac{1}{2}$: $\varphi(\tfrac{1}{2}) = 1 - \varphi(\tfrac{1}{2})$,
so $\varphi(\tfrac{1}{2}) = \tfrac{1}{2} \notin \{0, 1\}$.
Lean: \texttt{no\_boolean\_neg\_retraction}.
\end{proof}

The result is stronger than ``no retraction'': no negation-preserving map
$[0,1] \to \{0,1\}$ exists at all, because the fixed point at $\tfrac{1}{2}$
has no image in $\{0,1\}$.
The Boolean algebra embeds into Cred as a subalgebra, but there is no
surjection back.

\textbf{Kleene quotient.}
The set $\{0, \tfrac{1}{2}, 1\}$ is \emph{not} closed under Cred operations:
$\tfrac{1}{2} \conjc \tfrac{1}{2} = \tfrac{1}{4} \notin \{0, \tfrac{1}{2}, 1\}$.
Intermediate values interact to produce other intermediate values, so the
three-valued algebra cannot live inside Cred as a subalgebra.
The collapse homomorphism (Theorem~\ref{thm:collapse}) is the necessary
quotient: it maps the entire open interval $(0,1)$ to $\tfrac{1}{2}$,
absorbing the products that leave $\{0, \tfrac{1}{2}, 1\}$.

\textbf{Uniqueness of the Kleene target.}
No analogous quotient exists for the other standard three-valued algebras.

\begin{proposition}[No Collapse to G\"odel Logic]
\label{prop:no-godel}
Let $\mathrm{neg}_G$ denote G\"odel negation on $\{0, \tfrac{1}{2}, 1\}$:
$\mathrm{neg}_G(0) = 1$ and $\mathrm{neg}_G(c) = 0$ for $c > 0$~\cite{godel1932}.
There is no function $\varphi : [0,1] \to \{0, \tfrac{1}{2}, 1\}$ such that
$\varphi(1-c) = \mathrm{neg}_G(\varphi(c))$ for all $c$.
\end{proposition}
\begin{proof}
G\"odel negation has no fixed point:
$\mathrm{neg}_G(0) = 1 \neq 0$,
$\mathrm{neg}_G(\tfrac{1}{2}) = 0 \neq \tfrac{1}{2}$,
$\mathrm{neg}_G(1) = 0 \neq 1$.
But $\varphi(\tfrac{1}{2}) = \varphi(1 - \tfrac{1}{2}) = \mathrm{neg}_G(\varphi(\tfrac{1}{2}))$
requires $\varphi(\tfrac{1}{2})$ to be a fixed point.
Lean: \texttt{no\_godel\_collapse}, \texttt{godel\_neg\_no\_fixed\_point}.
\end{proof}

\begin{proposition}[No Collapse to \L ukasiewicz Logic]
\label{prop:no-luk}
Let $\mathrm{conj}_{\text{\L}} (a, b) = \max(a + b - 1,\; 0)$ be the \L ukasiewicz
conjunction on $\{0, \tfrac{1}{2}, 1\}$~\cite{lukasiewicz1920}.
There is no function $\varphi : [0,1] \to \{0, \tfrac{1}{2}, 1\}$ preserving both
negation ($\varphi(1-c) = 1-\varphi(c)$) and conjunction
($\varphi(c_1 \cdot c_2) = \mathrm{conj}_{\text{\L}}(\varphi(c_1), \varphi(c_2))$).
\end{proposition}
\begin{proof}
Negation preservation at $c = \tfrac{1}{2}$:
$\varphi(\tfrac{1}{2}) = \varphi(1 - \tfrac{1}{2}) = 1 - \varphi(\tfrac{1}{2})$,
so $\varphi(\tfrac{1}{2}) = \tfrac{1}{2}$.
Pick $c = 1/\!\sqrt{2} \in (0,1)$; then $c \conjc c = c^2 = \tfrac{1}{2}$.
Conjunction preservation gives
$\varphi(\tfrac{1}{2}) = \mathrm{conj}_{\text{\L}}(\varphi(c), \varphi(c))$.
Since $\varphi(c) \in \{0, \tfrac{1}{2}, 1\}$:
\[
\mathrm{conj}_{\text{\L}}(0,0) = 0,\quad
\mathrm{conj}_{\text{\L}}(\tfrac{1}{2},\tfrac{1}{2}) = 0,\quad
\mathrm{conj}_{\text{\L}}(1,1) = 1
\]
None equals $\tfrac{1}{2} = \varphi(\tfrac{1}{2})$.
Lean: \texttt{no\_luk\_collapse}, \texttt{luk\_conj\_half\_half}.
\end{proof}

The pattern: the negation fixed point at $\tfrac{1}{2}$ propagates constraints
that only the Kleene algebra can absorb.
Classical logic and G\"odel logic fail because they lack a negation fixed point.
\L ukasiewicz logic has the same negation, but its conjunction
$\mathrm{conj}_{\text{\L}}(\tfrac{1}{2}, \tfrac{1}{2}) = 0$ disagrees with the
Kleene value $\min(\tfrac{1}{2}, \tfrac{1}{2}) = \tfrac{1}{2}$: the
self-conjunction of the fixed point must equal the fixed point, and only the
Kleene algebra satisfies this.

\begin{remark}[Belnap-Dunn Four-Valued Logic]
\label{rem:fde}
The Kleene algebra embeds as a subalgebra of the Belnap-Dunn four-valued
lattice~\cite{belnap1977,dunn1976}, via the inclusion
$\{0, \tfrac{1}{2}, 1\} \hookrightarrow \{\mathbf{f}, \mathbf{b}, \mathbf{t}, \mathbf{n}\}$
mapping $0 \mapsto \mathbf{f}$, $\tfrac{1}{2} \mapsto \mathbf{b}$, $1 \mapsto \mathbf{t}$.
The Cred collapse therefore factors through the Belnap-Dunn algebra.
A surjective homomorphism from Cred to the full four-valued algebra would need
to reach $\mathbf{n}$ (``neither true nor false''), which appears to have no
natural preimage in $[0,1]$.
\end{remark}

\subsection{Relationship to Probability}
\label{sec:probability}

Every probability space satisfies Cred's axioms: it has values in $[0,1]$,
the complement $P(\neg A) = 1 - P(A)$, and the chain rule
$P(A \mid B) \cdot P(B) = P(A \land B)$.
Probability is therefore a special case of Cred.

What probability adds:
\begin{itemize}
\item A sigma-algebra of events, providing structure for ``what can be measured.''
\item A countably additive measure, determining the joint $P(A \land B)$ from
the measure rather than requiring it as input.
\item Sigma-additivity: $P(A \lor B) = P(A) + P(B)$ for disjoint events.
\end{itemize}

These additions buy concrete benefits.
The joint is determined, not parameterized: $P(A \land B)$ is a consequence of
the measure, so conditioning needs only $A$ and $B$, not the joint as a separate
input.
Disjunction uses the actual joint via inclusion-exclusion:
$P(A \lor B) = P(A) + P(B) - P(A \land B)$.
Conditioning at zero can sometimes be resolved through
disintegration~\cite{chang1997}, yielding a specific (a.e.\ unique) value
determined by the measure structure.

What Cred drops: no measure, no sigma-algebra, no additivity, no event structure.
The product $c_1 \conjc c_2$ is multiplication, not a claim about the actual joint.
Conditioning requires the joint as external input.
At evidence zero, Cred leaves the conditional unconstrained rather than
undefined (as in standard probability) or resolved (as via disintegration).

\subsection{Comparison Table}

\begin{center}
\small
\begin{tabular}{lp{3.8cm}p{3.5cm}l}
\toprule
\textbf{System} & \textbf{Core axioms} & \textbf{Adds on top} & \textbf{At evidence $= 0$} \\
\midrule
\textbf{Cred} & $[0,1]$ + complement + chain rule & --- & Unconstrained \\
Probability & Cred's axioms & Measure + $\sigma$-additivity & Undefined$^*$ \\
Product fuzzy ($\Pi$) & $[0,1]$ + product t-norm + residuated impl. & Deductive calculus & $0 \to y = 1$ \\
Classical logic & $\{0,1\}$ + Boolean ops + material impl. & --- & $0 \to y = 1$ \\
RM3 & $\{0,\tfrac{1}{2},1\}$ + De Morgan ops + $\max$ impl. & Designated values & $0 \to y = 1$$^\dagger$ \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\smallskip
\noindent $^*$Undefined: the expression is ill-formed (division by zero).
Unconstrained: the expression is well-formed but any value satisfies it.

\noindent $^\dagger$RM3 has $0 \to b = 1$ in its truth table, but only values
$\{1, \tfrac{1}{2}\}$ are designated for valid inference, so explosion is blocked
at the level of consequence rather than at the truth-table level.

\smallskip
The table makes the hierarchy explicit.
Probability is a special case of Cred (same chain rule, additional structure).
Cred's algebra $\{\negc, \conjc, \disjc\}$ collapses to the Kleene algebra
shared by K$_3$, LP, and RM3.
Cred's conditioning, when instantiated as a binary arrow via the Kleene joint,
gives the product residuated implication for evidence $> 0$---distinct from
RM3's $\max(\negc a, b)$.
The relationship is: \emph{same connective algebra, different arrow}.
Product fuzzy logic commits to the residuated arrow (including $0 \to y = 1$);
Cred leaves the joint free and the evidence-zero row unconstrained.

\subsection{Related Work}

Cred sits in a web of prior work.
The hierarchy---Cred as a weakening of probability, distinct from product fuzzy
logic---organizes the relationships.

\textbf{Primitive Conditional Probability.}
R\'enyi~\cite{renyi1955,renyi1970} first axiomatized conditional probability as
primitive, with the chain rule as a defining constraint rather than division.
Popper~\cite{popper1959} independently argued that conditional probability is more
fundamental than absolute probability, developing ``Popper functions.''
Van Fraassen~\cite{vanfraassen1976} extended the Popper-R\'enyi approach.
Cred inherits the chain rule from this tradition but drops the measure space
entirely.

\textbf{Probability and Relevant Logic.}
Van Fraassen~\cite{vanfraassen1983} showed that conditional probability functions
can model relevant logic's consequence relation, connecting probabilistic and
relevance-based approaches.
Cred makes this connection algebraically explicit through the collapse homomorphism.

\textbf{Product Fuzzy Logic.}
H\'ajek~\cite{hajek1998} developed product logic ($\Pi$), which uses the product
t-norm with \emph{residuated} implication ($x \to y = \min(y/x, 1)$ for $x > 0$,
and $0 \to y = 1$); the residuated implication forces ex falso.
Klement, Mesiar, and Pap~\cite{klement2000} gave a comprehensive treatment of
triangular norms; the product t-norm, probabilistic sum, and standard complement
form the product De Morgan triplet.
Cred shares this triplet and, when conditioning is instantiated with the Kleene
joint $\min(a, b)$, \emph{recovers} the product residuated implication for
evidence $> 0$.
The divergence is solely at evidence $= 0$: product logic forces $0 \to y = 1$;
Cred conditioning is unconstrained.
On the finite set $\{0, \tfrac{1}{2}, 1\}$, the product residuated implication
coincides with the G\"odel implication; on $[0,1]$ they diverge.

\textbf{Fuzzy Logic and Conditional Probability.}
H\'ajek, Godo, and Esteva~\cite{hajek1995} studied conditional probability
within fuzzy logic and noted that the product residuated implication is
discontinuous.
As a workaround, they expressed $P(\psi \mid \varphi) \geq a$ via the
chain rule constraint $P(\varphi \land \psi) \geq a \cdot P(\varphi)$
rather than division---essentially the same move Cred makes.
They treated this as a technical device; Cred adopts it as the defining
design choice and explores its algebraic consequences.

\textbf{Cox-Jaynes Probability Logic.}
Cox's theorem~\cite{cox1961} proves that any consistent system for plausible
reasoning satisfying certain desiderata must be isomorphic to probability theory.
Jaynes~\cite{jaynes2003} developed this into ``Probability Theory: The Logic of
Science.''
Cox's desiderata require conditioning to be a \emph{function} (unique output for
each input). Cred relaxes this: conditioning is a \emph{relation} that becomes
non-functional when evidence $= 0$.

\textbf{Conditionals and Triviality.}
Adams~\cite{adams1965,adams1998} developed probabilistic semantics for conditionals
where $P(A \to B) = P(B|A)$.
Lewis~\cite{lewis1976} proved that treating conditionals as propositions with this
property leads to triviality.
Cred sidesteps triviality by treating conditioning as a structural operation,
not as a credence assigned to a conditional proposition.

\textbf{Relevant and Paraconsistent Logic.}
Anderson and Belnap~\cite{anderson1975} developed relevant logics R and E,
rejecting ex falso via syntactic relevance conditions.
Dunn~\cite{dunn1976} provided algebraic semantics connecting relevant logic to
De Morgan lattices.
RM3~\cite{meyer1972} is a three-valued relevant logic.
Da Costa~\cite{dacosta1974} developed paraconsistent logics tolerating contradiction
without explosion.
Priest~\cite{priest1979} introduced the Logic of Paradox (LP).
Brady~\cite{brady2006} and Weber~\cite{weber2010} developed paraconsistent set
theory with unrestricted comprehension.
The collapse homomorphism (Section~\ref{sec:collapse}) connects Cred to the
Kleene algebra~\cite{kleene1952} that underlies K$_3$, LP, and RM3.
These three logics share the same truth tables for $\{\negc, \conjc, \disjc\}$;
the collapse is to this shared algebra, not to any one logic's consequence relation.

\textbf{Many-Valued Logics.}
\L ukasiewicz~\cite{lukasiewicz1920} introduced three-valued and infinite-valued logics.
Kleene~\cite{kleene1952} developed strong and weak three-valued logics.
G\"odel~\cite{godel1932} introduced three-valued intuitionistic logic.
Belnap~\cite{belnap1977} introduced a four-valued logic for reasoning with
incomplete and contradictory information.
Zadeh~\cite{zadeh1965} introduced fuzzy sets with $[0,1]$ membership degrees.
Section~\ref{sec:scope} shows that no analogous collapse exists for
\L ukasiewicz, G\"odel, or classical algebras, making the Kleene algebra the
unique natural three-valued quotient of Cred.

\textbf{Copulas and Fuzzy Implications.}
Grzegorzewski~\cite{grzegorzewski2013} defined \emph{probabilistic
implications} $I_C(a,b) = C(a,b)/a$ from a copula $C$ and computed that
the Fr\'echet upper bound (min copula) yields the Goguen (product
residuated) implication.
This is the copula-side construction; Cred arrives at the same arrow from
the chain-rule side by setting joint $= \min(a,b)$.
The Bayes consistency criterion---that $f(a,b) \cdot a = f(b,a) \cdot b$,
equivalently that the induced joint is symmetric---and its use to
\emph{diagnose} why RM3 and G\"odel arrows fail (they are not
copula-derived) appears to be new.
Lawvere~\cite{lawvere1973} showed that residuation generalizes directed
distance in enriched categories; this principle applies to the product
residuated via the additive generator $-\log x$~\cite{klement2000}.

\textbf{Measure-Theoretic Conditioning.}
In measure theory, conditioning on null sets uses disintegration~\cite{chang1997},
yielding a specific (a.e.\ unique) value determined by the measure structure.
Fritz~\cite{fritz2020} treats conditioning categorically via Markov categories.
Cred drops the measure, so disintegration does not apply; at evidence zero, the
conditional is unconstrained rather than determined.

\textbf{Cred's Contributions.}
The product De Morgan triplet is standard; primitive conditioning is due to
R\'enyi and Popper; the connection to relevant logic is van Fraassen's;
the chain rule constraint in a fuzzy setting appears in H\'ajek et al.;
copula-derived implications are due to Grzegorzewski.
Given this prior art, Cred contributes:
\begin{enumerate}[label=(\arabic*)]
\item Identifying the chain rule as the sole structural axiom needed beyond
$[0,1]$ and complement---a weakening of probability that retains conditioning
while dropping the measure.
\item Adopting chain-rule conditioning as a design choice (not a workaround),
replacing residuated implication in the product De Morgan triplet.
\item The explicit collapse homomorphism to the Kleene algebra (underlying
K$_3$, LP, and RM3), with impossibility results for \L ukasiewicz, G\"odel,
and classical targets, and Lean verification.
\item Bayes consistency as a selection criterion: the chain rule applied from
both directions uniquely selects the product residuated implication among
standard arrows, equivalently characterizing Bayes-consistent arrows as
copula-derived, with RM3 and G\"odel as counterexamples (machine-checked).
\item Systematic treatment of fixed points with machine-checked proofs.
\end{enumerate}

%=============================================================================
\section{Lean Formalization}
\label{sec:lean}
%=============================================================================

The core algebra is formalized in Lean~4 with Mathlib.
The formalization comprises approximately 1120 lines covering:

\begin{itemize}
\item \textbf{Types:} \texttt{Credence} (value in $[0,1]$), \texttt{Conditioning}
(packaging conditional with chain rule proof), \texttt{ThreeVal} (three-valued type).
\item \textbf{Operations:} \texttt{neg}, \texttt{conj}, \texttt{disj},
\texttt{collapse}.
\item \textbf{Theorems:} Algebraic properties, conditioning behavior, fixed points,
collapse homomorphism.
\end{itemize}

\textbf{Key definitions (Lean syntax):}
\begin{verbatim}
structure Credence where
  val : R
  nonneg : 0 <= val
  le_one : val <= 1

structure Conditioning (joint evidence : Credence) where
  condCred : Credence
  chainRule : conj condCred evidence = joint
\end{verbatim}

The chain rule uses \texttt{conj} (multiplication on credence values).
The \texttt{joint} parameter is supplied as input, representing $\cred(A \land B)$.

\textbf{Key theorems:}
\begin{itemize}
\item \texttt{conditioning\_zero\_any}: any credence satisfies chain rule when evidence $= 0$
\item \texttt{conditioning\_unique}: uniqueness when evidence $> 0$
\item \texttt{liar\_fixed\_point}: $\negc \tfrac{1}{2} = \tfrac{1}{2}$
\item \texttt{neg\_fixed\_point\_unique}: $\tfrac{1}{2}$ is the unique negation fixed point
\item \texttt{collapse\_neg}, \texttt{collapse\_conj}, \texttt{collapse\_disj}: homomorphism proofs
\item \texttt{conj\_disj\_not\_distrib}: non-distributivity counterexample
\item \texttt{conj\_idempotent\_iff}: idempotence only at $\{0,1\}$
\item \texttt{boolean\_neg\_closed}, \texttt{boolean\_conj\_closed},
\texttt{boolean\_disj\_closed}: $\{0,1\}$ is a subalgebra
\item \texttt{three\_val\_conj\_not\_closed}: $\{0,\tfrac{1}{2},1\}$ is not a subalgebra
\item \texttt{no\_boolean\_neg\_retraction}: no negation-preserving map $[0,1] \to \{0,1\}$
\item \texttt{no\_godel\_collapse}: no negation-preserving map to G\"odel algebra
\item \texttt{no\_luk\_collapse}: no negation-and-conjunction-preserving map to \L ukasiewicz
\item \texttt{godel\_impl}, \texttt{rm3\_impl}: implication tables with comparison theorems
\item \texttt{rm3\_not\_bayes\_consistent}: RM3 implication fails Bayes consistency
\item \texttt{godel\_bayes\_consistent}: G\"odel/product residuated is Bayes-consistent on $\{0,\tfrac{1}{2},1\}$
\item \texttt{frechet\_upper}, \texttt{frechet\_lower}: Fr\'echet-Hoeffding bounds
\item \texttt{frechet\_conditioning\_exists}: bidirectional conditioning for any joint in bounds
\end{itemize}

Verification: \texttt{cd lean \&\& lake build}.
Toolchain: Lean 4.16.0 with Mathlib 4.16.0.

%=============================================================================
\section{Conclusion}
\label{sec:conclusion}
%=============================================================================

Cred is probability without the measure: values in $[0,1]$, the complement, and the
chain rule as the sole structural axiom.
Every probability space satisfies these axioms; Cred drops the sigma-algebra
and sigma-additivity.
When evidence has credence zero, the chain rule is satisfied by any value,
leaving the conditional unconstrained---avoiding ``from impossibility, anything
follows'' by the same chain rule that probability uses, without the measure
theory that would resolve it.

The product and its De Morgan dual provide the natural algebra on $[0,1]$,
shared with product fuzzy logic.
The three-valued collapse establishes a surjective homomorphism to the Kleene
algebra---the shared foundation of K$_3$, LP, and RM3---for the connective
fragment $\{\negc, \conjc, \disjc\}$, and no analogous collapse exists for
\L ukasiewicz, G\"odel, or classical targets.
Conditioning, when instantiated as a binary arrow via the Kleene joint,
recovers the product residuated implication for positive evidence; the sole
divergence from product fuzzy logic is at evidence zero, where Cred is
unconstrained rather than forced to~$1$.
Applying the chain rule from both directions (Bayes' theorem) selects the
product residuated implication as the unique Bayes-consistent arrow among
standard logic implications: RM3's arrow fails even on three values, and
G\"odel's fails on the continuum.
The product residuated implication is copula-derived---it arises from
conditioning under the Fr\'echet upper bound (maximum positive
dependence)---and Bayes consistency is equivalent to copula-derivability,
explaining why non-copula arrows fail.
The negation fixed point at $\tfrac{1}{2}$ provides a natural algebraic value
for self-negating sentences.
All core results are machine-checked in Lean~4.

Future work includes extending Cred to first-order with graded quantifiers and
exploring connections to paraconsistent set theory.

%=============================================================================
% Bibliography
%=============================================================================

\begin{thebibliography}{99}

\bibitem{adams1965}
Adams, E.W.
\newblock The logic of conditionals.
\newblock \emph{Inquiry}, 8:166--197, 1965.

\bibitem{adams1998}
Adams, E.W.
\newblock \emph{A Primer of Probability Logic}.
\newblock CSLI Publications, Stanford, 1998.

\bibitem{anderson1975}
Anderson, A.R. and Belnap, N.D.
\newblock \emph{Entailment: The Logic of Relevance and Necessity}, Vol.~I.
\newblock Princeton University Press, 1975.

\bibitem{belnap1977}
Belnap, N.D.
\newblock A useful four-valued logic.
\newblock In Dunn, J.M. and Epstein, G., editors, \emph{Modern Uses of
Multiple-Valued Logic}, pages 5--37. Reidel, 1977.

\bibitem{brady2006}
Brady, R.
\newblock \emph{Universal Logic}.
\newblock CSLI Lecture Notes, vol.~109. CSLI Publications, 2006.

\bibitem{chang1997}
Chang, J.T. and Pollard, D.
\newblock Conditioning as disintegration.
\newblock \emph{Statistica Neerlandica}, 51(3):287--317, 1997.

\bibitem{cox1961}
Cox, R.T.
\newblock \emph{The Algebra of Probable Inference}.
\newblock Johns Hopkins University Press, 1961.

\bibitem{dacosta1974}
da Costa, N.C.A.
\newblock On the theory of inconsistent formal systems.
\newblock \emph{Notre Dame Journal of Formal Logic}, 15(4):497--510, 1974.

\bibitem{definetti1937}
de Finetti, B.
\newblock La pr\'evision: ses lois logiques, ses sources subjectives.
\newblock \emph{Annales de l'Institut Henri Poincar\'e}, 7:1--68, 1937.

\bibitem{dunn1976}
Dunn, J.M.
\newblock Intuitive semantics for first-degree entailments and coupled trees.
\newblock \emph{Philosophical Studies}, 29:149--168, 1976.

\bibitem{fritz2020}
Fritz, T.
\newblock A synthetic approach to Markov kernels, conditional independence
and theorems on sufficient statistics.
\newblock \emph{Advances in Mathematics}, 370, Article 107239, 2020.

\bibitem{godel1932}
G\"odel, K.
\newblock Zum intuitionistischen Aussagenkalk\"ul.
\newblock \emph{Anzeiger der Akademie der Wissenschaften in Wien},
69:65--66, 1932.

\bibitem{grzegorzewski2013}
Grzegorzewski, P.
\newblock Probabilistic implications.
\newblock \emph{Fuzzy Sets and Systems}, 226:53--66, 2013.

\bibitem{hajek1995}
H\'ajek, P., Godo, L., and Esteva, F.
\newblock Fuzzy logic and probability.
\newblock In \emph{Proceedings of the 11th Conference on Uncertainty in
Artificial Intelligence (UAI)}, pages 237--244, 1995.

\bibitem{hajek1998}
H\'ajek, P.
\newblock \emph{Metamathematics of Fuzzy Logic}.
\newblock Trends in Logic, vol.~4. Kluwer, 1998.

\bibitem{jaynes2003}
Jaynes, E.T.
\newblock \emph{Probability Theory: The Logic of Science}.
\newblock Cambridge University Press, 2003.

\bibitem{kleene1952}
Kleene, S.C.
\newblock \emph{Introduction to Metamathematics}.
\newblock North-Holland, 1952.

\bibitem{klement2000}
Klement, E.P., Mesiar, R., and Pap, E.
\newblock \emph{Triangular Norms}.
\newblock Trends in Logic, vol.~8. Kluwer, 2000.

\bibitem{kripke1975}
Kripke, S.
\newblock Outline of a theory of truth.
\newblock \emph{Journal of Philosophy}, 72(19):690--716, 1975.

\bibitem{lawvere1973}
Lawvere, F.W.
\newblock Metric spaces, generalized logic, and closed categories.
\newblock \emph{Rendiconti del Seminario Matematico e Fisico di Milano},
43:135--166, 1973.

\bibitem{lewis1976}
Lewis, D.
\newblock Probabilities of conditionals and conditional probabilities.
\newblock \emph{Philosophical Review}, 85:297--315, 1976.

\bibitem{lukasiewicz1920}
\L ukasiewicz, J.
\newblock O logice tr\'ojwarto\'sciowej.
\newblock \emph{Ruch Filozoficzny}, 5:170--171, 1920.

\bibitem{meyer1972}
Meyer, R.K. and Routley, R.
\newblock Algebraic analysis of entailment I.
\newblock \emph{Logique et Analyse}, 15:407--428, 1972.

\bibitem{nelsen2006}
Nelsen, R.B.
\newblock \emph{An Introduction to Copulas}.
\newblock Springer Series in Statistics, 2nd edition. Springer, 2006.

\bibitem{popper1959}
Popper, K.
\newblock \emph{The Logic of Scientific Discovery}.
\newblock Hutchinson, 1959.

\bibitem{priest1979}
Priest, G.
\newblock The logic of paradox.
\newblock \emph{Journal of Philosophical Logic}, 8:219--241, 1979.

\bibitem{ramsey1931}
Ramsey, F.P.
\newblock Truth and probability.
\newblock In Braithwaite, R.B., editor, \emph{The Foundations of Mathematics
and other Logical Essays}, pages 156--198. Kegan Paul, 1931.

\bibitem{renyi1955}
R\'enyi, A.
\newblock On a new axiomatic theory of probability.
\newblock \emph{Acta Mathematica Academiae Scientiarum Hungaricae}, 6:285--335, 1955.

\bibitem{renyi1970}
R\'enyi, A.
\newblock \emph{Foundations of Probability}.
\newblock Holden-Day, 1970.

\bibitem{vanfraassen1976}
van Fraassen, B.C.
\newblock Probabilities of conditionals.
\newblock In Harper, W.L. and Hooker, C.A., editors, \emph{Foundations of
Probability Theory, Statistical Inference, and Statistical Theories of
Science}, Vol.~1, pages 261--308. Reidel, 1976.

\bibitem{vanfraassen1983}
van Fraassen, B.C.
\newblock Gentlemen's wagers: Relevant logic and probability.
\newblock \emph{Philosophical Studies}, 43:47--61, 1983.

\bibitem{weber2010}
Weber, Z.
\newblock Transfinite numbers in paraconsistent set theory.
\newblock \emph{Review of Symbolic Logic}, 3(1):71--92, 2010.

\bibitem{zadeh1965}
Zadeh, L.A.
\newblock Fuzzy sets.
\newblock \emph{Information and Control}, 8:338--353, 1965.

\end{thebibliography}

\end{document}
