% The Credence Algebra: A Primer
% Part 5: rigorous entry point to the Cred series
%
% References to Lean formalization: lean/Cred/Core, lean/Cred/Collapse,
% lean/Cred/Cond, lean/Cred/Bridge, lean/Cred/Fixpoint

\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
    pdfauthor={Christopher Albert},
    pdftitle={The Credence Algebra: A Primer}
}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}
\emergencystretch=3em
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes}
\usepackage{graphicx}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

% Custom commands
\newcommand{\cred}{\mathsf{cred}}
\newcommand{\negc}{\mathord{\sim}}
\newcommand{\conjc}{\otimes}
\newcommand{\disjc}{\sqcup}
\newcommand{\condc}[2]{({#1} \mid {#2})}
\newcommand{\Cond}{\mathsf{Cond}}
\newcommand{\Sol}{\mathsf{Sol}}
\newcommand{\Kl}{\mathbb{K}}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Nat}{\mathbb{N}}

\title{The Credence Algebra: A Primer}

\author{Christopher Albert\\
\small Institute for Theoretical and Computational Physics\\
\small Graz University of Technology}
\date{June 2026}

\begin{document}

\maketitle

\begin{abstract}
This primer is a self-contained development of the credence algebra
used throughout the Cred series: the unit interval $[0,1]$ with
complement $\negc c = 1 - c$, product conjunction
$a \conjc b = ab$, and the De Morgan dual disjunction
$a \disjc b = a + b - ab$.
We prove the De Morgan triplet laws, locate the idempotents and the
unique negation fixed point $\tfrac{1}{2}$, and measure the exact
failure of distributivity.
The map sending $0$ to $0$, $1$ to $1$, and the interior to
$\tfrac{1}{2}$ is shown to be a surjective homomorphism onto the
Kleene lattice $\{0, \tfrac{1}{2}, 1\}$; no analogous map exists onto
classical, G\"odel, or \L ukasiewicz targets.
Designation then yields the consequence relations K$_3$ and LP and
their graded refinements, with explosion failing below threshold
$\tfrac{1}{2}$ and excluded middle holding up to $\tfrac{3}{4}$.
Conditioning enters as the solution set
$\Cond(j,e) = \{c : c \conjc e = j\}$ of the chain rule, which is a
singleton, the full interval, or empty; on crisp values it diverges
from the material conditional exactly at impossible evidence.
Finally, self-referential sentences are read as fixed-point equations,
and the product residuum is shown to be incompatible with contraction,
which blocks Curry's paradox.
Every result is elementary and proved in full; all are machine-checked
in Lean~4 with Mathlib.
The primer is the entry point to
Parts~1--4~\cite{part1,part2}, which develop the congruence
classification, the probability bridge, and the foundational calculus.
\end{abstract}

%=============================================================================
\section{Introduction}
\label{sec:intro}
%=============================================================================

A \emph{credence} is a real number in $[0,1]$ recording degree of
belief~\cite{definetti1937,jaynes2003}.
The Cred series studies one specific algebra on these values: the
\emph{product De Morgan triplet}~\cite{klement2000,hajek1998},
consisting of the complement $\negc c = 1 - c$, the product
$a \conjc b = ab$, and the dual disjunction
$a \disjc b = a + b - ab$,
together with \emph{chain-rule conditioning}, the constraint
$\cred(A \mid B) \cdot \cred(B) = \cred(A \land B)$ taken as
primitive rather than derived from division.
This primer develops the structure from scratch.
Every notion gets a formula, every claim a proof; the reader needs
only real arithmetic.

Three levels organize the material.
The \emph{value level} (Sections~\ref{sec:algebra}
and~\ref{sec:fixed}) is the algebra itself: operations, order,
idempotents, fixed points.
The \emph{consequence level} (Sections~\ref{sec:collapse}
and~\ref{sec:consequence}) asks which value assignments propagate
designated status from premises to conclusions; the three-element
quotient $\{0, \tfrac{1}{2}, 1\}$ and the logics K$_3$ and LP live
here.
The \emph{conditioning level} (Section~\ref{sec:conditioning}) asks
which conditional credences are consistent with a given joint and
evidence; the answer is a solution set, not a connective.
Section~\ref{sec:fixpoint} applies all three levels to
self-referential sentences.

Two design choices drive everything.
First, conjunction is the \emph{product}, not the minimum, so the
interior $(0,1)$ is closed under all three operations
(Theorem~\ref{thm:collapse-hom}) and the only idempotents are the
endpoints (Proposition~\ref{prop:idempotent}).
Second, conditioning is the \emph{chain rule}, so impossible evidence
constrains nothing: $c \cdot 0 = 0$ holds for every $c$, and no ex
falso is built in (Theorem~\ref{thm:trichotomy}).
The interaction of the two choices is the recurring theme: the
negation fixed point $\tfrac{1}{2}$ forces the Kleene lattice as the
collapse target, the unconstrained zero-evidence case separates
conditioning from the material conditional, and the product residuum
refuses contraction.

Each result carries its Lean~4 name once, in the form
``(machine-checked, \texttt{name})''; the proofs in the text stand on
their own.
Appendix~\ref{app:lean} locates the names in the repository.
This primer states no new theorems: everything here is either
classical~\cite{kleene1952,hajek1998,klement2000} or proved in
Parts~1--4~\cite{part1,part2}.
What is new is the arrangement: one document, full proofs, no
prerequisites.

%=============================================================================
\section{The Value Algebra}
\label{sec:algebra}
%=============================================================================

\subsection{Operations}

\begin{definition}[Credence]
\label{def:credence}
A \emph{credence} is a real number $c \in [0,1]$.
The value $0$ is \emph{impossibility}, the value $1$ is
\emph{certainty}.
(Machine-checked as the structure \texttt{Credence}, a real with
proofs $0 \leq c$ and $c \leq 1$.)
\end{definition}

\begin{definition}[Negation]
\label{def:neg}
The \emph{negation} of $c$ is
\[
\negc c = 1 - c.
\]
Since $c \in [0,1]$ implies $1 - c \in [0,1]$, negation maps
credences to credences.
\end{definition}

\begin{definition}[Conjunction]
\label{def:conj}
The \emph{conjunction} of $a$ and $b$ is the product
\[
a \conjc b = a \cdot b.
\]
For $a, b \in [0,1]$ we have $0 \leq ab \leq a \leq 1$, so $\conjc$
maps $[0,1]^2$ to $[0,1]$.
\end{definition}

The product is the product t-norm of fuzzy
logic~\cite{klement2000,hajek1998}; Goguen~\cite{goguen1969} used it
as the conjunction of his logic of inexact concepts.
It is an operation on credence \emph{values}.
It does not compute the joint credence of two propositions: that
requires dependence information and is supplied separately in
Section~\ref{sec:conditioning}.

\begin{definition}[Disjunction]
\label{def:disj}
The \emph{disjunction} of $a$ and $b$ is the De Morgan dual of
conjunction:
\[
a \disjc b
  = \negc(\negc a \conjc \negc b)
  = 1 - (1-a)(1-b)
  = a + b - ab.
\]
Since $(1-a)(1-b) \in [0,1]$, the dual stays in $[0,1]$.
\end{definition}

The formula $a + b - ab$ is the \emph{probabilistic sum}.
Figure~\ref{fig:surfaces} plots both binary operations over the unit
square.
The triple $(\negc, \conjc, \disjc)$ is the product De Morgan
triplet.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.46\textwidth,
  view={125}{28},
  domain=0:1, y domain=0:1,
  samples=21, samples y=21,
  xlabel={$a$}, ylabel={$b$},
  zmin=0, zmax=1,
  zlabel={$a \conjc b$},
  title={$a \conjc b = ab$}]
\addplot3[surf] {x*y};
\end{axis}
\end{tikzpicture}
\hfill
\begin{tikzpicture}
\begin{axis}[
  width=0.46\textwidth,
  view={125}{28},
  domain=0:1, y domain=0:1,
  samples=21, samples y=21,
  xlabel={$a$}, ylabel={$b$},
  zmin=0, zmax=1,
  zlabel={$a \disjc b$},
  title={$a \disjc b = a + b - ab$}]
\addplot3[surf] {x+y-x*y};
\end{axis}
\end{tikzpicture}
\caption{The two binary operations on $[0,1]^2$.
Left: the product conjunction, zero on the boundary edges
$a = 0$ and $b = 0$ and equal to the identity along $b = 1$.
Right: the De Morgan dual, one along $a = 1$ and $b = 1$ and equal to
the identity along $b = 0$.
Both surfaces map the open square into the open interval $(0,1)$,
the closure property behind Theorem~\ref{thm:collapse-hom}.}
\label{fig:surfaces}
\end{figure}

\begin{definition}[Order]
\label{def:order}
Credences carry the order of $\Real$: $a \leq b$ as real numbers.
This is a partial (indeed total) order with least element $0$ and
greatest element $1$.
\end{definition}

\begin{example}[Worked values]
\label{ex:values}
Let $a = 0.6$ and $b = 0.7$.
Then
$\negc a = 0.4$,
$a \conjc b = 0.42$, and
$a \disjc b = 0.6 + 0.7 - 0.42 = 0.88$.
The dual route agrees:
$\negc(\negc a \conjc \negc b) = 1 - (0.4)(0.3) = 0.88$.
Note $a \conjc b \leq \min(a,b)$ and
$a \disjc b \geq \max(a,b)$, with strict inequality here.
\end{example}

\subsection{The De Morgan triplet laws}

\begin{proposition}[Triplet laws]
\label{prop:laws}
For all $a, b, c \in [0,1]$:
\begin{enumerate}[label=(\alph*),nosep]
\item $\negc\negc a = a$; \quad $\negc 0 = 1$; \quad $\negc 1 = 0$
  \hfill (involution, boundary)
\item $a \conjc b = b \conjc a$; \quad
  $(a \conjc b) \conjc c = a \conjc (b \conjc c)$; \quad
  $a \conjc 1 = a$; \quad $a \conjc 0 = 0$
  \hfill (commutative monoid)
\item $a \disjc b = b \disjc a$; \quad
  $(a \disjc b) \disjc c = a \disjc (b \disjc c)$; \quad
  $a \disjc 0 = a$; \quad $a \disjc 1 = 1$
  \hfill (dual monoid)
\item $\negc(a \conjc b) = \negc a \disjc \negc b$; \quad
  $\negc(a \disjc b) = \negc a \conjc \negc b$
  \hfill (De Morgan)
\item $a \leq b \iff \negc b \leq \negc a$; \quad
  $a \leq a' \Rightarrow a \conjc b \leq a' \conjc b$
  and $a \disjc b \leq a' \disjc b$
  \hfill (order)
\end{enumerate}
\end{proposition}

\begin{proof}
(a) $1 - (1 - a) = a$; the boundary values are immediate.
(b) Commutativity and associativity are those of real
multiplication; $a \cdot 1 = a$ and $a \cdot 0 = 0$.
(c) In the form $a \disjc b = 1 - (1-a)(1-b)$, all four identities
transport from (b) through the substitution $x \mapsto 1 - x$:
for associativity,
$(a \disjc b) \disjc c = 1 - (1-a)(1-b)(1-c) = a \disjc (b \disjc c)$.
(d) $\negc(a \conjc b) = 1 - ab$ and
$\negc a \disjc \negc b = (1-a) + (1-b) - (1-a)(1-b) = 1 - ab$;
the second law follows by applying (a) to the first.
(e) $a \leq b \iff 1 - b \leq 1 - a$; multiplication by $b \geq 0$
preserves $\leq$; monotonicity of $\disjc$ follows from that of
$\conjc$ through (d) and the antitonicity of $\negc$.
(Machine-checked, \texttt{neg\_neg}, \texttt{neg\_zero},
\texttt{neg\_one}, \texttt{conj\_comm}, \texttt{conj\_assoc},
\texttt{conj\_one}, \texttt{conj\_zero}, \texttt{disj\_comm},
\texttt{disj\_assoc}, \texttt{disj\_zero}, \texttt{disj\_one},
\texttt{de\_morgan\_conj}, \texttt{de\_morgan\_disj},
\texttt{neg\_le\_neg\_iff}.)
\end{proof}

The structure is two commutative monoids exchanged by an involution.
It is not a lattice and not a De Morgan algebra: the operations are
not idempotent, and distributivity fails.
Both failures are quantitative, and both have exact formulas.

\begin{proposition}[Idempotents]
\label{prop:idempotent}
$a \conjc a = a$ if and only if $a \in \{0,1\}$, and likewise
$a \disjc a = a$ if and only if $a \in \{0,1\}$.
\end{proposition}

\begin{proof}
$a^2 = a$ rewrites to $a(a-1) = 0$, so $a = 0$ or $a = 1$.
Dually, $a \disjc a = 2a - a^2$, and $2a - a^2 = a$ rewrites to
$a(1-a) = 0$.
(Machine-checked, \texttt{conj\_idempotent\_iff}.)
\end{proof}

\begin{proposition}[Distributivity defect]
\label{prop:defect}
For all $a, b, c \in [0,1]$:
\begin{align}
(a \conjc b) \disjc (a \conjc c) - a \conjc (b \disjc c)
  &= abc\,(1-a) \;\geq\; 0,
\label{eq:defect-conj}\\
a \disjc (b \conjc c) - (a \disjc b) \conjc (a \disjc c)
  &= a\,(1-a)(1-b)(1-c) \;\geq\; 0.
\label{eq:defect-disj}
\end{align}
In particular $\conjc$ sub-distributes over $\disjc$ and $\disjc$
sub-distributes over $\conjc$, with equality in
\eqref{eq:defect-conj} if and only if $a \in \{0,1\}$ or $bc = 0$,
and in \eqref{eq:defect-disj} if and only if $a \in \{0,1\}$ or
$b = 1$ or $c = 1$.
\end{proposition}

\begin{proof}
For \eqref{eq:defect-conj}:
$a \conjc (b \disjc c) = a(b + c - bc) = ab + ac - abc$, while
$(a \conjc b) \disjc (a \conjc c) = ab + ac - a^2bc$;
the difference is $abc - a^2bc = abc(1-a)$.
For \eqref{eq:defect-disj}:
expanding $(a + b - ab)(a + c - ac)$ and subtracting from
$a + bc - abc$ gives
$a(1-a)(1 - b - c + bc) = a(1-a)(1-b)(1-c)$.
Both products of factors in $[0,1]$ are nonnegative, and each
vanishes exactly when one factor does.
\end{proof}

\begin{example}[Numeric counterexample to distributivity]
\label{ex:nondistrib}
Take $a = b = c = \tfrac{1}{2}$.
Then
\[
a \conjc (b \disjc c)
  = \tfrac{1}{2}\bigl(\tfrac{1}{2} + \tfrac{1}{2} - \tfrac{1}{4}\bigr)
  = \tfrac{3}{8}
  = \tfrac{6}{16},
\qquad
(a \conjc b) \disjc (a \conjc c)
  = \tfrac{1}{4} + \tfrac{1}{4} - \tfrac{1}{16}
  = \tfrac{7}{16},
\]
and the gap $\tfrac{1}{16}$ is the defect
$abc(1-a) = \tfrac{1}{8} \cdot \tfrac{1}{2}$ from
Proposition~\ref{prop:defect}.
(The counterexample is machine-checked,
\texttt{conj\_disj\_not\_distrib}; the closed-form defect is
elementary and verified here by hand.)
\end{example}

\subsection{Spread and certainty}

Applying conjunction to a credence and its own negation yields a
quantity with a classical name: the Bernoulli variance.

\begin{definition}[Spread and certainty]
\label{def:spread}
The \emph{spread} of $c$ is $c \conjc \negc c = c(1-c)$; the
\emph{certainty} of $c$ is its De Morgan dual,
$c \disjc \negc c = 1 - c(1-c)$.
\end{definition}

\begin{proposition}[Quarter cap, three-quarter floor]
\label{prop:spread}
For all $c \in [0,1]$,
\[
c \conjc \negc c
  = \tfrac{1}{4} - \bigl(c - \tfrac{1}{2}\bigr)^2
  \leq \tfrac{1}{4},
\qquad
c \disjc \negc c
  = \tfrac{3}{4} + \bigl(c - \tfrac{1}{2}\bigr)^2
  \geq \tfrac{3}{4},
\]
with equality in both exactly at $c = \tfrac{1}{2}$, and
$c \disjc \negc c = 1$ exactly at $c \in \{0,1\}$.
\end{proposition}

\begin{proof}
Complete the square: $c(1-c) = \tfrac{1}{4} - (c - \tfrac{1}{2})^2$.
The certainty statement is the complement.
Equality holds where the square vanishes; the value $1$ requires
$c(1-c) = 0$.
(Machine-checked, \texttt{spread\_le\_quarter},
\texttt{certainty\_ge\_three\_quarters},
\texttt{certainty\_eq\_one\_iff}.)
\end{proof}

Figure~\ref{fig:spread} plots both curves.
The two constants $\tfrac{1}{4}$ and $\tfrac{3}{4}$ return in
Section~\ref{sec:consequence} as the exact thresholds for
noncontradiction and excluded middle.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.62\textwidth, height=0.45\textwidth,
  xmin=0, xmax=1, ymin=0, ymax=1.05,
  xlabel={$c$},
  legend pos=outer north east,
  legend cell align=left,
  domain=0:1, samples=101]
\addplot[thick] {x*(1-x)};
\addlegendentry{spread $c \conjc \negc c$}
\addplot[thick, dashed] {1 - x*(1-x)};
\addlegendentry{certainty $c \disjc \negc c$}
\addplot[dotted] {0.25};
\addlegendentry{$\tfrac{1}{4}$}
\addplot[dotted] {0.75};
\addlegendentry{$\tfrac{3}{4}$}
\end{axis}
\end{tikzpicture}
\caption{Spread $c(1-c)$ and certainty $1 - c(1-c)$.
The spread is capped at $\tfrac{1}{4}$, the certainty floored at
$\tfrac{3}{4}$, both extremal at the negation fixed point
$c = \tfrac{1}{2}$ (Proposition~\ref{prop:spread}).
These two constants reappear as the explosion and excluded-middle
thresholds in Section~\ref{sec:consequence}.}
\label{fig:spread}
\end{figure}

%=============================================================================
\section{Fixed Points and Equilibria}
\label{sec:fixed}
%=============================================================================

\begin{lemma}[Unique negation fixed point]
\label{lem:fixed}
$\negc c = c$ if and only if $c = \tfrac{1}{2}$.
\end{lemma}

\begin{proof}
$1 - c = c$ rewrites to $2c = 1$.
(Machine-checked, \texttt{liar\_fixed\_point} and
\texttt{neg\_fixed\_point\_unique}.)
\end{proof}

The lemma is one line, and most of the primer depends on it.
On the two-element set $\{0,1\}$ the same equation $\negc c = c$ has
no solution; the interval supplies exactly one, at the midpoint.
Figure~\ref{fig:negation} shows the geometry.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.55\textwidth, height=0.55\textwidth,
  xmin=0, xmax=1, ymin=0, ymax=1,
  xlabel={$c$},
  legend pos=outer north east,
  legend cell align=left,
  domain=0:1, samples=101]
\addplot[thick] {1-x};
\addlegendentry{$\negc c = 1 - c$}
\addplot[dashed] {x};
\addlegendentry{identity}
\addplot[dotted, thick] {x*x};
\addlegendentry{$c \conjc c = c^2$}
\addplot[dash dot] {2*x - x*x};
\addlegendentry{$c \disjc c = 2c - c^2$}
\addplot[only marks, mark=*] coordinates {(0.5,0.5)};
\node at (axis cs:0.5,0.5) [anchor=south west]
  {$\bigl(\tfrac{1}{2},\tfrac{1}{2}\bigr)$};
\end{axis}
\end{tikzpicture}
\caption{The negation line $\negc c = 1 - c$ meets the identity only
at $c = \tfrac{1}{2}$ (Lemma~\ref{lem:fixed}).
Self-conjunction $c^2$ lies strictly below the identity on $(0,1)$
and self-disjunction $2c - c^2$ strictly above
(Proposition~\ref{prop:dynamics}); all three curves agree only at the
equilibria of Proposition~\ref{prop:equilibria}.}
\label{fig:negation}
\end{figure}

\begin{proposition}[Interior dynamics]
\label{prop:dynamics}
For $0 < c < 1$,
\[
c \conjc c < c < c \disjc c,
\]
and the powers $c^{n}$ decrease strictly in $n$.
\end{proposition}

\begin{proof}
$c^2 - c = c(c-1) < 0$ and $(2c - c^2) - c = c(1-c) > 0$ for
interior $c$.
For the powers, $c^{n+1} = c^n \cdot c < c^n$ since $0 < c < 1$ and
$c^n > 0$.
(Machine-checked, \texttt{self\_conj\_lt}, \texttt{self\_disj\_gt},
\texttt{pow\_strictly\_decreasing}.)
\end{proof}

\begin{proposition}[Equilibria]
\label{prop:equilibria}
$c \conjc c = c$ or $\negc c = c$ holds if and only if
$c \in \{0, \tfrac{1}{2}, 1\}$.
\end{proposition}

\begin{proof}
Combine Proposition~\ref{prop:idempotent} (idempotents $\{0,1\}$)
with Lemma~\ref{lem:fixed} (fixed point $\tfrac{1}{2}$).
(Machine-checked, \texttt{equilibria\_characterization}.)
\end{proof}

The three equilibria $\{0, \tfrac{1}{2}, 1\}$ are the candidate
carrier of any small quotient of the algebra.
The next section shows that they carry exactly one.

%=============================================================================
\section{The Kleene Collapse}
\label{sec:collapse}
%=============================================================================

\subsection{The Kleene lattice}

\begin{definition}[Kleene lattice]
\label{def:kleene}
The \emph{Kleene lattice} is the three-element set
$\Kl = \{0, \tfrac{1}{2}, 1\}$ with operations
\[
\mathrm{neg}(x) = 1 - x,
\qquad
\mathrm{conj}(x,y) = \min(x,y),
\qquad
\mathrm{disj}(x,y) = \max(x,y).
\]
\end{definition}

This is the value structure shared by Kleene's strong three-valued
logic K$_3$~\cite{kleene1952}, Priest's Logic of Paradox
LP~\cite{priest1979,asenjo1966}, and RM3~\cite{meyer1972}; the three
logics differ only in designation (Section~\ref{sec:consequence}).
On $\Kl$ the De Morgan laws
$1 - \min(x,y) = \max(1-x, 1-y)$ hold, $\min$ and $\max$ are
idempotent, and each distributes over the other.

\begin{definition}[Collapse]
\label{def:collapse}
The \emph{collapse} $\kappa : [0,1] \to \Kl$ is
\[
\kappa(c) =
\begin{cases}
0 & \text{if } c = 0,\\
\tfrac{1}{2} & \text{if } 0 < c < 1,\\
1 & \text{if } c = 1.
\end{cases}
\]
\end{definition}

\begin{theorem}[Collapse homomorphism]
\label{thm:collapse-hom}
The collapse is surjective and preserves all three operations:
for all $a, b \in [0,1]$,
\[
\kappa(\negc a) = \mathrm{neg}(\kappa(a)),
\qquad
\kappa(a \conjc b) = \min(\kappa(a), \kappa(b)),
\qquad
\kappa(a \disjc b) = \max(\kappa(a), \kappa(b)).
\]
\end{theorem}

\begin{proof}
Surjectivity: $\kappa(0) = 0$, $\kappa(\tfrac{1}{2}) = \tfrac{1}{2}$,
$\kappa(1) = 1$.

\emph{Negation.}
$\negc$ swaps $0$ and $1$ and maps the interior to itself
($0 < c < 1$ iff $0 < 1 - c < 1$); $\mathrm{neg}$ acts the same way
on $\Kl$.

\emph{Conjunction.}
If $a = 0$ or $b = 0$ then both sides are $0$
(the product vanishes; $\min$ with $0$ is $0$).
If $a = 1$ then $a \conjc b = b$ and
$\min(1, \kappa(b)) = \kappa(b)$; symmetrically for $b = 1$.
In the remaining case $a, b \in (0,1)$:
then $0 < ab \leq a < 1$, so
$\kappa(a \conjc b) = \tfrac{1}{2} =
\min(\tfrac{1}{2}, \tfrac{1}{2})$.

\emph{Disjunction.}
If $a = 1$ or $b = 1$ then $a \disjc b = 1$ and the maximum is $1$.
If $a = 0$ then $a \disjc b = b$, matching
$\max(0, \kappa(b))$; symmetrically for $b = 0$.
For $a, b \in (0,1)$ write $a \disjc b = 1 - (1-a)(1-b)$;
since $(1-a)(1-b) \in (0,1)$, the value lies in $(0,1)$ and both
sides are $\tfrac{1}{2}$.
(Machine-checked, \texttt{collapse\_neg}, \texttt{collapse\_conj},
\texttt{collapse\_disj}, \texttt{collapse\_surjective}.)
\end{proof}

\begin{remark}[Quotient, not subalgebra]
\label{rem:not-subalgebra}
The set $\{0, \tfrac{1}{2}, 1\}$ is not closed under $\conjc$:
$\tfrac{1}{2} \conjc \tfrac{1}{2} = \tfrac{1}{4}$ leaves it
(machine-checked, \texttt{three\_val\_conj\_not\_closed}).
The Kleene lattice is therefore a \emph{quotient} of the credence
algebra along $\kappa$, not a subalgebra.
By contrast $\{0,1\}$ \emph{is} a subalgebra: Boolean values are
closed under $\negc$, $\conjc$, $\disjc$
(machine-checked, \texttt{boolean\_conj\_closed} and its negation and
disjunction companions).
Classical logic embeds upward into the algebra; three-valued logic is
reached downward by collapsing.
Figure~\ref{fig:tower} shows both maps.
\end{remark}

\subsection{Impossibility of other targets}

The Kleene target is forced.
Each alternative fails on the fixed point of Lemma~\ref{lem:fixed}.

\begin{proposition}[No collapse to classical logic]
\label{prop:no-classical}
No map $\varphi : [0,1] \to \{0,1\}$ satisfies
$\varphi(\negc c) = 1 - \varphi(c)$ for all $c$.
\end{proposition}

\begin{proof}
At $c = \tfrac{1}{2}$, Lemma~\ref{lem:fixed} gives
$\varphi(\tfrac{1}{2}) = 1 - \varphi(\tfrac{1}{2})$, which has no
solution in $\{0,1\}$.
(Machine-checked, \texttt{no\_boolean\_neg\_retraction}.)
\end{proof}

\begin{proposition}[No collapse to G\"odel negation]
\label{prop:no-godel}
Equip $\Kl$ with the G\"odel negation~\cite{godel1932}
$\neg_G(0) = 1$ and $\neg_G(x) = 0$ for $x > 0$.
No map $\varphi : [0,1] \to \Kl$ satisfies
$\varphi(\negc c) = \neg_G(\varphi(c))$ for all $c$.
\end{proposition}

\begin{proof}
$\neg_G$ has no fixed point: $\neg_G(0) = 1 \neq 0$,
$\neg_G(\tfrac{1}{2}) = 0 \neq \tfrac{1}{2}$,
$\neg_G(1) = 0 \neq 1$.
But $\varphi(\tfrac{1}{2}) = \varphi(\negc\tfrac{1}{2}) =
\neg_G(\varphi(\tfrac{1}{2}))$ would be one.
(Machine-checked, \texttt{no\_godel\_collapse}.)
\end{proof}

\begin{proposition}[No collapse to \L ukasiewicz conjunction]
\label{prop:no-luk}
Equip $\Kl$ with the Kleene negation and the \L ukasiewicz
conjunction~\cite{lukasiewicz1920}
$x \conjc_{\text{\L}} y = \max(x + y - 1, 0)$.
No map $\varphi : [0,1] \to \Kl$ preserves both.
\end{proposition}

\begin{proof}
Negation preservation forces $\varphi(\tfrac{1}{2})$ to be a fixed
point of the Kleene negation on $\Kl$, hence
$\varphi(\tfrac{1}{2}) = \tfrac{1}{2}$.
Set $c = \sqrt{1/2}$, so that $c \conjc c = \tfrac{1}{2}$.
Then
$\tfrac{1}{2} = \varphi(c \conjc c)
  = \varphi(c) \conjc_{\text{\L}} \varphi(c)$.
But $x \conjc_{\text{\L}} x = \max(2x - 1, 0)$ takes the values
$0, 0, 1$ on $x = 0, \tfrac{1}{2}, 1$, never $\tfrac{1}{2}$.
(Machine-checked, \texttt{no\_luk\_collapse}.)
\end{proof}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[node distance=9mm and 16mm, >=Stealth]
% Hasse diagram of the Kleene lattice
\node (k0) {$0$};
\node (kh) [above=of k0] {$\tfrac{1}{2}$};
\node (k1) [above=of kh] {$1$};
\draw (k0) -- (kh) -- (k1);
\draw[{Stealth}-{Stealth}, dashed, bend left=60] (k0) to
  node[left, font=\small] {$\mathrm{neg}$} (k1);
\path (kh) edge [->, dashed, loop right] (kh);
\node [below=3mm of k0, font=\small]
  {Kleene lattice $\Kl$: order, $\mathrm{neg}$};
% Tower
\begin{scope}[xshift=62mm]
\node (I)  {$[0,1]$};
\node (K) [right=of I] {$\{0,\tfrac{1}{2},1\}$};
\node (B) [right=of K] {$\{0,1\}$};
\draw[->>] (I) to node[above, font=\small] {$\kappa$} (K);
\draw[->] (K) to node[above, font=\small] {$\times$}
  node[below, font=\small] {no $\mathrm{neg}$-map} (B);
\draw[{Hooks[right]}-{Stealth}, bend left=28] (B) to
  node[below, font=\small] {subalgebra} (I);
\node [below=12mm of K, font=\small, text width=58mm, align=center]
  {Tower: $\kappa$ is a surjective homomorphism
   (Theorem~\ref{thm:collapse-hom}); no negation-preserving map
   reaches $\{0,1\}$ (Proposition~\ref{prop:no-classical});
   $\{0,1\}$ embeds back as a subalgebra
   (Remark~\ref{rem:not-subalgebra}).};
\end{scope}
\end{tikzpicture}
\caption{Left: the Kleene lattice as a Hasse diagram, with negation
swapping $0$ and $1$ and fixing $\tfrac{1}{2}$.
Right: the collapse tower.
The second leg is crossed out: negation has no fixed point in
$\{0,1\}$, so the tower stops at three elements.}
\label{fig:tower}
\end{figure}

\begin{remark}[Uniqueness]
\label{rem:uniqueness}
Part~1~\cite{part1} sharpens these impossibility results to a
classification: among all three-element quotients compatible with
$\negc$ and $\conjc$, the Kleene lattice with partition
$\{0\}, (0,1), \{1\}$ is the only one, and no two-element quotient
exists.
The primer needs only the existence direction
(Theorem~\ref{thm:collapse-hom}) and
Proposition~\ref{prop:no-classical}.
\end{remark}

%=============================================================================
\section{Designation and Consequence}
\label{sec:consequence}
%=============================================================================

A valuation assigns values; a logic additionally says which values
license assertion.
That choice is \emph{designation}, and it is independent of the
algebra~\cite{belnap1977,priest2008}.

\begin{definition}[Designation]
\label{def:designation}
On $\Kl$, the \emph{K$_3$ designated set} is $\{1\}$ and the
\emph{LP designated set} is $\{\tfrac{1}{2}, 1\}$.
(Machine-checked, \texttt{isDesignatedK3},
\texttt{isDesignatedLP}.)
\end{definition}

\begin{definition}[Graded consequence]
\label{def:graded}
Fix a threshold $t \in (0,1]$ and a set $\alpha$ of atoms.
A \emph{valuation} is a map $v : \alpha \to [0,1]$.
The premises $\Gamma \subseteq \alpha$ \emph{entail} the conclusion
$\varphi \in \alpha$ at threshold $t$, written
$\Gamma \vDash_t \varphi$, when every valuation with $v(p) \geq t$
for all $p \in \Gamma$ also has $v(\varphi) \geq t$.
(Machine-checked, \texttt{gradedConsequence}.)
\end{definition}

Designation on $\Kl$ and thresholds on $[0,1]$ match across the
collapse: K$_3$ designation corresponds to $t = 1$ (certainty) and
LP designation to positivity; the formal bridges are proved in
Part~2~\cite{part2}.
Here we work directly on $[0,1]$ and prove the two facts a newcomer
needs: contradictions do not explode, and excluded middle holds only
up to a sharp threshold.

\begin{theorem}[No explosion below $\tfrac{1}{2}$]
\label{thm:no-explosion}
Let $0 < t \leq \tfrac{1}{2}$.
There is a valuation $v$ with
\[
v(A) \geq t,
\qquad
\negc v(A) \geq t,
\qquad
v(B) < t.
\]
Hence the premises $A$ and $\negc A$ do not entail an unrelated $B$
at any threshold up to $\tfrac{1}{2}$.
\end{theorem}

\begin{proof}
Take $v(A) = \tfrac{1}{2}$ and $v(B) = 0$.
Then $v(A) = \negc v(A) = \tfrac{1}{2} \geq t$, while
$v(B) = 0 < t$.
(Machine-checked, \texttt{graded\_no\_explosion}; the LP instance on
$\Kl$ is \texttt{lp\_no\_explosion}.)
\end{proof}

\begin{proposition}[Vacuity above $\tfrac{1}{2}$]
\label{prop:vacuity}
For $t > \tfrac{1}{2}$, no valuation satisfies both $v(A) \geq t$ and
$\negc v(A) \geq t$; the explosion sequent then holds vacuously.
\end{proposition}

\begin{proof}
$v(A) + \negc v(A) = 1$, so both values at least $t$ would force
$1 \geq 2t > 1$.
(Machine-checked, \texttt{graded\_explosion\_vacuous}.)
\end{proof}

The boundary is exactly $\tfrac{1}{2}$, the negation fixed point:
below it a contradiction is expressible and harmless, above it a
contradiction cannot be stated at full strength.
Excluded middle has the complementary boundary.

\begin{corollary}[Excluded-middle window]
\label{cor:tnd}
The formula $A \disjc \negc A$ is valid at threshold $t$, in the
sense that $v(A) \disjc \negc v(A) \geq t$ for every valuation, if
and only if $t \leq \tfrac{3}{4}$.
\end{corollary}

\begin{proof}
By Proposition~\ref{prop:spread},
$c \disjc \negc c \geq \tfrac{3}{4}$ for every $c$, with value
exactly $\tfrac{3}{4}$ at $c = \tfrac{1}{2}$.
So validity holds for $t \leq \tfrac{3}{4}$, and the valuation
$v(A) = \tfrac{1}{2}$ refutes it for $t > \tfrac{3}{4}$.
(Machine-checked, \texttt{threshold\_excluded\_middle\_iff}.)
\end{proof}

\begin{proposition}[K$_3$ has no tautologies, LP keeps excluded middle]
\label{prop:k3-lp}
On $\Kl$:
(i) the constant valuation $w \equiv \tfrac{1}{2}$ assigns
$\tfrac{1}{2}$ to every formula built from $\mathrm{neg}$, $\min$,
$\max$, so no formula is K$_3$-valid;
(ii) $\max(x, 1-x) \in \{\tfrac{1}{2}, 1\}$ for every $x \in \Kl$,
so $A \disjc \negc A$ is LP-valid.
\end{proposition}

\begin{proof}
(i) The singleton $\{\tfrac{1}{2}\}$ is closed under all three
operations: $1 - \tfrac{1}{2} = \tfrac{1}{2}$ and
$\min, \max$ of equal arguments are that argument.
By induction every formula evaluates to $\tfrac{1}{2}$, which is not
K$_3$-designated.
(ii) Check the three cases: $x = 0$ and $x = 1$ give $1$,
$x = \tfrac{1}{2}$ gives $\tfrac{1}{2}$; both are LP-designated.
(Machine-checked, \texttt{k3\_no\_tautology},
\texttt{lp\_excluded\_middle}; the Lean statement of (i) is for
atomic conclusions, and the inductive extension is the closure
computation displayed here.)
\end{proof}

\begin{example}[The map of thresholds]
\label{ex:thresholds}
Reading Theorem~\ref{thm:no-explosion},
Proposition~\ref{prop:vacuity}, and Corollary~\ref{cor:tnd} together:
for $0 < t \leq \tfrac{1}{2}$ contradictions are expressible and
explosion fails; for $\tfrac{1}{2} < t \leq \tfrac{3}{4}$
contradictory premises are unsatisfiable while excluded middle still
holds; for $t > \tfrac{3}{4}$ excluded middle fails as well.
The two boundary numbers are the cap and floor of
Figure~\ref{fig:spread}.
\end{example}

%=============================================================================
\section{Conditioning as a Solution Set}
\label{sec:conditioning}
%=============================================================================

\subsection{The chain rule and the admissible set}

Conditional credence is governed by one equation, taken as primitive
in the tradition of R\'enyi~\cite{renyi1955} and
Popper~\cite{popper1959}.

\begin{definition}[Chain rule]
\label{def:chain}
A \emph{conditioning witness} for the pair
$(j, e) \in [0,1]^2$, with $j$ the \emph{joint} credence
$\cred(A \land B)$ and $e$ the \emph{evidence} credence $\cred(B)$,
is a credence $c$ satisfying the \emph{chain rule}
\[
c \conjc e = j,
\qquad\text{in propositional notation}\quad
\condc{A}{B} \conjc \cred(B) = \cred(A \land B).
\]
(Machine-checked as the structure \texttt{Conditioning}.)
\end{definition}

\begin{definition}[Admissible set]
\label{def:cond}
The \emph{admissible set} of the pair $(j,e)$ is the solution set of
the chain rule,
\[
\Cond(j, e) = \{\, c \in [0,1] : c \conjc e = j \,\}.
\]
\end{definition}

\begin{proposition}[Membership criterion]
\label{prop:mem}
$c \in \Cond(j,e)$ if and only if $c \conjc e = j$, if and only if
there is a conditioning witness for $(j,e)$ whose conditional value
is $c$.
\end{proposition}

\begin{proof}
The first equivalence is Definition~\ref{def:cond}; the second packs
the equation into the witness and back.
(Machine-checked, \texttt{mem\_cond\_iff}.)
\end{proof}

\begin{theorem}[Trichotomy]
\label{thm:trichotomy}
For $j, e \in [0,1]$:
\begin{enumerate}[label=\textup{(\arabic*)},nosep]
\item if $e > 0$ and $j \leq e$, then
  $\Cond(j,e) = \{j/e\}$, a singleton;
\item if $e = 0$ and $j = 0$, then $\Cond(0,0) = [0,1]$, the full
  interval;
\item if $j > e$ (which includes $e = 0$ with $j > 0$), then
  $\Cond(j,e) = \emptyset$.
\end{enumerate}
Consequently $\Cond(j,e) \neq \emptyset$ if and only if $j \leq e$.
\end{theorem}

\begin{proof}
(1) For $e > 0$ the real equation $ce = j$ has the unique solution
$c = j/e$, and $j \leq e$ places it in $[0,1]$.
(2) $c \cdot 0 = 0$ holds for every $c$: the equation imposes no
constraint.
(3) Any solution satisfies
$j = ce \leq 1 \cdot e = e$, so $j > e$ admits none.
For the final equivalence: if $j \leq e$ and $e > 0$ apply (1); if
$j \leq e = 0$ then $j = 0$ and (2) applies.
(Machine-checked, \texttt{cond\_singleton\_of\_pos},
\texttt{cond\_zero\_zero\_univ}, \texttt{cond\_nonempty\_iff};
pointwise freedom at zero evidence is
\texttt{conditioning\_zero\_any}, uniqueness at positive evidence is
\texttt{conditioning\_unique}.)
\end{proof}

Case (2) is the design choice of the series.
Probability leaves $P(A \mid B)$ undefined at $P(B) = 0$; residuated
fuzzy implications force the value $1$ there
(ex falso, see Section~\ref{sec:fixpoint});
the chain rule leaves the value \emph{unconstrained}.
Dubois and Prade's conditional possibility instead defaults to $1$ at
zero evidence~\cite{dubois1988}; the admissible set records the
freedom rather than resolving it.

\begin{example}[Admissible sets at evidence $\tfrac{1}{2}$]
\label{ex:cond-table}
Fix $e = \tfrac{1}{2}$ and vary the joint:
\begin{center}
\begin{tabular}{ccl}
\toprule
$j$ & $\Cond(j, \tfrac{1}{2})$ & case in
  Theorem~\ref{thm:trichotomy} \\
\midrule
$0$ & $\{0\}$ & (1): $c = 0/\tfrac{1}{2}$ \\
$\tfrac{1}{4}$ & $\{\tfrac{1}{2}\}$
  & (1): $c = \tfrac{1}{4}/\tfrac{1}{2}$ \\
$\tfrac{1}{2}$ & $\{1\}$ & (1): $c = 1$ \\
$\tfrac{3}{4}$ & $\emptyset$ & (3): $j > e$, incoherent inputs \\
\bottomrule
\end{tabular}
\end{center}
Positive evidence pins the conditional to one value; the joint
selects which.
At $e = 0$ the entire column collapses to the two degenerate rows
$\Cond(0,0) = [0,1]$ and $\Cond(j,0) = \emptyset$ for $j > 0$.
\end{example}

\subsection{Divergence from the material conditional}

On crisp values the algebra is classical
(Remark~\ref{rem:not-subalgebra}), with one exception, and the
exception is exactly the zero-evidence case.

\begin{definition}[Crisp embedding and material conditional]
\label{def:material}
Write $\iota(\mathsf{false}) = 0$ and $\iota(\mathsf{true}) = 1$ for
the crisp embedding.
The \emph{material conditional} of evidence $e$ and joint $j$ in
$\{\mathsf{true}, \mathsf{false}\}$ is
$\mathrm{mat}(j,e) = \lnot e \lor j$; as a credence value it is
$\negc \iota(e) \disjc \iota(j)$.
\end{definition}

\begin{theorem}[Material divergence]
\label{thm:material}
Let $j, e$ be crisp and coherent
($j = \mathsf{true}$ implies $e = \mathsf{true}$).
Then
\[
\Cond\bigl(\iota(j), \iota(e)\bigr)
  = \{\iota(\mathrm{mat}(j,e))\}
\quad\iff\quad
e = \mathsf{true}.
\]
\end{theorem}

\begin{proof}
If $e = \mathsf{true}$: $c \conjc 1 = c$, so
$\Cond(\iota(j), 1) = \{\iota(j)\}$, and
$\mathrm{mat}(j, \mathsf{true}) = j$.
If $e = \mathsf{false}$: coherence forces $j = \mathsf{false}$, and
Theorem~\ref{thm:trichotomy}(2) gives $\Cond(0,0) = [0,1]$, which is
not the singleton $\{1\} = \{\iota(\mathrm{mat}(j,\mathsf{false}))\}$.
(Machine-checked, \texttt{material\_divergence}; the
characterization of full underdetermination on crisp inputs is
\texttt{cond\_underdetermined\_iff}.)
\end{proof}

\begin{example}[The moon and the sum]
\label{ex:moon}
Let $B$ be ``the moon is made of cheese'' and $A$ be ``$2+2=5$'',
with $\cred(B) = 0$, $\cred(A) = 0$, and joint
$\cred(A \land B) = 0$.
Read as a material conditional, ``if $B$ then $A$'' has value
\[
\negc \cred(B) \disjc \cred(A \land B) = 1 \disjc 0 = 1:
\]
designated under every designation of
Section~\ref{sec:consequence}.
This is the ex falso row: the connective computes its value from the
marginals, and a false antecedent forces the output $1$ regardless of
the consequent; on the Kleene quotient the same row appears in the
RM3 implication table (machine-checked, \texttt{rm3\_ex\_falso}).
Read as conditioning, the sentence asks for $\Cond(0,0)$, which by
Theorem~\ref{thm:trichotomy}(2) is all of $[0,1]$: every conditional
credence is admissible and none is forced
(machine-checked, \texttt{cred\_no\_ex\_falso}).
The material reading designates the sentence because vacuous truth is
hard-coded into the connective; the conditioning reading reports that
impossible evidence supports nothing, including the sentence itself.
Theorem~\ref{thm:material} says these two readings agree everywhere
else.
\end{example}

%=============================================================================
\section{Fixed-Point Equations and the Curry Block}
\label{sec:fixpoint}
%=============================================================================

\subsection{Self-reference as an equation}

A self-referential sentence specifies its own credence as a function
of itself.
The algebra treats it as an equation and reports the solutions.

\begin{definition}[Solution set]
\label{def:solutions}
For $\Phi : [0,1] \to [0,1]$, the \emph{solution set} is
\[
\Sol(\Phi) = \{\, c \in [0,1] : \Phi(c) = c \,\},
\]
the fixed-point set of $\Phi$.
(Machine-checked, \texttt{solutions},
\texttt{solutions\_eq\_fixedPoints}.)
\end{definition}

\begin{proposition}[Liar and truth-teller]
\label{prop:liar-tt}
$\Sol(\negc) = \{\tfrac{1}{2}\}$ and
$\Sol(\mathrm{id}) = [0,1]$.
On the crisp fragment $\{0,1\}$ the liar equation $\lnot b = b$ has
no solution.
\end{proposition}

\begin{proof}
The first claim is Lemma~\ref{lem:fixed}; the second is immediate
since every $c$ satisfies $c = c$; the third is the two-case check
$\lnot\mathsf{true} \neq \mathsf{true}$,
$\lnot\mathsf{false} \neq \mathsf{false}$.
(Machine-checked, \texttt{solutions\_neg\_eq\_singleton},
\texttt{solutions\_id\_eq\_univ},
\texttt{bool\_neg\_solutions\_empty}.)
\end{proof}

The liar (``this sentence is false'') pins one value; the
truth-teller (``this sentence is true'') pins none.
The truth-teller's solution set coincides with the zero-evidence
admissible set:
\[
\Sol(\mathrm{id}) = [0,1] = \Cond(0,0)
\]
(machine-checked,
\texttt{truth\_teller\_same\_shape\_as\_zero\_evidence}).
An uninformative equation and impossible evidence leave the same
freedom, reached from different levels.

\subsection{The product residuum and contraction}

Curry's paradox needs an internal conditional with three properties:
modus ponens, conditional proof, and contraction.
The algebra grants any two and refuses the third.

\begin{definition}[Product residuum]
\label{def:resid}
The \emph{product residuum} is
\[
a \to b =
\begin{cases}
1 & \text{if } a = 0,\\
\min\bigl(b/a,\, 1\bigr) & \text{if } a > 0.
\end{cases}
\]
(Machine-checked, \texttt{prodResid}.)
\end{definition}

This is the residuated implication of H\'ajek's product
logic~\cite{hajek1998}.
Note $0 \to b = 1$: the residuum hard-codes ex falso, in contrast to
$\Cond$ (Theorem~\ref{thm:trichotomy}).

\begin{definition}[MP, CP, contraction]
\label{def:mp-cp}
For $f : [0,1]^2 \to [0,1]$:
\begin{enumerate}[label=(\roman*),nosep]
\item \emph{Modus ponens}, $\mathrm{MP}(f)$:\quad
  $f(a,b) \conjc a \leq b$ for all $a, b$.
\item \emph{Conditional proof}, $\mathrm{CP}(f)$:\quad
  $c \conjc a \leq b$ implies $c \leq f(a,b)$, for all $a, b, c$.
\item \emph{Contraction}, $\mathrm{Contr}(f)$:\quad
  $f(a, f(a,b)) \leq f(a,b)$ for all $a, b$.
\end{enumerate}
\end{definition}

\begin{lemma}[MP and CP pin the residuum]
\label{lem:resid-unique}
If $f$ satisfies $\mathrm{MP}(f)$ and $\mathrm{CP}(f)$, then for all
$a > 0$ and all $b$,
$f(a,b) = \min(b/a, 1) = (a \to b)$.
\end{lemma}

\begin{proof}
MP and CP together say $c \conjc a \leq b \iff c \leq f(a,b)$:
the forward direction is CP, and from $c \leq f(a,b)$ monotonicity of
$\conjc$ (Proposition~\ref{prop:laws}(e)) gives
$c \conjc a \leq f(a,b) \conjc a \leq b$.
So $f(a,b)$ is the greatest $c$ with $ca \leq b$.
For $a > 0$ that maximum is $\min(b/a, 1)$:
it satisfies $\min(b/a,1)\,a \leq (b/a)a = b$, and any $c$ with
$ca \leq b$ has $c \leq b/a$ and $c \leq 1$.
(Machine-checked, \texttt{residuation\_unique},
\texttt{prodResid\_unique\_positive}.)
\end{proof}

\begin{theorem}[Curry block]
\label{thm:curry}
No $f : [0,1]^2 \to [0,1]$ satisfies $\mathrm{MP}(f)$,
$\mathrm{CP}(f)$, and $\mathrm{Contr}(f)$ simultaneously.
\end{theorem}

\begin{proof}
Suppose $f$ satisfies all three.
By Lemma~\ref{lem:resid-unique},
\[
f\bigl(\tfrac{1}{2}, \tfrac{1}{4}\bigr)
  = \min\bigl(\tfrac{1/4}{1/2},\, 1\bigr) = \tfrac{1}{2},
\qquad
f\bigl(\tfrac{1}{2}, \tfrac{1}{2}\bigr)
  = \min(1, 1) = 1.
\]
Contraction at $(a,b) = (\tfrac{1}{2}, \tfrac{1}{4})$ demands
\[
f\bigl(\tfrac{1}{2},\, f(\tfrac{1}{2}, \tfrac{1}{4})\bigr)
  = f\bigl(\tfrac{1}{2}, \tfrac{1}{2}\bigr)
  = 1
  \;\leq\;
  f\bigl(\tfrac{1}{2}, \tfrac{1}{4}\bigr)
  = \tfrac{1}{2},
\]
a contradiction.
(Machine-checked, \texttt{curry\_block}; the residuum's own
contraction failure at the same point is
\texttt{prod\_resid\_no\_contraction}.)
\end{proof}

\begin{remark}[Curry fixed points]
\label{rem:curry-fixed}
The Curry sentence ``if this sentence is true then $B$'' becomes the
equation $c = (c \to b)$ under the residuum.
For $b > 0$ it has the solution $c = \sqrt{b}$:
since $\sqrt{b} > 0$,
$\sqrt{b} \to b = \min(b/\sqrt{b}, 1) = \min(\sqrt{b}, 1) = \sqrt{b}$
(machine-checked, \texttt{curry\_fixed\_point\_positive}).
For $b = 0$ there is none: $c = 0$ gives $0 \to 0 = 1 \neq 0$, and
$c > 0$ gives $c \to 0 = 0 \neq c$
(machine-checked, \texttt{curry\_no\_fixed\_point\_zero}).
The paradoxical step is never the fixed point; it is contraction,
and Theorem~\ref{thm:curry} removes it.
\end{remark}

\begin{example}[Residuum values]
\label{ex:resid}
$\tfrac{4}{5} \to \tfrac{1}{5} = \min(\tfrac{1}{4}, 1) =
\tfrac{1}{4}$:
strong evidence, weak joint, small conditional.
$\tfrac{1}{5} \to \tfrac{4}{5} = \min(4, 1) = 1$:
when $b \geq a$ the residuum saturates.
Both values agree with the singleton case of
Theorem~\ref{thm:trichotomy} when the joint is $\min(a,b)$, the
maximally dependent choice; the two notions part company only at
$a = 0$, where the residuum says $1$ and $\Cond$ says everything.
\end{example}

%=============================================================================
\section{Scope and Further Reading}
\label{sec:scope}
%=============================================================================

The primer proved:
the De Morgan triplet laws with their exact distributivity defect
(Propositions~\ref{prop:laws} and~\ref{prop:defect});
the unique negation fixed point and the equilibria
(Lemma~\ref{lem:fixed}, Proposition~\ref{prop:equilibria});
the Kleene collapse and the impossibility of classical, G\"odel, and
\L ukasiewicz targets
(Theorem~\ref{thm:collapse-hom},
Propositions~\ref{prop:no-classical}--\ref{prop:no-luk});
the explosion and excluded-middle thresholds
(Theorem~\ref{thm:no-explosion}, Corollary~\ref{cor:tnd});
the conditioning trichotomy and the material divergence
(Theorems~\ref{thm:trichotomy} and~\ref{thm:material});
and the Curry block (Theorem~\ref{thm:curry}).

What the primer does not do:
it does not classify the congruences of the triplet (the Kleene
partition is the unique three-element quotient, and the four-class
case is open; Part~1~\cite{part1});
it does not develop the formula-level bridges between threshold
consequence and K$_3$/LP, Bayesian and Jeffrey updating, or the
copula uniqueness theorem (Part~2~\cite{part2});
and it does not treat graded predicates, quantifiers, the labelled
sequent calculus, or proof certificates (Parts~3 and~4).
Each of those developments starts from the algebra assembled here.

\section*{Acknowledgements}

Anthropic's Claude was used as an assistant for both the Lean
formalization and the manuscript preparation.

%=============================================================================
% Bibliography
%=============================================================================

\bibliographystyle{plain}
\bibliography{../cred}

%=============================================================================
% APPENDIX
%=============================================================================
\appendix

%=============================================================================
\section{Lean Formalization}
\label{app:lean}
%=============================================================================

All results are machine-checked in Lean~4 with Mathlib; the names
cited in the text resolve in the following modules of the repository.

\begin{itemize}[nosep]
\item \texttt{Cred/Core/Value.lean}: operations, order, triplet laws,
  idempotents, fixed points, spread and certainty, interior dynamics
  (Sections~\ref{sec:algebra} and~\ref{sec:fixed}).
\item \texttt{Cred/Collapse/ThreeVal.lean} and
  \texttt{Cred/Collapse/Hom.lean}: the Kleene lattice, the collapse
  homomorphism, the no-go results for Boolean, G\"odel, and
  \L ukasiewicz targets, the RM3 implication table
  (Section~\ref{sec:collapse}, Example~\ref{ex:moon}).
\item \texttt{Cred/Core/Consequence.lean}: designation, graded
  consequence, the no-explosion and vacuity results
  (Section~\ref{sec:consequence}).
\item \texttt{Cred/Threshold.lean}: the sharp excluded-middle window
  (Corollary~\ref{cor:tnd}).
\item \texttt{Cred/Cond/Admissible.lean}: the chain rule, the
  admissible set, the trichotomy
  (Section~\ref{sec:conditioning}).
\item \texttt{Cred/Bridge/Crisp.lean}: the crisp embedding and the
  material divergence (Theorem~\ref{thm:material}).
\item \texttt{Cred/Fixpoint.lean} and \texttt{Cred/Bridge/Curry.lean}:
  solution sets, liar and truth-teller, the residuum, the Curry block
  (Section~\ref{sec:fixpoint}).
\end{itemize}

Verification: \texttt{cd lean \&\& lake build}.
Toolchain: Lean 4.16.0 with Mathlib 4.16.0.

\end{document}
