% The Irreducible Commitment: Prior, Seed, and Supplied Conditioning
% Part 4: the bridge between Bayesian conditioning, the chain-rule primitive,
% and the bootstrap commitment, made precise.
%
% References to Lean formalization: lean/Cred/Cond/Admissible.lean,
% lean/Cred/Update.lean, lean/Cred/Coherence.lean, lean/Cred/Cromwell.lean,
% lean/Cred/Lindley.lean, lean/Cred/Valuation.lean, lean/Cred/ToyModel.lean

\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
    pdfauthor={Christopher Albert},
    pdftitle={The Irreducible Commitment: Prior, Seed, and Supplied Conditioning}
}
\usepackage[capitalize]{cleveref}
\usepackage[margin=1in]{geometry}
\emergencystretch=3em
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes}
\usepackage{graphicx}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

% Custom commands
\newcommand{\cred}{\mathsf{cred}}
\newcommand{\negc}{\mathord{\sim}}
\newcommand{\conjc}{\otimes}
\newcommand{\disjc}{\sqcup}
\newcommand{\condc}[2]{({#1} \mid {#2})}
\newcommand{\Cond}{\mathsf{Cond}}
\newcommand{\CondSet}{\mathsf{CondSet}}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Nat}{\mathbb{N}}

\title{The Irreducible Commitment:\\
Prior, Seed, and Supplied Conditioning}

\author{Christopher Albert\\
\small Institute for Theoretical and Computational Physics\\
\small Graz University of Technology}
\date{June 2026}

\begin{document}

\maketitle

\begin{abstract}
We make precise the bridge between Bayesian conditioning, the chain-rule
conditioning primitive of Parts~1--3, and the opening commitment that any
updating system must supply before it can run.
The formal core: the chain rule
$\cred(A \mid B) \conjc \cred(B) = \cred(A \land B)$ defines the
admissible set $\Cond(j,e) = \{c \in [0,1] : c \cdot e = j\}$, which
obeys a trichotomy (singleton $\{j/e\}$ for $e > 0$, all of $[0,1]$ for
$e = j = 0$, empty for $e = 0 < j$); Cromwell's rule (credences $0$ and
$1$ are absorbing under coherent conditioning); sequential update equals
batch update for a fixed joint model; a contraction model in which the
seed determines every finite stage yet washes out in the limit; and a
credal-set characterization in which ranging the supplied joint over a
class of priors yields exactly the interval-valued posteriors of
imprecise probability, degenerating to the full interval at zero
evidence.
The Bayesian model-comparison boundary (Occam factor, Bartlett's
improper-prior indeterminacy, Lindley's paradox) is derived exactly for
a Gaussian location model.
All credence-algebra results are machine-checked in Lean~4 with Mathlib;
the Gaussian computations are classical.
The synthesis is honest in scope: the prior, the foundational seed, and
the supplied joint play the same structural role, an input the machinery
needs and cannot generate, but the correspondence is an organizing
analogy, not a theorem; we claim no resolution of Bayesian foundations.
\end{abstract}

%=============================================================================
\section{Introduction}
\label{sec:intro}
%=============================================================================

Bayesian updating needs a prior before the first datum
arrives~\cite{ramsey1931,definetti1937,jaynes2003}; a formal foundation
needs axioms before the first proof runs~\cite{simpson2009}; chain-rule
conditioning needs a supplied joint credence before a conditional has a
value~\cite{part1}.
This paper makes the third instance fully precise and uses it to state
exactly what the first two share with it.

The setting is the credence algebra of Parts~1--3~\cite{part1,part2}:
values in $[0,1]$ with negation $\negc c = 1 - c$, product conjunction
$c_1 \conjc c_2 = c_1 c_2$, and De Morgan disjunction
$c_1 \disjc c_2 = c_1 + c_2 - c_1 c_2$.
Conditioning is primitive, governed by the chain rule
$\cred(A \mid B) \conjc \cred(B) = \cred(A \land B)$ in the tradition of
R\'enyi~\cite{renyi1955}, Popper~\cite{popper1959}, and
de Finetti~\cite{definetti1936}, without division and without a measure
space.
The joint credence is an input, not a function of the marginals; the
algebra does not manufacture dependence.

\textbf{Main results.}
\Cref{thm:trichotomy} classifies the admissible set
$\Cond(j,e) = \{c \in [0,1] : c \cdot e = j\}$:
a singleton for positive evidence, the full interval for $e = j = 0$,
empty otherwise.
\Cref{thm:cromwell} proves Cromwell's rule inside the algebra: a prior
of $0$ or $1$ is unmoved by any coherent evidence, and zero evidence is
absorbing; the same annihilator identity $c \cdot 0 = 0$ that yields
no-ex-falso~\cite{part2} yields the unrecoverable zero.
\Cref{thm:seq-batch} proves that sequential and batch updates agree for
a fixed joint model: coherence is path independence in the data
grouping, not convergence.
\Cref{thm:contraction} gives a contraction model separating the two:
the seed determines every finite stage exactly, yet the orbit converges
to the target from every seed.
\Cref{thm:condset-positive,thm:condset-zero} characterize the credal
set obtained by ranging the supplied joint over a class of priors, and
\Cref{thm:bartlett,thm:lindley} derive the Bartlett and Lindley limits
exactly for a Gaussian location model: the Occam factor in the marginal
likelihood $m(x) = \int p(x \mid \theta)\, d\pi(\theta)$ vanishes as the
prior widens, so a diffuse-enough seed loses every comparison.

\textbf{Scope.}
The credence-algebra results are machine-checked in Lean~4 with Mathlib
(\cref{app:lean}); the Gaussian computations of
\cref{sec:modelcomparison} are classical and not formalized.
The three-way correspondence of \cref{sec:synthesis} (prior, seed,
supplied joint) is an organizing analogy between systems, stated as
remarks; we claim no functor between them and no contribution to the
justification of Bayesian inference itself.
The M\"unchhausen trilemma~\cite{albert1968} and Neurath's
boat~\cite{neurath1932} frame the discussion; they prove nothing here.

\textbf{Outline.}
\Cref{sec:chainrule} defines the chain rule and classifies the
admissible set.
\Cref{sec:valuations} isolates the supplied joint.
\Cref{sec:update} defines Bayesian and Jeffrey updates and their
zero-evidence underdetermination.
\Cref{sec:coherence} proves sequential-equals-batch coherence.
\Cref{sec:cromwell} proves Cromwell's rule and its no-ex-falso tie.
\Cref{sec:contraction} gives the contraction model.
\Cref{sec:modelcomparison} develops credal sets and the model-comparison
boundary.
\Cref{sec:synthesis} states the synthesis.
\Cref{app:lean} maps results to the Lean formalization.

%=============================================================================
\section{The Chain Rule and the Admissible Set}
\label{sec:chainrule}
%=============================================================================

\begin{definition}[Credence algebra~\cite{part1}]
\label[definition]{def:credence}
A \emph{credence} is a real number $c \in [0,1]$.
Negation, conjunction, and disjunction are
\[
\negc c = 1 - c, \qquad
c_1 \conjc c_2 = c_1 c_2, \qquad
c_1 \disjc c_2 = \negc(\negc c_1 \conjc \negc c_2) = c_1 + c_2 - c_1 c_2 .
\]
\end{definition}

\begin{definition}[Conditioning structure]
\label[definition]{def:conditioning}
A \emph{conditioning structure} for a pair $(j, e) \in [0,1]^2$ of
\emph{joint} and \emph{evidence} credences is a credence
$c \in [0,1]$ satisfying the chain rule
\[
c \conjc e = j .
\]
In propositional notation, with $j = \cred(A \land B)$ and
$e = \cred(B)$: $\condc{A}{B} \conjc \cred(B) = \cred(A \land B)$.
(Machine-checked, \texttt{Conditioning}.)
\end{definition}

The joint $j$ is supplied as a parameter; it is not computed from
marginals.
\Cref{sec:valuations} returns to this point.

\begin{definition}[Admissible set]
\label[definition]{def:cond}
The \emph{admissible set} of conditional credences for $(j,e)$ is
\[
\Cond(j, e) = \{ c \in [0,1] : c \cdot e = j \} .
\]
(Machine-checked, \texttt{Cond}.)
\end{definition}

\begin{proposition}[Membership]
\label[proposition]{prop:membership}
$c \in \Cond(j,e)$ if and only if $c$ is a conditioning structure
for $(j,e)$.
\end{proposition}
\begin{proof}
Both sides assert the equation $c \cdot e = j$.
(Machine-checked, \texttt{mem\_cond\_iff}.)
\end{proof}

\begin{lemma}[Coherence bound]
\label[lemma]{lem:coherence-bound}
If $\Cond(j,e) \neq \emptyset$ then $j \leq e$.
\end{lemma}
\begin{proof}
Let $c \cdot e = j$ with $c \in [0,1]$.
Then $j = c \cdot e \leq 1 \cdot e = e$.
(Machine-checked, \texttt{conditioning\_implies\_joint\_le\_evidence}.)
\end{proof}

\begin{theorem}[Trichotomy of the admissible set]
\label{thm:trichotomy}
For all $j, e \in [0,1]$:
\begin{enumerate}[label=\textup{(\alph*)},nosep]
\item if $e > 0$ and $j \leq e$, then
  $\Cond(j,e) = \{ j/e \}$, a singleton;
\item if $e = j = 0$, then $\Cond(0,0) = [0,1]$, the full interval;
\item if $j > e$ (in particular $e = 0 < j$), then
  $\Cond(j,e) = \emptyset$.
\end{enumerate}
Consequently $\Cond(j,e) \neq \emptyset$ if and only if $j \leq e$.
\end{theorem}
\begin{proof}
(a)~For $e > 0$ the equation $c \cdot e = j$ is equivalent to
$c = j/e$, and $j \leq e$ gives $j/e \in [0,1]$, so exactly one
admissible value exists.
(b)~For every $c \in [0,1]$, $c \cdot 0 = 0$, so the defining equation
holds identically.
(c)~If $c \cdot e = j$ then $j \leq e$ by
\cref{lem:coherence-bound}, contradicting $j > e$.
The nonemptiness criterion combines (a)--(c): for $j \leq e$ either
$e > 0$ (case (a)) or $e = 0$, which forces $j = 0$ (case (b)).
(Machine-checked, \texttt{cond\_singleton\_of\_pos},
\texttt{cond\_zero\_zero\_univ}, \texttt{cond\_nonempty\_iff},
\texttt{cond\_zero\_evidence\_ne\_empty}.)
\end{proof}

\begin{corollary}[Zero-evidence underdetermination]
\label[corollary]{cor:zero-underdetermined}
For every $c \in [0,1]$ there is a conditioning structure for $(0,0)$
with conditional credence $c$: impossible evidence imposes no
constraint, and there is no ex falso.
Moreover, any conditioning structure with evidence $0$ forces $j = 0$.
\end{corollary}
\begin{proof}
The first claim is \cref{thm:trichotomy}(b) read through
\cref{prop:membership}; the second is \cref{lem:coherence-bound} with
$e = 0$.
(Machine-checked, \texttt{conditioning\_zero\_any},
\texttt{conditioning\_zero\_forces\_joint\_zero}.)
\end{proof}

\begin{proposition}[Existence and uniqueness at positive evidence]
\label[proposition]{prop:existence-uniqueness}
For $e > 0$, a conditioning structure for $(j,e)$ exists if and only if
$j \leq e$, and is then unique with value $j/e$.
\end{proposition}
\begin{proof}
Immediate from \cref{thm:trichotomy}(a) and (c).
(Machine-checked, \texttt{conditioning\_mk},
\texttt{conditioning\_unique}.)
\end{proof}

\subsection{Fr\'echet Bounds on the Supplied Joint}

The chain rule, applied in both directions, constrains which joints can
be supplied for given marginals $a = \cred(A)$ and $b = \cred(B)$; the
constraint is the Fr\'echet--Hoeffding interval~\cite{nelsen2006}.

\begin{proposition}[Fr\'echet bounds]
\label[proposition]{prop:frechet}
Let $a, b, j \in [0,1]$.
\begin{enumerate}[label=\textup{(\alph*)},nosep]
\item \emph{Upper bound:} if conditioning structures for $(j,a)$ and
  $(j,b)$ both exist, then $j \leq \min(a, b)$.
\item \emph{Lower bound:} if the complement coherence
  $1 - a - b + j \geq 0$ holds (the credence of
  $\negc A \land \negc B$ is nonnegative), then $j \geq a + b - 1$.
\item \emph{Realization:} if $a, b > 0$, every
  $j \in [0, \min(a,b)]$ admits conditioning structures in both
  directions.
\end{enumerate}
\end{proposition}
\begin{proof}
(a)~Apply \cref{lem:coherence-bound} to each direction:
$j \leq a$ and $j \leq b$.
(b)~Rearrange $1 - a - b + j \geq 0$.
(c)~For $j \leq \min(a,b)$ with $a, b > 0$,
\cref{prop:existence-uniqueness} produces the conditionals $j/a$ and
$j/b$.
(Machine-checked, \texttt{frechet\_upper}, \texttt{frechet\_lower},
\texttt{frechet\_conditioning\_exists}.)
\end{proof}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[scale=5.2, every node/.style={font=\small}]
% Panel (a): the admissible set Cond(j,e)
\node[anchor=west, font=\small\bfseries] at (-0.08, 0.13)
  {(a) Admissible set $\Cond(j,e) \subseteq [0,1]$};
% Case e > 0
\begin{scope}[yshift=0cm]
\draw[->] (-0.02,0) -- (1.08,0);
\foreach \x/\l in {0/0, 1/1}
  \draw (\x,0.015) -- (\x,-0.015) node[below] {$\l$};
\fill (0.6,0) circle (0.012);
\node[above] at (0.6,0.015) {$j/e$};
\node[anchor=west] at (1.12,0) {$e>0$,\ $j\le e$: singleton};
\end{scope}
% Case e = j = 0
\begin{scope}[yshift=-0.20cm]
\draw[->] (-0.02,0) -- (1.08,0);
\foreach \x/\l in {0/0, 1/1}
  \draw (\x,0.015) -- (\x,-0.015) node[below] {$\l$};
\draw[line width=2.2pt] (0,0) -- (1,0);
\node[anchor=west] at (1.12,0) {$e=j=0$: all of $[0,1]$};
\end{scope}
% Case e = 0 < j
\begin{scope}[yshift=-0.40cm]
\draw[->] (-0.02,0) -- (1.08,0);
\foreach \x/\l in {0/0, 1/1}
  \draw (\x,0.015) -- (\x,-0.015) node[below] {$\l$};
\node[anchor=west] at (1.12,0) {$j>e$ (e.g.\ $e=0<j$): $\emptyset$};
\end{scope}
% Panel (b): admissible joints (Frechet)
\begin{scope}[yshift=-0.72cm]
\node[anchor=west, font=\small\bfseries] at (-0.08, 0.13)
  {(b) Admissible joints for marginals $a=0.7$, $b=0.5$};
\draw[->] (-0.02,0) -- (1.08,0);
\foreach \x/\l in {0/0, 1/1}
  \draw (\x,0.015) -- (\x,-0.015) node[below] {$\l$};
\draw[line width=2.2pt] (0,0) -- (0.5,0);
\draw (0.5,0.03) -- (0.5,-0.03) node[below] {$\min(a,b)$};
\draw (0.2,0.03) -- (0.2,-0.03) node[below] {$a{+}b{-}1$};
\node[anchor=west] at (1.12,0)
  {chain rule: $j \in [0,\min(a,b)]$};
\node[anchor=west, text width=5.2cm] at (1.12,-0.18)
  {\footnotesize complement coherence sharpens\\
   the lower end to $\max(a{+}b{-}1,0)$};
\end{scope}
\end{tikzpicture}
\caption{The conditioning region.
\textbf{(a)} The trichotomy of \cref{thm:trichotomy}: positive evidence
pins a single conditional $j/e$; zero evidence with zero joint admits
the whole interval; an incoherent pair admits nothing.
\textbf{(b)} The Fr\'echet bounds of \cref{prop:frechet}: the chain rule
alone confines the supplied joint to $[0, \min(a,b)]$, drawn for
$a = 0.7$, $b = 0.5$; complement coherence raises the floor to
$\max(a + b - 1, 0) = 0.2$.}
\label{fig:conditioning-region}
\end{figure}

%=============================================================================
\section{The Supplied Joint: Valuations}
\label{sec:valuations}
%=============================================================================

A valuation interprets propositions as credences.
The three valuation classes below differ only in how much dependence
structure they carry, and the difference locates the seed exactly.

\begin{definition}[Complement-preserving valuation]
\label[definition]{def:cpval}
A \emph{complement-preserving valuation} on a set $\alpha$ of
propositions with a negation map $\neg : \alpha \to \alpha$ is a map
$v : \alpha \to [0,1]$ with
\[
v(\neg A) = \negc v(A) = 1 - v(A) \quad \text{for all } A \in \alpha .
\]
(Machine-checked, \texttt{CpValuation}.)
\end{definition}

\begin{definition}[Independent valuation]
\label[definition]{def:indepval}
An \emph{independent valuation} is a complement-preserving valuation
that is additionally a homomorphism for the connectives:
\[
v(A \land B) = v(A) \conjc v(B), \qquad
v(A \lor B) = v(A) \disjc v(B) .
\]
(Machine-checked, \texttt{IndepValuation}.)
\end{definition}

\begin{definition}[Joint valuation]
\label[definition]{def:jointval}
A \emph{joint valuation} is a complement-preserving valuation equipped
with a supplied joint $J : \alpha \times \alpha \to [0,1]$ satisfying
the marginal bounds
\[
J(A, B) \leq v(A), \qquad J(A, B) \leq v(B)
\quad \text{for all } A, B \in \alpha .
\]
(Machine-checked, \texttt{JointValuation}.)
\end{definition}

\begin{proposition}[Joint valuations satisfy the Fr\'echet upper bound]
\label[proposition]{prop:jointval-frechet}
For every joint valuation, $J(A,B) \leq \min(v(A), v(B))$; and whenever
$v(B) > 0$, the conditional $\condc{A}{B}$ is uniquely determined as
$J(A,B) / v(B)$.
\end{proposition}
\begin{proof}
The bound is the conjunction of the two marginal bounds; the
conditional is \cref{prop:existence-uniqueness} applied to
$(J(A,B),\, v(B))$.
(Machine-checked, \texttt{joint\_frechet\_upper},
\texttt{JointValuation.conditioning}.)
\end{proof}

\begin{proposition}[Independence trivializes conditioning]
\label[proposition]{prop:indep-trivial}
Let $v$ be an independent valuation and $v(B) > 0$.
Then the unique conditional credence of $A$ on $B$ with the
homomorphic joint $J(A,B) = v(A) \conjc v(B)$ is
\[
\condc{A}{B} = \frac{v(A)\, v(B)}{v(B)} = v(A) :
\]
evidence never moves any credence.
\end{proposition}
\begin{proof}
By \cref{prop:existence-uniqueness} the conditional is
$J(A,B)/v(B) = v(A) v(B) / v(B) = v(A)$.
\end{proof}

\begin{remark}[The joint is the seed]
\label[remark]{rem:joint-seed}
\Cref{prop:indep-trivial} is the algebraic form of the irreducibility
claim.
If the joint is computed from the marginals by the connective algebra,
conditioning is trivial; Part~1 shows that the only truth-functional
joint that is symmetric, idempotent, and Bayes-consistent is
$\min(a,b)$, and that no truth-functional choice reproduces general
dependence~\cite{part1}.
Nontrivial updating therefore requires the joint to be \emph{supplied}:
$J$ in \cref{def:jointval} is an input the algebra cannot generate,
exactly as a prior is an input Bayes' rule cannot
generate~\cite{jaynes2003} and an axiom base is an input a proof
calculus cannot generate~\cite{simpson2009}.
\end{remark}

%=============================================================================
\section{Bayesian and Jeffrey Update}
\label{sec:update}
%=============================================================================

\begin{definition}[Bayesian update]
\label[definition]{def:bayes-update}
For $j \leq e$ and $e > 0$, the \emph{Bayesian update} is the unique
conditioning structure of \cref{prop:existence-uniqueness}:
\[
\mathrm{bayes}(j, e) = \frac{j}{e} \in [0,1] .
\]
(Machine-checked, \texttt{bayesianUpdate},
\texttt{bayesianUpdate\_val}.)
\end{definition}

\begin{proposition}[Characterization of the Bayesian update]
\label[proposition]{prop:bayes-char}
For $j \leq e$ and $e > 0$:
\begin{enumerate}[label=\textup{(\alph*)},nosep]
\item $\mathrm{bayes}(j,e) \conjc e = j$ (chain rule);
\item if $c \conjc e = j$ for any $c \in [0,1]$, then
  $c = \mathrm{bayes}(j,e)$ (uniqueness);
\item for a joint valuation $v$ with $v(B) > 0$, the posterior
  $v'(A) = \mathrm{bayes}(J(A,B),\, v(B))$ satisfies the chain rule at
  every proposition $A$.
\end{enumerate}
\end{proposition}
\begin{proof}
(a)~$(j/e) \cdot e = j$.
(b)~\Cref{thm:trichotomy}(a): the admissible set is the singleton
$\{j/e\}$.
(c)~Apply (a) with $j = J(A,B)$, $e = v(B)$; the marginal bound of
\cref{def:jointval} supplies $j \leq e$.
(Machine-checked, \texttt{bayesianUpdate\_chain\_rule},
\texttt{bayesianUpdate\_unique},
\texttt{bayesianUpdateValuation\_chain\_rule}.)
\end{proof}

\begin{definition}[Jeffrey update~\cite{jeffrey1965}]
\label[definition]{def:jeffrey}
Given conditional credences $c_B = \condc{A}{B}$ and
$c_{\neg B} = \condc{A}{\neg B}$ and a new evidence credence
$q \in [0,1]$ for $B$, the \emph{Jeffrey update} is the mixture
\[
\mathrm{jeffrey}(c_B, c_{\neg B}, q) = c_B\, q + c_{\neg B}\, (1 - q) .
\]
(Machine-checked, \texttt{jeffreyUpdate}, \texttt{jeffreyUpdate\_val}.)
\end{definition}

\begin{proposition}[Jeffrey update: range and endpoints]
\label[proposition]{prop:jeffrey-props}
For all $c_B, c_{\neg B}, q \in [0,1]$:
\begin{enumerate}[label=\textup{(\alph*)},nosep]
\item $\mathrm{jeffrey}(c_B, c_{\neg B}, q) \in [0,1]$;
\item $\mathrm{jeffrey}(c_B, c_{\neg B}, 1) = c_B$
  (certain evidence: the Bayesian case);
\item $\mathrm{jeffrey}(c_B, c_{\neg B}, 0) = c_{\neg B}$
  (learning $\neg B$ with certainty).
\end{enumerate}
\end{proposition}
\begin{proof}
(a)~The mixture is a convex combination of $c_B, c_{\neg B} \in [0,1]$:
$0 \leq c_B q + c_{\neg B}(1-q) \leq q + (1-q) = 1$.
(b)~At $q = 1$ the second term vanishes and the first is $c_B$.
(c)~Symmetrically at $q = 0$.
(Machine-checked, \texttt{jeffreyUpdate\_certain},
\texttt{jeffreyUpdate\_zero}.)
\end{proof}

\begin{theorem}[Both update rules are underdetermined at zero evidence]
\label{thm:update-underdetermined}
If the prior evidence credence is $0$, then:
\begin{enumerate}[label=\textup{(\alph*)},nosep]
\item every posterior $c \in [0,1]$ is consistent with the chain rule,
  so the Bayesian update has no unique extension to $e = 0$;
\item the Jeffrey mixture contains the unconstrained term
  $c_B \in \Cond(0,0) = [0,1]$, so the Jeffrey update is likewise
  underdetermined.
\end{enumerate}
\end{theorem}
\begin{proof}
(a)~\Cref{cor:zero-underdetermined}: $\Cond(0,0) = [0,1]$, so no value
of the posterior is distinguished by the defining equation.
(b)~The input $c_B = \condc{A}{B}$ of \cref{def:jeffrey} is itself a
conditioning structure on evidence $B$; with $\cred(B) = 0$ it ranges
over all of $[0,1]$ by (a), and distinct choices give distinct mixtures
whenever $q > 0$.
(Machine-checked, \texttt{bayesian\_underdetermined\_at\_zero},
\texttt{jeffrey\_underdetermined\_at\_zero}.)
\end{proof}

%=============================================================================
\section{Coherence: Sequential Equals Batch}
\label{sec:coherence}
%=============================================================================

Fix a joint model and condition on nested evidence in one step or in
two.
The intermediate cancels; this is the multiplicative coherence of the
chain rule, and it is all that ``coherence'' means here.

\begin{theorem}[Sequential equals batch]
\label{thm:seq-batch}
Let $s \leq m \leq b$ in $[0,1]$ with $m, b > 0$ be the credences of
nested events $A \subseteq B \subseteq C$ under a fixed joint model
($s = \cred(A)$, $m = \cred(B)$, $b = \cred(C)$, with joints given by
the nesting).
Then
\[
\mathrm{bayes}(s, b)
= \mathrm{bayes}(s, m) \conjc \mathrm{bayes}(m, b),
\qquad\text{i.e.}\qquad
\frac{s}{b} = \frac{s}{m} \cdot \frac{m}{b} .
\]
\end{theorem}
\begin{proof}
Since $m \neq 0$ and $b \neq 0$,
$\dfrac{s}{m} \cdot \dfrac{m}{b} = \dfrac{s\,m}{m\,b} = \dfrac{s}{b}$,
and each factor lies in $[0,1]$ by the nesting $s \leq m \leq b$, so
both sides are the same credence.
(Machine-checked, \texttt{sequential\_eq\_batch}.)
\end{proof}

\begin{corollary}[The sequential posterior is coherent]
\label[corollary]{cor:seq-chain-rule}
Under the hypotheses of \cref{thm:seq-batch}, the composed sequential
update satisfies the same chain rule as the batch update:
\[
\bigl( \mathrm{bayes}(s,m) \conjc \mathrm{bayes}(m,b) \bigr)
\conjc b = s .
\]
\end{corollary}
\begin{proof}
Rewrite the left side by \cref{thm:seq-batch} as
$\mathrm{bayes}(s,b) \conjc b$ and apply the chain rule of
\cref{prop:bayes-char}(a).
(Machine-checked, \texttt{sequential\_chain\_rule}.)
\end{proof}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[>=Stealth, node distance=2.0cm and 5.6cm,
  every node/.style={font=\small}]
\node (prior) {prior state $(s, m, b)$};
\node (batch) [right=of prior] {$\dfrac{s}{b}$};
\node (step)  [below=of prior]
  {$\Bigl(\dfrac{s}{m},\ \dfrac{m}{b}\Bigr)$};
\draw[->] (prior) -- node[above] {batch: condition $A$ on $C$} (batch);
\draw[->] (prior) -- node[left, align=center]
  {sequential:\\ condition $A$ on $B$,\\ then $B$ on $C$} (step);
\draw[->] (step) -- node[below right] {$\conjc$\ (product)} (batch);
\end{tikzpicture}
\caption{The commuting computation of \cref{thm:seq-batch} for nested
events $A \subseteq B \subseteq C$ with credences
$s \leq m \leq b$, $m, b > 0$.
Both routes from the prior state produce the same posterior
$s/b$; the intermediate credence $m$ cancels in the product
$(s/m) \conjc (m/b)$.}
\label{fig:commuting}
\end{figure}

\begin{remark}[Coherence is not convergence]
\label[remark]{rem:coherence-not-convergence}
\Cref{thm:seq-batch} holds the joint model fixed and says nothing
about which prior wins as data accumulate.
It licenses regrouping a data stream into one effective observation; it
does not move the prior.
Convergence of posteriors under accumulating evidence is a separate
theorem with separate hypotheses (Bernstein--von
Mises~\cite{vandervaart1998}, merging of opinions~\cite{blackwell1962})
and has sharp counterexamples in infinite-dimensional
models~\cite{diaconis1986}.
\Cref{sec:contraction} isolates the distinction in a model small enough
to prove everything.
\end{remark}

%=============================================================================
\section{Cromwell's Rule and the No-Ex-Falso Tie}
\label{sec:cromwell}
%=============================================================================

Cromwell's rule advises never to assign prior credence $0$ or $1$ to a
contingent proposition, because conditioning cannot revise either
value~\cite{jaynes2003}.
In the credence algebra this is a theorem about the annihilator
$c \cdot 0 = 0$, and it is the \emph{same} fact that blocks ex falso.

\begin{theorem}[Cromwell's rule]
\label{thm:cromwell}
Let $a = \cred(A)$ be a prior and $e = \cred(B) > 0$ an evidence
credence with supplied joint $j = \cred(A \land B)$ obeying the
Fr\'echet bounds of \cref{prop:frechet} (chain-rule upper bound and
complement coherence).
\begin{enumerate}[label=\textup{(\alph*)},nosep]
\item \emph{Zero is absorbing:} if $a = 0$ then $j = 0$ and the
  posterior is $\mathrm{bayes}(0, e) = 0$ for every $e > 0$.
\item \emph{One is absorbing:} if $a = 1$ then $j = e$ and the
  posterior is $\mathrm{bayes}(e, e) = 1$ for every $e > 0$.
\item \emph{Zero evidence absorbs:} $c \conjc 0 = 0$ for every
  $c \in [0,1]$; no conditional revises zero evidence to a positive
  joint, and any conditioning structure on evidence $0$ has joint $0$.
\end{enumerate}
\end{theorem}
\begin{proof}
(a)~The upper Fr\'echet bound gives $j \leq \min(0, e) = 0$, so
$j = 0$ and the unique conditional is $0/e = 0$.
(b)~The upper bound gives $j \leq \min(1, e) = e$; complement
coherence gives $j \geq a + e - 1 = e$; hence $j = e$ and the unique
conditional is $e/e = 1$.
(c)~$c \cdot 0 = 0$ identically; positivity of the product would
contradict this, and $j = c \cdot 0 = 0$ for any structure on evidence
$0$.
(Machine-checked components, \texttt{cromwell\_zero\_absorbing},
\texttt{cromwell\_no\_positive\_revision},
\texttt{cromwell\_fixed\_point}; parts (a) and (b) assemble them with
\cref{prop:frechet} and are not separate Lean lemmas.)
\end{proof}

\begin{theorem}[The Cromwell / no-ex-falso tie]
\label{thm:cromwell-tie}
For every $c \in [0,1]$, both of the following hold simultaneously:
\[
c \conjc 0 = 0
\qquad\text{and}\qquad
c \in \Cond(0,0) .
\]
One equation, two readings: the product erases $c$ (a zero joint can
never be revised upward: Cromwell), and precisely because the product
erases $c$, the equation constrains $c$ not at all (impossible evidence
forces no conclusion: no ex falso).
\end{theorem}
\begin{proof}
Both conjuncts restate $c \cdot 0 = 0$:
the first directly, the second through \cref{def:cond}.
(Machine-checked, \texttt{cromwell\_no\_ex\_falso\_tie};
the same pairing phrased at the update boundary is
\texttt{bayesian\_cromwell\_tie}.)
\end{proof}

\begin{proposition}[The update excludes the absorbing point]
\label[proposition]{prop:update-positive}
The Bayesian update of \cref{def:bayes-update} is defined only for
$e > 0$; its positivity precondition excludes zero evidence outright,
so Cromwell's rule appears in the update mechanism as a side condition,
not as a convention about $0/0$.
\end{proposition}
\begin{proof}
$\mathrm{bayes}(j,e) = j/e$ presupposes $e \neq 0$; if $e = 0$ the
hypothesis $0 < e$ of \cref{def:bayes-update} fails.
(Machine-checked, \texttt{bayesianUpdate\_requires\_positive\_evidence}.)
\end{proof}

\begin{remark}[Contrast with residuation]
\label[remark]{rem:residuation-contrast}
Product residuation $a \to b = \sup\{c : ac \leq b\}$ fills the
zero-evidence cell with $1$ (ex falso) for every left-continuous
t-norm~\cite{hajek1998}; classical conditional probability leaves it
undefined~\cite{cox1961}; the chain rule leaves it unconstrained
(\cref{cor:zero-underdetermined}).
All three options agree at positive evidence~\cite{part1}.
None of the three is neutral; the choice at $e = 0$ is part of the
opening commitment.
\end{remark}

%=============================================================================
\section{The Seed at Finite and Infinite Time: a Contraction Model}
\label{sec:contraction}
%=============================================================================

The following one-dimensional model separates, with exact constants,
the two claims that \cref{rem:coherence-not-convergence} insists on
keeping apart: the seed determines every finite stage, and the seed
washes out in the limit, except at the absorbing boundary.

\begin{definition}[Averaging refinement map]
\label[definition]{def:toymap}
For a target $t \in [0,1]$, define $f_t : [0,1] \to [0,1]$ by
\[
f_t(x) = \frac{x + t}{2} .
\]
The orbit from seed $x_0 \in [0,1]$ is $x_n = f_t^{\,n}(x_0)$.
(Machine-checked, \texttt{toyA}.)
\end{definition}

\begin{theorem}[Exact geometric contraction]
\label{thm:contraction}
For all $t, x_0 \in [0,1]$ and $n \in \Nat$,
\[
f_t^{\,n}(x_0) - t = \Bigl(\tfrac{1}{2}\Bigr)^{n} (x_0 - t),
\qquad\text{hence}\qquad
\bigl| f_t^{\,n}(x_0) - t \bigr|
= \Bigl(\tfrac{1}{2}\Bigr)^{n} \, |x_0 - t| .
\]
\end{theorem}
\begin{proof}
Induction on $n$.
Base: $f_t^{\,0}(x_0) - t = x_0 - t$.
Step: $f_t(x) - t = \frac{x + t}{2} - t = \frac{x - t}{2}$, so applying
$f_t$ once more halves the signed distance; the displayed identity
follows, and taking absolute values gives the second form.
(Machine-checked, \texttt{toyA\_iterate\_dist},
\texttt{toyA\_contraction\_bound}.)
\end{proof}

\begin{corollary}[Seed matters at every finite stage]
\label[corollary]{cor:seed-matters}
For all seeds $x_0 \neq x_0'$ and every $n \in \Nat$,
\[
f_t^{\,n}(x_0) - f_t^{\,n}(x_0')
= \Bigl(\tfrac{1}{2}\Bigr)^{n} (x_0 - x_0') \neq 0 :
\]
no finite number of refinement steps erases the opening commitment.
\end{corollary}
\begin{proof}
Subtract the two instances of the identity in \cref{thm:contraction};
the difference is a nonzero multiple of $x_0 - x_0'$.
(Machine-checked instance, \texttt{toyA\_seed\_matters}.)
\end{proof}

\begin{corollary}[Seed washes out in the limit]
\label[corollary]{cor:washout}
For every seed $x_0 \in [0,1]$,
$f_t^{\,n}(x_0) \to t$ as $n \to \infty$, and the limit satisfies the
self-consistency equation $f_t(t) = t$.
\end{corollary}
\begin{proof}
By \cref{thm:contraction} the distance to $t$ is
$2^{-n}|x_0 - t| \to 0$.
Fixed point: $f_t(t) = (t + t)/2 = t$.
(Machine-checked, \texttt{toyA\_contraction\_tendsto},
\texttt{toyA\_target\_fixed}, \texttt{toyA\_limit\_fixed}.)
\end{proof}

\begin{theorem}[The absorbing boundary is exempt]
\label{thm:absorbing-orbit}
Under the product map $g_e(x) = x \conjc e$ (one multiplicative update
per step, any fixed $e \in [0,1]$), the zero seed is absorbing:
\[
g_e^{\,n}(0) = 0 \quad \text{for all } n \in \Nat ,
\]
and at every step the iterate is a witness of the unconstrained
zero-evidence conditioning $\Cond(0,0)$.
\end{theorem}
\begin{proof}
Induction on $n$: $g_e(0) = 0 \cdot e = 0$, and if $g_e^{\,n}(0) = 0$
then $g_e^{\,n+1}(0) = g_e(0) = 0$.
Each iterate equals $0 \in [0,1] = \Cond(0,0)$ by
\cref{thm:trichotomy}(b).
(Machine-checked, \texttt{toyA\_absorbing\_zero},
\texttt{toyA\_absorbing\_zero\_no\_ex\_falso}.)
\end{proof}

\begin{remark}[What the model does and does not show]
\label[remark]{rem:toy-analogy}
\Cref{cor:washout} is the local, exactly solvable analogue of the
washing out of priors (Bernstein--von Mises~\cite{vandervaart1998},
merging of opinions~\cite{blackwell1962}); \cref{thm:absorbing-orbit}
is the iterated face of Cromwell's rule.
The analogy is illustrative, not structural: real posterior convergence
holds under measure-theoretic hypotheses that can fail in
infinite-dimensional models~\cite{diaconis1986}, while the contraction
here is unconditional because the model is one-dimensional and linear.
\end{remark}

%=============================================================================
\section{Credal Sets and the Model-Comparison Boundary}
\label{sec:modelcomparison}
%=============================================================================

When the seed itself is uncertain, the honest posterior is a set, not a
number~\cite{walley1991,berger1985,levi1980}.
This section first characterizes the set obtained by ranging the
supplied joint over a class of priors, then derives, exactly, the
classical model-comparison boundary where a too-diffuse seed loses
every comparison.

\subsection{Credal Classes over the Supplied Joint}

\begin{definition}[Induced posterior set]
\label[definition]{def:condset}
For a class $S \subseteq [0,1]$ of joint credences and a fixed evidence
$e \in [0,1]$, the \emph{induced posterior set} is
\[
\CondSet(S, e) = \bigcup_{j \in S} \Cond(j, e) .
\]
(Machine-checked, \texttt{condSet}.)
\end{definition}

\begin{proposition}[Monotonicity]
\label[proposition]{prop:condset-monotone}
If $S \subseteq T$ then $\CondSet(S,e) \subseteq \CondSet(T,e)$:
enlarging the prior class can only enlarge the posterior set.
\end{proposition}
\begin{proof}
A union over a larger index set contains the union over the smaller.
(Machine-checked, \texttt{condSet\_monotone\_joint}.)
\end{proof}

\begin{theorem}[Positive evidence: the posterior set is the image of
the feasible class]
\label{thm:condset-positive}
For $e > 0$,
\[
\CondSet(S, e)
= \Bigl\{ \tfrac{j}{e} \;:\; j \in S,\ j \leq e \Bigr\} .
\]
In particular, if $\CondSet(S,e) = [0,1]$ then for every
$c \in [0,1]$ there is a feasible $j \in S$ with $j \leq e$ and
$j/e = c$; and the full feasible class realizes everything:
$\CondSet(\{j : j \leq e\},\, e) = [0,1]$.
\end{theorem}
\begin{proof}
By \cref{thm:trichotomy}, $\Cond(j,e)$ is $\{j/e\}$ when $j \leq e$
and $\emptyset$ when $j > e$, so the union collects exactly the values
$j/e$ over the feasible subclass.
For the surjectivity claim, membership of $c$ in the union produces a
feasible $j$ with $c \cdot e = j$, hence $j/e = c$.
For the last claim, given $c \in [0,1]$ take $j = c \cdot e$;
then $j \leq e$ and $j/e = c$.
(Machine-checked, \texttt{condSet\_positive\_evidence},
\texttt{condSet\_univ\_implies\_surjective},
\texttt{condSet\_full\_feasible\_eq\_univ}.)
\end{proof}

\begin{theorem}[Zero evidence: the posterior set is maximal]
\label{thm:condset-zero}
If $0 \in S$, then $\CondSet(S, 0) = [0,1]$; and for $j \neq 0$,
$\Cond(j, 0) = \emptyset$.
A prior class containing the degenerate joint, conditioned on
zero-credence evidence, constrains the posterior not at all.
\end{theorem}
\begin{proof}
$\Cond(0,0) = [0,1]$ by \cref{thm:trichotomy}(b), and this term alone
already exhausts $[0,1]$ inside the union; for $j \neq 0$,
\cref{lem:coherence-bound} makes the equation $c \cdot 0 = j$
unsolvable.
(Machine-checked, \texttt{condSet\_zero\_evidence\_of\_zero\_mem},
the empty case via \cref{thm:trichotomy}(c).)
\end{proof}

\begin{remark}[Imprecise probability]
\label[remark]{rem:imprecise}
\Cref{thm:condset-positive,thm:condset-zero} are the algebraic face of
the credal-set stance: where the prior class underdetermines the
update, the posterior is the induced interval, and the degenerate case
returns the vacuous set $[0,1]$, exactly as imprecise probability
prescribes for null events~\cite{walley1991,troffaes2014}.
The next subsection shows the same set-valued degeneration arising in
orthodox Bayesian model comparison.
\end{remark}

\subsection{Marginal Likelihood and the Occam Factor}

The remaining results of this section are classical computations, not
Lean-formalized; they are included because they make the cost of the
seed exact.
Throughout, $\varphi(u) = (2\pi)^{-1/2} e^{-u^2/2}$ is the standard
normal density and $\Phi$ its distribution function.

\begin{definition}[Marginal likelihood and Bayes factor]
\label[definition]{def:bayes-factor}
For a model with parameter $\theta$, prior $\pi$, likelihood
$p(x \mid \theta)$, and data $x$, the \emph{marginal likelihood} is
\[
m(x) = \int p(x \mid \theta) \, d\pi(\theta) ,
\]
and the \emph{Bayes factor} between models $M_1, M_0$ is
$B_{10}(x) = m_1(x) / m_0(x)$~\cite{jeffreys1939}.
\end{definition}

\begin{lemma}[Flat-prior marginal likelihood, Gaussian location model]
\label[lemma]{lem:gaussian-ml}
Let $M_0$: $x \sim \mathcal{N}(0,1)$, and $M_1$:
$g \sim \mathrm{Uniform}[0, W]$, $x \mid g \sim \mathcal{N}(g, 1)$.
Then
\[
m_1(x) = \frac{1}{W} \bigl( \Phi(x) - \Phi(x - W) \bigr),
\qquad
B_{10}(x) = \frac{\sqrt{2\pi}}{W}\, e^{x^2/2}
  \bigl( \Phi(x) - \Phi(x - W) \bigr) .
\]
For $x \in [0, W]$ the maximized likelihood is
$p(x \mid \hat g) = \varphi(0) = (2\pi)^{-1/2}$ at $\hat g = x$, and
\[
m_1(x) = p(x \mid \hat g) \cdot \Omega(x, W),
\qquad
\Omega(x, W)
= \frac{\sqrt{2\pi}}{W} \bigl( \Phi(x) - \Phi(x - W) \bigr)
\leq \frac{\sqrt{2\pi}}{W} ,
\]
exhibiting the Occam factor $\Omega$ as the fraction of prior width the
fitted parameter uses.
\end{lemma}
\begin{proof}
Substituting $u = x - g$,
\[
m_1(x) = \frac{1}{W} \int_0^W \varphi(x - g)\, dg
= \frac{1}{W} \int_{x-W}^{x} \varphi(u)\, du
= \frac{1}{W} \bigl( \Phi(x) - \Phi(x - W) \bigr) .
\]
Dividing by $m_0(x) = \varphi(x) = (2\pi)^{-1/2} e^{-x^2/2}$ gives
$B_{10}$.
For $x \in [0,W]$ the likelihood $g \mapsto \varphi(x - g)$ attains its
maximum $\varphi(0)$ at $g = x \in [0,W]$, and
$m_1(x) / \varphi(0) = \sqrt{2\pi}\, m_1(x) = \Omega(x,W)$; the bound
uses $\Phi(x) - \Phi(x-W) \leq 1$.
\end{proof}

\begin{theorem}[Bartlett's indeterminacy: the diffuse seed loses]
\label{thm:bartlett}
For every fixed datum $x$,
\[
B_{10}(x) \leq \frac{\sqrt{2\pi}\, e^{x^2/2}}{W}
\qquad\text{and}\qquad
\lim_{W \to \infty} B_{10}(x) = 0 :
\]
as the prior width grows without bound, the comparison turns against
$M_1$ regardless of the data, and in the improper limit the Bayes
factor has no determinate positive value~\cite{bartlett1957}.
\end{theorem}
\begin{proof}
From \cref{lem:gaussian-ml},
$B_{10}(x) = \sqrt{2\pi}\, e^{x^2/2}\,
(\Phi(x) - \Phi(x-W)) / W
\leq \sqrt{2\pi}\, e^{x^2/2} / W \to 0$ as $W \to \infty$, since the
$\Phi$ difference is bounded by $1$ while the numerator is fixed.
\end{proof}

\begin{theorem}[Lindley's paradox: the fine-tuned trajectory]
\label{thm:lindley}
Take $n$ i.i.d.\ observations under the models of
\cref{lem:gaussian-ml}, with sufficient statistic
$\bar{x} \sim \mathcal{N}(g, 1/n)$.
Along the trajectory on which the standardized statistic
$z = \sqrt{n}\, \bar{x}$ is held fixed as $n \to \infty$,
\[
B_{10}(n)
= \frac{\sqrt{2\pi}}{W \sqrt{n}}\; e^{z^2/2}\,
  \bigl( \Phi(z) - \Phi(z - \sqrt{n}\, W) \bigr)
\;\longrightarrow\; 0 ,
\]
while the frequentist tail probability $2(1 - \Phi(|z|))$ stays
constant: the Bayesian verdict reverses against $M_1$ on a trajectory
along which the significance test rejects $M_0$ at a fixed
level~\cite{lindley1957}.
\end{theorem}
\begin{proof}
The marginal likelihoods are computed as in \cref{lem:gaussian-ml}
with $\varphi$ replaced by the $\mathcal{N}(g, 1/n)$ density of
$\bar{x}$: substituting $u = \sqrt{n}(\bar{x} - g)$,
\[
m_1(\bar{x})
= \frac{1}{W} \int_0^W \sqrt{n}\, \varphi\bigl(\sqrt{n}(\bar{x}-g)\bigr) dg
= \frac{1}{W}\bigl( \Phi(\sqrt{n}\,\bar{x})
  - \Phi(\sqrt{n}(\bar{x} - W)) \bigr),
\]
and $m_0(\bar{x}) = \sqrt{n}\, \varphi(\sqrt{n}\, \bar{x})
= \sqrt{n/2\pi}\; e^{-n \bar{x}^2 / 2}$.
With $z = \sqrt{n}\, \bar{x}$ fixed, the ratio is the displayed
expression; it is bounded by
$\sqrt{2\pi}\, e^{z^2/2} / (W \sqrt{n}) \to 0$.
The frequentist statistic depends on $z$ alone, hence is constant on
the trajectory.
\end{proof}

\begin{remark}[The implicit Occam factor of the significance test]
\label[remark]{rem:implicit-omega}
The likelihood-ratio statistic of the significance test is
$e^{z^2/2}$, which is the first factor of $B_{10}$ in
\cref{thm:lindley} with the Occam factor replaced by $1$.
The Lindley reversal is exactly the gap between $\Omega = 1$ and the
$\Omega \sim \sqrt{2\pi}/(W\sqrt{n})$ that a proper prior supplies;
and the trajectory of constant $z$ is a measure-zero path in the data,
off which $\bar{x}$ tracks $g$ and both verdicts agree
eventually~\cite{jeffreys1939,berger1985}.
\end{remark}

\begin{remark}[Set-valued comparison under an unfixed seed]
\label[remark]{rem:set-valued-bf}
Range the prior width over a class $W \in [W_1, W_2]$ and
\cref{lem:gaussian-ml} returns an interval of Bayes factors,
monotone in $1/W$; in the improper limit the interval degenerates.
This is the statistical image of
\cref{thm:condset-positive,thm:condset-zero}: where the seed class
underdetermines the comparison, the admissible answer is a set, and
the degenerate seed yields the vacuous answer.
The structural match is exact at the level of the two trichotomies; we
do not claim the Gaussian model embeds in the credence algebra.
\end{remark}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[>=Stealth, xscale=1.6, yscale=2.6,
  every node/.style={font=\small}]
% axes
\draw[->] (-0.3,0) -- (6.6,0) node[below] {$g$};
\draw[->] (0,-0.05) -- (0,1.15) node[left] {density};
% flat prior of width W
\draw[dashed, thick] (0,0.16) -- (6,0.16);
\draw[dashed] (0,0) -- (0,0.16);
\draw[dashed] (6,0) -- (6,0.16);
\node[above] at (5.1,0.16) {prior $1/W$};
\draw (6,0.03) -- (6,-0.03) node[below] {$W$};
\draw (0,0.03) -- (0,-0.03) node[below] {$0$};
% likelihood bump centered at hat g = 2.2, sd 0.4
\draw[thick, domain=0.7:3.7, samples=120, smooth]
  plot (\x, {exp(-(\x-2.2)^2/0.32)});
\node[above] at (2.2,1.0) {likelihood $p(x \mid g)$};
\draw (2.2,0.03) -- (2.2,-0.03) node[below] {$\hat g$};
% effective width annotation
\draw[<->] (1.7,0.55) -- (2.7,0.55);
\node[above] at (2.2,0.55) {$\approx \sqrt{2\pi}\,\sigma$};
% shaded overlap: prior mass under the bump
\fill[gray, fill opacity=0.35, domain=0.7:3.7, samples=120, smooth]
  plot (\x, {min(exp(-(\x-2.2)^2/0.32), 0.16)}) -- (0.7,0) -- cycle;
% Omega annotation
\node[anchor=west, text width=5.0cm] at (4.0,0.75)
  {Occam factor:\\[2pt]
   $\displaystyle \Omega \approx \frac{\sqrt{2\pi}\,\sigma}{W}$\\[4pt]
   prior mass spent on the\\ data-supported region};
\draw[->] (4.0,0.62) to[bend right=15] (2.6,0.14);
\end{tikzpicture}
\caption{The Occam factor of \cref{lem:gaussian-ml}.
A flat prior of width $W$ spends density $1/W$ everywhere; the
likelihood occupies an effective width $\approx \sqrt{2\pi}\,\sigma$
around the fit $\hat g$ ($\sigma = 1$ in the lemma, $1/\sqrt{n}$ after
$n$ observations).
The marginal likelihood is the best fit times the shaded fraction
$\Omega \approx \sqrt{2\pi}\,\sigma / W$ of prior mass the fit uses;
widening the seed shrinks $\Omega$ and, by \cref{thm:bartlett},
eventually loses the comparison regardless of the data.}
\label{fig:occam}
\end{figure}

%=============================================================================
\section{The Irreducible Commitment}
\label{sec:synthesis}
%=============================================================================

The theorems assemble into one structural statement about the credence
algebra, plus an analogy that we keep explicitly labeled as one.

\begin{theorem}[The supplied pair is irreducible in the credence algebra]
\label{thm:irreducible}
Nontrivial chain-rule conditioning requires the pair $(j, e)$ as an
input that the connective algebra does not determine:
\begin{enumerate}[label=\textup{(\alph*)},nosep]
\item computing $j$ from the marginals by the connective $\conjc$
  trivializes updating (\cref{prop:indep-trivial});
\item the admissible conditional is a function of the supplied pair
  alone, and only on the region $e > 0$, $j \leq e$
  (\cref{thm:trichotomy});
\item the boundary behavior is forced, not chosen: credences $0$ and
  $1$ absorb (\cref{thm:cromwell}), and the zero-evidence cell is
  unconstrained rather than explosive (\cref{thm:cromwell-tie});
\item changing the data grouping does not change the posterior
  (\cref{thm:seq-batch}); changing the supplied pair does
  (\cref{cor:seed-matters}).
\end{enumerate}
\end{theorem}
\begin{proof}
Each clause is the cited result; the conjunction is the claim.
\end{proof}

\begin{remark}[The dictionary, as analogy]
\label[remark]{rem:dictionary}
\Cref{tab:dictionary} aligns the three systems.
Each row matches structural roles, not mathematical objects; no functor
or bi-interpretation is claimed.
The first two columns record standard
results~\cite{jaynes2003,simpson2009,debruijn1970,turing1939,feferman1962};
the third column is proved in this paper and in
Parts~1--3~\cite{part1,part2}.
\end{remark}

\begin{table}[ht]
\centering
\small
\begin{tabular}{p{0.29\linewidth}p{0.29\linewidth}p{0.32\linewidth}}
\toprule
\textbf{Probability} & \textbf{Foundations} & \textbf{Credence algebra} \\
\midrule
prior $\pi$ &
axiom base, trusted kernel &
supplied pair $(j, e)$ (\cref{def:jointval}) \\[3pt]
marginal likelihood, Occam factor $\Omega$
(\cref{lem:gaussian-ml}) &
axiom strength, reverse mathematics~\cite{simpson2009} &
shape of $\Cond(j,e)$ (\cref{thm:trichotomy}) \\[3pt]
posterior convergence, merging~\cite{vandervaart1998,blackwell1962} &
iterated reflection~\cite{turing1939,feferman1962} &
contraction to the fixed point
(\cref{cor:washout}) \\[3pt]
Cromwell's rule: $0$, $1$ unrevisable~\cite{jaynes2003} &
G\"odel and Tarski limits~\cite{godel1931} &
absorbing boundary, no ex falso
(\cref{thm:cromwell,thm:cromwell-tie}) \\[3pt]
imprecise-probability posterior sets~\cite{walley1991,berger1985} &
independence: a family of models &
credal set $\CondSet(S,e)$
(\cref{thm:condset-positive,thm:condset-zero}) \\
\bottomrule
\end{tabular}
\caption{The dictionary.
Rows match structural roles across three systems; only the third
column is formalized here.
The first row is the irreducible commitment in each vocabulary.}
\label{tab:dictionary}
\end{table}

\begin{remark}[Trilemma and boat]
\label[remark]{rem:trilemma}
That the opening commitment cannot be derived inside the system it
grounds is the regress horn of the M\"unchhausen
trilemma~\cite{albert1968}; that it can nonetheless be revised in
flight is Neurath's boat~\cite{neurath1932}; that adopting it is a
practical choice rather than an internal theorem is Carnap's
external question~\cite{carnap1950}.
We add nothing to these; the theorems above only show that, within the
credence algebra, the commitment can be \emph{located} (the supplied
pair), \emph{priced} (the Occam factor, the width of $\Cond$), and
\emph{bounded} (the trichotomy and the absorbing boundary).
\end{remark}

\begin{remark}[The de Bruijn discipline]
\label[remark]{rem:debruijn}
Mechanization relocates trust; it does not eliminate it.
Every theorem of
\cref{sec:chainrule,sec:valuations,sec:update,sec:coherence,sec:cromwell,sec:contraction}
and the credal-set results of \cref{sec:modelcomparison} are checked by
the Lean kernel against Mathlib's axioms, which is the de Bruijn
criterion: a small trusted core re-checks each proof, so the reader
trusts the checker, not the author~\cite{debruijn1970}.
Gentzen's consistency proof~\cite{gentzen1936} is the canonical
precedent for such relocation: the unprovable statement is exchanged
for a stronger principle held outside the system.
The relocation here is to a smaller, public, re-runnable core, not to
no core at all.
\end{remark}

%=============================================================================
\section{Conclusion}
\label{sec:conclusion}
%=============================================================================

The chain rule $c \conjc e = j$, taken as the defining identity of
conditioning, yields a complete and elementary account of the opening
commitment in the credence algebra.
The admissible set obeys an exact trichotomy
(\cref{thm:trichotomy}); the boundary values absorb and the same
annihilator blocks ex falso (\cref{thm:cromwell,thm:cromwell-tie});
coherence is path independence in the data grouping and nothing more
(\cref{thm:seq-batch}); the seed governs every finite stage and washes
out in the limit, except at the absorbing boundary
(\cref{thm:contraction,thm:absorbing-orbit}); and ranging the seed over
a class produces exactly the set-valued posteriors of imprecise
probability, with the Gaussian model-comparison boundary
(\cref{thm:bartlett,thm:lindley}) as the classical statistical face of
the same degeneration.

Every framework that conditions must commit to something at evidence
zero, and to a seed before the first update: probability leaves the
cell undefined and demands a prior; residuated implication fills the
cell with $1$ and demands a t-norm; the chain rule leaves the cell
unconstrained and demands the pair $(j,e)$.
None of these is neutral.
The defensible position, and the only one the theorems support, is to
make the one unavoidable commitment explicit, minimal, and checked:
here it is $c \cdot 0 = 0$ together with a named, supplied pair,
machine-verified down to the kernel.

\section*{Acknowledgements}

Anthropic's Claude was used as an assistant for both the Lean
formalization and the manuscript preparation.

%=============================================================================
% Bibliography
%=============================================================================

\bibliographystyle{plain}
\bibliography{../cred}

%=============================================================================
% APPENDICES
%=============================================================================
\appendix

%=============================================================================
\section{Lean Formalization}
\label{app:lean}
%=============================================================================

All credence-algebra results are formalized in Lean~4 with Mathlib;
verification is \texttt{cd lean \&\& lake build} with toolchain
Lean~4.16.0 and Mathlib~4.16.0, zero \texttt{sorry}.
The Gaussian computations of \cref{sec:modelcomparison}
(\cref{lem:gaussian-ml,thm:bartlett,thm:lindley}) are classical and not
formalized.

\textbf{File map.}
\begin{itemize}[nosep]
\item \texttt{Cred/Cond/Admissible.lean}: the \texttt{Conditioning}
  structure, the admissible set \texttt{Cond}, the trichotomy, and the
  Fr\'echet bounds (\cref{sec:chainrule}).
\item \texttt{Cred/Valuation.lean}: \texttt{CpValuation},
  \texttt{IndepValuation}, \texttt{JointValuation}
  (\cref{sec:valuations}).
\item \texttt{Cred/Update.lean}: Bayesian and Jeffrey updates and
  zero-evidence underdetermination (\cref{sec:update}).
\item \texttt{Cred/Coherence.lean}: sequential-equals-batch coherence
  (\cref{sec:coherence}).
\item \texttt{Cred/Cromwell.lean}: Cromwell's rule and the
  no-ex-falso tie (\cref{sec:cromwell}).
\item \texttt{Cred/ToyModel.lean}: the contraction model
  (\cref{sec:contraction}).
\item \texttt{Cred/Lindley.lean}: credal classes over the supplied
  joint (\cref{sec:modelcomparison}).
\end{itemize}

\textbf{Design choices.}
\texttt{Credence} is a structure carrying explicit \texttt{nonneg} and
\texttt{le\_one} proofs.
\texttt{Conditioning joint evidence} is a dependent type packaging the
conditional credence with its chain-rule proof, so existence and
uniqueness statements are statements about inhabitants of a type.
\texttt{Cond j e} is the plain solution set
$\{c \mid c \conjc e = j\}$; the equivalence of the two views is
\texttt{mem\_cond\_iff}.
The Bayesian update \texttt{bayesianUpdate} takes the positivity
hypothesis $0 < e$ as an argument, so the absorbing point is excluded
by type discipline rather than by a runtime convention
(\cref{prop:update-positive}).

\medskip
\noindent
\texttt{structure Conditioning (joint evidence : Credence) where}\\
\null\quad \texttt{condCred : Credence}\\
\null\quad \texttt{chainRule : condCred}~$\conjc$~\texttt{evidence = joint}

\medskip
\noindent
\texttt{def Cond (j e : Credence) : Set Credence := \{c | c}~$\conjc$~\texttt{e = j\}}

\medskip
\textbf{Key theorems (this paper).}
\begin{itemize}[nosep]
\item Trichotomy: \texttt{cond\_singleton\_of\_pos},
  \texttt{cond\_zero\_zero\_univ}, \texttt{cond\_nonempty\_iff}
\item Zero evidence: \texttt{conditioning\_zero\_any},
  \texttt{conditioning\_zero\_forces\_joint\_zero}
\item Fr\'echet: \texttt{frechet\_upper}, \texttt{frechet\_lower},
  \texttt{frechet\_conditioning\_exists}
\item Update: \texttt{bayesianUpdate\_chain\_rule},
  \texttt{bayesianUpdate\_unique}, \texttt{jeffreyUpdate\_certain},
  \texttt{jeffreyUpdate\_zero},
  \texttt{bayesian\_underdetermined\_at\_zero}
\item Coherence: \texttt{sequential\_eq\_batch},
  \texttt{sequential\_chain\_rule}
\item Cromwell: \texttt{cromwell\_zero\_absorbing},
  \texttt{cromwell\_fixed\_point}, \texttt{cromwell\_no\_ex\_falso\_tie}
\item Contraction: \texttt{toyA\_iterate\_dist},
  \texttt{toyA\_contraction\_tendsto}, \texttt{toyA\_absorbing\_zero}
\item Credal sets: \texttt{condSet\_monotone\_joint},
  \texttt{condSet\_positive\_evidence},
  \texttt{condSet\_full\_feasible\_eq\_univ},
  \texttt{condSet\_zero\_evidence\_of\_zero\_mem}
\end{itemize}

\end{document}
